### 2014-04-08 paper/discussion.tex
added sample input and paper files

\section{Additional results analysis}\label{additional}
In this section we examine some secondary quantities derived
from the hydrodynamic simulations above. To keep this discussion
concise, we will use selected simulations from above for illustration.

\subsection{Three-dimensionality}
A simple measure of three-dimensionality of the flow is to compare vertical 
to horizontal motion. Since we are interested in non-axisymmetric
perturbations to the gap edge, we first Fourier transform the
meridional momentum densities

\begin{align}
  (v_{Rm}, v_{zm} ) \equiv\int_0^{2\pi}\rho\times(u_R,
  u_z)\exp{(-\ii m\phi)}d\phi. 
\end{align}
We define the three dimensionality as $\Theta_m(z/H)$, where 
\begin{align}
  \Theta_m^2 \equiv \frac{\avg{|v_{zm}|^2}}{\avg{|v_{zm}|^2}
    +\avg{|v_{Rm}|^2}},  
\end{align}
and $\avg{\cdot}$ denotes a radial average. 
Admittedly, this is a crude measure, and exact values of $\Theta_m$
varies somewhat with details of the average. However, we have
experimented with different averaging domains and 
found the features described below are robust.

The top panel in Fig. \ref{compare_vprofiles_3d008} shows $\Theta_m$
for Cases 1--3 at $t=40P_0$. The radial average is taken over
$r-r_p\in[3,7]r_h$. These are all vortex modes (see
Fig. \ref{vortex8_polar_dens} and
Fig. \ref{vortex4_vortex10_overall}). The flow becomes increasingly
three-dimensional away from the midplane but 
$\Theta_m = O(10^{-1})$ is small. In an averaged sense the flow is
mostly horizontal. At the end of the  
simulation for Case 1, an azimuthally extended vortex dominates the
flow, for which we measured $\Theta_1\sim
0.2$---0.3. Thus, although vertical motion can become an appreciable
fraction of horizontal motion, the former never dominates.

$\Theta_m$ for Cases 4---7 are shown in in bottom panel of
Fig. \ref{compare_vprofiles_3d008}. The radial average is performed
over $r-r_p\in[2,6]r_h$ 
because the global spirals in Cases 6---7 significantly protrude the
gap edge. The snapshot is taken at $t=50P_0$ for Cases 4---5, at
$t=40P_0$ for Case 6 and at $t=30P_0$, so that the
vortices and spirals have comparable over-densities at the
gap edge. It also reflects the fact that spiral modes are more
unstable than vortex modes and develop earlier  
\citep{lin11b}. $\Theta_m \sim 0.2$ is again not particularly
large, but the spiral modes are distinctly more three-dimensional than
vortex modes. This is likely due to additional vertical
acceleration provided by the strong self-gravity in those cases.

%
% can't do detail comparision between cases because instability have
% different growth rates. different 'stages' in instability

% Case 6 show 3 edge disturbances but this is likely to be a mixture
% of spiral and vortex modes, rather than the m=3 spiral mode. This is
% because 
% we clearly see m=2 spiral in Case 7 (Q_0=1.5), and to see higher m spiral
% modes one should increase the strength of self-gravity (e.g. lin11b
% observe m=3 spirals with Q_0=1.2) 
%

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.7cm 0cm
    0cm]{figures/compare_vprofiles_3d008}  
  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm
    0.2cm]{figures/compare_vprofiles_3d010}  
  \caption{Three-dimensionality of the non-axisymmetric
    flow near the outer gap edge. Top: Cases 1---3 (vortex modes). 
    Bottom: Cases 4---5 (vortex modes) and Cases 6---7 (spiral
    modes). The azimuthal wavenumber $m$ is chosen to match the number
    of vortices or large-scale spirals observed. 
    \label{compare_vprofiles_3d008}}
\end{figure}

\subsection{Vortensity field}\label{vortensity}
A fundamental distinction between the linear vortex and edge mode
instability is their association with local vortensity minimum and
maximum, respectively. In this section we compare vortensity fields of 
discs with vortex modes (Case 2) and edge modes (Case 7). 
%We use Case 2 
%and Case 7 as examples, respectively. 
More specifically, we examine the relative
perturbation to the vertical component of vortensity,  
\begin{align}
  \Delta\eta_z\equiv \frac{\eta_z - \eta_z(t=0)}{\eta_z(t=0)},
\end{align}
where
\begin{align}
  \eta_z \equiv \frac{\bm{\hat{z}}\cdot\nabla\times\bm{u}}{\rho}. 
\end{align}

Fig. \ref{vortex4_vortex1_vortxy} compares $\Delta\eta_z$ in the
midplane when vortices and spirals develop. For planetary gaps, 
vortensity maxima and minima are both located near the gap edges with 
characteristic separation of the local scale-height. The vortensity
ring at the inner gap edge ($r-r_p\simeq - 2r_h$) remain
well-defined. The vortex instability is associated with the local
vortensity minimum near the outer gap edge --- seen as localised closed
contour lines centred about $r - r_p \sim 4r_h$. The vortensity ring at
$r-r_p\sim +2r_h$ becomes distorted as a \emph{consequence} of
large-scale vortex formation just exterior to it. By contrast, the
edge-spiral mode is associated with the local vortensity
maximum. Their development inherently disrupts the vortensity
rings. This is seen in the right panel as the outer ring is broken up.

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0cm 0cm 1.9cm 
    0.9cm]{figures/vortex4_vortxy_008}\includegraphics[scale=.425,clip=true,trim=2.3cm    
    0.0cm 0cm 
    0.9cm]{figures/vortex1_vortxy_006}
  \caption{Relative perturbation to the vertical component of
    vortensity at the midplane in a disc with the vortex instability
    (left, Case 2 at $t=40P_0$) and the 
    spiral instability (right, Case 7 at $t=30P_0$). Negative
    perturbations in the region $r-r_p\in[2,6]r_h$ are outlined by white lines. 
    Dotted horizontal lines indicate azimuthal cuts taken in
    Fig. \ref{vortex4_vortex1_vortRZ}.   
    \label{vortex4_vortex1_vortxy}}
\end{figure}

The vertical structures also differ. Fig. \ref{vortex4_vortex1_vortRZ}
compares $\Delta\eta_z$ at azimuths coinciding with 
a vortex or the edge disturbance of the spiral mode. Both
instabilities involve $\Delta\eta_z<0$. It is clear that 
the spiral mode has stronger vertical dependence. Its 
region of $\Delta\eta_z<0$ becomes thinner away from
the midplane. In the vortex case this region remains 
about the same width and  $\Delta\eta_z$ is approximately uniform
within it. 
%This is qualitatively
%consistent with increased (downards) vertical motion. 
While $\mathrm{min}(\Delta\eta_z)$ is of comparable
magnitude, the vortensity ring at $r-r_p=2r_h$is much weaker and
thinner in the spiral case ($\Delta\eta_z$ being a factor $\sim 4$
smaller than the vortex case).

\begin{figure}
  \centering
  \includegraphics[scale=.47,clip=true,trim=0cm 1.32cm .0cm
    .6cm]{figures/vortex4_vortRZ_008}\\\includegraphics[scale=.47,clip=true,trim=0cm 
    0.cm .0cm
    .6cm]{figures/vortex1_vortRZ_006}
  \caption{Relative vertical vortensity perturbation
    associated with the vortex instability (top) and spiral
    instability (bottom). The slices are taken at azimuths shown by
    white dotted lines in Fig. \ref{vortex4_vortex1_vortxy}. Negative
    perturbations in the region $r-r_p\in[2,5.5]r_h$ are outlined by
    white lines. \label{vortex4_vortex1_vortRZ}} 
\end{figure}

\subsection{Disc-planet torques}
The presence of non-axisymmetric disturbances at gap edges is
expected to significantly affect disc-planet torques. It has been 
confirmed in 2D simulations that both vortex and spiral modes lead to
oscillatory torques of either sign \citep{li05,lin11b}. 
%Limited 
%numerical resolution the current 3D calculations prevent accurate torque measurements, 
It this section we measure the disc-on-planet torques in several of the above simulations
to confirm the main features found in 2D.

We calculate the specific torque acting on the planet due to a mass
element as 
\begin{align}
d\bm{T}(\bm{r}) &=
\frac{\bm{r}_p\times\bm{r}G\rho(\bm{r})d^3\bm{r}}{d_p^3}f(\bm{r},\bm{r}_p),\\
f(\bm{r},\bm{r}_p)&\equiv 1
-\exp{\left(-\frac{1}{2}\left|\frac{\bm{r}-\bm{r}_p}{\epsilon_c
      r_h}\right|^2\right)}.  
\end{align} 
The tapering function $f$ reduce contributions from close to the
planet, thereby reducing numerical artifacts arising from this region
because of the diverging potential and limited resolution. We set  
the parameter $\epsilon_c=1$ so that tapering does not significantly
reduce contributions from the instabilities, since they develop
at $\gtrsim 2r_h$ away from the planet's orbital radius.

Fig. \ref{torque3} shows the disc-on-planet torques in Case 3 and Case 5, which develop
the $m=5$ and $m=6$ vortex modes, respectively. These plots are qualitatively similar to
2D simulations \citep[e.g.][]{li05}. The torques oscillate on orbital 
time-scales and its instantaneous values can be of either sign. However, 
upon averaging over the simulation we find the total torques are negative in both cases. This
means inwards migration is still favoured. 
%The planet is expected to interact

We extended Case 5 to $t=135P_0$ and find the vortices 
have similar over-densities as at $t=50P_0$. However, 
Fig. \ref{torque3} show the torque oscillation amplitudes decrease
towards the end of Case 5 compared to $t\in[40,80]P_0$.  
At $t=50P_0$ the vortices are located in $r-r_p\in[3.5,5.5]r_h$ but by
$t=135P_0$ they are located in $r-r_p\in[4,6]r_h$. 
Given that $t\in[40,80]P_0$ is only $20P_0$ to $50P_0$ after the planet  
potential has been fully introduced, gap-formation is probably 
ongoing during this time. We expect torque amplitudes to be larger
during  gap-formation since the vortices lie closer to the planet. 
%On the
%other hand, gap-widening in the presence of vortices was also seen in  
%2D simulations on long timescales \citep{lin11a}  
%So there may also be
%contributions from radial redistribution of material due to the
%vortices. 
% This might be
%brought about by the of the density bump at the outer gap edge
%\citep{meheut12b}

%The original
%density bump at the outer gap edge ($r-r_p\simeq 4r_h$) shifts outward
%between these times. 
%This is unlike in the fixed-orbit 2D simulation by \cite{li05}, where
%the oscillation amplitude does not decrease over several hundred
%orbits. 
%The different setups (especially self-gravity and numerical resolution) 
%prevent a detail comparison

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.cm 0cm
    0cm]{figures/vortex10_torque3_tqex}  
  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm 0.2cm]{figures/vortex2b_torque3_tqex} 
  \caption{Instantaneous disc-on-planet torques in simulations where the vortex mode develops. 
    Top: Case 3. Bottom: Case 5. Note that Case 5 has been extended to $t=135P_0$. \label{torque3}}
\end{figure}

Next we examine disc-planet torques in the presence of the spiral
modes. The top panel of Fig. \ref{torque3_spiral} shows the
instantaneous disc-on-planet torques. Contributions from the
inner disc ($r < r_p$) and outer disc ($r > r_p$) are plotted
separately for comparison with Fig. 18b in \cite{lin11b}, which is 
similar to the present plot. Large oscillations in the outer torque
due to edge mode spirals cause the total torque to be positive or
negative at a given instant.

Unlike the vortex modes, Fig. \ref{torque3_spiral} shows that spiral
modes can lead to a positive running-time averaged torques (bottom panel). 
The average torques become more positive with time after spiral modes develop, 
and with increasing self-gravity (which increases the instability strength). 
%The torque is more positive with
%increasing strength of the spiral instability.
This was also observed in high-resolution 2D simulations in
\cite{lin11b}. There it was suggested that the creation of large   
`voids' in between spiral arms decreases the time-averaged density in  
the planet-induced wakes, thereby reducing associated torque
magnitudes. Since the outer planetary wake normally provide a negative
torque, the spiral modes make the total torque more positive. 
The similarity between 2D and 3D results indicate that
outwards migration induced by spiral modes, which was seen in 2D by
\cite{lin11b,lin12b}, will also operate in 3D.

%It was later shown that the spiral instability can supply material to
%go on horseshoe turns ahead of the planet, which also provide a
%positive torque \citep{lin12b}. This is because the spiral modes 
%are associated with local vortensity maximum which are located just
%inside the gap edge in the co-orbital region.

%This is because the
%spirals are associated with local vortensity maximum, which is just
%inside the outer gap edge. Thus over-densities can be supplied by the
%instability to go on horseshoe orbits

%Since spiral modes are associated with disturbances at the local
%vortensity maximum just inside the outer gap edge, its influence on
%disc-planet torques is expected to be more si

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.825cm 0cm
    0cm]{figures/vortex1_torque2_tqex}  
  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm 0.2cm]{figures/torque3_tqex} 
  \caption{Disc-on-planet torques in the presence of spiral modes
    associated with the outer gap edge. Top: total torque (solid),
    torque from the inner disc (dotted) and from the outer disc
    (dashed) in Case 7. Bottom: time-averaged torques in Case 6
    (solid) and Case 7 (dotted).
\label{torque3_spiral}}
\end{figure}

### 2014-04-08 paper/model.tex
added sample input and paper files

\section{Disc-planet models}\label{model}
The system is a two-dimensional, self-gravitating gas disc of mass
$M_d$ rotating about a central star of mass $M_*$. Cylindrical
co-ordinates $(r,\phi,z)$ centred on the star is adopted. The frame is   
non-rotating. Computational units are such that
$G=M_*=\mathcal{R}=\mu=1$ where $G$ is the gravitational constant,
$\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight. %% \cite{lin12}

The disc is governed by the standard fluid equations  
\begin{align}\label{3d_gov_eq}
  &\frac{\p\Sigma}{\p t}+\nabla\cdot(\Sigma \bv)=0, \\
  & \frac{\p\bv}{\p t}+\bv\cdot\nabla\bv= -\frac{1}{\rho}\nabla p 
  - \nabla{\Phi} + \bm{f}_\nu,\\
  & \frac{\p e}{\p t} + \nabla\cdot(e\bv) = -p\nabla\cdot\bv +
  \mathcal{H} - \mathcal{C}, \\
  &\nabla^2\Phi_d = 4\pi G \Sigma \delta(z),\label{poisson}
\end{align}
where $\Sigma$ is the surface density, $\bv = (v_r,v_\phi)$ the fluid
velocity, $p$ is the pressure, $e=p/(\gamma-1)$ is the energy
density and the adiabatic index $\gamma=1.4$ is assumed constant.

The total potential $\Phi$ includes the stellar potential, planet potential
(described below), and the disc potential $\Phi_d$ which satisfies the
Poisson equation \ref{poisson}. Indirect potentials are also included
to account for the non-inertial reference frame. In the momentum
equations, $\bm{f}_\nu$ represent viscous forces whose magnitude is 
characterised by a constant kinematic viscosity parameter $\nu$.

\subsection{Heating and cooling}
In the energy equation, the heating term $\mathcal{H}$ is defined as 
\begin{align}
  \mathcal{H} \equiv Q^+ - Q^+_i\frac{\Sigma}{\Sigma_i}, 
\end{align}
where $Q^+$ represents viscous heating and subscript $i$ denotes
evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
\begin{align}
  \mathcal{C} \equiv \frac{1}{t_c}\left(e -
  e_i\frac{\Sigma}{\Sigma_i}\right),  
\end{align}
where $t_c = \beta\Omega_k^{-1}$ is the cooling time,
$\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
input parameter. Although idealised, this cooling prescription allows one 
to explore the full range of thermodynamic response of the disc in a 
systematic way.

Note that the energy source terms have been chosen to be absent 
at $t=0$. This allows the steady state energy equation to be satisfied
provided the flow is axisymmetric and $v_r$ is negligible compared to $v_\phi$.  
% If the disc is axisymmetric, steady and with only azimuthal flow, then the 
%energy equation is trivially satisfied. 
%This allows one to specify equilibrium discs (see below)
%in the same way as in the case where the energy equation is replaced by a fixed temperature 
%profile (i.e. $c_\mathrm{iso}^2\equiv p/\Sigma$ is prescribed). 
The cooling term attempts 
to restore the initial energy density (and therefore temperature) profile.

\subsection{Disc model and initial condition}
The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with $\rout =
25\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
surface density profile

\begin{align}\label{initial_density}
   \Sigma_i(r) = \Sigma_0\left(\frac{r}{r_\mathrm{in}}\right)^{-s}
    \left[1 - \sqrt{\frac{r_\mathrm{in}}{r + H_i(\rin)}}\,\right] 
\end{align}
where the power-law index $s=3/2$, $H(r) = \ciso\Omega_k $ defines the disc scale-height 
where $\ciso=\sqrt{p/\Sigma}$ isothermal sound-speed. The disc aspect-ratio is defined as $h\equiv H/r$ and initially
$h_i=0.05$. The surface density scale $\Sigma_0$ is set by  
\begin{align}
\Sigma_0 = \frac{h_i\rout\Omega_k^2(\rout)}{\pi G Q_o},
\end{align}
where $Q_o$ is the thin-disc Keplerian Toomre parameter at the outer boundary initially.

The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
pressure, stellar gravity and disc gravity. The initial radial velocity is $v_{r i}=3\nu/r$ 
with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi i}\simeq r\Omega_k$ 
and $\hat{\nu}\lesssim O(10^{-5})$, so that $|v_{r i}/v_{\phi i}|\ll1$ and the initial 
flow is effectively only in the azimuthal direction.

\subsection{Planet potential}\label{planet_config}
The planet potential is given by
\begin{align}
\Phi_p = -\frac{GM_p}{\sqrt{|\bm{r} - \bm{r}_p|^2 + \epsilon_p^2}},
\end{align}
where $M_p=qM_*$ is the planet mass ($q$ is the planet-to-star mass ratio), $\bm{r}_p=(r_p,\phi_p)$ is the planet's 
position in the disc and $\epsilon_p=0.5r_h$ is a softening length with $r_h=(q/3)^{1/3}r_p$ being the Hill radius. 
%Hill radius chosen so eps doesn't change with time (scale-height does)
In most simulations the planet is held on a fixed circular orbit, so that $r_p(t) = r_p(0) \equiv r_{p0}$. This also
defines the time unit $P_0\equiv 2\pi/\Omega_k(r_{p0})$ used to describe results.

In this study only one planet mass is considered, with $q=10^{-3}$. This corresponds to a Jupiter-mass
planet if $M_*=M_{\sun}$. The initial orbital radius is $r_{p0} = 10\rin$.

\section{Numerical experiments}\label{method}
The self-gravitating disc-planet system is evolved using the 
\texttt{FARGO-ADSG} code \citep{baruteau08, baruteau08b}. This is a modified version 
of the original \texttt{FARGO} code \citep{masset00a} to include the energy 
equation and disc gravity. The code employs finite-difference scheme similar 
to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
algorithm to circumvent the time-step restriction set by the inner disc boundary. 
The disc is divided into $(N_r,N_\phi)$ zones in the radial and azimuthal directions, 
respectively. The grid spacing is logarithmic in radius and uniform in azimuth.

\subsection{Control parameters}
The disc parameters are $Q_o$, which characterizes the strength of self-gravity; 
and $\beta$ which sets the cooling time. Two levels of self-gravity is considered: 
a light disc with $Q_o=8$ and a heavy disc with $Q_o=1.5$. The cooling 
parameter $\beta$ is chosen indirectly  through the parameter
$\tilde{\beta}$ such that

\begin{align}
  t_c(r_p+x_s) = \beta\Omega_k^{-1}(r_p+x_s) = \tilde{\beta} t_{\mathrm{lib}}(r_p+x_s), 
\end{align}
where $x_s$ is the distance from the planet, and $t_\mathrm{lib}$ is
the time interval between successive encounters of a fluid element and
the planet's azimuth. Assuming Keplerian orbital frequencies and $x_s\ll r_p$
gives $t_\mathrm{lib}\simeq 4\pi r_p/(3\Omega_{kp} x_s)$, where
$\Omega_{kp} = \Omega_k(r_p)$. Therefore   
\begin{align}\label{betatilde}
  \beta = \tilde{\beta} \frac{4\pi r_p}{3x_s} \left(1  - \frac{3x_s}{2r_p}\right), 
\end{align}
where $x_s\ll r_p$ was used again. We use $x_s = 2r_h$ in
Eq. \ref{betatilde}. For a planet mass with $q=10^{-3}$,
Eq. \ref{betatilde} then gives $\beta \simeq 23.9\tilde{\beta}$.

%giant planet gap is 2.5rh
%2rh is CR of G.E.I.

\subsection{Simulation procedure}
The disc is evolved without a planet for $t\in[0,20]P_0$. The planet potential is smoothly switched on
between $t\in[20,30]P_0$. In a separate set of calculations the planet
potential is turned off after $t>t_\mathrm{off}$, immediately  
afterwhich the surface density, energy density and velocity fields are
azimuthally averaged and then subject to random surface density  
perturbations of magnitude $10^{-2}$.

\subsection{Diagonistic measures}
%% The Toomre parameter is 
%% \begin{align}\label{toomre}
%% Q \equiv \frac{\kappa c_s}{\pi G \Sigma},
%% \end{align}
%% where $\kappa^2 = r^{-3}\p(rv_\phi)/\p r$ is the square of the epicyclic frequency and $c_s = \sqrt{\gamma} \ciso $ is the adiabatic sound-speed. 
%% Note that Eq. \ref{toomre} differs from the Keplerian Toomre parameter used to label the disc models. 
%The potential vorticity (PV), or vortensity, is 
%\begin{align}
%\eta \equiv \frac{\kappa^2}{2\Omega\Sigma}.
%\end{align}
%Note that $\eta$ is related to $Q$ through $\kappa$.

According to previous 2D studies, the key quantitiy in describing the
stability of radially structured non-barotropic discs is the
\emph{generalized vortensity} (GV): 
\begin{align}
  \tilde{\eta}\equiv \left(\frac{\kappa^2}{2\Omega\Sigma}\right)S^{-2/\gamma}, 
\end{align}
where $S \equiv p/\Sigma^\gamma$ is defined to be the entropy. The
quantity in brackets, $\eta\equiv \kappa^2/2\Omega\Sigma$, is the
usual vortensity or potential vorticity (PV).

%It will be convenient to analyse results in terms of perturbations with respect to a smooth background. Let 
%\begin{align}
%\Delta f = \frac{f-\avg{f}}{\avg{f}}
%\end{align}
%denote the perturbation of a flow quantity $f$

### 2014-04-08 paper/paper.tex
added sample input and paper files

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{epsfig,rotate,graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{subfigure}
\usepackage{lscape}
\usepackage{bm}

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle_\phi}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\DeclareMathOperator{\erf}{erf}

\title[Gaps in non-isothermal discs]{Gap formation and stability in 
  non-isothermal self-gravitating discs}

\author[Lin]{ Min-Kai Lin
  \thanks{E-mail: mklin924@cita.utoronto.ca} \\ 
Canadian Institute for Theoretical Astrophysics,  
60 St. George Street, Toronto, ON, M5S 3H8, Canada \\
}

\begin{document}

\maketitle
\begin{abstract}
Hydrodynamic simulations of self-gravitating disc-planet interactions are presented. The energy equation is solved with a simple cooling law 
such that $\beta = t_c\Omega_k$ is constant, where $t_c$ is the cooling time and $\Omega_k$ is the Keplerian frequency. 
The simulations are customized to examine the stability of gaps opened by massive planets as a function of $\beta$.

\end{abstract}

\begin{keywords}
planetary systems: formation --- planetary systems:
protoplanetary discs
\end{keywords}

\section{Introduction}\label{intro}

\input{model}
\input{results}
%\input{discussion}
%\input{summary}

%\section*{Acknowledgments}

\bibliographystyle{mn2e}
\bibliography{ref}

\appendix
%\input{appendix}
%\input{appendix2}

\end{document}

### 2014-04-08 paper/results.tex
added sample input and paper files

\section{Gaps in low mass viscous discs}

We first describe results for weakly self-gravitating
discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
gravitationally stable \citep{lin11b}. We also impose a dimensionless
kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
vortex instablity \citep{valborro07}. These parameter values
are typical for disc-planet simulations and yield stable/steady 
gap profiles. This will be useful reference cases to  
understand the unstable cases considered later.

Fig. \ref{lvisc_steady_gap} shows the steady state gap profiles in terms of the relative
surface density and aspect-ratio at $t=200P_0$. Three levels of cooling are considered:
$\tbeta=0.1,\,1.0$ and $10.0$ (i.e. $t_c\Omega_k\simeq2.4,\,24$ and $240$). We refer to them as
fast, moderate and slow cooling cases, respectively.

%we checked that open bc doesn't affect gap structure
%checked that t=200 plots are similar to t=100

%surface density plot

%aspect-ratio plot

%general vortensity plot

\begin{figure}
  \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/compare_profiles_dens020.ps}\\
  \includegraphics[scale=.42]{figures/compare_profiles_h020.ps}
  \caption{Steady-state gap profiles in a low mass viscous disc. The
    surface density perturbation (top) and disc aspect-ratio (bottom)
    are shown as a function of the cooling parameter:  
    $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
    moderate cooling) and $\tbeta=10$ (dashed, slow
    cooling). \label{lvisc_steady_gap}}  
\end{figure}

\begin{figure}
  \includegraphics[scale=.42]{figures/compare_profiles_gvort020.ps}
  \caption{Gap structure in a low mass viscous disc, in terms of the
    perturbed generalized vortensity as a function of the cooling
    parameter. \label{lvisc_steady_gvort}} 
\end{figure}

\section{Gaps in massive discs}
We first examine disc models with $Q_o=1.5$. We set the physical
viscosity $\hat{\nu}=10^{-9}$, so that the only energy source is
through shock-heating (via artificial viscosity) and the $\mathcal{C}$
function when $e\Sigma<e_i\Sigma_i$. The numerical resolution is
$N_r\times N_\phi = 512\times 1024$.

%Since we are primarily concerened with gap stability, it is important
%to first examine the gap structured opened by the planet as a function
%of cooling. 
Fig. \ref{gvort1d_q1d5} compares the gap structure in terms of the GV profile for three
levels of cooling: $\tilde{\beta}=0.1$, $\tilde{\beta}=1$ and
$\tbeta=10$. For convenience we will refer to these cases as fast,
intermediate and slow cooling, respectively. The snapshot is taken at
$t=30P_0$, just after the planet is fully introduced.

In terms of the GV profile, the qualitative features of the planetary
gap --- localized GV extrema at the gap edges --- remain unchanged
despite two orders of magnitude difference in the cooling rate.

\begin{figure}
%  \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/dens1d_q1d5_003_global.ps}\\
%  \includegraphics[scale=.42]{figures/gvort1d_q1d5_003_global.ps}
  \caption{Gap structure in terms of the perturbed surface density
    (top) and perturbed generalized
    vortensity (bottom), as a function of the cooling parameter:  
    $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
    intermediate cooling) and $\tbeta=10$ (dashed, slow
    cooling). \label{gvort1d_q1d5}} 
\end{figure}

\begin{figure*}
%  \includegraphics[scale=.55,clip=true,trim=0cm 0cm 0cm
%    0cm]{figures/noniso0_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
%    0cm]{figures/noniso1_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
%    0cm]{figures/noniso2_HR_dens004}
  \caption{Gap instability in the heavy disc model, as a function of
    the cooling parameter: $\tbeta=0.1$ (left), $\tbeta=1$ (middle)
    and $\tbeta=10$ (right). The relative surface density
    perturbation is shown. \label{polarxy_q1d5}} 
\end{figure*}

### 2014-04-08 paper/summary.tex
added sample input and paper files

\section{Summary and discussion}\label{summary}
We have performed customised numerical simulations of  
three-dimensional self-gravitating discs, in which an embedded 
satellite or planet has opened a gap. We explicitly verified in 
3D the main results on gap stability previously obtained from 
2D calculations \citep{koller03,li05,valborro07,meschiari08}, in
particular those by \cite{lin11a,lin11b}.

Planetary gaps are potentially unstable because of the existence of 
vortensity extrema generated by planet-induced shocks. Disc-satellite
interaction also occur in other systems such as stars in black hole
accretion discs \citep{kocsis11,mckernan11,baruteau11b}.  Furthermore, 
gaps opened by satellites are just one type of structured disc. 
Other examples 
include dead zone boundaries mentioned in \S\ref{intro} and
transition discs \citep{regaly12}.

Thus, although we were motivated by previous works on planetary gap 
stability, and hence considered disc-planet systems, we expect our
results to be applicable to discs with radial structure of other    
origin, provided the vortensity profiles involve stationary points and
therefore prone to the same instabilities.

\subsection{Confirmed 2D results}
We demonstrated the development of the vortex instability
at the outer gap edge opened by a giant planet in 3D 
discs. We began with a non-self-gravitating disc, in which a few vortices
develop then quickly merge. The quasi-steady state is a single
azimuthally extended vortex at the gap edge. This evolution is similar
to the 2D simulation in \cite{valborro07}.

We also showed the effect of self-gravity on the vortex
instability observed in 2D \citep{lyra09,lin11a} persists in 3D.
We observe more vortices as the strength of self-gravity is increased
by increasing the density scale. These vortices resist merging
on dynamical timescales. In our disc models with $Q_0=3$, the 
multi-vortex configuration lasts until the end of the
simulations.

As expected from 2D linear theory \citep{lin11b}, 
vortex modes are suppressed in our massive disc models. Instead, a 
global spiral instability develops which is associated with the local
vortensity maximum just inside the outer gap edge. These are distinct
from vortex modes since self-gravity is essential.

Our limited numerical resolution does not permit accurate disc-planet
torque measurements, but 
the qualitative effect of the vortex and spiral instabilities,
previously studied in 2D, have been reproduced in 3D --- oscillatory
torques of either sign and the tendency for spiral modes to provide on
average a positive torque. 
The similarity to 2D results is not surprising since for  
giant planets $r_h\gtrsim H$, so the razor-thin disc approximation is
expected to be valid as far as disc-planet interaction is
concerned. Furthermore, vertical self-gravity increases the midplane
density while reducing that in the atmosphere (Appendix
\ref{vertsg_mod}), so that given a fixed temperature profile the 2D
approximation is even better for self-gravitating discs.

 It is worth mentioning here that in shearing sheet
  simulations,  \cite{mamatsashvili09} found 
  self-gravity to favour vortices of smaller scale. They initialized
  a local patch of an unstructured disc with random velocity perturbations.
  Their gravito-turbulent discs are dominated by small vortices limited by
  the local Jeans length (which is smaller than the scale-height). Without self-gravity, 
  they merge to form larger vortices. 
  These observations are similar to the above results for the vortex mode.  
  However, the setups are quite different 
  as we consider radially structured, laminar global discs. Our
  large-scale vortices develop from a linear instability and have horizontal sizes
  of a few scale-heights. Thus, confirmation of the above results in 3D is only valid
  for the edge instabilities considered in this paper.

%torques (expected because giant planets are 'thicker' than disc)

\subsection{Effects of 3D}
In our simulations the dominant three-dimensional effect is
vertical self-gravity on the density field. In the
non-self-gravitating limit, the relative density perturbation
associated with a vortex is columnar with weak vertical dependence. This is
consistent with with 3D linear and nonlinear simulations
\citep{meheut10, meheut12a,  meheut12b, lin12}.

As the strength of self-gravity is increased, vortices become 
more vertically stratified --- they are condensed towards the
midplane. For moderately self-gravitating discs, the vortex
midplane density enhancement can be twice that near the upper
boundary.  The spiral modes display significant vertical structure
near the gap edge, while the density waves they launch in the outer
disc is columnar.  The latter is probably due to the chosen equation of state (see below).

The effect of vertical self-gravity on the vortex mode is seen even in
our least massive disc model with $Q_0=8$. One can consider an
initially smooth disc which is justified to be
non-self-gravitating. However, this approximation may become less good
with the creation of a vortensity minimum, because it
can also be a local minimum in the Toomre $Q$
(Eq. \ref{toomreq_vortensity}).  That is, the Toomre parameter is
decreased with the development of local radial structure (such as a density bump). 
The non-self-gravitating approximation worsens further when the vortex
instability associated with $\mathrm{min}(\eta)$ develops, because the   
vortices are regions of enhanced density, especially if they merge
into a single large vortex. Thus, the
non-self-gravitating approximation is not guaranteed to hold 
in the perturbed state even if it does in the initial disc.

It is interesting to note that linear calculations of the vortex
instability in vertically isothermal , non-self-gravitating discs 
show that the vertical velocity vanishes near the vortex
centroid  \citep{meheut12a,lin12}, but this is not observed in
nonlinear calculations \citep[][ and in the present
  simulations]{meheut12a}.  This contrasts to their anti-cyclonic
horizontal flow, which can be computed in linear theory and seen in
hydrodynamic simulations \citep{li00,li01}.  This suggests that vertical 
motions in this case may be associated with secondary processes
\citep[e.g][]{goodman87}.

Although we observe somewhat complicated vertical flow for both
vortex and spiral instabilities, the vertical Mach number is
at most a few per cent in magnitude. Also, the magnitude of
vertical motion is at most $\sim 20\%$  of the radial motion on
average. This suggests that the disturbances at the gap edge is 
roughly two-dimensional in the present disc models. This would be 
consistent with early studies which find instabilities associated with
co-rotation singularities are two-dimensional
\citep{papaloizou85,goldreich86}.  Recent 3D linear 
  calculations also find that, even with a vertical temperature gradient, 
  the vortex mode (without self-gravity) is
  largely 2D near the vortensity minimum \citep{lin12}.

\subsection{Caveats and outlooks}\label{caveats}

In order to provide 3D examples of previous 2D results, a range of 
disc models had to be simulated, each for many dynamical 
timescales with full self-gravity. To maintain reasonable
computational cost, numerical resolution in the $r\phi$ plane is much
reduced compared to razor-thin disc simulations ($\sim  4$---6 cells
per $H$ compared to $\sim 16$ in 2D).  Despite this, our
plots in the $r\phi$ plane closely resemble those obtained from 2D
simulations.

On the other hand, the low resolutions adopted here are unlikely to
capture elliptic instabilities which may destroy 
vortices in 3D \citep{lesur09b,lesur10}. 
This might not be a serious
issue when the condition for the vortex instability is maintained 
(by a planet in the present context). Also, the vortex 
grows on dynamical timescales whereas the elliptic instability takes
much longer \citep{lesur09b}. The initial development of vortices is not expected to be 
suppressed by the elliptic instability \citep[as found
  by][]{meheut12a}.

The vortex instability produces smaller vortices with increasing
self-gravity. According to \cite{lesur09b}, vortices with small
aspect-ratios ($\lesssim 4$) are strongly unstable in 3D, but note that their  
model is a local patch of a smooth disc without self-gravity, 
whereas we considered a gap edge in a global self-gravitating disc.

Inclusion of self-gravity may change the stability properties of
vortices. In particular, we found that a vortex can flatten somewhat under its own
weight. \cite{lithwick09} suggested vertical gravity helps to 
stabilise vortices in 3D. Vertical \emph{self}-gravity 
can enhance this effect.  \citeauthor{lithwick09} 
found in local 3D simulations that `tall' vortices are unstable 
whereas `short' vortices survive as in 2D \citep{godon99}. 
We may expect the more stratified vortices formed in self-gravitating 
discs to be more stable than those in non-self-gravitating discs \citep{barranco05}. 
The elliptic instability is also weakened by stratification \citep{lesur09b}, 
so vertical self-gravity may also be stabilising in this respect.

%Improved numerical models, including the energy equation, should be 
%simulated to explore these speculations. 
% 
%vortex instability on dynmical timescales. elliptic instability is
%weaker growth

We have adopted the locally isothermal EOS for simplicity and direct
comparison with previous 2D results. This EOS limits the 3D 
structure of density waves compared to thermally 
stratified discs \citep[e.g][]{lin90,lubow98,ogilvie99}, which can 
cause refraction of waves out of the midplane.  
The locally isothermal EOS represents the limit of efficient cooling
\citep{boss98}. This might apply in optically thin
regions of a disk\footnote{For example, \cite{cossins10} find optical
depths $\tau < 0.2$ beyond $\sim100\mathrm{AU}$ in their models of protoplanetary
discs.}, but is violated if high densities develop, such as 
self-gravitating clumps \citep{pickett00}. Our edge instabilities only 
reach moderate over-densities. Nevertheless, enhanced
vertical stratification of the edge disturbances observed in our
simulations are likely exaggerated by the EOS. Clearly, it is necessary to
extend models of edge instabilities in self-gravitating discs to
include an appropriate energy equation.

Another important issue is vertical boundary conditions. We simulated
a thin disc and imposed a reflecting upper boundary to prevent mass
loss from above. This setup  may enhance the two-dimensionality
of the problem. The vortex instability tends to involve the 
entire column of fluid, especially in the weak self-gravity regime. It
is therefore a global instability in $z$. We suspect the spiral mode
is less affected by vertical boundary conditions because the
instability tends to concentrate material at the midplane. Future
simulations will consider varying the vertical domain size and upper
disc boundary conditions.

Our torque measurements indicate migration will be significantly
affected by the vortex and spiral instabilities. Preliminary 3D
simulations with a freely migrating planet have been performed. We
recover the vortex-planet scattering and the spiral-induced outward
migration described by \cite{lin10,lin12b}.   
The disc-planet torque is determined by the density field, and
the above instabilities have density perturbations that either
have weak vertical dependence or concentrated at the midplane. Thus, we
believe that at present, 2D simulations are more advantageous for studies
focusing on migration, because high resolution is feasible and needed.

### 2014-05-30 paper/model.tex
tex updates

@@
 The disc is evolved without a planet for $t\in[0,20]P_0$. The planet potential is smoothly switched on
-between $t\in[20,30]P_0$. In a separate set of calculations the planet
-potential is turned off after $t>t_\mathrm{off}$, immediately  
-afterwhich the surface density, energy density and velocity fields are
-azimuthally averaged and then subject to random surface density  
-perturbations of magnitude $10^{-2}$.
+between $t\in[20,30]P_0$.
+%%  In a separate set of calculations the planet
+%% potential is turned off after $t>t_\mathrm{off}$, immediately  
+%% afterwhich the surface density, energy density and velocity fields are
+%% azimuthally averaged and then subject to random surface density  
+%% perturbations of magnitude $10^{-2}$.
 
 \subsection{Diagonistic measures}
-%% The Toomre parameter is 
-%% \begin{align}\label{toomre}
-%% Q \equiv \frac{\kappa c_s}{\pi G \Sigma},
+%% According to previous 2D studies, the key quantitiy in describing the
+%% stability of radially structured non-barotropic discs is the
+%% \emph{generalized vortensity} (GV): 
+%% \begin{align}
+%%   \tilde{\eta}\equiv \left(\frac{\kappa^2}{2\Omega\Sigma}\right)S^{-2/\gamma}, 
 %% \end{align}
-%% where $\kappa^2 = r^{-3}\p(rv_\phi)/\p r$ is the square of the epicyclic frequency and $c_s = \sqrt{\gamma} \ciso $ is the adiabatic sound-speed. 
-%% Note that Eq. \ref{toomre} differs from the Keplerian Toomre parameter used to label the disc models. 
-%The potential vorticity (PV), or vortensity, is 
-%\begin{align}
-%\eta \equiv \frac{\kappa^2}{2\Omega\Sigma}.
-%\end{align}
-%Note that $\eta$ is related to $Q$ through $\kappa$.
-
-According to previous 2D studies, the key quantitiy in describing the
-stability of radially structured non-barotropic discs is the
-\emph{generalized vortensity} (GV): 
-\begin{align}
-  \tilde{\eta}\equiv \left(\frac{\kappa^2}{2\Omega\Sigma}\right)S^{-2/\gamma}, 
-\end{align}
-where $S \equiv p/\Sigma^\gamma$ is defined to be the entropy. The
-quantity in brackets, $\eta\equiv \kappa^2/2\Omega\Sigma$, is the
-usual vortensity or potential vorticity (PV).
-
-%It will be convenient to analyse results in terms of perturbations with respect to a smooth background. Let 
-%\begin{align}
-%\Delta f = \frac{f-\avg{f}}{\avg{f}}
-%\end{align}
-%denote the perturbation of a flow quantity $f$
+%% where $S \equiv p/\Sigma^\gamma$ is defined to be the entropy. The
+%% quantity in brackets, $\eta\equiv \kappa^2/2\Omega\Sigma$, is the
+%% usual vortensity or potential vorticity (PV).

### 2014-05-30 paper/results.tex
tex updates

@@
-\section{Gaps in low mass viscous discs}
+\section{Gaps in low mass discs (RL)}
 
-We first describe results for weakly self-gravitating
-discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
-gravitationally stable \citep{lin11b}. We also impose a dimensionless
-kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
-vortex instablity \citep{valborro07}. These parameter values
-are typical for disc-planet simulations and yield stable/steady 
-gap profiles. This will be useful reference cases to  
-understand the unstable cases considered later.
+%% We first describe results for weakly self-gravitating
+%% discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
+%% gravitationally stable \citep{lin11b}. We also impose a dimensionless
+%% kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
+%% vortex instablity \citep{valborro07}. These parameter values
+%% are typical for disc-planet simulations and yield stable/steady 
+%% gap profiles. This will be useful reference cases to  
+%% understand the unstable cases considered later.
 
-Fig. \ref{lvisc_steady_gap} shows the steady state gap profiles in terms of the relative
-surface density and aspect-ratio at $t=200P_0$. Three levels of cooling are considered:
-$\tbeta=0.1,\,1.0$ and $10.0$ (i.e. $t_c\Omega_k\simeq2.4,\,24$ and $240$). We refer to them as
-fast, moderate and slow cooling cases, respectively.
+%% Fig. \ref{lvisc_steady_gap} shows the steady state gap profiles in terms of the relative
+%% surface density and aspect-ratio at $t=200P_0$. Three levels of cooling are considered:
+%% $\tbeta=0.1,\,1.0$ and $10.0$ (i.e. $t_c\Omega_k\simeq2.4,\,24$ and $240$). We refer to them as
+%% fast, moderate and slow cooling cases, respectively.
 
@@
 
-\begin{figure}
-  \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/compare_profiles_dens020.ps}\\
-  \includegraphics[scale=.42]{figures/compare_profiles_h020.ps}
-  \caption{Steady-state gap profiles in a low mass viscous disc. The
-    surface density perturbation (top) and disc aspect-ratio (bottom)
-    are shown as a function of the cooling parameter:  
-    $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
-    moderate cooling) and $\tbeta=10$ (dashed, slow
-    cooling). \label{lvisc_steady_gap}}  
-\end{figure}
+%% \begin{figure}
+%%   \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/compare_profiles_dens020.ps}\\
+%%   \includegraphics[scale=.42]{figures/compare_profiles_h020.ps}
+%%   \caption{Steady-state gap profiles in a low mass viscous disc. The
+%%     surface density perturbation (top) and disc aspect-ratio (bottom)
+%%     are shown as a function of the cooling parameter:  
+%%     $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
+%%     moderate cooling) and $\tbeta=10$ (dashed, slow
+%%     cooling). \label{lvisc_steady_gap}}  
+%% \end{figure}
 
-\begin{figure}
-  \includegraphics[scale=.42]{figures/compare_profiles_gvort020.ps}
-  \caption{Gap structure in a low mass viscous disc, in terms of the
-    perturbed generalized vortensity as a function of the cooling
-    parameter. \label{lvisc_steady_gvort}} 
-\end{figure}
+%% \begin{figure}
+%%   \includegraphics[scale=.42]{figures/compare_profiles_gvort020.ps}
+%%   \caption{Gap structure in a low mass viscous disc, in terms of the
+%%     perturbed generalized vortensity as a function of the cooling
+%%     parameter. \label{lvisc_steady_gvort}} 
+%% \end{figure}
 
-\section{Gaps in massive discs}
-We first examine disc models with $Q_o=1.5$. We set the physical
-viscosity $\hat{\nu}=10^{-9}$, so that the only energy source is
-through shock-heating (via artificial viscosity) and the $\mathcal{C}$
-function when $e\Sigma<e_i\Sigma_i$. The numerical resolution is
-$N_r\times N_\phi = 512\times 1024$.
+\section{Gaps in massive discs (MKL)}
+%% We first examine disc models with $Q_o=1.5$. We set the physical
+%% viscosity $\hat{\nu}=10^{-9}$, so that the only energy source is
+%% through shock-heating (via artificial viscosity) and the $\mathcal{C}$
+%% function when $e\Sigma<e_i\Sigma_i$. The numerical resolution is
+%% $N_r\times N_\phi = 512\times 1024$.
 
-%Since we are primarily concerened with gap stability, it is important
-%to first examine the gap structured opened by the planet as a function
-%of cooling. 
-Fig. \ref{gvort1d_q1d5} compares the gap structure in terms of the GV profile for three
-levels of cooling: $\tilde{\beta}=0.1$, $\tilde{\beta}=1$ and
-$\tbeta=10$. For convenience we will refer to these cases as fast,
-intermediate and slow cooling, respectively. The snapshot is taken at
-$t=30P_0$, just after the planet is fully introduced.
+%% %Since we are primarily concerened with gap stability, it is important
+%% %to first examine the gap structured opened by the planet as a function
+%% %of cooling. 
+%% Fig. \ref{gvort1d_q1d5} compares the gap structure in terms of the GV profile for three
+%% levels of cooling: $\tilde{\beta}=0.1$, $\tilde{\beta}=1$ and
+%% $\tbeta=10$. For convenience we will refer to these cases as fast,
+%% intermediate and slow cooling, respectively. The snapshot is taken at
+%% $t=30P_0$, just after the planet is fully introduced.
 
-In terms of the GV profile, the qualitative features of the planetary
-gap --- localized GV extrema at the gap edges --- remain unchanged
-despite two orders of magnitude difference in the cooling rate.
+%% In terms of the GV profile, the qualitative features of the planetary
+%% gap --- localized GV extrema at the gap edges --- remain unchanged
+%% despite two orders of magnitude difference in the cooling rate.
 
-\begin{figure}
-%  \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/dens1d_q1d5_003_global.ps}\\
-%  \includegraphics[scale=.42]{figures/gvort1d_q1d5_003_global.ps}
-  \caption{Gap structure in terms of the perturbed surface density
-    (top) and perturbed generalized
-    vortensity (bottom), as a function of the cooling parameter:  
-    $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
-    intermediate cooling) and $\tbeta=10$ (dashed, slow
-    cooling). \label{gvort1d_q1d5}} 
-\end{figure}
+%% \begin{figure}
+%% %  \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/dens1d_q1d5_003_global.ps}\\
+%% %  \includegraphics[scale=.42]{figures/gvort1d_q1d5_003_global.ps}
+%%   \caption{Gap structure in terms of the perturbed surface density
+%%     (top) and perturbed generalized
+%%     vortensity (bottom), as a function of the cooling parameter:  
+%%     $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
+%%     intermediate cooling) and $\tbeta=10$ (dashed, slow
+%%     cooling). \label{gvort1d_q1d5}} 
+%% \end{figure}
 
-\begin{figure*}
-%  \includegraphics[scale=.55,clip=true,trim=0cm 0cm 0cm
-%    0cm]{figures/noniso0_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
-%    0cm]{figures/noniso1_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
-%    0cm]{figures/noniso2_HR_dens004}
-  \caption{Gap instability in the heavy disc model, as a function of
-    the cooling parameter: $\tbeta=0.1$ (left), $\tbeta=1$ (middle)
-    and $\tbeta=10$ (right). The relative surface density
-    perturbation is shown. \label{polarxy_q1d5}} 
-\end{figure*}
+%% \begin{figure*}
+%% %  \includegraphics[scale=.55,clip=true,trim=0cm 0cm 0cm
+%% %    0cm]{figures/noniso0_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
+%% %    0cm]{figures/noniso1_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
+%% %    0cm]{figures/noniso2_HR_dens004}
+%%   \caption{Gap instability in the heavy disc model, as a function of
+%%     the cooling parameter: $\tbeta=0.1$ (left), $\tbeta=1$ (middle)
+%%     and $\tbeta=10$ (right). The relative surface density
+%%     perturbation is shown. \label{polarxy_q1d5}} 
+%% \end{figure*}

### 2014-06-16 paper/results.tex
Added preliminary findings of longterm behaviour and short term as well as figures

@@
 %% fast, moderate and slow cooling cases, respectively.
+
+Long term simulation of vortices were done for $\tilde{\beta}=0.1,1.0$ and
+ $\hat{\nu}=10^{-6},10^{-8},10^{-9}$ up to a total of 2000 orbits. Values
+ of higher $\tilde{\beta}$ resulted generally in longer vortex lifetimes.
+ The inviscid cases also were characterized by a sudden death of vortices
+ while the higher viscosity case showed a gradual decrease in mode amplitudes
+ attributed to the vortices.
+
+Short term simulations of the linear phase of instability growth also distinguished
+ the different effects of cooling rates. Simulations with higher
+ $\tilde{\beta}$ were characterized by slower growth rates centered at the
+ $m=2$ mode, lower density pertibations and vortex locations, and a more unitary aspect
+ ratio for the vortices.
+
+ \begin{figure}
+   \includegraphics[scale=.42]{figures/stability_vis9betalow.ps}
+   \caption{Mode intensity relative to $m=0$ for $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=0.1$.}
+ \label{stability_vis9lowb}
+ \end{figure}
+
+\begin{figure}
+   \includegraphics[scale=.42]{figures/stability_vis9betamed.ps}
+   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=1.0$. }
+ \label{stability_vis9medb)}
+ \end{figure}
+
+\begin{figure}
+   \includegraphics[scale=.42]{figures/stability_vis6betalow.ps}
+   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=0.1$. }
+ \label{stability_vis6lowb)}
+ \end{figure}
+
+\begin{figure}
+   \includegraphics[scale=.42]{figures/stability_vis6betamed.ps}
+   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$. }
+ \label{stability_vis6medb)}
+ \end{figure}
+Same as \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$. 
+\begin{figure}
+   \includegraphics[scale=.60]{figures/analysis_sigma50lowb.ps}
+   \caption{Density pertibation of disk at time=$50t_p$ with $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=0.1$. }
+ \label{shortterm_lowb)}
+ \end{figure}
+
+\begin{figure}
+   \includegraphics[scale=.60]{figures/analysis_sigma50highb.ps}
+   \caption{Density pertibation of disk at time=$50t_p$ with $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=10.0$. }
+ \label{shortterm_highb)}
+ \end{figure}
 

### 2014-06-23 paper/results.tex
brief notes on what information should be included

@@
 
+\subsection{Gap structure}
+{\bf plots of the gap structure before instability sets in, e.g. at
+  t=30. surface density perturbation, gen. vortensity perturbation,
+  aspect-ratio. compare diff. beta
+}
+
+\subsection{Linear stability}
+{\bf present `planet-off' simulations. compare growthrate and dominant
+  m as function of beta. although focus is linear stability
+  properties, can also show non-linear evolution for completeness (to
+  be compared with planet-on sims. later)
+}
+
+\subsection{Non-linear evolution}
+           {\bf describe a few planet-on sims. as func. of beta. lifetime. 
+           }
 %we checked that open bc doesn't affect gap structure
@@
 
-\section{Gaps in massive discs (MKL)}
+%%\section{Gaps in massive discs (MKL)}
 %% We first examine disc models with $Q_o=1.5$. We set the physical

### 2014-07-03 poster/poster.tex
poster template setup

\documentclass[final,hyperref={pdfpagelabels=false}]{beamer}
\usepackage{grffile}
\mode<presentation>{\usetheme{I6pd2}}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{amsmath,amsthm, amssymb, latexsym}
%\usepackage{times}\usefonttheme{professionalfonts}  % obsolete
%\usefonttheme[onlymath]{serif}
\boldmath
\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}
\usepackage{ragged2e} 
\usepackage{bm}
% change list indention level
% \setdefaultleftmargin{3em}{}{}{}{}{}

%\usepackage{snapshot} % will write a .dep file with all dependencies, allows for easy bundling

\usepackage{array,booktabs,tabularx}
\newcolumntype{Z}{>{\centering\arraybackslash}X} % centered tabularx columns
\newcommand{\pphantom}{\textcolor{ta3aluminium}} % phantom introduces a vertical space in p formatted table columns??!!
\newcommand{\p}{\partial}
\newcommand{\bu}{\bm{u}}
\newcommand{\bv}{\bm{v}}

\listfiles

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{figures/}}

\title{\huge Gap formation and stability in non-isothermal disks}
\author{Min-Kai Lin and Robert Les}
\institute[CITA]{Canadian Institute for Theoretical Astrophysics, 60
  St George Street, Toronto, M5S 3H8, Canada}
%% \date[Sep. 8th, 2009]{Sep. 8th, 2009}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newlength{\columnheight}
\setlength{\columnheight}{105cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{frame}
  \begin{columns}
    % ---------------------------------------------------------%
    % Set up a column 
    \begin{column}{.49\textwidth}
      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
        \begin{minipage}[T]{.95\textwidth}  % tweaks the width, makes a new \textwidth
          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
            % ---------------------------------------------------------%
            % fill each column with content            
            \begin{block}{{\Large Introduction}}
              \justifying
                  {\large              
                    {\bf
                      stuff
                    }
                  }
            \end{block}
            \vfill

            \begin{block}{{\Large The generalized potential vorticity}}
              \justifying

            \end{block}

            \begin{block}{{\Large Disk-planet model with simple cooling}}

            \end{block}
            \vfill

            \begin{block}{{\Large Linear stability}}
              \justifying

            \end{block}

            \vfill            
            \begin{block}{\Large{Non-linear vortex evolution}}
              \justifying

            \end{block}
          }
        \end{minipage}
      \end{beamercolorbox}
    \end{column}
    % ---------------------------------------------------------%
    % end the column

    % ---------------------------------------------------------%
    % Set up a column 
    \begin{column}{.49\textwidth}
      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
            % ---------------------------------------------------------%
            % fill each column with content

            \begin{block}{{\Large Conclusions so far}}\justifying
              {\large{\bf

                }
              }
            \end{block}
          }
          % ---------------------------------------------------------%
          % end the column
        \end{minipage}
      \end{beamercolorbox}
    \end{column}
    % ---------------------------------------------------------%
    % end the column
  \end{columns}
  \vskip1ex
  %\tiny\hfill\textcolor{ta2gray}{Created with \LaTeX \texttt{beamerposter}  \url{http://www-i6.informatik.rwth-aachen.de/~dreuw/latexbeamerposter.php}}
  %% \tiny\hfill{Created with \LaTeX \texttt{beamerposter}  \url{http://www-i6.informatik.rwth-aachen.de/~dreuw/latexbeamerposter.php} \hskip1em}
\end{frame}
\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Local Variables: 
%%% mode: latex
%%% TeX-PDF-mode: t
%%% End:

### 2014-07-03 poster/makepdf.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+latex poster.tex
+dvips -o poster.ps poster.dvi
+ps2pdf -sPAPERSIZE=a0 poster.ps
+echo done

### 2014-07-03 poster/poster.tex
updated poster tex

@@
 
-\title{\huge Gap formation and stability in non-isothermal disks}
+\title{\huge Gap formation and stability in non-isothermal
+  disks\\ \Large [2014 CITA Summer Student Program]}
 \author{Min-Kai Lin and Robert Les}
@@
                     {\bf
-                      stuff
                     }
@@
             \vfill
-
-            \begin{block}{{\Large The generalized potential vorticity}}
-              \justifying
-
-            \end{block}
 
@@
 
-            \begin{block}{{\Large Linear stability}}
+            \begin{block}{{\Large The generalized potential vorticity}}
               \justifying
-
+              {\bf generalized vortensity profile due to planet for
+                different beta. also density profiles.}
             \end{block}
@@
             \vfill
-            \begin{block}{\Large{Non-linear vortex evolution}}
+
+            \begin{block}{{\Large Linear stability}}
               \justifying
-
+                  {\bf plot of growth rate v.s m for different beta,
+                    or table. 2D density plot showing development of
+                    vortex, for different beta}
             \end{block}
+            \vfill
           }
@@
 
+            \begin{block}{\Large{Non-linear vortex evolution}}
+              \justifying
+                  {\bf `lifetime' plot of m=1 for diff beta, snapshots
+                    (before, during, after vortex decay). quote rossby
+                  number when vortices in quasi-steady state}
+            \end{block}
+            \vfill
             \begin{block}{{\Large Conclusions so far}}\justifying
@@
               }
+              \vfill
             \end{block}

### 2014-07-21 poster/poster.tex
Added plot files and updated poster
Added comments describing current results
poster-making

@@
 \usepackage{array,booktabs,tabularx}
-\newcolumntype{Z}{>{\centering\arraybackslash}X} % centered tabularx columns
+\newcolumntype{R}{>{\centering\arraybackslash}X} % right justified tabularx columns
 \newcommand{\pphantom}{\textcolor{ta3aluminium}} % phantom introduces a vertical space in p formatted table columns??!!
@@
 \setlength{\columnheight}{105cm}
+\setbeamertemplate{caption}{\centering\insertcaption\par}
 
@@
   \begin{columns}
+
+    %Aspect ratio: For higher $\tilde{\beta}$ values the disk aspect ratio increases non-monotonicly.
+Values for $\tilde{\beta}=0.1,1.0,10.0$ are $h=0.05,0.055,0.08$.
+The values are not constant over time and make roughly parabolic shapes from intial planet formation to vortex death.
+The aspect ratio of the vortices themselves seem to be constantly near $h=0.05$ regardless of the cooling paramter.
+Also $\tilde{\beta}=5.0$ which has longest lifetime (2000+ orbits) corresponds to approximately $h=0.06$ which is similar result as Fu et al.
+
+    %Mach Number: Mach numbers generally in the vortex region of disk averaged aziumthally are $<0.4$ during semi-stable vortex state.
+Just as vortex dies out mach numbers drop to $M=0.2$ which are from the shock.
+Mach number near vortices during semi-stable state are near 0.2 from 2D plots.
+
+    %Reynold's Stress:The averaged reynold's stress of the disk is usually of order $\alpha=10^{-2} - 10^{-3}$.
+During vortex death $\alpha=10^{-2}$ consistantly for all $\tilde{\beta}$ than drops to lower bound after vortex dissipation.
+Vortices themselves seem to have almost no characteristic reynolds stress associated with them from 2D plots.
+
+    %Gap profile:After vortex death sees a distortion in outer gap depth and the outer gap edge slowly and continuously extends outwards.
+Gap edge in semi-steady state is around 5 hill radii (gap edges not very clear since current resolution isn't good).
+Changes in gap depth start before vortex dissaption and indicate that matter is moving inwards towards the planet.
+Supports notion that vortex distorts background profile.
+This effect increases largely during the dissipation phase
+
+    %Vortex:The overdensiy of the vortex itself can reach 10 times the intial density of the disk.
+Vortex evolves from a large vortex to a highly dense small vortex.
+As it dies off the vortex distorts in shape most probably from interaction with shocks.Disapation time is around 50 orbits.
+
+    %Rossby Numbers:Rossby numbers during semi-stable states are approximately around $Ro=-0.2$.
+Just after vortex amplitude drop, Rossby numbers increase to around $Ro=-0.4$ because of spreading of angular momentum from the once densly packed spinning vortex.
+
+    %Non-linear Planet off:Long Term planet-off simulations do not show these drops in vortex amplitude.
+The vortices are of order of magnitude less in max amplitude and over time slowly decay assumingly by viscous friction.
+This supports fact that planet helps drive the interactions we are seeing.
+Results in progress.
+
     % ---------------------------------------------------------%
@@
             \begin{block}{{\Large Disk-planet model with simple cooling}}
-
+             We numerically evolve the 2D fluid equations including a gap-opening giant planet, treated as an external potential on fixed circular orbit. We include in the energy equation
+        a simple source term in the form
+	\begin{align*}
+	\frac{\partial e}{\partial t} = \cdots + \frac{e-e_{t=0}}{t_\mathrm{cool}},
+	\end{align*}
+        and $t_\mathrm{cool} = \beta \Omega_k^{-1}$, where $\Omega_k$ is the Keplerian frequency. We choose $\beta$ indirectly 
+        through $\tilde{\beta}$ such that
+        \begin{align*}
+        t_\mathrm{cool, gap\,edge} = \tilde{\beta} t_\mathrm{lib, gap\,edge},
+        \end{align*}
+        where $t_\mathrm{lib, gap\,edge}$ is the time interval between successive encounters of a fluid element, 
+        located at the gap edge, and the planet's azimuth. Thus $\tilde{\beta}\ll 1$ implies that such a fluid returns to its initial temperature
+        before passing by the planet again. 
         \end{block}
@@
                     different beta. also density profiles.}
+
+                  \begin{figure}
+                    \centering
+                    \hfill
+                    \begin{minipage}{0.49\textwidth}
+                      \includegraphics[width=\textwidth]{Posterfig_Density}
+                    \end{minipage}
+                    \hfill
+                    \begin{minipage}{0.49\textwidth}
+                      \includegraphics[width=\textwidth]{Posterfig_Vortensity}
+                    \end{minipage}
+                    \hfill
+                  \end{figure}
+
+                  \begin{figure}
+                    \centering
+                    \includegraphics[width=0.5\textwidth]{Posterfig_HR}
+                  \end{figure}
+
             \end{block}
@@
                     vortex, for different beta}
+
+                   \begin{figure}
+                    \centering
+                    \hfill
+                    \begin{minipage}{0.45\textwidth}
+                      \includegraphics[width=\textwidth]{Posterfig_lowb}
+                    \end{minipage}
+                    \hfill
+                    \begin{minipage}{0.45\textwidth}
+                      \includegraphics[width=\textwidth]{Posterfig_medb}
+                    \end{minipage}
+                    \hfill
+                  \end{figure}
+
+                   \begin{table}
+                     \hfill
+                     \begin{minipage}{\textwidth}
+                       \begin{tabularx}{\textwidth}{l*{10}{R}} %\toprule \addlinespace[10pt]
+                         \multicolumn{11}{c}{$\tilde{\beta}=0.1$} \\ \midrule
+                         m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
+                         $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
+                       \end{tabularx}
+                     \end{minipage}
+
+                     \vspace{1in}
+
+                     \begin{minipage}{\textwidth}
+                       \begin{tabularx}{\textwidth}{l*{5}{R}} %\toprule \addlinespace[10pt]
+                         \multicolumn{6}{c}{$\tilde{\beta}=1.0$} \\ \midrule
+                         m                    & 1 & 2 & 3 & 4 & 5  \\ 
+                         $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
+                       \end{tabularx}
+                     \end{minipage}
+                     \hfill
+                   \end{table}
+
             \end{block}
@@
 
+            \begin{block}{\Large{Linear stability Cont.}}
+              \justifying
+              \begin{figure}
+                \centering
+                \includegraphics[width=\textwidth]{Posterfig_Growth}
+              \end{figure}
+            \end{block}
+
+            \vfill
+
             \begin{block}{\Large{Non-linear vortex evolution}}
@@
                   number when vortices in quasi-steady state}
+
+                  \begin{figure}
+                    \centering
+                    \includegraphics[width=\textwidth]{Posterfig_Lifetime_b}
+                  \end{figure}
+
+                  \begin{figure}
+                    \centering
+                    \hfill
+                    \begin{minipage}{0.3\textwidth}
+                      \includegraphics[width=\textwidth]{Posterfig_Before}
+                      \caption{$Ro=-0.16$}
+                    \end{minipage}
+                    \hfill
+                    \begin{minipage}{0.3\textwidth}
+                      \includegraphics[width=\textwidth]{Posterfig_During}
+                      \caption{$Ro=-0.43$}
+                    \end{minipage}
+                    \hfill
+                    \begin{minipage}{0.3\textwidth}
+                      \includegraphics[width=\textwidth]{Posterfig_After}
+                      \caption{$Ro=-0.10$}
+                    \end{minipage}
+                    \hfill
+                  \end{figure}
+
             \end{block}
             \vfill
+
             \begin{block}{{\Large Conclusions so far}}\justifying

### 2014-07-22 poster/poster.tex
minor update to tex

@@
 
-            \begin{block}{{\Large Disk-planet model with simple cooling}}
-             We numerically evolve the 2D fluid equations including a gap-opening giant planet, treated as an external potential on fixed circular orbit. We include in the energy equation
+            \begin{block}{{\Large Disk-planet model with simple
+                  cooling}}
+              \justifying
+             We numerically evolve the 2D fluid equations including a
+             gap-opening giant planet, treated as an external
+             potential on fixed circular orbit. We include in the
+             energy equation  
         a simple source term in the form 
@@
 
-            \begin{block}{{\Large The generalized potential vorticity}}
+        \begin{block}{{\Large Planet gaps as a function of cooling
+              rate}}
           \justifying
-                  {\bf generalized vortensity profile due to planet for
-                    different beta. also density profiles.}
+          % {\bf generalized vortensity profile due to planet for
+          %   different beta. also density profiles.}
 
@@
             \centering
-                    \hfill
-                    \begin{minipage}{0.49\textwidth}
-                      \includegraphics[width=\textwidth]{Posterfig_Density}
-                    \end{minipage}
-                    \hfill
-                    \begin{minipage}{0.49\textwidth}
-                      \includegraphics[width=\textwidth]{Posterfig_Vortensity}
-                    \end{minipage}
-                    \hfill
+%            \hfill
+%            \begin{minipage}{0.49\textwidth}
+              \includegraphics[width=.49\textwidth]{Posterfig_Density}\\
+%           \end{minipage}
+%            \hfill
+%            \begin{minipage}{0.49\textwidth}
+              \includegraphics[width=.49\textwidth]{Posterfig_Vortensity}
+%            \end{minipage}
+%            \hfill
           \end{figure}
 
-                  \begin{figure}
-                    \centering
-                    \includegraphics[width=0.5\textwidth]{Posterfig_HR}
-                  \end{figure}
+          %\begin{figure}
+          %  \centering
+          %  \includegraphics[width=0.5\textwidth]{Posterfig_HR}
+          %\end{figure}
 

### 2014-07-23 poster/poster.tex
poster making, 70%

@@
   disks\\ \Large [2014 CITA Summer Student Program]}
-\author{Min-Kai Lin and Robert Les}
+\author{Min-Kai Lin and Robert Les$^\dagger$}
 \institute[CITA]{Canadian Institute for Theoretical Astrophysics, 60
-  St George Street, Toronto, M5S 3H8, Canada}
+  St George Street, Toronto, M5S 3H8, Canada\\{\footnotesize$^\dagger$student: all
+  simulations, data analysis and figures}}
+
 %% \date[Sep. 8th, 2009]{Sep. 8th, 2009}
@@
                     {\bf
+                      Unstable disk-planet, leading to the
+                      formation of large-scale vortices, has been
+                      suggested to explain observations of transition
+                      disks with lopsided dust
+                      distributions. While theoretical disk-planet
+                      models employed locally isothermal disks, 
+                      the original linear theory of this vortex
+                      instability was developed for adiabataic disks. We
+                      generalize the study of planetary gap edge
+                      instabilities to non-isothermal disks by
+                      including an energy equation with variable
+                      cooling rate, and study its effect on the
+                      formation and evolution of gap-edge vortices. 
                     }
@@
 	\end{align*}
-        and $t_\mathrm{cool} = \beta \Omega_k^{-1}$, where $\Omega_k$ is the Keplerian frequency. We choose $\beta$ indirectly 
+        and $t_\mathrm{cool} = \beta \Omega_k^{-1}$, where $\Omega_k$
+        is the Keplerian frequency. The energy density
+        $e=p/(\gamma-1)$ where $p$ is the pressure field and
+        $\gamma=1.4$ is the adiabatic index. We choose $\beta$ indirectly 
         through $\tilde{\beta}$ such that
@@
         \end{align*}
-        where $t_\mathrm{lib, gap\,edge}$ is the time interval between successive encounters of a fluid element, 
+        where $t_\mathrm{lib, gap\,edge}$ is the time interval between
+        successive encounters of a fluid element 
         located at the gap edge, and the planet's azimuth. Thus $\tilde{\beta}\ll 1$ implies that such a fluid returns to its initial temperature
@@
 
-        \begin{block}{{\Large Planet gaps as a function of cooling
+        \begin{block}{{\Large Planetary gap profiles and cooling
               rate}}
           \justifying
-          % {\bf generalized vortensity profile due to planet for
-          %   different beta. also density profiles.}
-
+          %Profiles during initial stages of gap formation $\to$
+          \begin{minipage}[t]{0.49\textwidth}
             \begin{figure}
-            \centering
               %            \hfill
-%            \begin{minipage}{0.49\textwidth}
-              \includegraphics[width=.49\textwidth]{Posterfig_Density}\\
-%           \end{minipage}
-%            \hfill
-%            \begin{minipage}{0.49\textwidth}
-              \includegraphics[width=.49\textwidth]{Posterfig_Vortensity}
-%            \end{minipage}
-%            \hfill
+              \includegraphics[width=.99\textwidth,clip=true,trim=0.5cm
+                2cm 0cm 1cm]{Posterfig_Density}\\
+              \includegraphics[width=.99\textwidth,clip=true,trim=0.5cm
+                0.5cm 0cm 1cm]{Posterfig_Vortensity}
             \end{figure}
-
+          \end{minipage}
+          \begin{minipage}[t]{0.49\textwidth}
+            \begin{itemize}
+            \item Top panel: shallower gaps with increasing $t_\mathrm{cool}$ 
+              but no strong dependence 
+            \item Lower panel shows the generalized vortensity
+              \begin{align*}
+                \tilde{\eta} = \frac{\omega_z}{\Sigma}\times S^{-2/\gamma},
+              \end{align*}
+              where $\omega_z$ is the vertical vorticity and $S\equiv
+              p/\Sigma^\gamma$ is the entropy. 
+            \item Extrema in $\tilde{\eta}$ associated with
+              instability (Lovelace et al. 1999, Li et al. 2000)
+            \item Entropy-production across planet-induced shocks
+              weakens extrema in $\tilde{\eta}$, so
+             increasing $t_\mathrm{cool}$ may stabilize the
+              system by making gap edges less sharp
+%              \item A sufficiently deep minima in $\tilde{\eta}$ can
+%                cause instability
+%              \item 
+            \end{itemize}
+          \end{minipage}
           %\begin{figure}
@@
           %\end{figure}
-
         \end{block}
-
         \vfill
-
-            \begin{block}{{\Large Linear stability}}
+        \begin{block}{{\Large Numerical linear stability simulations}}
           \justifying
-                  {\bf plot of growth rate v.s m for different beta,
-                    or table. 2D density plot showing development of
-                    vortex, for different beta}
-
+          We switch off the planet potential after it is fully
+          introduced and a partial gap is already formed. An azimuthal
+          average is performed, and the system is subject to random
+          density perturbations.\\   
+          \vspace{-1.5cm}
+%                  {\bf plot of growth rate v.s m for different beta,
+%                    or table. 2D density plot showing development of
+%                    vortex, for different beta}                   
+                    \hfill
+                    \begin{minipage}[t]{0.49\textwidth}
                       \begin{figure}
                         \centering
-                    \hfill
-                    \begin{minipage}{0.45\textwidth}
-                      \includegraphics[width=\textwidth]{Posterfig_lowb}
+                        \includegraphics[width=.95\textwidth,clip=false,trim=0.01cm
+                          0cm .1cm 0cm]{Posterfig_lowb}
+                      \end{figure}
                     \end{minipage}
-                    \hfill
-                    \begin{minipage}{0.45\textwidth}
-                      \includegraphics[width=\textwidth]{Posterfig_medb}
+                                        %\hfill
+                    \begin{minipage}[t]{0.49\textwidth}
+                      \begin{figure}
+                        \includegraphics[width=.95\textwidth,clip=false,trim=0cm
+                        0cm .1cm 0cm]{Posterfig_medb}
+                      \end{figure}
                     \end{minipage}    
-                    \hfill
-                  \end{figure}
-
-                   \begin{table}
-                     \hfill
-                     \begin{minipage}{\textwidth}
-                       \begin{tabularx}{\textwidth}{l*{10}{R}} %\toprule \addlinespace[10pt]
-                         \multicolumn{11}{c}{$\tilde{\beta}=0.1$} \\ \midrule
-                         m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
-                         $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
-                       \end{tabularx}
-                     \end{minipage}
-
-                     \vspace{1in}
-
-                     \begin{minipage}{\textwidth}
-                       \begin{tabularx}{\textwidth}{l*{5}{R}} %\toprule \addlinespace[10pt]
-                         \multicolumn{6}{c}{$\tilde{\beta}=1.0$} \\ \midrule
-                         m                    & 1 & 2 & 3 & 4 & 5  \\ 
-                         $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
-                       \end{tabularx}
-                     \end{minipage}
-                     \hfill
-                   \end{table}
+                    \begin{itemize}
+                    \item Slowering cooling rate increases
+                      \emph{linear} stability and favors lower-$m$ modes
+                    \end{itemize}
+%                   \begin{table}
+%                     \hfill
+%                     \begin{minipage}{\textwidth}
+%                       \begin{tabularx}{\textwidth}{l*{10}{R}} %\toprule \addlinespace[10pt]
+%                         \multicolumn{11}{c}{$\tilde{\beta}=0.1$} \\ \midrule
+%                         m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
+%                         $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
+%                       \end{tabularx}
+%                     \end{minipage}
+%
+%                     \vspace{1in}
+%
+%                     \begin{minipage}{\textwidth}
+%                       \begin{tabularx}{\textwidth}{l*{5}{R}} %\toprule \addlinespace[10pt]
+%                         \multicolumn{6}{c}{$\tilde{\beta}=1.0$} \\ \midrule
+%                         m                    & 1 & 2 & 3 & 4 & 5  \\ 
+%                         $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
+%                       \end{tabularx}
+%                     \end{minipage}
+%                     \hfill
+%                   \end{table}
 
@@
 
-            \begin{block}{\Large{Linear stability Cont.}}
+ %           \begin{block}{\Large{Linear stability Cont.}}
+ %             \justifying
+ %             \begin{figure}
+ %               \centering
+ %               \includegraphics[width=\textwidth]{Posterfig_Growth}
+ %             \end{figure}
+ %           \end{block}
+ %
+ %           \vfill
+
+            \begin{block}{\Large{Non-linear evolution of gap edge
+                  vortices}}
               \justifying
-              \begin{figure}
-                \centering
-                \includegraphics[width=\textwidth]{Posterfig_Growth}
-              \end{figure}
-            \end{block}
-
-            \vfill
-
-            \begin{block}{\Large{Non-linear vortex evolution}}
-              \justifying
-                  {\bf `lifetime' plot of m=1 for diff beta, snapshots
-                    (before, during, after vortex decay). quote rossby
-                  number when vortices in quasi-steady state}
-
-                  \begin{figure}
-                    \centering
-                    \includegraphics[width=\textwidth]{Posterfig_Lifetime_b}
-                  \end{figure}
+%                  {\bf `lifetime' plot of m=1 for diff beta, snapshots
+%                    (before, during, after vortex decay). quote rossby
+%                  number when vortices in quasi-steady state}
 
@@
             \end{block}
+
+            \begin{block}{\Large{Vortex life-time and cooling rate}}
+              \justifying
+
+                  \begin{figure}
+                    \centering
+                    \includegraphics[width=\textwidth]{lifetime_regime1}\\
+                    \includegraphics[width=\textwidth]{lifetime_regime2}
+                  \end{figure}
+                  \end{block}
+
             \vfill
 
-            \begin{block}{{\Large Conclusions so far}}\justifying
+            \begin{block}{{\Large Discussion}}\justifying
               {\large{\bf

### 2014-07-23 poster/poster.tex
poster making, 90%

@@
         \end{align*}
-        where $t_\mathrm{lib, gap\,edge}$ is the time interval between
-        successive encounters of a fluid element 
-        located at the gap edge, and the planet's azimuth. Thus $\tilde{\beta}\ll 1$ implies that such a fluid returns to its initial temperature
-        before passing by the planet again. 
+\vspace{-1.8cm}
+        \begin{itemize}
+        \item  $t_\mathrm{lib, gap\,edge}$: time interval between
+          successive encounters of a fluid element located at the gap
+          edge, and the planet's azimuth
+        \item $\tilde{\beta}\ll 1 \Rightarrow$ fluid at the gap edge returns to its initial temperature
+          between going through planet-induced shocks 
+        \end{itemize}
         \end{block}
@@
           We switch off the planet potential after it is fully
-          introduced and a partial gap is already formed. An azimuthal
+          introduced and a partial gap is formed. An azimuthal
           average is performed, and the system is subject to random
-          density perturbations.\\   
-          \vspace{-1.5cm}
+          density perturbations. Fourier analysis of the density field
+          gives the dominant mode number $m$ and growth rate $\gamma$.  
+          \\   
+          \vspace{-2cm}
 %                  {\bf plot of growth rate v.s m for different beta,
@@
                         \centering
-                        \includegraphics[width=.95\textwidth,clip=false,trim=0.01cm
+                        \includegraphics[width=.9\textwidth,clip=false,trim=0.01cm
                           0cm .1cm 0cm]{Posterfig_lowb}
@@
                       \begin{figure}
-                        \includegraphics[width=.95\textwidth,clip=false,trim=0cm
+                        \includegraphics[width=.9\textwidth,clip=false,trim=0cm
                         0cm .1cm 0cm]{Posterfig_medb}
@@
 %                  number when vortices in quasi-steady state}
-
+              Generic evolution independent of $t_\mathrm{cool}$:
+              \begin{itemize}
+              \item Single vortex in quasi-steady state from vortex merging
+              \item Vorticity source from disk-planet interaction
+                $\to$ vortex intensifies
+              \item Vortex dissipates quickly after its Rossby number
+                $Ro$ minimizes, due to shocks and/or smoothing out the
+                gap edge  
+              \end{itemize}
                   \begin{figure}
@@
 
-            \begin{block}{\Large{Vortex life-time and cooling rate}}
+            \begin{block}{\Large{Vortex lifetimes and cooling rate}}
               \justifying
-
+              \begin{itemize}
+              \item Vortices typically live longer with increasing cooling time:
+              \end{itemize}
               \begin{figure}
                 \centering
-                    \includegraphics[width=\textwidth]{lifetime_regime1}\\
-                    \includegraphics[width=\textwidth]{lifetime_regime2}
+                \includegraphics[width=\textwidth,clip=true,trim=0cm
+                  1.5cm 0cm 1cm]{lifetime_regime1}
               \end{figure}
+              \vspace{-1cm}
+              \begin{itemize}
+              \item However, for very long $t_\mathrm{cool}$ vortex
+                lifetimes shorten again:
+              \end{itemize}
+              \begin{figure}
+                \centering
+                \includegraphics[width=\textwidth,clip=true,trim=0cm
+                  0cm 0cm 1cm]{lifetime_regime2}
+              \end{figure}
+              \vspace{-2cm}
+              \begin{itemize}
+              \item A cooling time of $\tilde{\beta}\sim 5$ maximizes
+                vortex lifetime (nearly adiabatic)
+              \end{itemize}
             \end{block}
-
             \vfill
-
-            \begin{block}{{\Large Discussion}}\justifying
+            \begin{block}{{\Large Summary}}\justifying
               {\large{\bf
-
+                  The stability of planetary gaps in gaseous
+                  protoplanetary disks is affected by disk
+                  thermodynamics. A slower cooling rate, which permits
+                  more heating, tends to stabilize the system, but
+                  also enables longer vortex lifetimes unless the
+                  cooling rate is extremely long. Our simulations
+                  suggest there exists an optimal $t_\mathrm{cool}$
+                  that maximizes vortex lifetimes, a result consistent
+                  with recent isothermal simulations by Fu et
+                  al. (2014). 
                 }

### 2014-07-24 poster/poster.tex
minor poster changes

@@
 \title{\huge Gap formation and stability in non-isothermal
-  disks\\ \Large [2014 CITA Summer Student Program]}
+  protoplanetary disks \large [2014 CITA Summer Student Program]}
 \author{Min-Kai Lin and Robert Les$^\dagger$}
@@
         \end{align*}
-\vspace{-1.8cm}
+\vspace{-1.6cm}
         \begin{itemize}
-        \item  $t_\mathrm{lib, gap\,edge}$: time interval between
-          successive encounters of a fluid element located at the gap
-          edge, and the planet's azimuth
-        \item $\tilde{\beta}\ll 1 \Rightarrow$ fluid at the gap edge returns to its initial temperature
-          between going through planet-induced shocks 
+        \item  $t_\mathrm{lib, gap\,edge}=2\pi/|\Omega_k(r_p)-\Omega_k(r_\mathrm{edge})|$: time between
+          successive encounters of a fluid element at the gap
+          edge and the planet's azimuth
+        \item $\tilde{\beta}\ll 1 \Rightarrow$ fluid at the gap edge
+          returns to its initial temperature inbetween going through planet-induced shocks 
         \end{itemize}
@@
           average is performed, and the system is subject to random
-          density perturbations. Fourier analysis of the density field
-          gives the dominant mode number $m$ and growth rate $\gamma$.  
+          density perturbations. We use Fourier analysis to find
+          the dominant mode number $m$ and growth rate $\gamma$.  
           \\   
@@
               \item Single vortex in quasi-steady state from vortex merging
-              \item Vorticity source from disk-planet interaction
+              \item Continous vorticity source from disk-planet interaction
                 $\to$ vortex intensifies
@@
                   also enables longer vortex lifetimes unless the
-                  cooling rate is extremely long. Our simulations
+                  cooling time is extremely long. Our simulations
                   suggest there exists an optimal $t_\mathrm{cool}$
@@
                   with recent isothermal simulations by Fu et
-                  al. (2014). 
+                  al. (2014,ApJ,788,L41) with variable disk aspect-ratio. 
                 }

### 2014-07-24 poster/poster.tex
Updated lifetime plot and 2D linear plots
poster making

@@
                     {\bf
-                      Unstable disk-planet, leading to the
-                      formation of large-scale vortices, has been
+                      The formation of large-scale vortices at planetary gap edges has been
                       suggested to explain observations of transition
                       disks with lopsided dust
-                      distributions. While theoretical disk-planet
-                      models employed locally isothermal disks, 
-                      the original linear theory of this vortex
-                      instability was developed for adiabataic disks. We
+                      distributions. Thus far, theoretical disk-planet
+                      models have employed locally isothermal disks, while 
+                      the linear theory of the vortex-formation 
+                      instability was originally developed for adiabataic disks. We
                       generalize the study of planetary gap edge
@@
              gap-opening giant planet, treated as an external
-             potential on fixed circular orbit. We include in the
+             potential on a fixed circular orbit. We include in the
              energy equation  
@@
         \begin{itemize}
-        \item  $t_\mathrm{lib, gap\,edge}=2\pi/|\Omega_k(r_p)-\Omega_k(r_\mathrm{edge})|$: time between
+        \item  $t_\mathrm{lib, gap\,edge}=2\pi/|\Omega_k(r_p)-\Omega_k(r_\mathrm{edge})|$ is the time inbetween
           successive encounters of a fluid element at the gap
@@
           \justifying
-          We switch off the planet potential after it is fully
+          In these simulations we switch off the planet potential after it is fully
           introduced and a partial gap is formed. An azimuthal
           average is performed, and the system is subject to random
-          density perturbations. We use Fourier analysis to find
+          density perturbations. We then use Fourier analysis to find
           the dominant mode number $m$ and growth rate $\gamma$.  

### 2014-07-24 poster/poster.tex
poster done

@@
 %                  number when vortices in quasi-steady state}
-              Generic evolution independent of $t_\mathrm{cool}$:
+              We include the planet throughout. Generic evolution independent of $t_\mathrm{cool}$:
               \begin{itemize}
@@
               \item Vortex dissipates quickly after its Rossby number
-                $Ro$ minimizes, due to shocks and/or smoothing out the
+                $Ro$ minimizes, due to vortex-induced shocks and/or smoothing out the
                 gap edge  
@@
                       \includegraphics[width=\textwidth]{Posterfig_Before}
-                      \caption{$Ro=-0.16$}
+                      \caption{$Ro=-0.15$}
                     \end{minipage}
@@
                       \includegraphics[width=\textwidth]{Posterfig_After}
-                      \caption{$Ro=-0.10$}
+                      \caption{$Ro=-0.05$}
                     \end{minipage}
@@
               \item A cooling time of $\tilde{\beta}\sim 5$ maximizes
-                vortex lifetime (nearly adiabatic)
+                vortex lifetime (thick line)
               \end{itemize}

### 2014-07-24 poster/poster.tex
poster completed

@@
                     {\bf
-                      The formation of large-scale vortices at planetary gap edges has been
+                      Large-scale vortex formation at planetary gap edges has been
                       suggested to explain observations of transition
                       disks with lopsided dust
-                      distributions. Thus far, theoretical disk-planet
+                      distributions. Thus far, theoretical
                       models have employed locally isothermal disks, while 
                       the linear theory of the vortex-formation 
-                      instability was originally developed for adiabataic disks. We
+                      instability was originally developed for adiabatic disks. We
                       generalize the study of planetary gap edge
@@
         \begin{itemize}
-        \item  $t_\mathrm{lib, gap\,edge}=2\pi/|\Omega_k(r_p)-\Omega_k(r_\mathrm{edge})|$ is the time inbetween
+        \item  $t_\mathrm{lib, gap\,edge}=2\pi/|\Omega_k(r_p)-\Omega_k(r_\mathrm{edge})|$ is the time in between
           successive encounters of a fluid element at the gap
@@
         \item $\tilde{\beta}\ll 1 \Rightarrow$ fluid at the gap edge
-          returns to its initial temperature inbetween going through planet-induced shocks 
+          returns to its initial temperature in between going through planet-induced shocks 
         \end{itemize}
@@
                     \begin{itemize}
-                    \item Slowering cooling rate increases
+                    \item Slower cooling rate increases
                       \emph{linear} stability and favors lower-$m$ modes
@@
               \item Single vortex in quasi-steady state from vortex merging
-              \item Continous vorticity source from disk-planet interaction
+              \item Continuous vorticity source from disk-planet interaction
                 $\to$ vortex intensifies

### 2014-07-27 paper/model.tex
brief notes on what information should be included
updated paper

@@
 \section{Disc-planet models}\label{model}
-The system is a two-dimensional, self-gravitating gas disc of mass
-$M_d$ rotating about a central star of mass $M_*$. Cylindrical
+The system is a two-dimensional non-self-gravitating gas disc orbiting
+a central star of mass $M_*$.  
+%self-gravitating gas disc of mass
+%$M_d$ rotating about a central star of mass $M_*$. 
+Cylindrical
 co-ordinates $(r,\phi,z)$ centred on the star is adopted. The frame is   
@@
   & \frac{\p e}{\p t} + \nabla\cdot(e\bv) = -p\nabla\cdot\bv +
-  \mathcal{H} - \mathcal{C}, \\
-  &\nabla^2\Phi_d = 4\pi G \Sigma \delta(z),\label{poisson}
+  \mathcal{H} - \mathcal{C}, %\\
+%  &\nabla^2\Phi_d = 4\pi G \Sigma \delta(z),\label{poisson}
 \end{align}
@@
 The total potential $\Phi$ includes the stellar potential, planet potential
-(described below), and the disc potential $\Phi_d$ which satisfies the
-Poisson equation \ref{poisson}. Indirect potentials are also included
+(described below)%, and the disc potential $\Phi_d$ which satisfies the
+%Poisson equation \ref{poisson}. 
+and indirect potentials are also included
 to account for the non-inertial reference frame. In the momentum
 equations, $\bm{f}_\nu$ represent viscous forces whose magnitude is 
-characterised by a constant kinematic viscosity parameter $\nu$.
+characterised by a constant kinematic viscosity parameter
+$\nu$. However, we will be considering effectively inviscid discs by
+adopting small values of $\nu$ .
 
@@
 The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with $\rout =
-25\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
+40\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
 surface density profile
@@
 \end{align}
-where the power-law index $s=3/2$, $H(r) = \ciso\Omega_k $ defines the disc scale-height 
+where the power-law index $s=2$, $H(r) = \ciso\Omega_k $ defines the disc scale-height 
 where $\ciso=\sqrt{p/\Sigma}$ isothermal sound-speed. The disc aspect-ratio is defined as $h\equiv H/r$ and initially
-$h_i=0.05$. The surface density scale $\Sigma_0$ is set by  
-\begin{align}
-\Sigma_0 = \frac{h_i\rout\Omega_k^2(\rout)}{\pi G Q_o},
-\end{align}
-where $Q_o$ is the thin-disc Keplerian Toomre parameter at the outer boundary initially.
+$h=0.05$. For a non-self-gravitating disc, the surface density scale
+$\Sigma_0$ is arbitrary.
+
+%The surface density scale $\Sigma_0$ is set by  
+%\begin{align}
+%\Sigma_0 = \frac{h_i\rout\Omega_k^2(\rout)}{\pi G Q_o},
+%\end{align}
+%where $Q_o$ is the thin-disc Keplerian Toomre parameter at the outer boundary initially.
 
@@
 \end{align}
-where $M_p=qM_*$ is the planet mass ($q$ is the planet-to-star mass ratio), $\bm{r}_p=(r_p,\phi_p)$ is the planet's 
-position in the disc and $\epsilon_p=0.5r_h$ is a softening length with $r_h=(q/3)^{1/3}r_p$ being the Hill radius. 
+where $M_p$ is the planet mass and we fix $q\equiv M_p/M_*=10^{-3}$
+throughout this work. This corresponds to a Jupiter-mass planet if $M_*=M_{\sun}$. 
+The planet's position in the disc 
+ $\bm{r}_p=(r_p,\phi_p)$  and $\epsilon_p=0.5r_h$ is a softening
+length with $r_h=(q/3)^{1/3}r_p$ being the Hill radius.  
 %Hill radius chosen so eps doesn't change with time (scale-height does)
-In most simulations the planet is held on a fixed circular orbit, so that $r_p(t) = r_p(0) \equiv r_{p0}$. This also
-defines the time unit $P_0\equiv 2\pi/\Omega_k(r_{p0})$ used to describe results.
-
-In this study only one planet mass is considered, with $q=10^{-3}$. This corresponds to a Jupiter-mass
-planet if $M_*=M_{\sun}$. The initial orbital radius is $r_{p0} = 10\rin$.
+The planet is held on a fixed circular orbit with $ r_p = 10\rin$. This also
+defines the time unit $P_0\equiv 2\pi/\Omega_k(r_p)$ used to describe results. 
+%In this study only one planet mass is considered, with $q=10^{-3}$. 
+%The initial orbital radius is
 
@@
 of the original \texttt{FARGO} code \citep{masset00a} to include the energy 
-equation and disc gravity. The code employs finite-difference scheme similar 
+equation. The code employs finite-difference scheme similar 
 to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
@@
 
-\subsection{Control parameters}
-The disc parameters are $Q_o$, which characterizes the strength of self-gravity; 
-and $\beta$ which sets the cooling time. Two levels of self-gravity is considered: 
-a light disc with $Q_o=8$ and a heavy disc with $Q_o=1.5$. The cooling 
-parameter $\beta$ is chosen indirectly  through the parameter
+\subsection{Parametrized cooling}
+In this work we only vary one control parameter: the cooling
+time. 
+%The disc parameters are $Q_o$, which characterizes the strength of self-gravity; 
+%and $\beta$ which sets the cooling time. Two levels of self-gravity is considered: 
+%a light disc with $Q_o=8$ and a heavy disc with $Q_o=1.5$. 
+The cooling parameter $\beta$ is chosen indirectly  through the parameter
 $\tilde{\beta}$ such that
@@
 
-\subsection{Simulation procedure}
-The disc is evolved without a planet for $t\in[0,20]P_0$. The planet potential is smoothly switched on
-between $t\in[20,30]P_0$.
+%\subsection{Simulation procedure}
+%The disc is evolved without a planet for $t\in[0,20]P_0$. The planet potential is smoothly switched on
+%between $t\in[20,30]P_0$.
 %%  In a separate set of calculations the planet
@@
 \subsection{Diagonistic measures}
+
+\subsubsection{Generalized potential vorticity}
+
+The generalized potential vorticity (PV, or vortensity) is defined as
+\begin{align}
+  \tilde{\eta} = \frac{\kappa^2}{2\Omega\Sigma}\times S^{-2/\gamma}, 
+\end{align}
+where $\kappa^2 = r^{-3}\partial_r(r^4\Omega^2)$ is the square of the
+epicycle frequency, $\Omega=v_\phi/r$ is the angular speed, and
+$S\equiv p/\Sigma^\gamma$ is the entropy. The first factor is the
+usual potential vorticity.
+
+The generalized PV appears in the description of the linear stability
+of radially-structured adiabatic discs \citep{lovelace99,li00}, where
+the authors show an extremum in $\teta$ may lead to a dynamical
+instability. In a barotropic disc, the entropy factor is absent
+and the important quantity is the PV.
+
+\subsubsection{Fourier modes}
+
+\subsubsection{Rossby number}
+
+%{\bf {MKL.} describe vortensity and generalized vortensity, and its
+%  relevance to stability. fourier mode amplitude definition and how
+%  growth rates are obtained. rossby number. density
+%  perturbations. (see lin 2014)
+%}
+
 %% According to previous 2D studies, the key quantitiy in describing the

### 2014-07-27 paper/paper.tex
tex updates
updated paper

@@
 \newcommand{\tbeta}{\tilde{\beta}}
+\newcommand{\teta}{\tilde{\eta}}
+\newcommand{\tcool}{t_\mathrm{cool}}
 \DeclareMathOperator{\erf}{erf}
@@
 \title[Gaps in non-isothermal discs]{Gap formation and stability in 
-  non-isothermal self-gravitating discs}
+  non-isothermal protoplanetary discs}
 
-\author[Lin]{ Min-Kai Lin
-  \thanks{E-mail: mklin924@cita.utoronto.ca} \\ 
-Canadian Institute for Theoretical Astrophysics,  
+\author[Les and Lin]{Robert Les
+  $^1$\thanks{robert.les@mail.utoronto.ca} and Min-Kai Lin $^1$
+  \thanks{ mklin924@cita.utoronto.ca} \\ 
+$^1$Canadian Institute for Theoretical Astrophysics,  
 60 St. George Street, Toronto, ON, M5S 3H8, Canada \\
@@
 \begin{abstract}
-Hydrodynamic simulations of self-gravitating disc-planet interactions are presented. The energy equation is solved with a simple cooling law 
-such that $\beta = t_c\Omega_k$ is constant, where $t_c$ is the cooling time and $\Omega_k$ is the Keplerian frequency. 
-The simulations are customized to examine the stability of gaps opened by massive planets as a function of $\beta$.
+%Hydrodynamic simulations of self-gravitating disc-planet interactions are presented. 
+Several observations of transition discs show lopsided
+dust-distributions, which is suggestive of a dust-trapping mechanism at work. A
+popular explanation is the formation of a large-scale vortex at the
+edge of a gap opened by a giant planet. While theoretical models
+support this picture, numerical simluations have thus far
+employed locally isothermal discs. On the other hand, the theory of the
+vortex-forming instability was originally developed for adiabatic
+discs. We generalise the study of planetary gap stability to non-isothermal
+discs using customised numerical simulations of disc-planet
+systems. We include in the energy equation a cooling term with cooling
+timescale $t_\mathrm{cool}=\beta\Omega_k^{-1}$, where $\Omega_k$ is
+the Keplerian frequency, and examine the effect of $\beta$ on the
+stability of gap edges.
+
+%The simulations are customized to examine the stability of gaps opened by massive planets as a function of $\beta$.
 

### 2014-07-27 paper/results.tex
updated paper

--- 2014-06-16 paper/results.tex
@@
-\section{Gaps in low mass discs (RL)}
+\section{Numerical linear stability simulations}
+We first consider the effect of a cooling time on the \emph{linear} stability
+of planetary gaps. There are two competing effects. Increasing
+$t_\mathrm{cool}$ leads to more heating and hence a larger sound-speed
+$c_s$, which favours instability \citep{li00} since the vortex
+instability is pressre-driven. However, gap-opening requires $r_h>H$,
+so increasing the sound-speed, and hence $H$, leads to a shallower
+gap (see below) and expected to be more stable.
+
+In the following simulations, the planet is introduced at $t=20P_0$ and
+its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
+planet potential and azimuthally average the surface density, energy
+and velocity fields. We then perturb the density field and continue
+evolve the disc. %more details on perturbation: m=1 to 10, amplitude?  
+%resolution?
+%switch off at t=30 because in full disk-planet sims, instability sets
+%in soon after planet introduced
+
+\subsection{Gap structure}
+{\bf plots of the gap structure before instability sets in, e.g. at
+  t=30. surface density perturbation, gen. vortensity perturbation,
+  aspect-ratio. compare diff. beta. gaps become more shallow with
+  increasing cooling time (but not significantly - not enough time for
+  full gap formation). generalized vortensity: extrema smoothed out
+  with increasing cooling - expect more stability. aspect-ratio:
+  significant heating. increase in H so expect larger vortices
+}
+
+\subsection{Linear modes and growth rates}
+{\bf present `planet-off' simulations. one `fourier mode v.s. time'
+  plot to show growth of linear instability. 
+  compare growth rate and dominant
+  m as function of beta (table). 2D figs to contrast (also used to
+  show it's the minimum in generalized pv that goes unstable.    
+  result: increasing cooling time makes the gap
+  more stable, and favors lower m. note: the `basic state' should be
+  the system at t=30 after azimuthal average. linear results should
+  have small perturbations. 
+}
 
@@
 %% fast, moderate and slow cooling cases, respectively.
+
+\section{Non-linear evolution of
+  gap-edge vortices with finite cooling time} 
+{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
+  to plot vortex lifetime v.s. beta? table: 
+  averaged quantities over quasi-steady state: aspect-ratio (to
+  compare with fu at al), rossby number (vortex strength
+  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
+  only inviscid cases. describe evolution of one case. main
+  conclusion: longer vortex lifetime with increasing cooling time (up
+  to some optimal timescale). vortex death: induced-shock and/or
+  smoothing the gap edge. describe simulation setup, resolution?
+  should mention that results consistent with lower-resolution prelim
+  runs. maybe torques? 
+}
 
@@
 
-\begin{figure}
-   \includegraphics[scale=.42]{figures/stability_vis6betalow.ps}
-   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=0.1$. }
- \label{stability_vis6lowb)}
- \end{figure}
+%\begin{figure}
+%   \includegraphics[scale=.42]{figures/stability_vis6betalow.ps}
+%   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=0.1$. }
+% \label{stability_vis6lowb)}
+% \end{figure}
 
-\begin{figure}
-   \includegraphics[scale=.42]{figures/stability_vis6betamed.ps}
-   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$. }
- \label{stability_vis6medb)}
- \end{figure}
-Same as \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$. 
+%\begin{figure}
+%   \includegraphics[scale=.42]{figures/stability_vis6betamed.ps}
+%   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$. }
+% \label{stability_vis6medb)}
+% \end{figure}
+%Same as \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$.
+
 \begin{figure}

### 2014-08-07 paper/model.tex
paper update

@@
 Note that the energy source terms have been chosen to be absent 
-at $t=0$. This allows the steady state energy equation to be satisfied
-provided the flow is axisymmetric and $v_r$ is negligible compared to $v_\phi$.  
+at $t=0$. This allows the disc to be initialised in a steady state
+energy, provided the flow is axisymmetric and $v_r$ is negligible compared to $v_\phi$.  
 % If the disc is axisymmetric, steady and with only azimuthal flow, then the 
@@
 %profile (i.e. $c_\mathrm{iso}^2\equiv p/\Sigma$ is prescribed). 
-The cooling term attempts 
-to restore the initial energy density (and therefore temperature) profile.
+The energy source term attempts to restore the initial energy density (and
+therefore temperature) profile. In practice, this is a cooling term at
+the gap edge because disc-planet interaction leads to heating.
 
@@
 \section{Numerical experiments}\label{method}
-The self-gravitating disc-planet system is evolved using the 
+The disc-planet system is evolved using the 
 \texttt{FARGO-ADSG} code \citep{baruteau08, baruteau08b}. This is a modified version 

### 2014-08-07 paper/results.tex
paper update

@@
-\section{Numerical linear stability simulations}
-We first consider the effect of a cooling time on the \emph{linear} stability
-of planetary gaps. There are two competing effects. Increasing
-$t_\mathrm{cool}$ leads to more heating and hence a larger sound-speed
-$c_s$, which favours instability \citep{li00} since the vortex
-instability is pressre-driven. However, gap-opening requires $r_h>H$,
-so increasing the sound-speed, and hence $H$, leads to a shallower
-gap (see below) and expected to be more stable.
-
-In the following simulations, the planet is introduced at $t=20P_0$ and
+\section{Growth of non-axisymmetric modes without the influence of the
+  planet}
+%We first consider the linear stability of planetary gaps.  
+In this section, the planet is introduced at $t=20P_0$ and 
 its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
 planet potential and azimuthally average the surface density, energy
-and velocity fields. We then perturb the density field and continue
-evolve the disc. %more details on perturbation: m=1 to 10, amplitude?  
+and velocity fields. At this point the planet has carved a partial
+gap. We then perturb the density field and continue to 
+evolve the disc. This procedure allows us to analyse the growth of
+non-axisymmetric modes associated with the gap, but without
+complications from non-axisymmetry arising directly from disk-planet
+interaction.
+
+%The goal of these simulations is to compliment the disk-planet
+%simluations presented later. 
+These `planet-off' simulations are not linear stability calculations
+because our cooling term restores the initial temperature profile
+corresponding to $H/r=0.05$, rather than the heated gap edge.
+
+%disk-planet leads to heating 
+%tcool tries to return it to the initial smooth profile (axisymmetric)
+%if tcool > trwi -> temp returns to original value. heated gap not
+%captured
+
+%There are two competing effects. Increasing
+%$t_\mathrm{cool}$ leads to more heating and hence a larger sound-speed
+%$c_s$, which favours instability \citep{li00} since the vortex
+%instability is pressre-driven. However, gap-opening requires $r_h>H$,
+%so increasing the sound-speed, and hence $H$, leads to a shallower
+%gap (see below) and expected to be more stable.
+
+%tcool -> return to initial, t=0 temp. profile
+%whereas the `basic state' is the gap profile with structured temp profile
+
+%more details on perturbation: m=1 to 10, amplitude?  
 %resolution?
-%switch off at t=30 because in full disk-planet sims, instability sets
-%in soon after planet introduced
 
@@
 {\bf present `planet-off' simulations. one `fourier mode v.s. time'
-  plot to show growth of linear instability. 
+  plot to show growth of linear instability. do an adiabatic case (or
+  extremely long cooling time) to see the effect of heated gap edge
+  (i.e. temp doesn't go back to t=0 value too quickly)
   compare growth rate and dominant
@@
 
+\subsubsection{Nearly adiabatic disks}
+{\bf a case with extrmely long cooling time (or adiabatic run) in
+  order to look at effect of heated gap edge.  prelim result: not
+  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
+  purely adiabatic disk. change in $h$ not important for linear
+  perturbations (but important for setting up the basic state). 
+}
 %% We first describe results for weakly self-gravitating

### 2014-08-22 paper/model.tex
Paper update

@@
 The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with $\rout =
-40\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
+25\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
 surface density profile
@@
 with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi i}\simeq r\Omega_k$ 
-and $\hat{\nu}\lesssim O(10^{-5})$, so that $|v_{r i}/v_{\phi i}|\ll1$ and the initial 
+and $\hat{\nu}= 10^{-9}$, so that $|v_{r i}/v_{\phi i}|\ll1$ and the initial 
 flow is effectively only in the azimuthal direction.
+With this choice of physical viscosity $\hat{\nu}$ the only energy source is through shock-heating (via artificial viscosity) and the $\mathcal{C}$ function when $e\Sigma<e_i\Sigma_i$
 
@@
 \subsubsection{Fourier modes}
+Convential Rossby instabilities are characterized by exponetially growing pertibations originating from an axisymetric original state.
+Though this paper will not deal with formal linear instability scenarios, linear mode measurements will be useful to characterize azimuthal vortex numbers and overdensity values.
+
+The fourier transform of the density field is defined as
+\begin{align}\label{fouriertransform}
+  \Sigma_m(r,t) = \int_{0}^{2\pi} \Sigma(r,\phi,t) \, \mathrm{e}^{-\mathrm{i}m\phi+\sigma_m \mathrm{t}} \, \mathrm{d}\phi
+\end{align}
+Here $m$ is the aziumuthal wave frequency and $\sigma_m$ is the angular frequency of the $m \mathrm{th}$ mode which is complex value.
+Since $\sigma_m$ is a value dependent on $r$, a more useful diagnostic is the mode growth rate defined as
+\begin{align}\label{growth}
+  \frac{\partial \langle|\Sigma_m|\rangle_r }{\partial \mathrm{t}}= \gamma \langle|\Sigma_m|\rangle_r 
+\end{align}
+where $\gamma$ is the real part of the $\sigma_m$ and $\langle|\cdot|\rangle_r$ denotes the radial averaging of the absolute value over the region of interest.
+By using Eq.~\ref{growth} the growth rates of the linear unstable modes can be found from successive spatial Fourier transforms over an appropriate period of time.
 
 \subsubsection{Rossby number}
+The Rossby Number defined as
+\begin{align}
+  Ro = \frac{{\bf \hat{z}} \cdot \nabla \times \bv - \langle {\bf\hat{z}} \cdot \nabla \times \bv \rangle_{\phi}}{2\langle\Omega\rangle_{\phi}}, 
+\end{align}
+is a dimensionless number which measures the ratio of interial to Corriolis forces.
+Here $\langle \cdot \rangle_{\phi}$ denotes azimuthal the averaged quantity.
+Values of $Ro<0$ correspond to anti-cyclonic rotation with respect to the background motion and thus can be used to quantify area of a vortex and its intensity.
 

### 2014-08-22 paper/paper.tex
paper update
Paper update

@@
 
+\usepackage{array,booktabs,tabularx}
+\newcolumntype{R}{>{\centering\arraybackslash}X} % right justified tabularx columns
+
 \title[Gaps in non-isothermal discs]{Gap formation and stability in 
@@
 Several observations of transition discs show lopsided
-dust-distributions, which is suggestive of a dust-trapping mechanism at work. A
+dust-distributions. This is suggestive of a dust-trapping mechanism at work. A
 popular explanation is the formation of a large-scale vortex at the
@@
 stability of gap edges.
+For measurments reminsent of linear stability, gap edges stabilize with increasing $\beta$ as dominate modes and growth rates decrease in value.
+In the non-linear limit of planet-vortex interactions, large scale overdense vortices were found to last for 1000's of orbits in a quasi-steady state characterized by sudden disspation on the order of 100's or planet orbits.
+Vortex lifetimes were found to have a non-monoticly dependence on $\beta$ and an optimum value for cooling rate was found to maximize the lifetime.
 

### 2014-08-22 paper/results.tex
Paper update

@@
 
+Simulations were resolved with grid divisions of $(N_r,N_{\phi})=(1024,2048)$ for time up to $200P_0$ allowing high resolution and detection of high $m$ structures.
+Cases of $\tilde{\beta}=0.1,1.0,10.0$ were analyzed to provide a range of fast, moderate, and slowly cooled disks.
+
 %The goal of these simulations is to compliment the disk-planet
@@
 These `planet-off' simulations are not linear stability calculations
-because our cooling term restores the initial temperature profile
-corresponding to $H/r=0.05$, rather than the heated gap edge.
+since the cooling term in our energy equation restores the initial temperature profile corresponding to $H/r=0.05$, rather than the heated gap edge.
+Linear stability calculations and adiabatic cases will be further discussed in section~\ref{adiabatic_section}.
 
@@
 \subsection{Gap structure}
-{\bf plots of the gap structure before instability sets in, e.g. at
-  t=30. surface density perturbation, gen. vortensity perturbation,
-  aspect-ratio. compare diff. beta. gaps become more shallow with
-  increasing cooling time (but not significantly - not enough time for
-  full gap formation). generalized vortensity: extrema smoothed out
-  with increasing cooling - expect more stability. aspect-ratio:
-  significant heating. increase in H so expect larger vortices
+%{\bf plots of the gap structure before instability sets in, e.g. at
+%  t=30. surface density perturbation, gen. vortensity perturbation,
+%  aspect-ratio. compare diff. beta. gaps become more shallow with
+%  increasing cooling time (but not significantly - not enough time for
+%  full gap formation). generalized vortensity: extrema smoothed out
+%  with increasing cooling - expect more stability. aspect-ratio:
+%  significant heating. increase in H so expect larger vortices
+%}
+
+Since stability of the gap is of interest we analyze the intial state formed by the planet-disk intereactions.
+The intial gap profiles after the averaging procedure are shown in Fig \ref{intial1D}.
+Quicker cooling rates (lower $\tilde{\beta}$) result in deeper density gradients and more shallow gaps in the disk but never significant enough to form a clear gap in this time period.
+Generalized vortensity profiles show smoothening with larger cooling rates with the extrema becoming less pronounced.
+Charactereristicly for such disk-planet Rossby vortices the generalized vortensity minima correspond to density maxima in the disk.
+Larger $\tilde\beta$ values also as expeced result in higher disk aspect ratios $H/r$, which is a value directly correlated to temperature in our model.
+
+ For a planet to carve a gap in the surrounding material it has been shown that $q>H^3$ is needed \citep{crida06}.
+This indicates that for hotter disks, which have respectively higher $H$, it becomes more difficult by a cubic dependence for planet induced gaps to be formed.
+This is seen in our simulations as changes in $\tilde{\beta}$ result in changes of $h$ by less than $20\%$ yet gap depth almost doubles.
+Since generalized vortensity extrema are also correspondingly less pronounced and density gradients become less large we expect that the stability of gaps to increase with larger $\tilde{\beta}$.
+
+\begin{figure}
+  \includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
+    2cm 0cm 0cm]{figures/compare_sigma}
+  \includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
+    2cm 0cm 1cm]{figures/compare_gvortensity}
+\includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
+    0.5cm 0cm 1cm]{figures/compare_aspectratio}
+  \caption{Gap profiles for the intial partial gap opened before instability emerges at $t=30P_0$ for quick (solid line), moderate (dashed line), and slow cooling (broken line).
+Relative density pertibation (top), generalized vortensity pertibation (middle) and disk aspect ratio (bottom) are shown.
+\label{intial1D}}
+\end{figure}
+
+%\begin{tabularx}{0.4\textwidth}{l*{10}{R}} \toprule
+%  \multicolumn{11}{c}{$\tilde{\beta}=0.1$} \\ \midrule
+%  m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
+ % $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
+%\end{tabularx}
+
+%\begin{tabularx}{0.4\textwidth}{l*{5}{R}} \toprule
+%  \multicolumn{6}{c}{$\tilde{\beta}=1.0$} \\ \midrule
+%  m                    & 1 & 2 & 3 & 4 & 5  \\ 
+%  $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
+%\end{tabularx}
+
+\begin{table}
+  \centering
+  \caption{Dominant mode and growth rates for $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling) values during `planet-off' simulations \label{modetable}}
+  \hfill
+  \begin{minipage}{0.3\linewidth}
+    \begin{tabularx}{\textwidth}{l R} 
+      \multicolumn{2}{c}{$\tilde{\beta}=0.1$} \\ 
+      \toprule
+      m & $\gamma10^2/\Omega(r_p)$ \\
+      \midrule
+      6 & 7.3 \\
+      7 & 7.8 \\
+      8 & 7.9 \\
+      9 & 7.9 \\
+      10 & 6.8 \\ 
+      \bottomrule
+    \end{tabularx}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.3\linewidth}
+    \begin{tabularx}{\textwidth}{l R} 
+      \multicolumn{2}{c}{$\tilde{\beta}=1.0$} \\ 
+      \toprule
+      m & $\gamma10^2/\Omega(r_p)$ \\
+      \midrule
+      3 & 2.0 \\
+      4 & 2.2 \\
+      5 & 2.3 \\
+      6 & 1.6 \\
+      7 & 1.1 \\ 
+      \bottomrule
+    \end{tabularx}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.3\linewidth}
+    \begin{tabularx}{\textwidth}{l R} 
+      \multicolumn{2}{c}{$\tilde{\beta}=10.0$} \\ 
+      \toprule
+      m & $\gamma10^2/\Omega(r_p)$ \\
+      \midrule
+      1 & 1.1 \\
+      2 & 1.6 \\
+      3 & 1.7 \\
+      4 & 1.2 \\
+      5 & 0.1 \\ 
+      \bottomrule
+    \end{tabularx}
+  \end{minipage}
+  \hfill
+\end{table}
+
+\subsection{Linear modes and growth rates}\label{linear}
+%{\bf present `planet-off' simulations. one `fourier mode v.s. time'
+%  plot to show growth of linear instability. do an adiabatic case (or
+%  extremely long cooling time) to see the effect of heated gap edge
+%  (i.e. temp doesn't go back to t=0 value too quickly)
+%  compare growth rate and dominant
+%  m as function of beta (table). 2D figs to contrast (also used to
+%  show it's the minimum in generalized pv that goes unstable.    
+%  result: increasing cooling time makes the gap
+%  more stable, and favors lower m. note: the `basic state' should be
+%  the system at t=30 after azimuthal average. linear results should
+%  have small perturbations. 
+%}
+After the azimuthal averaging, the small pertibation excites rapidly growing instabilities by the RWI.
+We analyticly charcterize these modes with quantities $m$ and $\gamma$ as defined by Eq.~\ref{fouriertransform} and Eq.~\ref{growth}.
+Table \ref{modetable} contains the results of the mode growth rates for a variety of cooling rates during this semi-linear phase.
+Results indicate that as $\tilde{\beta}$ was increased from $ 0.1\rightarrow10.0$ the dominant azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the respective growth rate for mode decreased from $ \gamma/\Omega(r_p)=0.079 \rightarrow 0.017$.
+Snapshots of the instability in $r-\phi$ plane for the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
+
+These `planet-off' simulations, which are reminiscient of linear stability measurements, show that stability of gap edges increase with longer cooling rates (ie hotter gaps).
+This is as expected from the intial gap profiles and the gap forming criteria discussed in the previous section.
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
+    0cm 0cm 0cm]{figures/linear_stability}
+  \caption{Log plot of the azimuthal Fourier mode of disk surface density non-dimenionlized by the background mode, during the `planet-off' simulations for $\tilde{\beta}=10$.
+Colors correspond to different $m$ values.
+The m=3 component of the fourier transform clearly is most dominant during this growth phase of the instability and has corresponding $\gamma=0.017\Omega(r_p)$.
+\label{linearmodes}}
+\end{figure}
+
+\begin{figure*}
+  \centering
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_lowb}
   }
-
-\subsection{Linear modes and growth rates}
-{\bf present `planet-off' simulations. one `fourier mode v.s. time'
-  plot to show growth of linear instability. do an adiabatic case (or
-  extremely long cooling time) to see the effect of heated gap edge
-  (i.e. temp doesn't go back to t=0 value too quickly)
-  compare growth rate and dominant
-  m as function of beta (table). 2D figs to contrast (also used to
-  show it's the minimum in generalized pv that goes unstable.    
-  result: increasing cooling time makes the gap
-  more stable, and favors lower m. note: the `basic state' should be
-  the system at t=30 after azimuthal average. linear results should
-  have small perturbations. 
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_medb}
   }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_highb}
+  }  
+  \caption{$r-\phi$ plots of the generalized vortensity pertibation for cases of $\tilde{\beta}=0.1,1.0,10.0$ (left,middle,right) during the linear phases of growth.
+The number of visible vortices decrease as $\tilde{\beta}$ increases and snapshot times correspondingly increase to capture the appropriate phase of the growth.
+\label{2Dlinear}}
+\end{figure*}
 
 \subsubsection{Nearly adiabatic disks}
-{\bf a case with extrmely long cooling time (or adiabatic run) in
-  order to look at effect of heated gap edge.  prelim result: not
-  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
-  purely adiabatic disk. change in $h$ not important for linear
-  perturbations (but important for setting up the basic state). 
-}
+\label{adiabatic_section}
+%{\bf a case with extrmely long cooling time (or adiabatic run) in
+%  order to look at effect of heated gap edge.  prelim result: not
+%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
+%  purely adiabatic disk. change in $h$ not important for linear
+%  perturbations (but important for setting up the basic state). 
+%}
+During the `planet-off' simulations the analysis is not truly a linear stability measurement as we do not capture the effects of a heated disk on the instability growth.
+Since $t_{\mathrm{cool}}<t_{\mathrm{instability}}$ the disk manages to cool back to its intial value of $h=0.05$ before the instability emerges and thus we do not have a steady background state to form a linear stability problem.
+To compare our 'planet-off' analysis with a true linear stability analysis we ran a simulation  with $\tilde{\beta}=100.0$ which corresponds with an almost adiabtaic case.
+In this simulation the cooling rate is slow enough that the gap temperature profile changes only marginally over the timescale of the growth of the instability.
+The results show that the linear stability of the $\tilde{\beta}=100.0$ cases almost exactly equivilant to the case of $\tilde{\beta}=10.0$ in terms of mode growth rates and gap profiles.
+Since the disk heats up only to values $h\simeq0.06$ in the adiabatic case, the difference in growth rates between this and the original disk temperature of $h=0.05$ changes almost insignificantly as shown by linear theory \citep{li00}.
+This indicate that a continuously hot gap edge is not as important for development of vortices as the heating effects on the formation of the intail gap state.
+
 %% We first describe results for weakly self-gravitating
@@
   gap-edge vortices with finite cooling time} 
-{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
-  to plot vortex lifetime v.s. beta? table: 
-  averaged quantities over quasi-steady state: aspect-ratio (to
-  compare with fu at al), rossby number (vortex strength
-  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
-  only inviscid cases. describe evolution of one case. main
-  conclusion: longer vortex lifetime with increasing cooling time (up
-  to some optimal timescale). vortex death: induced-shock and/or
-  smoothing the gap edge. describe simulation setup, resolution?
-  should mention that results consistent with lower-resolution prelim
-  runs. maybe torques? 
-}
+%{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
+%  to plot vortex lifetime v.s. beta? table: 
+%  averaged quantities over quasi-steady state: aspect-ratio (to
+%  compare with fu at al), rossby number (vortex strength
+%  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
+%  only inviscid cases. describe evolution of one case. main
+%  conclusion: longer vortex lifetime with increasing cooling time (up
+%  to some optimal timescale). vortex death: induced-shock and/or
+%  smoothing the gap edge. describe simulation setup, resolution?
+%  should mention that results consistent with lower-resolution prelim
+%  runs. maybe torques? 
+%}
 
-Long term simulation of vortices were done for $\tilde{\beta}=0.1,1.0$ and
- $\hat{\nu}=10^{-6},10^{-8},10^{-9}$ up to a total of 2000 orbits. Values
- of higher $\tilde{\beta}$ resulted generally in longer vortex lifetimes.
- The inviscid cases also were characterized by a sudden death of vortices
- while the higher viscosity case showed a gradual decrease in mode amplitudes
- attributed to the vortices.
+Long term simulation of edge vortices were done for $\tilde{\beta}=0.1,0.5,1.0,5.0,10.0$ up to a total of $2000P_0$.
+Planets were left free to interact dynamiclly with the disk and vortices after $t=30P_0$ as apposed to previous section.
+These simulations were less resolved with $(N_r,N_{\phi})=(512,1024)$ in order for them to be computationally economical.
+An open boundry condition was used in the outer edge of the simulation region.
 
-Short term simulations of the linear phase of instability growth also distinguished
- the different effects of cooling rates. Simulations with higher
- $\tilde{\beta}$ were characterized by slower growth rates centered at the
- $m=2$ mode, lower density pertibations and vortex locations, and a more unitary aspect
- ratio for the vortices.
+After the semi-linear growth of vortices, vortex merger effects take hold and by 150$P_0$ for all $\tilde\beta$ values, only $m=1$ vortex modes exist.
+Plots of the amplitude of the $m=1$ mode for the different $\tilde\beta$ cases are shown in Fig~\ref{lifetimeplot}.
+These instabilities stay in quasi-steady states for longer than a 1000 of orbits in most cases, in which they grow in intensity and accumulate density.
+The growth of these vortices are supplied by the continuous generation of vorticity by the planet-disk interactions.
+Eventually the $m=1$ amplitudes are also characterized by sudden large drops of vortex intensity which happen over timescales of 20$P_0$.
+After these drops the instability is never seen to reform at such large over densities and density bumps are never larger than the planet induced shocks.
+We designate the time of these amplitude drops as the lifetime of the quasi-steady vortex.
+
+Rossby values have dynamics that are coupled with the lifetime and $m=1$ amplitude of the vortices.
+As the vortex forms it has a charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic motion.
+As the $m=1$ amplitude grows the relative spin of the vortex contiously grows along with it, reaching values of $Ro\approx-0.4$.
+As the vortex dies off as described above the associated Rossby number of the vortex accordingly quickly approaches 0.
+
+The sudden dissipation of these vortices seem to be a result of the vortices shocking the system.
+Measurements of the Mach number near the vortices show large increases in value when the vortex is dissipating which indicates that the system is more susceptible to shocks.
+Also gradients of surface density are seen to spike up in wake like features around the vortex during dissipation which is also indicative of shocks.
+Of note is also that the large overdensity the vortex creates during it's quasi-steady state distorts the background disk.
+So as the vortex grows in intensity it alters the intial gap structure that formulated its growth.
+
+Contradictory to our `planet-off' simulations, vortex lifetimes did not monotonicly depend on intial edge stability and from our simulations we find that there is $t_{\mathrm{coo}l}$ which optimizes the vortex lifetime as seen in Fig~\ref{lifetimeplot}.
+Increasing $\tilde\beta$ correspondingly increased the lifetime of the vortex up to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that lasted beyond the simulation time of 2000$P_0$.
+Cooling rates with $\tilde\beta>5.0$ showed lifetimes that were well below the max lifetime.
+By analyzing the aspect ratio of the corotation region around the vortex we find that longest lifetime $\tilde\beta=5.0$ disk corresponds to $h\approx0.06$ which is a result similarily found by \citet{fu14} to be the optimum lifetime for long-term simulations of isothermal disks with varying intial temperature profiles $h$.
+
+Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in the the non-montonic form of the lifetime.
+As the $\tilde\beta$ increases so does the stability of the gap edge as seen in section~\ref{linear}, however as $\tilde\beta$ increases so does the $c_{\mathrm{iso}}$ which is known from theory to effect the growth of instability \citep{li00}.
+Thus increasing $\tilde\beta$ increases the sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to a certain point in which the effects of the more stable gap edge becomes a significant effect.
+The correlation of growth rate and sound speed is non-negligible as in previous `planet-off' case as the disk now can reach values of $h\simeq0.08$ for very slow cooling and the importance of the effect can be imagined to accumulate over the now considerably longer timescale.
+These effects dictate when then the vortex reaches the critical value that causes disspation and hence the lifetime fo the vortex.
 
 \begin{figure}
-   \includegraphics[scale=.42]{figures/stability_vis9betalow.ps}
-   \caption{Mode intensity relative to $m=0$ for $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=0.1$.}
- \label{stability_vis9lowb}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+    0cm 0cm 1cm]{figures/longterm_stability}
+  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$ background mode for longterm non-linear simulations corrsponding to vortex lifetime.
+The longest living mode correpsonds to $\tilde\beta=5.0$ (the red line).
+\label{lifetimeplot}}
 \end{figure}
@@
 \begin{figure}
-   \includegraphics[scale=.42]{figures/stability_vis9betamed.ps}
-   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=1.0$. }
- \label{stability_vis9medb)}
+  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
+  \caption{2-D snapshot in polar form of relative density pertibation for $\tilde\beta=0.1$ vortex during quasi-steady state at $t=P_0$.
+Large scale non-axisyemetric overdensity can be seen.
+\label{Vortex2D}}
 \end{figure}
@@
 %Same as \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$.
-
-\begin{figure}
-   \includegraphics[scale=.60]{figures/analysis_sigma50lowb.ps}
-   \caption{Density pertibation of disk at time=$50t_p$ with $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=0.1$. }
- \label{shortterm_lowb)}
- \end{figure}
-
-\begin{figure}
-   \includegraphics[scale=.60]{figures/analysis_sigma50highb.ps}
-   \caption{Density pertibation of disk at time=$50t_p$ with $\hat{\nu}=10^{-9}$ and $\tilde{\beta}=10.0$. }
- \label{shortterm_highb)}
- \end{figure}
 

### 2014-09-03 paper/results.tex
Small edits

@@
 
-Simulations were resolved with grid divisions of $(N_r,N_{\phi})=(1024,2048)$ for time up to $200P_0$ allowing high resolution and detection of high $m$ structures.
+Simulations were resolved with grid divisions of $(N_r,N_{\phi})=(1024,2048)$ with $r_{\mathrm{out}}=25r_{\mathrm{in}}$ for time up to $200P_0$ allowing high resolution and detection of high $m$ structures.
 Cases of $\tilde{\beta}=0.1,1.0,10.0$ were analyzed to provide a range of fast, moderate, and slowly cooled disks.
@@
 
-Since stability of the gap is of interest we analyze the intial state formed by the planet-disk intereactions.
-The intial gap profiles after the averaging procedure are shown in Fig \ref{intial1D}.
-Quicker cooling rates (lower $\tilde{\beta}$) result in deeper density gradients and more shallow gaps in the disk but never significant enough to form a clear gap in this time period.
+Since stability of the gap is of interest we analyze the intial state formed by the planet-disk interactions.
+The intial gap profiles after the averaging procedure are shown in Fig \ref{intial1D} for varying $\tilde\beta$.
+Gaps formed are shallower for lower $\tilde\beta$ (faster cooling) but never significant enough to form a clear gap in this short timeperiod.
+Quicker cooling rates also result in steeper density gradients at the gap boundry and both the density maxima and density minima correspondingly increase and decrease with lower $\tilde\beta$.
+For all cases the gap width is $\approx 5r_{hill}$ which is set by torque balances in the disk \citep{crida06}.
+
  Generalized vortensity profiles show smoothening with larger cooling rates with the extrema becoming less pronounced.
 Charactereristicly for such disk-planet Rossby vortices the generalized vortensity minima correspond to density maxima in the disk.
-Larger $\tilde\beta$ values also as expeced result in higher disk aspect ratios $H/r$, which is a value directly correlated to temperature in our model.
+Larger $\tilde\beta$ values also as expected result in higher disk aspect ratios $H/r$, which is a value directly correlated to temperature in our model.
 
@@
 This is seen in our simulations as changes in $\tilde{\beta}$ result in changes of $h$ by less than $20\%$ yet gap depth almost doubles.
-Since generalized vortensity extrema are also correspondingly less pronounced and density gradients become less large we expect that the stability of gaps to increase with larger $\tilde{\beta}$.
+Since generalized vortensity extrema are also correspondingly less pronounced and density gradients become smaller we expect that the stability of gaps to increase with larger $\tilde{\beta}$.
 
@@
     0.5cm 0cm 1cm]{figures/compare_aspectratio}
-  \caption{Gap profiles for the intial partial gap opened before instability emerges at $t=30P_0$ for quick (solid line), moderate (dashed line), and slow cooling (broken line).
+  \caption{Gap profiles at $t=30P_0$ for the intial partial gap opened before instability emerges for quick (solid line), moderate (dashed line), and slow cooling (broken line).
 Relative density pertibation (top), generalized vortensity pertibation (middle) and disk aspect ratio (bottom) are shown.
@@
 Table \ref{modetable} contains the results of the mode growth rates for a variety of cooling rates during this semi-linear phase.
-Results indicate that as $\tilde{\beta}$ was increased from $ 0.1\rightarrow10.0$ the dominant azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the respective growth rate for mode decreased from $ \gamma/\Omega(r_p)=0.079 \rightarrow 0.017$.
+Results indicate that as $\tilde{\beta}$ was increased from $ 0.1\rightarrow10.0$ the dominant azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the respective growth rate decreased from $ \gamma/\Omega(r_p)=0.079 \rightarrow 0.017$.
 Snapshots of the instability in $r-\phi$ plane for the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
@@
     0cm 0cm 0cm]{figures/linear_stability}
-  \caption{Log plot of the azimuthal Fourier mode of disk surface density non-dimenionlized by the background mode, during the `planet-off' simulations for $\tilde{\beta}=10$.
+  \caption{Log plot of the azimuthal Fourier modes of disk surface density non-dimenionlized by the background mode during the `planet-off' simulations for $\tilde{\beta}=10$.
 Colors correspond to different $m$ values.
-The m=3 component of the fourier transform clearly is most dominant during this growth phase of the instability and has corresponding $\gamma=0.017\Omega(r_p)$.
+The $m=3$ component of the fourier transform is the fast growing mode during this growth phase of the instability and has corresponding $\gamma=0.017\Omega(r_p)$.
 \label{linearmodes}}
@@
   }  
-  \caption{$r-\phi$ plots of the generalized vortensity pertibation for cases of $\tilde{\beta}=0.1,1.0,10.0$ (left,middle,right) during the linear phases of growth.
-The number of visible vortices decrease as $\tilde{\beta}$ increases and snapshot times correspondingly increase to capture the appropriate phase of the growth.
+  \caption{Radial plots of the generalized vortensity pertibation for cases of $\tilde{\beta}=0.1,1.0,10.0$ (left,middle,right) during the linear phases of growth.
+The number of visible vortices decrease as $\tilde{\beta}$ increases and snapshot times correspondingly increase to capture the longer growth phases of the instabilities.
 \label{2Dlinear}}
@@
 Since the disk heats up only to values $h\simeq0.06$ in the adiabatic case, the difference in growth rates between this and the original disk temperature of $h=0.05$ changes almost insignificantly as shown by linear theory \citep{li00}.
-This indicate that a continuously hot gap edge is not as important for development of vortices as the heating effects on the formation of the intail gap state.
+This indicate that a continuously hot gap edge is not as important for development of vortices as the heating effects on the formation of the intial gap state.
 
@@
 These simulations were less resolved with $(N_r,N_{\phi})=(512,1024)$ in order for them to be computationally economical.
-An open boundry condition was used in the outer edge of the simulation region.
+An open boundry condition was used in the outer edge of $r_{\mathrm{out}}=45r_{\mathrm{in}}$ allowing significant room for linblad resonances of the vortices.
 
-After the semi-linear growth of vortices, vortex merger effects take hold and by 150$P_0$ for all $\tilde\beta$ values, only $m=1$ vortex modes exist.
+After the semi-linear growth of vortices, vortex merger effects take hold and by $150P_0$ for all $\tilde\beta$ values, only $m=1$ vortex modes exist.
 Plots of the amplitude of the $m=1$ mode for the different $\tilde\beta$ cases are shown in Fig~\ref{lifetimeplot}.
-These instabilities stay in quasi-steady states for longer than a 1000 of orbits in most cases, in which they grow in intensity and accumulate density.
+These instabilities stay in quasi-steady states for $>1000P_0$ in most cases, in which they grow in intensity and reaching overdenisities almost 10 times the intial local density.
 The growth of these vortices are supplied by the continuous generation of vorticity by the planet-disk interactions.
-Eventually the $m=1$ amplitudes are also characterized by sudden large drops of vortex intensity which happen over timescales of 20$P_0$.
+Eventually the $m=1$ amplitudes are also characterized by sudden large drops of vortex intensity which happen over timescales of $50P_0$.
 After these drops the instability is never seen to reform at such large over densities and density bumps are never larger than the planet induced shocks.
@@
 As the vortex forms it has a charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic motion.
-As the $m=1$ amplitude grows the relative spin of the vortex contiously grows along with it, reaching values of $Ro\approx-0.4$.
+As the $m=1$ amplitude grows the relative spin of the vortex continuously grows along with it, reaching values of $Ro\approx-0.4$.
 As the vortex dies off as described above the associated Rossby number of the vortex accordingly quickly approaches 0.
@@
 Contradictory to our `planet-off' simulations, vortex lifetimes did not monotonicly depend on intial edge stability and from our simulations we find that there is $t_{\mathrm{coo}l}$ which optimizes the vortex lifetime as seen in Fig~\ref{lifetimeplot}.
-Increasing $\tilde\beta$ correspondingly increased the lifetime of the vortex up to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that lasted beyond the simulation time of 2000$P_0$.
+Increasing $\tilde\beta$ correspondingly increased the lifetime of the vortex up to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that lasted beyond the simulation time of $2000P_0$.
 Cooling rates with $\tilde\beta>5.0$ showed lifetimes that were well below the max lifetime.
@@
 The correlation of growth rate and sound speed is non-negligible as in previous `planet-off' case as the disk now can reach values of $h\simeq0.08$ for very slow cooling and the importance of the effect can be imagined to accumulate over the now considerably longer timescale.
-These effects dictate when then the vortex reaches the critical value that causes disspation and hence the lifetime fo the vortex.
+These effects dictate when then the vortex reaches the critical value in which it shocks the system that subsequently causes disspation and hence the lifetime fo the vortex.
 
@@
   \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
-  \caption{2-D snapshot in polar form of relative density pertibation for $\tilde\beta=0.1$ vortex during quasi-steady state at $t=P_0$.
-Large scale non-axisyemetric overdensity can be seen.
+  \caption{Cartesian plot of relative density pertibation for $\tilde\beta=0.1$ vortex case during the quasi-steady state.
+The large scale non-axisyemetric $m=1$ overdensity can be seen.
 \label{Vortex2D}}

### 2014-09-11 paper/paper.tex
started intro

@@
 \section{Introduction}\label{intro}
+{
+\bf 
+observational motivation: asymmetric dust distributions. 
+theoretical motivations: generalize gap stability to include energy equation. 
+}
+
+Disk-planet interaction plays an important role in the theory of planet formation and has been invoked to explain 
+sub-structures obsererved in transition disks. It is well-known that a sufficiently massive planet opens a disk gap, and
+such structures have been detected. A more recent theoretical result is that planetary gaps are dynamically unstable under
+certain conditions. The so-called Rossby wave instability operates in radially structured disks, and leads 
+to vortex-formation.
+
+In this work, we extend the study of planetary gap stability against vortex formation to non-isothermal discs. 
+We include a fluid energy equation in the hydrodynamical equations describing disk-planet systems. We consider a simple
+cooling function and examine the effect cooling rate on the stability of planetary gaps.
+
+This paper is organised as follows.
 

### 2014-09-19 paper/model.tex
Small edits
edited linear discussion

--- 2014-08-07 paper/model.tex
@@
 where $\Sigma$ is the surface density, $\bv = (v_r,v_\phi)$ the fluid
-velocity, $p$ is the pressure, $e=p/(\gamma-1)$ is the energy
-density and the adiabatic index $\gamma=1.4$ is assumed constant.
+velocity, $p$ is the vertically-integrated pressure, $e=p/(\gamma-1)$ is the energy
+per unit area and the adiabatic index $\gamma=1.4$ is assumed
+constant.
 
 The total potential $\Phi$ includes the stellar potential, planet potential
-(described below)%, and the disc potential $\Phi_d$ which satisfies the
+(described below) %, and the disc potential $\Phi_d$ which satisfies the
 %Poisson equation \ref{poisson}. 
-and indirect potentials are also included
-to account for the non-inertial reference frame. In the momentum
+and indirect potentials to account for the non-inertial reference
+frame.%  \citep[see][for explicit expressions]{masset02}
+In the momentum 
 equations, $\bm{f}_\nu$ represent viscous forces whose magnitude is 
@@
 $\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
-input parameter. Although idealised, this cooling prescription allows one 
+input parameter. This cooling prescription allows one 
 to explore the full range of thermodynamic response of the disc in a 
-systematic way.
+systematic way: $\beta\ll1$ is a locally isothermal disc while
+$\beta\gg1$ is an adiabatic disk.
 
-Note that the energy source terms have been chosen to be absent 
-at $t=0$. This allows the disc to be initialised in a steady state
-energy, provided the flow is axisymmetric and $v_r$ is negligible compared to $v_\phi$.  
-% If the disc is axisymmetric, steady and with only azimuthal flow, then the 
-%energy equation is trivially satisfied. 
-%This allows one to specify equilibrium discs (see below)
-%in the same way as in the case where the energy equation is replaced by a fixed temperature 
-%profile (i.e. $c_\mathrm{iso}^2\equiv p/\Sigma$ is prescribed). 
-The energy source term attempts to restore the initial energy density (and
+Note that the energy source terms
+have been chosen to be absent at $t=0$, allowing the disc to be
+initialised close to steady state. The $\mathcal{C}$ function attempts to restore the initial energy density (and
 therefore temperature) profile. In practice, this is a cooling term at
@@
 \subsection{Disc model and initial condition}
-The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with $\rout =
-40\rin$ and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
+The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with
+$r_\mathrm{out}=25\mathrm{in}$ {\bf check: if all simulations use the
+  same domain, state it here} and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
 surface density profile
@@
 \begin{align}\label{initial_density}
-   \Sigma_i(r) = \Sigma_0\left(\frac{r}{r_\mathrm{in}}\right)^{-s}
+   \Sigma(r) = \Sigma_\mathrm{ref}\left(\frac{r}{r_\mathrm{in}}\right)^{-s}
     \left[1 - \sqrt{\frac{r_\mathrm{in}}{r + H_i(\rin)}}\,\right] 
@@
 $h=0.05$. For a non-self-gravitating disc, the surface density scale
-$\Sigma_0$ is arbitrary.
+$\Sigma_\mathrm{ref}$ is arbitrary.
 
@@
 The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
-pressure, stellar gravity and disc gravity. The initial radial velocity is $v_{r i}=3\nu/r$ 
-with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi i}\simeq r\Omega_k$ 
-and $\hat{\nu}\lesssim O(10^{-5})$, so that $|v_{r i}/v_{\phi i}|\ll1$ and the initial 
-flow is effectively only in the azimuthal direction.
+pressure, stellar gravity and disc gravity. The initial radial velocity is $v_{r}=3\nu/r$ 
+with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi}\simeq r\Omega_k$ 
+and $\hat{\nu}= 10^{-9}$, so that $|v_{r}/v_{\phi}|\ll1$ and the initial 
+flow is effectively only in the azimuthal direction.  With this choice
+of physical viscosity $\hat{\nu}$ the only heating source is through
+shock-heating (via artificial viscosity) and the $\mathcal{C}$
+function when $e\Sigma<e_i\Sigma_i$.
 
@@
 of the original \texttt{FARGO} code \citep{masset00a} to include the energy 
-equation. The code employs finite-difference scheme similar 
+equation. The code employs a finite-difference scheme similar 
 to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
@@
 In this work we only vary one control parameter: the cooling
-time. 
-%The disc parameters are $Q_o$, which characterizes the strength of self-gravity; 
+time. %The disc parameters are $Q_o$, which characterizes the strength of self-gravity; 
 %and $\beta$ which sets the cooling time. Two levels of self-gravity is considered: 
@@
 \end{align}
-where $x_s$ is the distance from the planet, and $t_\mathrm{lib}$ is
-the time interval between successive encounters of a fluid element and
-the planet's azimuth. Assuming Keplerian orbital frequencies and $x_s\ll r_p$
+where $x_s=2r_h$ is the approximate distance from the planet to its
+gap edge, and $t_\mathrm{lib}$ is the time interval between successive
+encounters of a fluid element at the gap edge and the planet's
+azimuth. That is, we measure the cooling time in units of the time
+taken for a gap edge fluid element to encounter planet-induced
+shocks.
+
+Assuming Keplerian orbital frequencies and $x_s\ll r_p$
 gives $t_\mathrm{lib}\simeq 4\pi r_p/(3\Omega_{kp} x_s)$, where
@@
 %2rh is CR of G.E.I. 
-
 %\subsection{Simulation procedure}
@@
 
-The generalized potential vorticity (PV, or vortensity) is defined as
+The generalized potential vorticity is defined as
 \begin{align}
@@
 $S\equiv p/\Sigma^\gamma$ is the entropy. The first factor is the
-usual potential vorticity.
+usual potential vorticity (PV, or vortensity).
 
@@
 \subsubsection{Fourier modes} 
+The Rossby wave instability is characterized by exponetially
+growing pertibations. Though in this paper we do not formally consider
+linear instability, modal analysis will be useful to
+analyse the growth of different wavenumbers, which is associated with
+the number of vortices initially formed by the RWI.
+
+%characterize 
+%azimuthal vortex numbers and overdensity values.
+
+The Fourier transform of the time-dependent surface density is
+\begin{align}\label{fouriertransform}
+  \Sigma_m(r,t) = \int_{0}^{2\pi}
+  \Sigma(r,\phi,t) \, \mathrm{e}^{-\mathrm{i}m\phi} \, \mathrm{d}\phi 
+\end{align} 
+where $m$ is the aziumuthal wave number. We define the growth rate
+$\gamma$ of the $m^\mathrm{th}$ component of the surface density
+through 
+% and $\sigma_m$ is the complex frequency
+%of the $m \mathrm{th}$ mode. Since $\sigma_m$ is a value dependent on
+%$r$, a more useful diagnostic is the mode growth rate defined as  
+\begin{align}\label{growth}
+  \frac{d \langle|\Sigma_m|\rangle_r }{dt}= \gamma \langle|\Sigma_m|\rangle_r 
+\end{align}
+where %$\gamma$ is the real part of the $\sigma_m$ and 
+$\langle|\cdot|\rangle_r$ denotes the average of the absolute value
+over a raidal region of interest. By using Eq.~\ref{growth} the growth
+rates of the unstable modes can be found from successive spatial
+Fourier transforms over an appropriate period of time.
 
 \subsubsection{Rossby number}
+The Rossby number
+\begin{align}
+  Ro = \frac{{\bf \hat{z}} \cdot \nabla \times \bv - \langle
+    {\bf\hat{z}} \cdot \nabla \times \bv
+    \rangle_{\phi}}{2\langle\Omega\rangle_{\phi}},  
+\end{align}
+is a dimensionless measure of relative vorticity. 
+% the ratio of interial to Corriolis forces. 
+Here $\langle \cdot \rangle_{\phi}$ denotes an azimuthal
+average. Values of $Ro<0$ correspond to anti-cyclonic rotation with
+respect to the background shear and thus can be used to identify
+vortices and quantify its intensity.
 

### 2014-09-19 paper/paper.tex
edited linear discussion

@@
 \author[Les and Lin]{Robert Les
-  $^1$\thanks{robert.les@mail.utoronto.ca} and Min-Kai Lin $^1$
-  \thanks{ mklin924@cita.utoronto.ca} \\ 
+  $^1$\thanks{robert.les@mail.utoronto.ca} and Min-Kai Lin $^{1,2}$
+  \thanks{ minkailin@email.arizona.edu} \\ 
 $^1$Canadian Institute for Theoretical Astrophysics,  
 60 St. George Street, Toronto, ON, M5S 3H8, Canada \\
+$^2$Department of Astronomy and Steward Observatory, University of
+Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA 
 }

### 2014-09-19 paper/results.tex
edited linear discussion

@@
   planet}
-%We first consider the linear stability of planetary gaps.  
 In this section, the planet is introduced at $t=20P_0$ and 
@@
 and velocity fields. At this point the planet has carved a partial
-gap. We then perturb the density field and continue to 
-evolve the disc. This procedure allows us to analyse the growth of
+gap. We then perturb the surface density for $r>r_p$ and continue to 
+evolve the disc. We impose  sinusoidal perturbations with 
+azimuthal wavenumbers $m\in[1,10]$. {\bf what are PERTMIN and PERTMAX values?}  
+This procedure allows us to analyse the growth of 
 non-axisymmetric modes associated with the gap, but without
-complications from non-axisymmetry arising directly from disk-planet
+complications from non-axisymmetry arising directly from disc-planet
 interaction.
 
-Simulations were resolved with grid divisions of $(N_r,N_{\phi})=(1024,2048)$ with $r_{\mathrm{out}}=25r_{\mathrm{in}}$ for time up to $200P_0$ allowing high resolution and detection of high $m$ structures.
-Cases of $\tilde{\beta}=0.1,1.0,10.0$ were analyzed to provide a range of fast, moderate, and slowly cooled disks.
+Note that these `planet-off' simulations are not linear stability
+calculations because since the cooling term in our energy equation
+restores the initial temperature profile corresponding to $H/r=0.05$,
+rather than the heated gap edge. Linear stability calculations and
+adiabatic cases will be further discussed in
+section~\ref{adiabatic_section}.
 
-%The goal of these simulations is to compliment the disk-planet
-%simluations presented later. 
-These `planet-off' simulations are not linear stability calculations
-since the cooling term in our energy equation restores the initial temperature profile corresponding to $H/r=0.05$, rather than the heated gap edge.
-Linear stability calculations and adiabatic cases will be further discussed in section~\ref{adiabatic_section}.
-
-%disk-planet leads to heating 
-%tcool tries to return it to the initial smooth profile (axisymmetric)
-%if tcool > trwi -> temp returns to original value. heated gap not
-%captured
-
-%There are two competing effects. Increasing
-%$t_\mathrm{cool}$ leads to more heating and hence a larger sound-speed
-%$c_s$, which favours instability \citep{li00} since the vortex
-%instability is pressre-driven. However, gap-opening requires $r_h>H$,
-%so increasing the sound-speed, and hence $H$, leads to a shallower
-%gap (see below) and expected to be more stable.
-
-%tcool -> return to initial, t=0 temp. profile
-%whereas the `basic state' is the gap profile with structured temp profile
-
-%more details on perturbation: m=1 to 10, amplitude?  
-%resolution?
+Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
+and we consider 
+%for a time up to $200P_0$ allowing high resolution and detection of high
+%$m$ structures.
+cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
+and slowly cooled discs.
 
 \subsection{Gap structure}
-%{\bf plots of the gap structure before instability sets in, e.g. at
-%  t=30. surface density perturbation, gen. vortensity perturbation,
-%  aspect-ratio. compare diff. beta. gaps become more shallow with
-%  increasing cooling time (but not significantly - not enough time for
-%  full gap formation). generalized vortensity: extrema smoothed out
-%  with increasing cooling - expect more stability. aspect-ratio:
-%  significant heating. increase in H so expect larger vortices
-%}
+%Since stability of the gap is of interest 
+We first examine the gap structure formed by planet-disc
+interaction as a function of the cooling time. The azimuthally-averaged 
+gap profiles are shown in Fig. \ref{intial1D} for varying
+$\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling)
+are deeper with steeper gradients at the gap edges. Faster cooling rates also 
+increase and decrease the surface density maxima and 
+minima increase, respectively. However, a clean gap does not form
+in this short time period.
 
-Since stability of the gap is of interest we analyze the intial state formed by the planet-disk interactions.
-The intial gap profiles after the averaging procedure are shown in Fig \ref{intial1D} for varying $\tilde\beta$.
-Gaps formed are shallower for lower $\tilde\beta$ (faster cooling) but never significant enough to form a clear gap in this short timeperiod.
-Quicker cooling rates also result in steeper density gradients at the gap boundry and both the density maxima and density minima correspondingly increase and decrease with lower $\tilde\beta$.
-For all cases the gap width is $\approx 5r_{hill}$ which is set by torque balances in the disk \citep{crida06}.
+%For all cases the gap width is $\approx 5r_{hill}$
+%which is set by torque balances in the disk \citep{crida06}.
 
- Generalized vortensity profiles show smoothening with larger cooling rates with the extrema becoming less pronounced.
-Charactereristicly for such disk-planet Rossby vortices the generalized vortensity minima correspond to density maxima in the disk.
-Larger $\tilde\beta$ values also as expected result in higher disk aspect ratios $H/r$, which is a value directly correlated to temperature in our model.
+Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
+i.e. higher temperatures. Heating mostly occur at the gap edges
+due to planet-induced spiral shocks. Decreasing the cooling time
+implies that this heat is retained in the disc. In the inviscid limit the gap
+opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
+\citep{crida06}. 
+% which is a value directly correlated to temperature in our model. 
+%For a planet to carve a gap in the surrounding material it has been
+%shown that $q>H^3$ is needed \citep{crida06}. 
+This indicates that for hotter discs (higher
+$h$), it becomes more difficult for a planet of fixed $q$ to open a
+gap. This explains the shallower gaps in surface density when
+$\tilde{\beta}$ is increased.
 
- For a planet to carve a gap in the surrounding material it has been shown that $q>H^3$ is needed \citep{crida06}.
-This indicates that for hotter disks, which have respectively higher $H$, it becomes more difficult by a cubic dependence for planet induced gaps to be formed.
-This is seen in our simulations as changes in $\tilde{\beta}$ result in changes of $h$ by less than $20\%$ yet gap depth almost doubles.
-Since generalized vortensity extrema are also correspondingly less pronounced and density gradients become smaller we expect that the stability of gaps to increase with larger $\tilde{\beta}$.
+%This is seen in our simulations as changes in $\tilde{\beta}$
+%result in changes of $h$ by less than $20\%$ yet gap depth almost
+%doubles.
+
+The important consequence of a heated gap edge is that the
+generalized vortensity profiles become smoother with increasing
+cooling times, with the extrema becoming less pronounced. Previous locally
+isothermal disc-planet simulations show the RWI associated with PV
+minima \citep{li05,lin10}. We can therefore expect the RWI to be associated with
+minima in the generalized vortensity (corresponding to local surface
+density maxima) in the non-isothermal case. Because the extrema are
+less sharp, the RWI is expected to be weaker and the gap to be more
+stable with longer cooling times.
+
+%Charactereristicly
+%for such disk-planet Rossby vortices the generalized vortensity
+%minima correspond to density maxima in the disk. 
+%Since generalized vortensity extrema are also
+% correspondingly less pronounced and density gradients become smaller
+% we expect that the stability of gaps to increase with larger
+% $\tilde{\beta}$.
 
 \begin{figure}
-  \includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
     2cm 0cm 0cm]{figures/compare_sigma}
-  \includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
-    2cm 0cm 1cm]{figures/compare_gvortensity}
-\includegraphics[width=.4\textwidth,clip=true,trim=0.5cm
-    0.5cm 0cm 1cm]{figures/compare_aspectratio}
-  \caption{Gap profiles at $t=30P_0$ for the intial partial gap opened before instability emerges for quick (solid line), moderate (dashed line), and slow cooling (broken line).
-Relative density pertibation (top), generalized vortensity pertibation (middle) and disk aspect ratio (bottom) are shown.
-\label{intial1D}}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+    2cm 0cm 1cm]{figures/compare_aspectratio}
+\includegraphics[width=\linewidth,clip=true,trim=0.5cm
+    0.5cm 0cm 1cm]{figures/compare_gvortensity}
+  \caption{Gap profiles at $t=30P_0$ for the intial partial gap opened
+    before instability emerges for fast (solid), moderate
+    (dashed), and slow cooling (dashed-dot). The relative surface density
+    pertibation (top), disc aspectratio (middle) and generalized
+    vortensity pertibation (bottom) are shown. \label{intial1D}}  
 \end{figure}
@@
   \centering
-  \caption{Dominant mode and growth rates for $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling) values during `planet-off' simulations \label{modetable}}
+  \caption{Dominant mode and growth rates for
+    $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling)
+    values during `planet-off' simulations \label{modetable}} 
   \hfill
@@
       \toprule
-      m & $\gamma10^2/\Omega(r_p)$ \\
+      $m$ & $10^2\gamma/\Omega(r_p)$ \\
       \midrule
@@
 
-\subsection{Linear modes and growth rates}\label{linear}
+{\bf
+\subsection{Axisymmetric stability}
+Check that the gap profiles are stable against axisymmetric
+stability. Can use analytical criteria given in section 3.1 of Li et
+al (2000, ApJ, 533, 1023). No plot needed, just statements. 
+}
+
+\subsection{Non-axisymmetric instability}\label{linear}
 %{\bf present `planet-off' simulations. one `fourier mode v.s. time'
@@
 %}
-After the azimuthal averaging, the small pertibation excites rapidly growing instabilities by the RWI.
-We analyticly charcterize these modes with quantities $m$ and $\gamma$ as defined by Eq.~\ref{fouriertransform} and Eq.~\ref{growth}.
-Table \ref{modetable} contains the results of the mode growth rates for a variety of cooling rates during this semi-linear phase.
-Results indicate that as $\tilde{\beta}$ was increased from $ 0.1\rightarrow10.0$ the dominant azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the respective growth rate decreased from $ \gamma/\Omega(r_p)=0.079 \rightarrow 0.017$.
-Snapshots of the instability in $r-\phi$ plane for the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
 
-These `planet-off' simulations, which are reminiscient of linear stability measurements, show that stability of gap edges increase with longer cooling rates (ie hotter gaps).
-This is as expected from the intial gap profiles and the gap forming criteria discussed in the previous section.
+%After the azimuthal averaging, the small pertibation excites rapidly
+%growing instabilities by the RWI.
+
+In all cases we observe growth of non-axisymmetric structures after the
+planet potential is switched off and the system subject to random
+perturbations. An example is shown in Fig. \ref{linearmodes} for 
+$\tilde{\beta}=10$. We charcterize these
+modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
+Eq.~\ref{fouriertransform}---\ref{growth}. Table \ref{modetable}
+lists the growth rates for the three cooling
+times, measured during linear growth.
 
@@
     0cm 0cm 0cm]{figures/linear_stability}
-  \caption{Log plot of the azimuthal Fourier modes of disk surface density non-dimenionlized by the background mode during the `planet-off' simulations for $\tilde{\beta}=10$.
-Colors correspond to different $m$ values.
-The $m=3$ component of the fourier transform is the fast growing mode during this growth phase of the instability and has corresponding $\gamma=0.017\Omega(r_p)$.
-\label{linearmodes}}
+  \caption{Evolution of azimuthal Fourier modes of disc surface
+    density, non-dimenionlized by the axisymmetric mode for the
+    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
+    to different $m$ values. The $m=3$ component is the fastest growing
+    mode during linear growth and has a corresponding
+    $\gamma=0.017\Omega(r_p)$. {\bf is the $\Sigma_0$ at
+      t=0?} \label{linearmodes}}  
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Table \ref{modetable} show that as
+$\tilde{\beta}$ was increased from $ 0.1\rightarrow10$ the dominant
+azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the
+respective growth rate decreased from $ \gamma/\Omega(r_p)=0.079
+\rightarrow 0.017$. However, despite two orders of magnitude increase in the
+cooling time, the instability remains dynamical with growth time
+$\lesssim 10P_0$. Snapshots of the instability in $r-\phi$ plane for
+the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
+
+These `planet-off' simulations show that gap edges become more stable with
+longer cooling times. This is expected because larger $\tilde{\beta}$
+result in hotter gap profiles at $t=30P_0$ with less pronounced
+generalized vortensity minima. Stabilization with increased
+cooling time is therefore due to a smoother basic state for the
+instability, as it is more difficult for the planet to open a gap if
+the disc is allowed to heat up.
+
+% which result in hotter gaps since the heating
+%due
+%This is as expected from the
+%intial gap profiles, because 
+% and the gap forming criteria discussed in the
+%previous section.
 
@@
   }  
-  \caption{Radial plots of the generalized vortensity pertibation for cases of $\tilde{\beta}=0.1,1.0,10.0$ (left,middle,right) during the linear phases of growth.
-The number of visible vortices decrease as $\tilde{\beta}$ increases and snapshot times correspondingly increase to capture the longer growth phases of the instabilities.
-\label{2Dlinear}}
+  \caption{Generalized vortensity perturbation (relative to $t=0$) for
+    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
+    the growth of non-axisymmetric modes. The planet potential has
+    been switched off.  The number of vortices
+    decrease as $\tilde{\beta}$ increases. Note that snapshots are
+    taken later because it takes longer for the vortices to grow and
+    become visible with increasing cooling time. 
+    %    correspondingly increase to capture the longer growth phases of
+    %    the instabilities. 
+    \label{2Dlinear} {\bf I'm curious to know if plots clearer if
+      perturbations are measured relative to t=30? remember t=30 is
+      our reference state for instability. But we'll probably keep
+      these for consistency (later non-linear plots are all measure
+      with respect to t=0).}} 
 \end{figure*}
 
-\subsubsection{Nearly adiabatic disks}
+\subsubsection{Nearly adiabatic discs}
 \label{adiabatic_section}
@@
 %}
-During the `planet-off' simulations the analysis is not truly a linear stability measurement as we do not capture the effects of a heated disk on the instability growth.
-Since $t_{\mathrm{cool}}<t_{\mathrm{instability}}$ the disk manages to cool back to its intial value of $h=0.05$ before the instability emerges and thus we do not have a steady background state to form a linear stability problem.
-To compare our 'planet-off' analysis with a true linear stability analysis we ran a simulation  with $\tilde{\beta}=100.0$ which corresponds with an almost adiabtaic case.
-In this simulation the cooling rate is slow enough that the gap temperature profile changes only marginally over the timescale of the growth of the instability.
-The results show that the linear stability of the $\tilde{\beta}=100.0$ cases almost exactly equivilant to the case of $\tilde{\beta}=10.0$ in terms of mode growth rates and gap profiles.
-Since the disk heats up only to values $h\simeq0.06$ in the adiabatic case, the difference in growth rates between this and the original disk temperature of $h=0.05$ changes almost insignificantly as shown by linear theory \citep{li00}.
-This indicate that a continuously hot gap edge is not as important for development of vortices as the heating effects on the formation of the intial gap state.
+The above `planet-off' simulations are not formally linear
+stability calculations, because the cooling time is shorter
+than the instability growth time, 
+%
+%as we do not capture the effects of a heated
+%disk on the instability growth. 
+$t_c<\gamma^{-1}$.  
+Thus the disk cools back to its intial value of $h=0.05$
+before or during the instability growth, implying we do not have a
+steady basic state to formulate a standard linear stability
+problem. %This means the disc temperture of $h=0.05$ in the the above cases
 
-%% We first describe results for weakly self-gravitating
-%% discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
-%% gravitationally stable \citep{lin11b}. We also impose a dimensionless
-%% kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
-%% vortex instablity \citep{valborro07}. These parameter values
-%% are typical for disc-planet simulations and yield stable/steady 
-%% gap profiles. This will be useful reference cases to  
-%% understand the unstable cases considered later.
+In order to perform a proper linear stability analysis and capture the
+effect of a heated gap edge during instability growth, we ran a simulation  with
+$\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
+%To compare our 'planet-off' analysis with a true
+%linear stability analysis 
+In this simulation the cooling rate is slow enough that the gap 
+temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
+only marginally over the instability growth timescale. %heat is
+                                %retained/maintained.
 
-%% Fig. \ref{lvisc_steady_gap} shows the steady state gap profiles in terms of the relative
-%% surface density and aspect-ratio at $t=200P_0$. Three levels of cooling are considered:
-%% $\tbeta=0.1,\,1.0$ and $10.0$ (i.e. $t_c\Omega_k\simeq2.4,\,24$ and $240$). We refer to them as
-%% fast, moderate and slow cooling cases, respectively.
+We find very similar gap profiles and mode growth rates for
+$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
+values $h\simeq0.06$ in the nearly adiabatic case. This is close to
+the original temperature of $h=0.05$, so growth rates are not expected
+to change significantly according to the linear theory presented in
+\citep{li00}.
+
+According to \cite{li00}, increasing $h$ increases linear growth rates
+of the RWI because it is pressure-driven. However, in the case 
+of disc-planet interaction, increasing $h$ has a stabilizing effect
+through the gap profile because it results in smoother gap
+edges. The fact that we observe smaller growth rates as $h$ is
+increased indicate that for planetary gaps, the importance of $h$ on
+the \emph{linear} RWI is through the setting the gap profile (i.e. basic
+state for the instability).
+
+%results in shallower gaps
+%which is expected to be
+
+%the difference in growth rates between this and the
+%original disc temperature of $h=0.05$ changes almost insignificantly
+%as shown by linear theory \citep{li00}. 
+%Although linear
+%theory predicts higher growth rates with increasing $h$
+
+%This indicate that a
+%continuously hot gap edge is not as important for development of
+%vortices as the heating effects on the formation of the intial gap
+%state.
+
+{\bf
+If there's time, repeat planet off simulations but keep the planet on
+until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
+are generally deeper (including the nearly adiabatic case). Although
+with the planet instability may already appear by $t=40P_0$, the
+azimtuhal average performed prior to perturbations will erase it. 
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Long term simulation of edge vortices were done for $\tilde{\beta}=0.1,0.5,1.0,5.0,10.0$ up to a total of $2000P_0$.
-Planets were left free to interact dynamiclly with the disk and vortices after $t=30P_0$ as apposed to previous section.
-These simulations were less resolved with $(N_r,N_{\phi})=(512,1024)$ in order for them to be computationally economical.
-An open boundry condition was used in the outer edge of $r_{\mathrm{out}}=45r_{\mathrm{in}}$ allowing significant room for linblad resonances of the vortices.
+Long term simulation of edge vortices were done for
+$\tilde{\beta}=0.1,0.5,1.0,5.0,10.0$ up to a total of
+$2000P_0$. Planets were left free to interact dynamiclly with the disk
+and vortices after $t=30P_0$ as apposed to previous section. These
+simulations were less resolved with $(N_r,N_{\phi})=(512,1024)$ in
+order for them to be computationally economical. An open boundry
+condition was used in the outer edge of
+$r_{\mathrm{out}}=45r_{\mathrm{in}}$ allowing significant room for
+linblad resonances of the vortices.
 
-After the semi-linear growth of vortices, vortex merger effects take hold and by $150P_0$ for all $\tilde\beta$ values, only $m=1$ vortex modes exist.
-Plots of the amplitude of the $m=1$ mode for the different $\tilde\beta$ cases are shown in Fig~\ref{lifetimeplot}.
-These instabilities stay in quasi-steady states for $>1000P_0$ in most cases, in which they grow in intensity and reaching overdenisities almost 10 times the intial local density.
-The growth of these vortices are supplied by the continuous generation of vorticity by the planet-disk interactions.
-Eventually the $m=1$ amplitudes are also characterized by sudden large drops of vortex intensity which happen over timescales of $50P_0$.
-After these drops the instability is never seen to reform at such large over densities and density bumps are never larger than the planet induced shocks.
-We designate the time of these amplitude drops as the lifetime of the quasi-steady vortex.
+After the semi-linear growth of vortices, vortex merger effects take
+hold and by $150P_0$ for all $\tilde\beta$ values, only $m=1$ vortex
+modes exist. Plots of the amplitude of the $m=1$ mode for the
+different $\tilde\beta$ cases are shown in
+Fig~\ref{lifetimeplot}. These instabilities stay in quasi-steady
+states for $>1000P_0$ in most cases, in which they grow in intensity
+and reaching overdenisities almost 10 times the intial local
+density. The growth of these vortices are supplied by the continuous
+generation of vorticity by the planet-disk interactions. Eventually
+the $m=1$ amplitudes are also characterized by sudden large drops of
+vortex intensity which happen over timescales of $50P_0$. After these
+drops the instability is never seen to reform at such large over
+densities and density bumps are never larger than the planet induced
+shocks. We designate the time of these amplitude drops as the lifetime
+of the quasi-steady vortex.
 
-Rossby values have dynamics that are coupled with the lifetime and $m=1$ amplitude of the vortices.
-As the vortex forms it has a charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic motion.
-As the $m=1$ amplitude grows the relative spin of the vortex continuously grows along with it, reaching values of $Ro\approx-0.4$.
-As the vortex dies off as described above the associated Rossby number of the vortex accordingly quickly approaches 0.
+Rossby values have dynamics that are coupled with the lifetime and
+$m=1$ amplitude of the vortices. As the vortex forms it has a
+charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic
+motion. As the $m=1$ amplitude grows the relative spin of the vortex
+continuously grows along with it, reaching values of
+$Ro\approx-0.4$. As the vortex dies off as described above the
+associated Rossby number of the vortex accordingly quickly approaches
+0.
 
-The sudden dissipation of these vortices seem to be a result of the vortices shocking the system.
-Measurements of the Mach number near the vortices show large increases in value when the vortex is dissipating which indicates that the system is more susceptible to shocks.
-Also gradients of surface density are seen to spike up in wake like features around the vortex during dissipation which is also indicative of shocks.
-Of note is also that the large overdensity the vortex creates during it's quasi-steady state distorts the background disk.
-So as the vortex grows in intensity it alters the intial gap structure that formulated its growth.
+The sudden dissipation of these vortices seem to be a result of the
+vortices shocking the system. Measurements of the Mach number near the
+vortices show large increases in value when the vortex is dissipating
+which indicates that the system is more susceptible to shocks. Also
+gradients of surface density are seen to spike up in wake like
+features around the vortex during dissipation which is also indicative
+of shocks. Of note is also that the large overdensity the vortex
+creates during it's quasi-steady state distorts the background
+disk. So as the vortex grows in intensity it alters the intial gap
+structure that formulated its growth.
 
-Contradictory to our `planet-off' simulations, vortex lifetimes did not monotonicly depend on intial edge stability and from our simulations we find that there is $t_{\mathrm{coo}l}$ which optimizes the vortex lifetime as seen in Fig~\ref{lifetimeplot}.
-Increasing $\tilde\beta$ correspondingly increased the lifetime of the vortex up to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that lasted beyond the simulation time of $2000P_0$.
-Cooling rates with $\tilde\beta>5.0$ showed lifetimes that were well below the max lifetime.
-By analyzing the aspect ratio of the corotation region around the vortex we find that longest lifetime $\tilde\beta=5.0$ disk corresponds to $h\approx0.06$ which is a result similarily found by \citet{fu14} to be the optimum lifetime for long-term simulations of isothermal disks with varying intial temperature profiles $h$.
+Contradictory to our `planet-off' simulations, vortex lifetimes did
+not monotonicly depend on intial edge stability and from our
+simulations we find that there is $t_{\mathrm{coo}l}$ which optimizes
+the vortex lifetime as seen in Fig~\ref{lifetimeplot}. Increasing
+$\tilde\beta$ correspondingly increased the lifetime of the vortex up
+to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that
+lasted beyond the simulation time of $2000P_0$. Cooling rates with
+$\tilde\beta>5.0$ showed lifetimes that were well below the max
+lifetime. By analyzing the aspect ratio of the corotation region
+around the vortex we find that longest lifetime $\tilde\beta=5.0$ disk
+corresponds to $h\approx0.06$ which is a result similarily found by
+\citet{fu14} to be the optimum lifetime for long-term simulations of
+isothermal disks with varying intial temperature profiles $h$.
 
-Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in the the non-montonic form of the lifetime.
-As the $\tilde\beta$ increases so does the stability of the gap edge as seen in section~\ref{linear}, however as $\tilde\beta$ increases so does the $c_{\mathrm{iso}}$ which is known from theory to effect the growth of instability \citep{li00}.
-Thus increasing $\tilde\beta$ increases the sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to a certain point in which the effects of the more stable gap edge becomes a significant effect.
-The correlation of growth rate and sound speed is non-negligible as in previous `planet-off' case as the disk now can reach values of $h\simeq0.08$ for very slow cooling and the importance of the effect can be imagined to accumulate over the now considerably longer timescale.
-These effects dictate when then the vortex reaches the critical value in which it shocks the system that subsequently causes disspation and hence the lifetime fo the vortex.
+Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
+the the non-montonic form of the lifetime. As the $\tilde\beta$
+increases so does the stability of the gap edge as seen in
+section~\ref{linear}, however as $\tilde\beta$ increases so does the
+$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
+instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
+sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
+a certain point in which the effects of the more stable gap edge
+becomes a significant effect. The correlation of growth rate and sound
+speed is non-negligible as in previous `planet-off' case as the disk
+now can reach values of $h\simeq0.08$ for very slow cooling and the
+importance of the effect can be imagined to accumulate over the now
+considerably longer timescale. These effects dictate when then the
+vortex reaches the critical value in which it shocks the system that
+subsequently causes disspation and hence the lifetime fo the vortex.
 
@@
     0cm 0cm 1cm]{figures/longterm_stability}
-  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$ background mode for longterm non-linear simulations corrsponding to vortex lifetime.
-The longest living mode correpsonds to $\tilde\beta=5.0$ (the red line).
-\label{lifetimeplot}}
+  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$
+    background mode for longterm non-linear simulations corrsponding
+    to vortex lifetime. The longest living mode correpsonds to
+    $\tilde\beta=5.0$ (the red line). \label{lifetimeplot}} 
 \end{figure}
@@
   \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
-  \caption{Cartesian plot of relative density pertibation for $\tilde\beta=0.1$ vortex case during the quasi-steady state.
-The large scale non-axisyemetric $m=1$ overdensity can be seen.
-\label{Vortex2D}}
+  \caption{Cartesian plot of relative density pertibation for
+    $\tilde\beta=0.1$ vortex case during the quasi-steady state. The
+    large scale non-axisyemetric $m=1$ overdensity can be
+    seen. \label{Vortex2D}} 
 \end{figure}

### 2014-09-22 paper/model.tex
results section edited by mkl

@@
 \subsection{Disc model and initial condition}
-The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with
-$r_\mathrm{out}=25\mathrm{in}$ {\bf check: if all simulations use the
-  same domain, state it here} and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
+The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ and
+$\phi\in[0,2\pi]$. The initial disc is axisymmetric with  
 surface density profile

### 2014-09-22 paper/results.tex
results section edited by mkl

@@
 Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
+with $r_\mathrm{out}=25r_\mathrm{in}$, 
 and we consider 
@@
 cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
-and slowly cooled discs.
+and slowly cooled discs.  {\bf boundary condition for these sims?}
 
@@
 values $h\simeq0.06$ in the nearly adiabatic case. This is close to
-the original temperature of $h=0.05$, so growth rates are not expected
-to change significantly according to the linear theory presented in
+the original temperature of $h=0.05$, so linear growth rates are not expected
+to change significantly %according to the linear theory presented in
 \citep{li00}.
@@
 {\bf
-If there's time, repeat planet off simulations but keep the planet on
+If there's time, do low resolution planet off simulations but keep the planet on
 until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
@@
 
-Long term simulation of edge vortices were done for
-$\tilde{\beta}=0.1,0.5,1.0,5.0,10.0$ up to a total of
-$2000P_0$. Planets were left free to interact dynamiclly with the disk
-and vortices after $t=30P_0$ as apposed to previous section. These
-simulations were less resolved with $(N_r,N_{\phi})=(512,1024)$ in
-order for them to be computationally economical. An open boundry
-condition was used in the outer edge of
-$r_{\mathrm{out}}=45r_{\mathrm{in}}$ allowing significant room for
-linblad resonances of the vortices.
+We now consider long term simulations of gap-edge vortices for
+$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
+throughout.  
+% up to a total of
+%$2000P_0$. 
+%Planets were left free to interact dynamiclly with the disk
+%and vortices after $t=30P_0$ as apposed to previous section.  
+We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
+in order for these simulations to be computationally feasible. We also use a larger
+disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
+effects on vortex evolution. We apply an open boundary at
+$r_\mathrm{out}$. {\bf mention lower resolution preliminary runs -
+  similar results to those presented here (?). want to demonstrate convergence}
 
-After the semi-linear growth of vortices, vortex merger effects take
-hold and by $150P_0$ for all $\tilde\beta$ values, only $m=1$ vortex
-modes exist. Plots of the amplitude of the $m=1$ mode for the
-different $\tilde\beta$ cases are shown in
-Fig~\ref{lifetimeplot}. These instabilities stay in quasi-steady
-states for $>1000P_0$ in most cases, in which they grow in intensity
-and reaching overdenisities almost 10 times the intial local
-density. The growth of these vortices are supplied by the continuous
-generation of vorticity by the planet-disk interactions. Eventually
-the $m=1$ amplitudes are also characterized by sudden large drops of
-vortex intensity which happen over timescales of $50P_0$. After these
-drops the instability is never seen to reform at such large over
-densities and density bumps are never larger than the planet induced
-shocks. We designate the time of these amplitude drops as the lifetime
-of the quasi-steady vortex.
+%An open boundry
+%condition was used in the outer edge of
+% allowing significant room for
+%linblad resonances of the vortices.
 
-Rossby values have dynamics that are coupled with the lifetime and
-$m=1$ amplitude of the vortices. As the vortex forms it has a
-charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic
-motion. As the $m=1$ amplitude grows the relative spin of the vortex
-continuously grows along with it, reaching values of
-$Ro\approx-0.4$. As the vortex dies off as described above the
-associated Rossby number of the vortex accordingly quickly approaches
-0.
+\subsection{Generic evolution}
 
-The sudden dissipation of these vortices seem to be a result of the
-vortices shocking the system. Measurements of the Mach number near the
-vortices show large increases in value when the vortex is dissipating
-which indicates that the system is more susceptible to shocks. Also
-gradients of surface density are seen to spike up in wake like
-features around the vortex during dissipation which is also indicative
-of shocks. Of note is also that the large overdensity the vortex
-creates during it's quasi-steady state distorts the background
-disk. So as the vortex grows in intensity it alters the intial gap
-structure that formulated its growth.
-
-Contradictory to our `planet-off' simulations, vortex lifetimes did
-not monotonicly depend on intial edge stability and from our
-simulations we find that there is $t_{\mathrm{coo}l}$ which optimizes
-the vortex lifetime as seen in Fig~\ref{lifetimeplot}. Increasing
-$\tilde\beta$ correspondingly increased the lifetime of the vortex up
-to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that
-lasted beyond the simulation time of $2000P_0$. Cooling rates with
-$\tilde\beta>5.0$ showed lifetimes that were well below the max
-lifetime. By analyzing the aspect ratio of the corotation region
-around the vortex we find that longest lifetime $\tilde\beta=5.0$ disk
-corresponds to $h\approx0.06$ which is a result similarily found by
-\citet{fu14} to be the optimum lifetime for long-term simulations of
-isothermal disks with varying intial temperature profiles $h$.
-
-Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
-the the non-montonic form of the lifetime. As the $\tilde\beta$
-increases so does the stability of the gap edge as seen in
-section~\ref{linear}, however as $\tilde\beta$ increases so does the
-$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
-instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
-sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
-a certain point in which the effects of the more stable gap edge
-becomes a significant effect. The correlation of growth rate and sound
-speed is non-negligible as in previous `planet-off' case as the disk
-now can reach values of $h\simeq0.08$ for very slow cooling and the
-importance of the effect can be imagined to accumulate over the now
-considerably longer timescale. These effects dictate when then the
-vortex reaches the critical value in which it shocks the system that
-subsequently causes disspation and hence the lifetime fo the vortex.
+The linear growth of the RWI and vortex-formation is followed by 
+vortex merging. For all $\tilde\beta$ values, only one vortex remains
+by $150P_0$.
+%i.e. only the $m=1$ mode remains. 
+{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
+  taken to reach one vortex as a function of cooling}
+%by $150P_0$ for all , only $m=1$ vortex
+%modes exist. 
+The evolution of the amplitude of the $m=1$ mode in surface density is
+shown in Fig~\ref{lifetimeplot} for   
+different $\tilde\beta$. {\bf should state (also for planet-off runs)
+  what the radial range is used for averaging the Fourier amplitudes}
 
@@
 
+In all cases the system remains in a quasi-steady state for
+$\gtrsim500P_0$ with a single vortex circulating 
+the outer gap edge at the local Keplerian  
+frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
+surface density perturbation in this state. During this stage, the 
+vortex grows in intensity. As the vortex forms it has a
+charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic
+motion. As the $m=1$ amplitude grows the relative spin of the vortex
+continuously grows along with it, reaching values of
+$Ro\approx-0.4$, corresponding to surface density almost 10 times the
+initial value. {\bf for what cooling are these values quoted for? do
+  these values depend on cooling?} Vortex growth is attributed to
+the continuous generation of vorticity by the planet-disc interactions. 
+{\bf is the evidence for this the fact that growth is not observed in
+  planet-off simulations?}
+
+{\bf note that there
+  are two origins for the surface density boost relative to $t=0$: 1) gap
+  formation, which increases the surface density at the location of
+  the instability (since they happen at surface density max., and 2)
+  vortex instability). what is the surface density boost relative to
+  the gap profile?}
+
+Interestingly, in most cases we observe a sudden drop in 
+the $m=1$ amplitude within the simulation time, which occurs over
+timescales of $50P_0$, i.e. much shorter than the duration of the quasi-steady
+state. {\bf is this dissipation timescale independent of cooling?}.  
+%Eventually
+%the $m=1$ amplitudes are also characterized by 
+After these characteristic drops, the instability is never seen to
+reform at such large overdensities and surface density bumps are never larger
+than that due to the planetary wakes. We designate the time of these
+amplitude drops as the lifetime 
+of the quasi-steady vortex.  
+{\bf question: does instability
+  actually happen again? if it does, but just weaker, that may be
+  because the first vortex has smoothed out the gap edge, so
+  instabilities following that would be weaker.}
+
 \begin{figure}
   \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
-  \caption{Cartesian plot of relative density pertibation for
-    $\tilde\beta=0.1$ vortex case during the quasi-steady state. The
-    large scale non-axisyemetric $m=1$ overdensity can be
-    seen. \label{Vortex2D}} 
+  \caption{Relative surface density perturbation for the
+    $\tilde\beta=0.1$ case during quasi-steady state with a single
+    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
+    similar. {\bf is this plot truncated? in the text it says the
+      domain goes to $r=45$?}
+    %The large scale non-axisyemetric $m=1$ overdensity can be
+    %seen. 
+    \label{Vortex2D}} 
 \end{figure}
 
-%\begin{figure}
-%   \includegraphics[scale=.42]{figures/stability_vis6betalow.ps}
-%   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=0.1$. }
-% \label{stability_vis6lowb)}
-% \end{figure}
+{\bf
+\subsection{Additional analysis on vortex dissipation}
+Conjecture: vortex death due to vortex shocking the system and
+dissipating its energy. discuss one case. here we want to explain or
+suggest why vortices die} 
+%Rossby values have dynamics that are coupled with the lifetime and
+%$m=1$ amplitude of the vortices.
 
-%\begin{figure}
-%   \includegraphics[scale=.42]{figures/stability_vis6betamed.ps}
-%   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$. }
-% \label{stability_vis6medb)}
-% \end{figure}
-%Same as \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$.
+As the vortex dies off as described above its associated Rossby number
+of quickly approaches $0$. Gradients of surface density are seen to
+spike up in wake-like features around the vortex during dissipation. 
+{\bf show
+  this. e.g. vortex before, during and after death. maybe show
+  snapshots of surface density perturbation (vortex) compared with
+  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
+  changes and shocks develop prior ot death. state whether this is
+  generic for all coolings}.  
+In addition, we also observed large increases in the Mach number near
+the vortex during dissipation, which indicate susceptibility to
+shocks. {\bf show mach number plot, or quote values. relevant mach
+  number is fluid flow speed with respect to 
+  vortex bulk motion around the disc. does the typical mach number
+  near the vortex just before its death depend on cooling?} 
+Thus, the sudden drop in vortex amplitude may  
+be a result of the development of associated spiral shocks dissipating
+its energy \citep[cf.][who 
+  suggested damping due to planet-induced shocks]{fu14}.
 
-%we checked that open bc doesn't affect gap structure
-%checked that t=200 plots are similar to t=100
+{\bf note that fu et al suggested vortex damping by the planet
+  shocks. here, we are suggesting vortex dissipates energy by shocking
+  the disk. our evidence is that we observe vortex growth before it
+  dissipates, and spiral shocks associated with the vortex always
+  develop before vortex death. once it starts shocking the disc it
+  loses energy. if the planet-induced shocks were responsible, death
+  should be more gradual. can we think of other evidences that it's
+  not the planet shock that's responsible?                    
+}
 
-%surface density plot
+Vortex growth during the quasi-steady state also acts to smooth out
+the gap edge. 
+%Of note is also that the large overdensity the vortex
+%creates during it's quasi-steady state distorts the background
+%disk. 
+{\bf show this (smoothening of gap edge due to vortex growth),
+  e.g. general vortensity evolution or surface density 
+  evolution}. 
+This is because the vortex has an associated viscosity {\bf quote a
+  typical value of alpha viscosity associated with vortex}. This acts
+against gap-opening by the planet, so that the condition for the RWI
+becomes less favourable as the vortex grows. {\bf self-limiting
+  behaviour?}
 
-%aspect-ratio plot
+\subsection{Vortex lifetimes as a function of cooling rate}
 
-%general vortensity plot
+{\bf describe dependency on
+cooling. Do we have enough data to have a `lifetime v.s. beta'
+  plot? here we want to explain why vortex lives longer with longer
+  cooling, but very long cooling results in shorter lifetimes. 
+  Consider the following quantities as a function of cooling:
+  initial strength of the single vortex; growth rate of single vortex
+  strength from its formation to just before death; ampiltude of
+  vortex just before death; disc temperature in the vortex region.
 
-%% \begin{figure}
-%%   \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/compare_profiles_dens020.ps}\\
-%%   \includegraphics[scale=.42]{figures/compare_profiles_h020.ps}
-%%   \caption{Steady-state gap profiles in a low mass viscous disc. The
-%%     surface density perturbation (top) and disc aspect-ratio (bottom)
-%%     are shown as a function of the cooling parameter:  
-%%     $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
-%%     moderate cooling) and $\tbeta=10$ (dashed, slow
-%%     cooling). \label{lvisc_steady_gap}}  
-%% \end{figure}
+  we expect the initial $m=1$ vortex (after merging) to be weaker with
+  increasing cooling time because the instability is weaker. (see
+  meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
+  amplitude after linear growth is correlated with the linear growth
+  rate).
 
-%% \begin{figure}
-%%   \includegraphics[scale=.42]{figures/compare_profiles_gvort020.ps}
-%%   \caption{Gap structure in a low mass viscous disc, in terms of the
-%%     perturbed generalized vortensity as a function of the cooling
-%%     parameter. \label{lvisc_steady_gvort}} 
-%% \end{figure}
+  suppose the vortex shocks when it reaches a certain amplitude (which
+  probably increases with longer cooling rate because it's more
+  difficult to shock a hotter disc). as we
+  increase the cooling time, the initial $m=1$ vortex is weaker, it
+  also grows more slowly because gap-opening is opposed by a hotter
+  disc. the above implies that it takes longer for the vortex reach
+  the shock-inducing amplitude and dissipate.
 
-%%\section{Gaps in massive discs (MKL)}
-%% We first examine disc models with $Q_o=1.5$. We set the physical
-%% viscosity $\hat{\nu}=10^{-9}$, so that the only energy source is
-%% through shock-heating (via artificial viscosity) and the $\mathcal{C}$
-%% function when $e\Sigma<e_i\Sigma_i$. The numerical resolution is
-%% $N_r\times N_\phi = 512\times 1024$.
+  but the above is not consistent with non-monotonic behaviour, which
+  predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
+  because we expect the initial vortex to be weakest (check) which
+  means it should take the longest to grow and shock. check
+  whether there's anything special about this case. one possibility is
+  that vortex growth during the quasi-steady state is actually faster
+  because of the raised temperature (li et al 2000 show increased
+  temperature favours RWI - possibiliy evidenced here by a stronger
+  vortex before death than other cases)
 
-%% %Since we are primarily concerened with gap stability, it is important
-%% %to first examine the gap structured opened by the planet as a function
-%% %of cooling. 
-%% Fig. \ref{gvort1d_q1d5} compares the gap structure in terms of the GV profile for three
-%% levels of cooling: $\tilde{\beta}=0.1$, $\tilde{\beta}=1$ and
-%% $\tbeta=10$. For convenience we will refer to these cases as fast,
-%% intermediate and slow cooling, respectively. The snapshot is taken at
-%% $t=30P_0$, just after the planet is fully introduced.
+  perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
+  non-monotonic behaviour
+}
 
-%% In terms of the GV profile, the qualitative features of the planetary
-%% gap --- localized GV extrema at the gap edges --- remain unchanged
-%% despite two orders of magnitude difference in the cooling rate.
+Fig. \ref{lifetimeplot} shows that increasing $\tilde{\beta}$
+increases the lifetime of the vortex up to a critical 
+value $\tilde{\beta}=5.0$ for which the vortex lasted beyond the
+simulation time of $2000P_0$. This is consistent with vortex
+dissipation when it induce shocks. Our `planet-off' simulations
+indicate weaker vortices are formed initially with increased cooling
+times. After merging, the growth of the single vortex, mediated by disc-planet
+interaction, is expected to be slower with increased cooling time
+because gap-opening becomes more difficult in a hotter
+disc. Furthermore, the increased sound-speed suggest the vortex should
+reach larger amplitudes in order to induce shocks. 
+These expectations
+imply that it takes longer for the vortex to induce shocks and
+dissipate, and hence longer lifetimes with increasing
+$\tilde{\beta}$.
 
-%% \begin{figure}
-%% %  \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/dens1d_q1d5_003_global.ps}\\
-%% %  \includegraphics[scale=.42]{figures/gvort1d_q1d5_003_global.ps}
-%%   \caption{Gap structure in terms of the perturbed surface density
-%%     (top) and perturbed generalized
-%%     vortensity (bottom), as a function of the cooling parameter:  
-%%     $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
-%%     intermediate cooling) and $\tbeta=10$ (dashed, slow
-%%     cooling). \label{gvort1d_q1d5}} 
-%% \end{figure}
+However, Fig. \ref{lifetimeplot} shows that increasing the cooling
+time further to $\tilde{\beta}=10$ actually results in a
+\emph{shorter} vortex lifetime. Notice for $\tilde{\beta}=10$ the vortex growth 
+during the quasi-steady state is faster than for
+$\tilde{\beta}=5$. This may be attributed to the higher temperature
+reached in the former case ($h\simeq 0.08$). {\bf temperature for
+  $\tilde{\beta}=5$?}, since \cite{li00} showed that the RWI grows
+faster with increased disc temperature. At long cooling times, this
+effect may dominate over the weaker gap-opening effect at higher $h$.
 
-%% \begin{figure*}
-%% %  \includegraphics[scale=.55,clip=true,trim=0cm 0cm 0cm
-%% %    0cm]{figures/noniso0_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
-%% %    0cm]{figures/noniso1_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
-%% %    0cm]{figures/noniso2_HR_dens004}
-%%   \caption{Gap instability in the heavy disc model, as a function of
-%%     the cooling parameter: $\tbeta=0.1$ (left), $\tbeta=1$ (middle)
-%%     and $\tbeta=10$ (right). The relative surface density
-%%     perturbation is shown. \label{polarxy_q1d5}} 
-%% \end{figure*}
+Thus, unlike our `planet-off' simulations we find that there is
+a cooling time $t_c$ which maximises the vortex lifetime. {\bf to make
+this comparison we need to discuss vortex lifetimes for planet-off
+simulations as well} 
+This
+non-monotonic behaviour was also reported by \cite{fu14}, who
+performed isothermal disc-planet simluations with different fixed
+aspectratios.   
+%Cooling rates with
+%$\tilde\beta>5.0$ showed lifetimes that were well below the max
+%lifetime. 
+By analyzing the region
+around the vortex {\bf define this region} we find the
+$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
+$h\approx0.06$. This value of $h$ was also found to maximize vortex
+lifetimes in the isothermal simulations of \cite{fu14}.
+
+%be the optimum lifetime for long-term simulations of
+%isothermal disks with varying intial temperature profiles $h$.
+
+%Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
+%the the non-montonic form of the lifetime. As the $\tilde\beta$
+%increases so does the stability of the gap edge as seen in
+%section~\ref{linear}, however as $\tilde\beta$ increases so does the
+%$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
+%instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
+%sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
+%a certain point in which the effects of the more stable gap edge
+%becomes a significant effect.
+
+%The correlation of growth rate and sound
+%speed is non-negligible as in previous `planet-off' case as the disk
+%now can reach values of $h\simeq0.08$ for very slow cooling and the
+%importance of the effect can be imagined to accumulate over the now
+%considerably longer timescale. These effects dictate when then the
+%vortex reaches the critical value in which it shocks the system that
+%subsequently causes disspation and hence the lifetime fo the vortex.

### 2014-09-22 comparisontest/src/SideEuler.c
Exampl Simulation

@@ -1,0 +1,602 @@
+#include "mp.h"
+
+extern boolean OpenInner, NonReflecting, OuterSourceMass, Evanescent;
+extern boolean SelfGravity, SGZeroMode, Adiabatic;
+extern Pair DiskOnPrimaryAcceleration;
+extern int dimfxy;
+real Hp0, Hg0, Ht0;
+
+real GasTotalMass (array)
+     PolarGrid *array;
+{
+  int i, j, ns;
+  real *density, total = 0.0, fulltotal=0.0;
+  ns = array->Nsec;
+  density = array->Field;
+  if (FakeSequential && (CPU_Rank > 0)) 
+    MPI_Recv (&total, 1, MPI_DOUBLE, CPU_Rank-1, 0, MPI_COMM_WORLD, &stat);
+  for (i = Zero_or_active; i < Max_or_active; i++) {
+    for (j = 0; j < ns; j++) {
+      total += Surf[i]*density[j+i*ns];
+    }
+  }
+  if (FakeSequential) {
+    if (CPU_Rank < CPU_Number-1)
+      MPI_Send (&total, 1, MPI_DOUBLE, CPU_Rank+1, 0, MPI_COMM_WORLD);
+  }
+  else
+    MPI_Allreduce (&total, &fulltotal, 1, MPI_DOUBLE, MPI_SUM, MPI_COMM_WORLD);
+  if (FakeSequential) {
+    MPI_Bcast (&total, 1, MPI_DOUBLE, CPU_Number-1, MPI_COMM_WORLD);
+    fulltotal = total;
+  }
+  return fulltotal;
+}
+
+real GasMomentum (Density, Vtheta)
+     PolarGrid *Density, *Vtheta;
+{
+  int i,j,ns;
+  real *density, *vtheta, total = 0.0, fulltotal=0.0;
+  ns = Density->Nsec;
+  density = Density->Field;
+  vtheta = Vtheta->Field;
+  if (FakeSequential && (CPU_Rank > 0)) 
+    MPI_Recv (&total, 1, MPI_DOUBLE, CPU_Rank-1, 2, MPI_COMM_WORLD, &stat);
+  for (i = Zero_or_active; i < Max_or_active; i++) {
+    for (j = 1; j < ns; j++) {
[…]

### 2014-09-22 comparisontest/src/Theo.c

@@ -1,0 +1,150 @@
+#include "mp.h"
+
+extern real ScalingFactor;
+
+/* Surface density */
+real Sigma(r)
+     real r;
+{
+  real cavity = 1.0;
+  real density, Hin, DR, newden, fac;
+
+  Hin = ASPECTRATIO*pow(RMIN,FLARINGINDEX+1.0);
+  SIGMA0 = ASPECTRATIO*pow(RMAX,-0.5)*pow(RMAX, -1.5);
+  SIGMA0 /=QOUT*PI*pow(RMAX, -SIGMASLOPE)*(1.0-pow(RMIN/(RMAX+Hin),0.5));
+
+  density = SIGMA0*pow(r,-SIGMASLOPE)*(1.0-pow(RMIN/(r+Hin),0.5));
+
+  return density;
+}
+
+void FillSigma() {
+  int i;
+  for (i = 0; i < NRAD; i++) {
+    SigmaMed[i] = Sigma(Rmed[i]);
+    SigmaInf[i] = Sigma(Rinf[i]);
+  }
+}
+
+void RefillSigma (Surfdens)
+     PolarGrid *Surfdens;
+{
+  int i, j, nr, ns, l;
+  real *field;
+  real moy;
+  nr = Surfdens->Nrad;
+  ns = Surfdens->Nsec;
+  field = Surfdens->Field;
+  for (i = 0; i < nr; i++) {
+    moy = 0.0;
+    for (j = 0; j < ns; j++) {
+      l = j+i*ns;
+      moy += field[l];
+    }
+    moy /= (real)ns;
+    SigmaMed[i] = moy;
+  }
+  SigmaInf[0] = SigmaMed[0];
+  for (i = 1; i < nr; i++) {
+    SigmaInf[i] = (SigmaMed[i-1]*(Rmed[i]-Rinf[i])+\
+		   SigmaMed[i]*(Rinf[i]-Rmed[i-1]))/\
+      (Rmed[i]-Rmed[i-1]);
+  }
+}
+
+/* Thermal energy */
+real Energy(r)
+     real r;
+{
+  real energy0;
+  if (ADIABATICINDEX == 1.0) {
+    fprintf (stderr, "The adiabatic index must differ from unity to initialize the gas internal energy. I must exit.\n");
+    prs_exit (1);
+  }
+  else
[…]

### 2014-09-22 comparisontest/src/Viscosity.c

@@ -1,0 +1,181 @@
+#include "mp.h"
+
+static PolarGrid *DRR, *DRP, *DPP;
+real FViscosity (rad)
+     real rad;
+{
+  real viscosity, rmin, rmax, scale, DR, fac;
+  int i = 0;
+  viscosity = VISCOSITY;
+  if (ViscosityAlpha) {
+    while (GlobalRmed[i] < rad) i++;
+    viscosity = ALPHAVISCOSITY*GLOBAL_bufarray[i]*	\
+      GLOBAL_bufarray[i]*pow(rad, 1.5);
+  }
+
+  return viscosity;
+}
+
+real AspectRatio (rad)
+     real rad;
+{
+  real aspectratio, rmin, rmax, scale;
+  aspectratio = ASPECTRATIO;
+  rmin = TRANSITIONRADIUS-TRANSITIONWIDTH*ASPECTRATIO;
+  rmax = TRANSITIONRADIUS+TRANSITIONWIDTH*ASPECTRATIO;
+  scale = 1.0+(PhysicalTime-PhysicalTimeInitial)*LAMBDADOUBLING;
+  rmin *= scale;
+  rmax *= scale;
+  if (rad < rmin) aspectratio *= TRANSITIONRATIO;
+  if ((rad >= rmin) && (rad <= rmax)) {
+    aspectratio *= exp((rmax-rad)/(rmax-rmin)*log(TRANSITIONRATIO));
+  }
+  return aspectratio;
+}
+
+void InitViscosity ()
+{
+  DivergenceVelocity = CreatePolarGrid(NRAD, NSEC, "DivV");
+  DRR                = CreatePolarGrid(NRAD, NSEC, "Drr");
+  DRP                = CreatePolarGrid(NRAD, NSEC, "Drp");
+  DPP                = CreatePolarGrid(NRAD, NSEC, "Dpp");
+  TAURR              = CreatePolarGrid(NRAD, NSEC, "TAUrr");
+  TAURP              = CreatePolarGrid(NRAD, NSEC, "TAUrp");
+  TAUPP              = CreatePolarGrid(NRAD, NSEC, "TAUpp");
+}
+
+void ComputeViscousTerms (RadialVelocity, AzimuthalVelocity, Rho)
[…]

### 2014-09-22 comparisontest/src/mp.h

@@ -1,0 +1,47 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <math.h>
+#include <ctype.h>
+#include "fondam.h"
+#include "types.h"
+#include "proto.h"
+/* PARALLEL LIBS */
+#ifdef _PARALLEL
+#include <mpi.h>
+#ifdef _FFTW
+#include <dfftw_mpi.h>
+#include <drfftw_mpi.h>
+#else
+#include "fftw_dummy.h"
+#endif
+#endif
+/* SEQUENTIAL LIBS */
+#ifndef _PARALLEL
+#include "mpi_dummy.h"
+#ifdef _FFTW
+#include <fftw.h>
+#include <rfftw.h>
+#else
+#include "fftw_dummy.h"
+#endif
+#endif
+
+#include "sg.h"
+#include "sgproto.h"
+#ifndef __LOCAL
+#include "param.h"
+#include "global_ex.h"
+#else
+#include "param_noex.h"
+#include "global.h"
+#endif
+#include <sys/times.h>
+#include <unistd.h>
+#include <string.h>
+#ifndef __ppc__
+#include <malloc.h>
+#endif
+#ifdef _TRAP_FPE
+#include <signal.h>
+#include <fenv.h>
+#endif

### 2014-09-22 comparisontest/src/proto.h

@@ -1,0 +1,140 @@
+void masterprint (const char *template, ...);
+void mastererr (const char *template, ...);
+real GetGlobalIFrac ();
+void prs_exit ();
+void *prs_malloc ();
+void erreur ();
+void message ();
+PolarGrid    *CreatePolarGrid();
+void MultiplyPolarGridbyConstant ();
+void DumpSources ();
+Force *AllocateForce ();
+void ComputeForce ();
+void FreeForce ();
+void UpdateLog ();
+void ReadfromFile ();
+void InitLabel ();
+void Initialization ();
+void var ();
+void ReadVariables ();
+void PrintUsage ();
+real TellNbOrbits ();
+real TellNbOutputs ();
+void TellEverything ();
+void GiveTimeInfo ();
+void InitSpecificTime ();
+void GiveSpecificTime ();
+void EmptyPlanetSystemFile ();
+void WritePlanetFile ();
+void WritePlanetSystemFile ();
+void WriteBigPlanetFile ();
+void WriteBigPlanetSystemFile ();
+real GetfromPlanetFile ();
+void RestartPlanetarySystem ();
+void WriteDiskPolar();
+void WriteDim ();
+void SendOutput ();
+void ComputeIndirectTerm ();
+void FillForcesArrays ();
+void AdvanceSystemFromDisk ();
+void AdvanceSystemRK5 ();
+void SolveOrbits ();
+real ConstructSequence ();
+void InitGasVelocities ();
+void AccreteOntoPlanets ();
+void FindOrbitalElements ();
+int FindNumberOfPlanets ();
+PlanetarySystem *AllocPlanetSystem ();
+void FreePlanetary ();
+PlanetarySystem *InitPlanetarySystem ();
+void ListPlanets ();
+real GetPsysInfo ();
+void RotatePsys ();
+void DerivMotionRK5 ();
+void TranslatePlanetRK5 ();
+void RungeKunta ();
+real GasTotalMass();
+real GasMomentum ();
+real GasTotalEnergy ();
+void DivisePolarGrid ();
+void InitComputeAccel ();
+Pair ComputeAccel ();
+void OpenBoundary ();
+void NonReflectingBoundary ();
+void ApplyOuterSourceMass ();
+void ApplyBoundaryCondition ();
+void ApplySubKeplerianBoundary ();
+void CorrectVtheta ();
+boolean DetectCrash ();
+void FillPolar1DArrays ();
+void InitEuler ();
+real min2 ();
+real max2 ();
[…]

### 2014-09-22 comparisontest/src/sgvelinit.c

@@ -1,0 +1,33 @@
+#include "mp.h"
+
+void init_azimutalvelocity_withSG (Vtheta)
+     PolarGrid *Vtheta;
+{
+  extern boolean SGZeroMode;
+  int i, j, l, ns, nr;
+  real r, invr;
+  real omega, omegakep;
+  real *vt;
+  vt = Vtheta->Field;
+  nr  = Vtheta->Nrad;
+  ns  = Vtheta->Nsec;
+  if ( !SGZeroMode ){
+    mpi_make1Dprofile (SG_Accr, GLOBAL_AxiSGAccr);
+   }  else
+    GLOBAL_AxiSGAccr = SG_Accr;
+  for(i = 0; i < GLOBALNRAD; i++){
+    printf("%f %f\n", Radii[i], GLOBAL_AxiSGAccr[i]);
+  }
+  for (i = 0; i < nr; i++) {
+    r = Rmed[i];
+    invr = 1./Rmed[i];
+    omegakep = sqrt(G*1.0*invr*invr*invr);
+    omega = sqrt( omegakep*omegakep*( 1.0 -				\
+				      (1.+SIGMASLOPE-2.0*FLARINGINDEX)*pow(ASPECTRATIO,2.0)*pow(r,2.0*FLARINGINDEX) ) - \
+		  invr*GLOBAL_AxiSGAccr[i+IMIN] );
+    for (j = 0; j < ns; j++) {
+      l = i*ns + j;
+      vt[l] = r*omega;
+    }
+  }
+}

### 2014-10-15 paper/paper.tex
intro added, abstract edit, ref update

--- 2014-08-22 paper/paper.tex
@@
 Several observations of transition discs show lopsided
-dust-distributions. This is suggestive of a dust-trapping mechanism at work. A
-popular explanation is the formation of a large-scale vortex at the
-edge of a gap opened by a giant planet. While theoretical models
-support this picture, numerical simluations have thus far
-employed locally isothermal discs. On the other hand, the theory of the
-vortex-forming instability was originally developed for adiabatic
-discs. We generalise the study of planetary gap stability to non-isothermal
+dust-distributions. A potential explanation is the formation of a
+large-scale vortex acting as a dust-trap at the edge of a gap opened
+by a giant planet. While numerical models of gap-edge vortices have
+thus far employed locally isothermal discs, the 
+theory of this vortex-forming or `Rossby wave' instability was
+originally developed for adiabatic discs.  
+We generalise the study of planetary gap stability to non-isothermal
 discs using customised numerical simulations of disc-planet
-systems. We include in the energy equation a cooling term with cooling
-timescale $t_\mathrm{cool}=\beta\Omega_k^{-1}$, where $\Omega_k$ is
+systems where the planet opens unstable gaps. 
+We include in the energy equation a simple cooling term with
+cooling timescale $t_\mathrm{cool}=\beta\Omega_k^{-1}$, where $\Omega_k$ is
 the Keplerian frequency, and examine the effect of $\beta$ on the
-stability of gap edges.
-For measurments reminsent of linear stability, gap edges stabilize with increasing $\beta$ as dominate modes and growth rates decrease in value.
-In the non-linear limit of planet-vortex interactions, large scale overdense vortices were found to last for 1000's of orbits in a quasi-steady state characterized by sudden disspation on the order of 100's or planet orbits.
-Vortex lifetimes were found to have a non-monoticly dependence on $\beta$ and an optimum value for cooling rate was found to maximize the lifetime.
+stability of gap edges and vortex lifetimes. We find increasing
+$\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
+dominant azimuthal wavenumber $m$ decreases.  
+%gap edges stabilize with increasing $\beta$ as dominate modes and
+%growth rates decrease in value. 
+We find a quasi-steady state consisting of one  
+large-scale overdense vortex circulating the outer gap edge, typically  
+lasting $O(10^3)$ orbits.  
+%characterized by
+%sudden disspation on the order of 100's or planet orbits.
+Vortex lifetimes were found to increase with cooling times up to an
+optimal value, after which vortex lifetimes decrease. This
+non-monotonic dependence is qualitatively consistent with 
+strictly isothermal simulations that differ in disc aspect ratios
+presented in Fu et al. (2014, ApJL, 788, L41).
 
-%The simulations are customized to examine the stability of gaps opened by massive planets as a function of $\beta$.
+%a non-monoticly dependence on $\beta$ and
+%an optimum value for cooling rate was found to maximize the lifetime.
+
+%The simulations are customized to examine the stability of gaps
+%opened by massive planets as a function of $\beta$.
 
@@
 \section{Introduction}\label{intro}
+%{
+%\bf 
+%observational motivation: asymmetric dust distributions. 
+%theoretical motivations: generalize gap stability to include energy equation. 
+%}
+The interaction between planets and protoplantary discs plays an
+important role in the theory of planet formation and disk 
+evolution. Disk-planet interaction may lead to the orbital migration
+of protoplanets and modify the structure of
+protoplanetary disks  \citep[see][for a recent review]{baruteau13}.  
+%\citep[][see \cite{baruteau13} for a recent
+%  review]{goldreich79, lin79, goldreich80,lin86}.
+
+A sufficiently massive planet can open a gap in a 
+gaseous protoplanetary disc \citep{pap_lin84,bryden99,crida06,fung14}, 
+while low mass planets may also open gaps if the disk viscosity is
+small enough \citep{li09,dong11,duffell13}. Support for such disk-planet
+interaction have begun to emerge in observations of circumstellar
+disks that reveal annular gaps
+\citep[e.g.][]{quanz13a,debes13,osorio14}, with possible evidence of
+companions within them \citep[e.g.][]{quanz13b,reggiani14}.
+
+%gaps can be unstable - vortices
+
+A recent theoretical development in the study of planetary gaps is
+their stability. When the disk viscosity is low and/or the planet mass
+is large, the presence of potential vorticity (PV, the ratio of
+vorticity to surface density) extrema can render planetary gaps
+dynamically unstable due to what is now referred to as the `Rossby
+wave instability' \citep[RWI,][]{lovelace99,li00}. This 
+eventually leads to vortex formation 
+\citep{li01,koller03,li05,valborro07}, which can significantly affect
+orbital migration of the planet  \citep{ou07,li09,lin10}.
+
+Vortex formation at gap edges may also have observable 
+consequences. Because disc vortices represent pressure maxima, they are
+able to collect dust particles 
+\citep{barge95,inaba06,lyra13}. Dust-trapping at gap-edge vortices
+have thus been suggested to explain asymmetric dust
+distributions observed in several transition disks
+\citep[e.g][]{marel13,isella13,perez14}.
+
+However, studies of Rossby vortices at planetary gap-edges have 
+adopted locally isothermal disks, %cite lin11, zhu14, lyra
+where the disk temperature is a fixed function of
+position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
+%isothermal valid for outer disk  
+On the other hand, the theory of the RWI was in fact
+developed for adiabatic disks \citep{li00}, which permits
+heating. %cite 
+In adiabatic disks, the relevant quantity for stability
+becomes a generalisation of the PV that include entropy variations
+\citep{lovelace99}.
+
+Gap-opening is associated with planet-induced spiral shocks. In an
+isothermal disk, PV-generation across these isothermal shocks leads to
+the RWI \citep{koller03,li05,valborro07,lin10}.    
+However, if cooling is inefficient and the shock is non-isothermal,
+then shock-heating may affect gap stability, since the
+relevant quantity is an entropy-modified PV (described below), and
+there is entropy-generation across the shocks.
+
+For example, previous
+linear simulations of the RWI found 
+that increasing the sound-speed favours instability \citep{li00,lin13}. In the context of
+planetary gaps, however, the increased temperature may also act to
+stabilise the disk by making gap-opening more difficult. It is 
+therefore of theoretical interest to clarify the effect of heating and
+cooling on the stability of planetary gaps.
+
+In this work, we extend the study of planetary gap stability against
+vortex formation to non-isothermal discs. We include in the fluid energy
+equation an one-parameter cooling prescription that allows us to probe
+disk thermodynamics ranging from nearly isothermal to nearly
+adiabatic.
+
+This paper is organised as follows. In \S\ref{model} we describe the
+equations governing the disc-planet system and initial conditions. Our
+numerical approach, including diagnositic measures, are given in
+\S\ref{method}. We present results from two sets of numerical
+experiments. In \S\ref{linear1} we use disc-planet interaction to set
+up discs with gaps, but study their stability without further
+influence from the planet. %stability through background state 
+We then examine the lifetime of gap-edge vortices in \S\ref{nonlinear}
+as a function of the imposed cooling rate. We conclude and summarize
+in \S\ref{summary} with a discussion of important caveats.
 
@@
 %\input{discussion}
-%\input{summary}
+\input{summary}
 

### 2014-10-15 paper/results.tex
intro added, abstract edit, ref update

@@
 \section{Growth of non-axisymmetric modes without the influence of the
-  planet}
+  planet}\label{linear1}
 In this section, the planet is introduced at $t=20P_0$ and 
@@
 \section{Non-linear evolution of
-  gap-edge vortices with finite cooling time} 
+  gap-edge vortices with finite cooling time}\label{nonlinear} 
 %{\bf main fig: vortex amplitude v.s. time for diff beta. enough data

### 2014-10-28 paper/results.tex
Updated Additional Analysis in results section

@@
 and velocity fields. At this point the planet has carved a partial
-gap. We then perturb the surface density for $r>r_p$ and continue to 
+gap and the RWI has not yet occured.
+ We then perturb the surface density for $r>r_p$ and continue to 
 evolve the disc. We impose  sinusoidal perturbations with 
-azimuthal wavenumbers $m\in[1,10]$. {\bf what are PERTMIN and PERTMAX values?}  
+azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
+ times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
 This procedure allows us to analyse the growth of 
@@
 Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
-with $r_\mathrm{out}=25r_\mathrm{in}$, 
-and we consider 
+with an open boundry condition at the outer disk edge $r_\mathrm{out}=25r_\mathrm{in}$. We consider 
 %for a time up to $200P_0$ allowing high resolution and detection of high
@@
 cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
-and slowly cooled discs.  {\bf boundary condition for these sims?}
+and slowly cooled discs. % {\bf boundary condition for these sims?}
 
@@
 
+The intial planet-disk interaction form bumps and grooves in the gap profiles
+ which can potentially be unstable due to axisymmetric instabilities. The
+ generalized local axisymmetric stability condition is the Solberg-Hoiland
+ criterion,
+\begin{equation}
+ \kappa^2+N^2 \geq 0 $ where $ N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r} \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right)
+\end{equation}
+where $N$ is the Brunt-V\"ais\"al\"a frequency for the disk due to entropy
+ variation which is non-zero for non-adiabatic models such as being considered.
+ At  $t=30P_0$ the outer gap edge of $r=2.5r_h$,  where the RWI is excited the
+ criterion value, is found to be approximately $0.002$ for all $\tilde\beta$
+ which is the local minimum in the gap region which indicates local
+ axisymmetric stability of gap. The Solberg-Hoiland criteria is similary
+ satisfied for the entire disk and for all times within the simulation length
+ with little change of values. Thus for all cases of $\tilde\beta$ the
+ planet-induced gaps formed are stable to axisymmertic instabilities.
+
 \subsection{Non-axisymmetric instability}\label{linear}
@@
 modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
-Eq.~\ref{fouriertransform}---\ref{growth}. Table \ref{modetable}
+Eq.~\ref{fouriertransform}---\ref{growth} with radial range of $[2r_h,5r_h]$. Table \ref{modetable}
 lists the growth rates for the three cooling
@@
   \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimenionlized by the axisymmetric mode for the
+    density, non-dimenionlized by the axisymmetric background keplerian mode
+ $\Sigma_0(t=0)$ for the
     `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
@@
     mode during linear growth and has a corresponding
-    $\gamma=0.017\Omega(r_p)$. {\bf is the $\Sigma_0$ at
-      t=0?} \label{linearmodes}}  
+    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
+% {\bf is the $\Sigma_0$ at t=0?} }  
 \end{figure}
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
   \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_lowb}
+    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
   }
@@
   \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_medb}
+    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
   }
@@
   \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_highb}
+    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
   }  
@@
       with respect to t=0).}} 
-\end{figure*}
+\end{figure}
 
@@
 
-{\bf
-If there's time, do low resolution planet off simulations but keep the planet on
-until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
-are generally deeper (including the nearly adiabatic case). Although
-with the planet instability may already appear by $t=40P_0$, the
-azimtuhal average performed prior to perturbations will erase it. 
-}
+%{\bf
+%If there's time, do low resolution planet off simulations but keep the planet on
+%until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
+%are generally deeper (including the nearly adiabatic case). Although
+%with the planet instability may already appear by $t=40P_0$, the
+%azimtuhal average performed prior to perturbations will erase it.}
 
@@
 effects on vortex evolution. We apply an open boundary at
-$r_\mathrm{out}$. {\bf mention lower resolution preliminary runs -
+$r_\mathrm{out}$.
+
+Preliminary simulations with same numerical scheme but lower resolution of
+ $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
+with the inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$ as well.
+Results and vortex behaviour are convergent with
+higher resolution simulations as discussed in this section.
+ {\bf mention lower resolution preliminary runs -
   similar results to those presented here (?). want to demonstrate convergence}
@@
 The linear growth of the RWI and vortex-formation is followed by 
-vortex merging. For all $\tilde\beta$ values, only one vortex remains
-by $150P_0$.
+vortex merging. Merging timescales are independent of $\tilde\beta$ value,
+ and by $60P_o$ only one vortex remains as the growth of these modes are now
+ accelerated by planetary influence.
 %i.e. only the $m=1$ mode remains.
-{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
-  taken to reach one vortex as a function of cooling}
-%by $150P_0$ for all , only $m=1$ vortex
-%modes exist. 
-The evolution of the amplitude of the $m=1$ mode in surface density is
-shown in Fig~\ref{lifetimeplot} for   
-different $\tilde\beta$. {\bf should state (also for planet-off runs)
-  what the radial range is used for averaging the Fourier amplitudes}
+
+The evolution of the amplitude of the $m=1$ surface density mode
+ over the radial range of $[2r_h,10r_h]$ is shown in Fig~\ref{lifetimeplot}
+ for different $\tilde\beta$.
+% {\bf should state (also for planet-off runs)
+ % what the radial range is used for averaging the Fourier amplitudes}
 
@@
 In all cases the system remains in a quasi-steady state for
-$\gtrsim500P_0$ with a single vortex circulating 
+$\gtrsim800P_0$ with a single vortex circulating 
 the outer gap edge at the local Keplerian  
@@
 vortex grows in intensity. As the vortex forms it has a
-charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic
+ value of $Ro\approx-0.1$ for all $\tilde\beta$
+ indicating anti-cyclonic
 motion. As the $m=1$ amplitude grows the relative spin of the vortex
 continuously grows along with it, reaching values of
-$Ro\approx-0.4$, corresponding to surface density almost 10 times the
-initial value. {\bf for what cooling are these values quoted for? do
-  these values depend on cooling?} Vortex growth is attributed to
+$Ro\approx-0.4$ characteristicly for all cooling rates. Relative surface
+ density pertibations at the center of vortices can reach values of 7 times the
+ intial density for all cases of $\tilde\beta$ in this quasi-steady state and
+ can be found as high as 9 times for the $\tilde\beta=1.0$ case.
+ The gap profiles apply a surface density boost in the vortex region as the
+ vortices are located at the gap edge maxima
+ but the values associated with this boost have relative density pertibations
+ of under 1 and thus are negligibe compared to the max amplitudes of the vortices.
+ Vortex growth is attributed to
 the continuous generation of vorticity by the planet-disc interactions. 
+This is supported by the evidence that in long term simulations where the
+planet is removed such as in section \ref{nonlinearplanetoff} the amplitude of
+ the vortices are not seen to contiunously grow without the planet influence.
+
+{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
+  taken to reach one vortex as a function of cooling}
+{\bf for what cooling are these values quoted for? do
+  these values depend on cooling?}
  {\bf is the evidence for this the fact that growth is not observed in
@@
 the $m=1$ amplitude within the simulation time, which occurs over
-timescales of $50P_0$, i.e. much shorter than the duration of the quasi-steady
-state. {\bf is this dissipation timescale independent of cooling?}.  
+timescales of $t\in[20P_0,70P_0]$, i.e. much shorter than the duration of the
+quasi-steady state.
+Dissipation timescales follow the basic trend of decreasing with increasing $\tilde\beta$ but more accurately correlated with Mach number values which will be discussed in following section
+{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
 %Eventually
@@
 of the quasi-steady vortex.
+The instability is not seen to emerge again after the intial vortex dissipation
+ due to the smoothening out of the gap edge during dissipation by the
+ transfer of density from the once tightly held vortex region to the surrounding
+ disk and gap.  
 {\bf question: does instability
@@
     vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
-    similar. {\bf is this plot truncated? in the text it says the
-      domain goes to $r=45$?}
+    similar.% {\bf is this plot truncated? in the text it says the
+      %domain goes to $r=45$?}
     %The large scale non-axisyemetric $m=1$ overdensity can be
@@
     \label{Vortex2D}} 
+\end{figure}
+
+\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
+Non-linear observations were also made for the `planet-off' simulations.
+After the linear growth phase of the vortices, vortex merging takes hold but
+ now on timescales of up to $150P_0$ as the vortices grow as linear
+ instabilities discussed previously. The vortex merging time is depedendent on
+ the growth rates of the modes and how long they take to saturate. Similar
+ quasi-steady vortices form but the long-term behaviour is significantly
+ different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
+ The quasi-steady vortices formed have significantly lower density pertibations
+ and now slowly decrease in amplitute in time as apposed to previously
+ where their growth was supported by planet interactions. The vortices simply
+ dissipate viscously and no longer undergo the quick dissipation that
+ characterized the dissipation as earlier. The dissipation rate is dependent
+ on the cooling rate with viscous dissipation decreasing with larger
+$\tilde\beta$.
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+    0cm 0cm 1.1cm]{figures/longterm_planetoff}
+  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
 \end{figure}
@@
 
-As the vortex dies off as described above its associated Rossby number
-of quickly approaches $0$. Gradients of surface density are seen to
+Gradients of surface density are seen to
 spike up in wake-like features around the vortex during dissipation. 
-{\bf show
-  this. e.g. vortex before, during and after death. maybe show
-  snapshots of surface density perturbation (vortex) compared with
-  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
-  changes and shocks develop prior ot death. state whether this is
-  generic for all coolings}.  
+Snapshots of relative density pertibation and density gradients during the
+quasi steady state, during disspation and just before death are displayed in
+Fig. \ref{shockplot} for $\tilde\beta=1.0$. Quasi-steady vortices have large 
+elliptical shapes, while dissipating vortices are denser and smaller in spacial
+extent.
+As the vortices grow in intensity wakes are seen to extend from them. These
+wakes correlate with large gradients in surface density and are first seen
+when vortices are in the later half of their quasi steady states. During this
+phase the average value of the gradients along the shock region are
+$|\nabla\Sigma| \approx 4 \Sigma$. Just before the vortices start dissipation
+this quantity sharply increases to a value of $ \approx6 \Sigma$ in which it
+ remains during dissipation and rest of vortex lifetime.  
+As the vortex dies off as described above its associated Rossby number quickly
+approaches $0$.
+
 In addition, we also observed large increases in the Mach number near
 the vortex during dissipation, which indicate susceptibility to
-shocks. {\bf show mach number plot, or quote values. relevant mach
-  number is fluid flow speed with respect to 
-  vortex bulk motion around the disc. does the typical mach number
-  near the vortex just before its death depend on cooling?} 
+shocks. Mach numbers with repect to the vortex bulk motion around the vortex
+ region of $|r-r_{vortex}|<2H$, where $H$ denotes the gas scale height, is
+ $\approx 0.8$ after the intial formation of the vortices for all
+ $\tilde\beta$. During the quasi-steady state the Mach number increases at a
+ steady rate with larger increases for lower $\tilde\beta$. Time values in
+ which Mach numbers reach values of 1 approximately correlate with the start 
+of dissipation and during dissipation values quickly increase to larger max values.
+ Dissipation timescales correlate with 
+Mach numbers reached during dissipation as the quickest dissipation occurs for 
+$\tilde\beta=0.5$ which is on order of $70P_0$ and reaches Mach number
+ $\approx1.2$ while the $\tilde\beta=0.1$ case which dissipates on scales o
+f $20P_0$ reaches Mach numbers $\approx1.4$
 Thus, the sudden drop in vortex amplitude may  
@@
   should be more gradual. can we think of other evidences that it's
-  not the planet shock that's responsible?                    
-}
+  not the planet shock that's responsible?}
+{\bf show
+  this. e.g. vortex before, during and after death. maybe show
+  snapshots of surface density perturbation (vortex) compared with
+  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
+  changes and shocks develop prior ot death. state whether this is
+  generic for all coolings}. 
+{\bf show mach number plot, or quote values. relevant mach
+  number is fluid flow speed with respect to 
+  vortex bulk motion around the disc. does the typical mach number
+  near the vortex just before its death depend on cooling?}
 
-Vortex growth during the quasi-steady state also acts to smooth out
-the gap edge. 
+Vortex growth during the quasi-steady state and dissipation also acts to smooth
+out the gap edge. To characterize the gap, we azimuthally and radially average
+the relative density pertibation within $r\in[r_{planet},r_{out}]$ where
+ $r_{out}=$ is defined as the outer gap radius defined as point where
+$\langle\Delta \Sigma(r_{out})/\Sigma(t=0)\rangle_{\phi}=0$. Thus a more 
+negative gap depth parameter correlates with a steeper gap edge. In
+ Fig. \ref{gapdepth} the gap depth parameter with respect to simulation time
+for $\tilde\beta=1.0$ is shown. The gap depth is seen to slowly increase as
+vortex reaches its max amplutude at $t=1000P_0$ which is followed by a sharp
+increase which correlates with the time of vortex dissipation.
 %Of note is also that the large overdensity the vortex
@@
 %disk.  
+The smoothening out of the gap edge during later half of quasi steady state
+ is because the vortex has an associated viscosity $O(10^{-2})$.
+%{\bf quote a typical value of alpha viscosity associated with vortex}.
+ This acts
+against gap-opening by the planet, so that the condition for the RWI
+becomes less favourable as the vortex grows.
+The sudden smoothening of gap egde during dissipation is due to the large
+influx of density from the vortex to the gap.
+{\bf self-limiting
+  behaviour?} 
 {\bf show this (smoothening of gap edge due to vortex growth),
   e.g. general vortensity evolution or surface density 
-  evolution}. 
-This is because the vortex has an associated viscosity {\bf quote a
-  typical value of alpha viscosity associated with vortex}. This acts
-against gap-opening by the planet, so that the condition for the RWI
-becomes less favourable as the vortex grows. {\bf self-limiting
-  behaviour?}
+  evolution}
 
+\begin{figure}
+\subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock1}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock2}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock3}
+  } \\[-0.98cm]
+\subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock4}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock5}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock6}
+  } 
+  \caption{Relative density pertibation of vortices (top) and density gradient
+ (bottom) for before (left), during (middle) and after dissipation (right).
+ Large density gradients are characteristics of shocks and become prominant
+ around vortex during dissipation} \label{shockplot}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figures/gapdepth}
+ \caption{Running average of relative density pertibation within the outer half
+of gap} \label{gapdepth}
+\end{figure}
 \subsection{Vortex lifetimes as a function of cooling rate}
@@
 Thus, unlike our `planet-off' simulations we find that there is
-a cooling time $t_c$ which maximises the vortex lifetime. {\bf to make
+a cooling time $t_c$ which maximises the vortex lifetime. In the long term
+ 'planet-off' simulations we find that the vortices dissipate very gradually and visciously. {\bf to make
 this comparison we need to discuss vortex lifetimes for planet-off

### 2014-11-09 paper/results.tex
Updated results and added figures

@@
 Note that these `planet-off' simulations are not linear stability
-calculations because since the cooling term in our energy equation
+calculations since the cooling term in our energy equation
 restores the initial temperature profile corresponding to $H/r=0.05$,
@@
  $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
-with the inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$ as well.
-Results and vortex behaviour are convergent with
+with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
+Vortex evolution and behaviour are convergent with
 higher resolution simulations as discussed in this section.
@@
  and by $60P_o$ only one vortex remains as the growth of these modes are now
- accelerated by planetary influence.
+ accelerated by planetary influence as apposed to `planet off' simulations.
 %i.e. only the $m=1$ mode remains.
@@
 continuously grows along with it, reaching values of
-$Ro\approx-0.4$ characteristicly for all cooling rates. Relative surface
- density pertibations at the center of vortices can reach values of 7 times the
- intial density for all cases of $\tilde\beta$ in this quasi-steady state and
- can be found as high as 9 times for the $\tilde\beta=1.0$ case.
- The gap profiles apply a surface density boost in the vortex region as the
- vortices are located at the gap edge maxima
- but the values associated with this boost have relative density pertibations
- of under 1 and thus are negligibe compared to the max amplitudes of the vortices.
+$Ro\approx-0.4$ characteristicly for all cooling rates.
+
+Surface density pertibations relative to the intial disk
+at the center of vortices can reach values of $\Delta\Sigma/\Sigma_0=7$ 
+ for all cases of $\tilde\beta$ in this quasi-steady state and
+can be found as high as $\Delta\Sigma/\Sigma_0= 9$ times for the 
+$\tilde\beta=10$ case.
+ The gap profiles formed also apply a surface density boost in the vortex 
+region as the vortices are located at the gap edge maxima.
+ The average value of relative density pertibation in the outer gap region of 
+$r=r_{p}+2.5r_h$
+has a value $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$ for all times and
+ cooling rates.
+Thus the surface density increases caused by gap formation are small 
+compared to the max amplitudes of the vortices.
  Vortex growth is attributed to
@@
 This is supported by the evidence that in long term simulations where the
-planet is removed such as in section \ref{nonlinearplanetoff} the amplitude of
- the vortices are not seen to contiunously grow without the planet influence.
+planet is removed, such as seen in Fig \ref{planetofflifetimeplot},
+ the amplitude of
+ the vortices are not seen to grow without the planet influence.
 
@@
 timescales of $t\in[20P_0,70P_0]$, i.e. much shorter than the duration of the
-quasi-steady state.
-Dissipation timescales follow the basic trend of decreasing with increasing $\tilde\beta$ but more accurately correlated with Mach number values which will be discussed in following section
-{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
+quasi-steady state. Dissipation timescales do not correlate with $\tilde\beta$
+ but instead are paramtrized by vortex Mach numbers.
+%{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
 %Eventually
@@
 of the quasi-steady vortex.
-The instability is not seen to emerge again after the intial vortex dissipation
- due to the smoothening out of the gap edge during dissipation by the
- transfer of density from the once tightly held vortex region to the surrounding
- disk and gap.  
-{\bf question: does instability
+After the vortices dissipate the RWI is not seen to be excited again and no
+vortices are seen to refrom in the simulations time.
+%The instability is not seen to emerge again after the intial vortex dissipation
+% due to the smoothening out of the gap edge during dissipation by the
+% transfer of density from the once tightly held vortex region to the surrounding disk and gap.  
+{\bf Discussed in next section:  question: does instability
   actually happen again? if it does, but just weaker, that may be
@@
  instabilities discussed previously. The vortex merging time is depedendent on
- the growth rates of the modes and how long they take to saturate. Similar
- quasi-steady vortices form but the long-term behaviour is significantly
+ the growth rates of the modes and saturation timescales with the
+slowest growing modes in $\tilde\beta=10$ taking the longest. Similar
+ quasi-steady vortices form as in the cases with continuous planet interaction
+but the long-term behaviour is significantly
  different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
@@
 where their growth was supported by planet interactions. The vortices simply
- dissipate viscously and no longer undergo the quick dissipation that
- characterized the dissipation as earlier. The dissipation rate is dependent
- on the cooling rate with viscous dissipation decreasing with larger
-$\tilde\beta$.
+dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.
+The vortices are never found to undergo the drastic sudden dissipation effects
+that characterize the planet induced growing cases do.
 
@@
     0cm 0cm 1.1cm]{figures/longterm_planetoff}
-  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
+  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' 
+simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
 \end{figure}
@@
 %$m=1$ amplitude of the vortices.
+
+\begin{figure}
+\subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock1}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock2}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock3}
+  } \\[-0.98cm]
+\subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock4}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock5}
+  }
+\hfill
+  \subfigure{
+    \includegraphics[width=0.3\linewidth]{figures/shock6}
+  } 
+  \caption{Relative density pertibation of vortices (top) and density gradient
+ (bottom) for before (left), during (middle) and after dissipation (right).
+ Large density gradients are characteristics of shocks and become prominant
+ around vortex during dissipation} \label{shockplot}
+\end{figure}
 
@@
 this quantity sharply increases to a value of $ \approx6 \Sigma$ in which it
- remains during dissipation and rest of vortex lifetime.  
+ remains for rest of vortex lifetime.  
 As the vortex dies off as described above its associated Rossby number quickly
-approaches $0$.
+approaches $0$ indicating a halting of cyclonic motion.
 
@@
 shocks. Mach numbers with repect to the vortex bulk motion around the vortex
- region of $|r-r_{vortex}|<2H$, where $H$ denotes the gas scale height, is
- $\approx 0.8$ after the intial formation of the vortices for all
- $\tilde\beta$. During the quasi-steady state the Mach number increases at a
- steady rate with larger increases for lower $\tilde\beta$. Time values in
- which Mach numbers reach values of 1 approximately correlate with the start 
-of dissipation and during dissipation values quickly increase to larger max values.
- Dissipation timescales correlate with 
-Mach numbers reached during dissipation as the quickest dissipation occurs for 
-$\tilde\beta=0.5$ which is on order of $70P_0$ and reaches Mach number
- $\approx1.2$ while the $\tilde\beta=0.1$ case which dissipates on scales o
-f $20P_0$ reaches Mach numbers $\approx1.4$
+ region of $|r-r_{vortex}|<2H$ {\bf (somewhat arbitrary, large enough to have bulk of vortex mach number increases quickly radially away from vortex)}
+%, where $H$ denotes the gas scale height, 
+are shown in Figure \ref{machplot} for various cooling rates over time.
+% $\approx 0.8$ after the intial formation of the vortices for all $\tilde\beta$. 
+During the quasi-steady state the Mach number increases at a
+steady rate with larger growth rate for higher $\tilde\beta$.
+In all cases Mach numbers are seen to spike up at a time which correlates with
+the start of vortex dissipation.
+% Time values in
+% which Mach numbers reach values of 1 approximately correlate with the start 
+%of dissipation and during dissipation values quickly increase to larger max values.
+The timescales of vortex dissipation also correlates with 
+Mach numbers reached during dissipation. The quickest dissipation occurs for 
+$\tilde\beta=0.5$ which is on order of $20P_0$ and reaches Mach number
+ $\approx1.4$ while the $\tilde\beta=0.1$ case which dissipates on scales o
+f $90P_0$ reaches Mach numbers $\approx1.2$.
 Thus, the sudden drop in vortex amplitude may  
@@
 against gap-opening by the planet, so that the condition for the RWI
-becomes less favourable as the vortex grows.
+becomes less favourable as the vortex grows and hence is why vortices are not
+seen to form again in simulation timescales.
 The sudden smoothening of gap egde during dissipation is due to the large
 influx of density from the vortex to the gap.
-{\bf self-limiting
+{\bf not sure what you mean by self-limiting here:self-limiting
   behaviour?} 
@@
 \begin{figure}
-\subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/shock1}
-  }
-\hfill
-  \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/shock2}
-  }
-\hfill
-  \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/shock3}
-  } \\[-0.98cm]
-\subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/shock4}
-  }
-\hfill
-  \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/shock5}
-  }
-\hfill
-  \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/shock6}
-  } 
-  \caption{Relative density pertibation of vortices (top) and density gradient
- (bottom) for before (left), during (middle) and after dissipation (right).
- Large density gradients are characteristics of shocks and become prominant
- around vortex during dissipation} \label{shockplot}
+    \includegraphics[width=\linewidth]{figures/mach}
+ \caption{Running plot of Mach number relative to vortex within $2H$ of the
+ vortex. Mach numbers increase at rates correlated with cooling rates than
+ spike up at times coinciding with the start of vortex dissipation as seen in
+ Fig \ref{lifetimeplot}} \label{machplot}
 \end{figure}
@@
 \end{figure}
+
 \subsection{Vortex lifetimes as a function of cooling rate}
@@
 
-Fig. \ref{lifetimeplot} shows that increasing $\tilde{\beta}$
+\begin{figure}
+    \includegraphics[width=\linewidth]{figures/betaplot}
+ \caption{Lifetime of vortices with repect to cooling paramter $\tilde\beta$ for both resolution of $(N_r,N_\phi)=(256,512)$ and $(512,1024)${\bf should I only show higher resolution points?} } \label{betaplot}
+\end{figure}
+
+Fig. \ref{lifetimeplot} and Fig. \ref{betaplot} shows that increasing 
+$\tilde{\beta}$
 increases the lifetime of the vortex up to a critical 
-value $\tilde{\beta}=5.0$ for which the vortex lasted beyond the
-simulation time of $2000P_0$. This is consistent with vortex
-dissipation when it induce shocks. Our `planet-off' simulations
+value $\tilde{\beta}=5.0$ for which the vortex lasted to a simulation time of
+ $1250P_0$ before starting to dissipate. This is consistent in time with the
+ vortex introducing shocks to the system.
+
+Our `planet-off' simulations
 indicate weaker vortices are formed initially with increased cooling
-times. After merging, the growth of the single vortex, mediated by disc-planet
+times. This is because it has been shown that vortex amplitude is dependent on 
+growth rate of the linear modes \citep{meheut2013}.
+At the beginning of the vortex quasi-steady state $t=100P_0$ we find
+that the relative surface density boost at center of vortex 
+is $\Delta\Sigma/\Sigma_0=2.5$ for $\tilde\beta=0.1$ 
+and decreases down to a minimum value of $\Delta\Sigma/\Sigma_0=1.48$ for 
+$\tilde\beta=10$.
+
+ After merging, the growth of the single vortex, mediated by disc-planet
 interaction, is expected to be slower with increased cooling time
@@
 during the quasi-steady state is faster than for
-$\tilde{\beta}=5$. This may be attributed to the higher temperature
-reached in the former case ($h\simeq 0.08$). {\bf temperature for
-  $\tilde{\beta}=5$?}, since \cite{li00} showed that the RWI grows
+$\tilde{\beta}=5$.
+ This may be attributed to the higher temperature
+reached in the latter case ($h\simeq 0.07$) in comparison with the former
+($h\simeq 0.06$)
+% {\bf temperature for $\tilde{\beta}=5$?}
+, since \cite{li00} showed that the RWI grows
 faster with increased disc temperature. At long cooling times, this
@@
 
+This effect can be seen in Mach number around the vortices where the
+growth of Mach number with respect to time is increasing with cooling rate.
+Thus the intially weaker formed vortices for higher cooling rate grow
+substantialy faster than the sound speed and can shock the system in earlier
+times.
+
 Thus, unlike our `planet-off' simulations we find that there is
 a cooling time $t_c$ which maximises the vortex lifetime. In the long term
- 'planet-off' simulations we find that the vortices dissipate very gradually and visciously. {\bf to make
-this comparison we need to discuss vortex lifetimes for planet-off
-simulations as well} 
+ 'planet-off' simulations we find that the vortices dissipate very gradually
+ and visciously dependent on cooling rate.
+% {\bf to make this comparison we need to discuss vortex lifetimes for planet-offsimulations as well} 
 This
@@
 %lifetime. 
-By analyzing the region
-around the vortex {\bf define this region} we find the
+By analyzing the radial region around the vortex within
+ $[2r_{h},10r_{h}]$
+% {\bf define this region}
+ we find the
 $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
-$h\approx0.06$. This value of $h$ was also found to maximize vortex
-lifetimes in the isothermal simulations of \cite{fu14}.
+$h\approx0.06$ on average throughout its lifetime. This value of $h$ was also
+ found to maximize vortex
+lifetimes in the isothermal simulations of \cite{fu14}. Note that $h$ is a fixed
+quantity in \cite{fu14} while in this work $h$ is a dynamical variable dependent
+ on the viscous heating and cooling rate as by choice of energy equation.
 
+These results were reproduced in lower resolution simulations as well. Vortex
+shocks were similarly seen during vortex dissipation and non-monotonic
+ dependence on lifetime with $\tilde\beta$ was similarly found.
+Vortex lifetimes are considerably longer in these runs with the maximum
+ lifetime longer than the simulation timescale of $2000P_o$ for 
+$\tilde\beta=5.0$ which again has $h\approx0.06$.
+This inconsistancy in timescales, may be due to resolution as a finer 
+grid spacing is required to accurately
+reproduce jump conditions along shocks and which cause vortex dissipation.
+ {\bf this is the best reason I could personally think of, for lower beta runs vortices last slightly longer but for higher values it is quite drasticly different.
+Not sure how numerical dissipation as you mentioned would be a factor in this case}
 %be the optimum lifetime for long-term simulations of

### 2014-11-18 paper/paper.tex
begin second edits - MKL

@@
 large-scale vortex acting as a dust-trap at the edge of a gap opened
-by a giant planet. While numerical models of gap-edge vortices have
-thus far employed locally isothermal discs, the 
+by a giant planet. Numerical models of gap-edge vortices have
+thus far employed locally isothermal discs, but the 
 theory of this vortex-forming or `Rossby wave' instability was
@@
 discs using customised numerical simulations of disc-planet
-systems where the planet opens unstable gaps. 
+systems where the planet open an unstable gap. 
 We include in the energy equation a simple cooling term with
@@
 strictly isothermal simulations that differ in disc aspect ratios
-presented in Fu et al. (2014, ApJL, 788, L41).
+presented in Fu et al..% (2014, ApJL, 788, L41).
 
@@
 The interaction between planets and protoplantary discs plays an
-important role in the theory of planet formation and disk 
-evolution. Disk-planet interaction may lead to the orbital migration
+important role in the theory of planet formation and disc 
+evolution. Disc-planet interaction may lead to the orbital migration
 of protoplanets and modify the structure of
-protoplanetary disks  \citep[see][for a recent review]{baruteau13}.  
+protoplanetary discs  \citep[see][for a recent review]{baruteau13}.  
 %\citep[][see \cite{baruteau13} for a recent
@@
 gaseous protoplanetary disc \citep{pap_lin84,bryden99,crida06,fung14}, 
-while low mass planets may also open gaps if the disk viscosity is
-small enough \citep{li09,dong11,duffell13}. Support for such disk-planet
+while low mass planets may also open gaps if the disc viscosity is
+small enough \citep{li09,dong11,duffell13}. Support for such disc-planet
 interaction have begun to emerge in observations of circumstellar
-disks that reveal annular gaps
+discs that reveal annular gaps
 \citep[e.g.][]{quanz13a,debes13,osorio14}, with possible evidence of
@@
 A recent theoretical development in the study of planetary gaps is
-their stability. When the disk viscosity is low and/or the planet mass
+their stability. When the disc viscosity is low and/or the planet mass
 is large, the presence of potential vorticity (PV, the ratio of
@@
 \citep{li01,koller03,li05,valborro07}, which can significantly affect
-orbital migration of the planet  \citep{ou07,li09,lin10}.
+orbital migration of the planet  \citep{ou07,li09,yu10,lin10}.
 
@@
 have thus been suggested to explain asymmetric dust
-distributions observed in several transition disks
+distributions observed in several transition discs
 \citep[e.g][]{marel13,isella13,perez14}.
@@
 However, studies of Rossby vortices at planetary gap-edges have 
-adopted locally isothermal disks, %cite lin11, zhu14, lyra
-where the disk temperature is a fixed function of
+adopted locally isothermal discs, %cite lin11, zhu14, lyra
+where the disc temperature is a fixed function of
 position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
-%isothermal valid for outer disk  
+%isothermal valid for outer disc  
 On the other hand, the theory of the RWI was in fact
-developed for adiabatic disks \citep{li00}, which permits
+developed for adiabatic discs \citep{li00}, which permits
 heating. %cite 
-In adiabatic disks, the relevant quantity for stability
+In adiabatic discs, the relevant quantity for stability
 becomes a generalisation of the PV that include entropy variations
@@
 Gap-opening is associated with planet-induced spiral shocks. In an
-isothermal disk, PV-generation across these isothermal shocks leads to
+isothermal disc, PV-generation across these isothermal shocks leads to
 the RWI \citep{koller03,li05,valborro07,lin10}.    
@@
 linear simulations of the RWI found  
-that increasing the sound-speed favours instability \citep{li00,lin13}. In the context of
+that increasing the sound-speed favours instability \citep{li00,lin13}.  
+In the context of
 planetary gaps, however, the increased temperature may also act to
-stabilise the disk by making gap-opening more difficult. It is 
+stabilise the disc by making gap-opening more difficult. It is 
 therefore of theoretical interest to clarify the effect of heating and
@@
 equation an one-parameter cooling prescription that allows us to probe
-disk thermodynamics ranging from nearly isothermal to nearly
+disc thermodynamics ranging from nearly isothermal to nearly
 adiabatic.

### 2014-11-18 paper/results.tex
begin second edits - MKL

@@
 complications from non-axisymmetry arising directly from disc-planet
-interaction.
+interaction (i.e. planet-induced wakes).
 
@@
 Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
-with an open boundry condition at the outer disk edge $r_\mathrm{out}=25r_\mathrm{in}$. We consider 
+with open boundaries at $r=r_\mathrm{in}$ and
+$r_\mathrm{out}=25r_\mathrm{in}$. We consider   
 %for a time up to $200P_0$ allowing high resolution and detection of high
@@
 are deeper with steeper gradients at the gap edges. Faster cooling rates also 
-increase and decrease the surface density maxima and 
-minima increase, respectively. However, a clean gap does not form
-in this short time period.
+increase (decrease) the surface density maxima (minima). However, a
+clean gap does not form in this short time period.
 
@@
 $\tilde{\beta}$ is increased. 
+%However, notice the gap structure
+%changes most significantly for $\beta =0.1 \to 1.0$. For $\beta \geq
+%1$
 
@@
 The important consequence of a heated gap edge is that the
-generalized vortensity profiles become smoother with increasing
+generalised vortensity profiles, $\tilde{\eta}$, becomes smoother with increasing
 cooling times, with the extrema becoming less pronounced. Previous locally
@@
 minima \citep{li05,lin10}. We can therefore expect the RWI to be associated with
-minima in the generalized vortensity (corresponding to local surface
+minima in the generalised vortensity (corresponding to local surface
 density maxima) in the non-isothermal case. Because the extrema are
@@
 
-{\bf
 \subsection{Axisymmetric stability}
-Check that the gap profiles are stable against axisymmetric
-stability. Can use analytical criteria given in section 3.1 of Li et
-al (2000, ApJ, 533, 1023). No plot needed, just statements. 
-}
-
-The intial planet-disk interaction form bumps and grooves in the gap profiles
+%Check that the gap profiles are stable against axisymmetric
+%stability. Can use analytical criteria given in section 3.1 of Li et
+%al (2000, ApJ, 533, 1023). No plot needed, just statements. 
+%}
+The intial planet-disc interaction form bumps and grooves in the gap profiles
 which can potentially be unstable due to axisymmetric instabilities. The
- generalized local axisymmetric stability condition is the Solberg-Hoiland
+generalised local axisymmetric stability condition is the Solberg-Hoiland
 criterion,  
-\begin{equation}
- \kappa^2+N^2 \geq 0 $ where $ N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r} \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right)
-\end{equation}
-where $N$ is the Brunt-V\"ais\"al\"a frequency for the disk due to entropy
- variation which is non-zero for non-adiabatic models such as being considered.
- At  $t=30P_0$ the outer gap edge of $r=2.5r_h$,  where the RWI is excited the
- criterion value, is found to be approximately $0.002$ for all $\tilde\beta$
- which is the local minimum in the gap region which indicates local
- axisymmetric stability of gap. The Solberg-Hoiland criteria is similary
- satisfied for the entire disk and for all times within the simulation length
- with little change of values. Thus for all cases of $\tilde\beta$ the
- planet-induced gaps formed are stable to axisymmertic instabilities.
+\begin{align}
+  \kappa^2+N^2 \geq 0 
+\end{align}
+where
+\begin{align}
+ N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
+ \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
+     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
+\end{align}
+is the square of the Brunt-V\"ais\"al\"a frequency.  
+%for the disk due to entropy
+%variation which is non-zero for non-adiabatic models such as being
+%considered. 
+At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
+(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
+$\sim 0.002$ (code units) for all $\tilde\beta$. The
+Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. The Solberg-Hoiland
+criteria is similary satisfied for the entire disc throughout the
+simulations. % with little change of values. 
+Thus for all values of $\tilde\beta$ the planet-induced gaps are
+stable to axisymmertic instabilities.
 
@@
 
-In all cases we observe growth of non-axisymmetric structures after the
-planet potential is switched off and the system subject to random
-perturbations. An example is shown in Fig. \ref{linearmodes} for 
+We now examine the evolution of the gap for $t>30P_0$, with the
+planet potential switched off, but with an added surface density
+perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
+10$, we observe exponential growth of non-axisymmetric
+structures. % after the 
+%planet potential is switched off and the system subject to random
+%perturbations.  
+An example is shown in Fig. \ref{linearmodes} for 
 $\tilde{\beta}=10$. We charcterize these
 modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
-Eq.~\ref{fouriertransform}---\ref{growth} with radial range of $[2r_h,5r_h]$. Table \ref{modetable}
-lists the growth rates for the three cooling
-times, measured during linear growth.
+Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
+averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
+lists the growth rates measured during 
+linear growth for 5 values of $m$ centred around the maximum.
 
@@
   \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimenionlized by the axisymmetric background keplerian mode
+    density, non-dimenionlized by the initial axisymmetric background mode 
     $\Sigma_0(t=0)$ for the
@@
 Table \ref{modetable} show that as
-$\tilde{\beta}$ was increased from $ 0.1\rightarrow10$ the dominant
-azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the
-respective growth rate decreased from $ \gamma/\Omega(r_p)=0.079
+$\tilde{\beta}$ is increases from $ 0.1\rightarrow10$ the dominant
+azimuthal Fourier mode decreases from $ m=9\rightarrow3$ and the
+respective growth rate decreases from $ \gamma/\Omega(r_p)=0.079
 \rightarrow 0.017$. However, despite two orders of magnitude increase in the
-cooling time, the instability remains dynamical with growth time
-$\lesssim 10P_0$. Snapshots of the instability in $r-\phi$ plane for
+cooling time, the instability remains dynamical with characteristic  growth time
+$\lesssim 10P_0$. Snapshots of the instability in for  
 the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
@@
     decrease as $\tilde{\beta}$ increases. Note that snapshots are
-    taken later because it takes longer for the vortices to grow and
-    become visible with increasing cooling time. 
+    taken later for increasing $\tilde{\beta}$ because it takes longer
+    for the vortices to grow and become visible with increasing cooling time. 
     % correspondingly increase to capture the longer growth phases of
     % the instabilities. 
-    \label{2Dlinear} {\bf I'm curious to know if plots clearer if
-      perturbations are measured relative to t=30? remember t=30 is
-      our reference state for instability. But we'll probably keep
-      these for consistency (later non-linear plots are all measure
-      with respect to t=0).}} 
+    \label{2Dlinear} 
+%{\bf I'm curious to know if plots clearer if
+%      perturbations are measured relative to t=30? remember t=30 is
+%      our reference state for instability. But we'll probably keep
+%      these for consistency (later non-linear plots are all measure
+%      with respect to t=0).}
+  } 
 \end{figure}
@@
 $t_c<\gamma^{-1}$.  
-Thus the disk cools back to its intial value of $h=0.05$
-before or during the instability growth, implying we do not have a
-steady basic state to formulate a standard linear stability
+Thus the disk cools back to its intial temperature corresponding to
+$h=0.05$ before or during the instability growth, so we do not
+have a steady basic state to formulate a standard linear stability 
 problem. %This means the disc temperture of $h=0.05$ in the the above cases
@@
 edges. The fact that we observe smaller growth rates as $h$ is
-increased indicate that for planetary gaps, the importance of $h$ on
-the \emph{linear} RWI is through the setting the gap profile (i.e. basic
-state for the instability).
+increased indicates that for planetary gaps, the importance of $h$ on
+the \emph{linear} RWI is through the setting the gap profile, i.e. basic
+state for the instability (as opposed to the linear response).
 
@@
 
+\subsection{Non-linear evolution without planet
+  influence} \label{nonlinearplanetoff} 
+{\bf typical rossby numbers for these `planet-off' vortices? do
+  stronger vortices (beta=0.1) dissipate faster? if we want to say the
+  vortices dissipate viscously, then need to quote a viscosity
+  (e.g. reynolds stress) and compute a viscous time and compare to the
+  observed decay time. Or attribute to numerical viscosity. It might
+  be simplest to just say the vortices decay without the planet. This
+  section is mainly used to contrast with planet-on simulations, so
+  don't need to go into detail. 
+}
+
+We also extended these `planet-off' simulations well into the non-linear
+regime. After the linear growth phase of the vortices, vortex merging
+takes hold on timescales of up to $150P_0$, until there is one vortex
+left. We find the vortex merging
+time is depedendent on the growth rates of the modes and saturation
+timescales, with the slowest growing modes in $\tilde\beta=10$ taking
+the longest to merge.
+
+Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
+density amplitude, which represents the post-merger single vortex. For
+completeness we also ran intermediate cases with $\tilde{\beta}=0.5$
+and $5.0$. The vortices simply decay, on a timescale of $O(10^3)$
+orbits. We will see in the that section that this behaviour is very
+different to when the planet potential is kept on. Interestingly,
+though, vortices arising from the stronger instabilities (shorter
+cooling times or lower $\tilde{\beta}$) decays faster.
+
+%Similar
+%quasi-steady vortices form as in the cases with continuous planet interaction
+%but the long-term behaviour is significantly
+%different without the planet as seen in 
+%The quasi-steady vortices formed have significantly lower density pertibations
+%and now slowly decrease in amplitute in time as apposed to previously
+%where their growth was supported by planet interactions. 
+%The vortices simply
+%dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.
+
+%Non-linear observations were also made for the `planet-off'
+%simulations.
+% as the vortices grow as linear
+% instabilities discussed previously.
+
+%The vortices are never found to undergo the drastic sudden dissipation effects
+%that characterize the planet induced growing cases do.
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+  0cm 0cm 1.1cm]{figures/longterm_planetoff}
+  \caption{Long term simulations without the planet potential after
+    the gap is set up. The $m=1$ surface density component at the
+    outer gap edge is shown.
+  } \label{planetofflifetimeplot}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We now consider long term simulations of gap-edge vortices for
+We now examine long-term simulations of gap-edge vortices for
 $\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
@@
 %and vortices after $t=30P_0$ as apposed to previous section.  
-We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
-in order for these simulations to be computationally feasible. We also use a larger
+We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order for these
+simulations to be computationally feasible. We also use a larger
 disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
-effects on vortex evolution. We apply an open boundary at
-$r_\mathrm{out}$.
+effects on vortex evolution, and apply an open boundaries at
+$r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
+simulations with $(N_r,N_{\phi})=(256,512)$ were initially carried
+out, which show similar behaviour and trends as the high-resolution
+runs reported below.
 
-Preliminary simulations with same numerical scheme but lower resolution of
- $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
-with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
-Vortex evolution and behaviour are convergent with
-higher resolution simulations as discussed in this section.
- {\bf mention lower resolution preliminary runs - 
-similar results to those presented here (?). want to demonstrate convergence}
+%where also done for the above $\tilde\beta$ values
+%with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
+%Vortex evolution and behaviour are convergent with
+%higher resolution simulations as discussed in this section.
+% {\bf mention lower resolution preliminary runs - 
+%similar results to those presented here (?). want to demonstrate convergence}
 
@@
 \subsection{Generic evolution} 
-
 The linear growth of the RWI and vortex-formation is followed by 
-vortex merging. Merging timescales are independent of $\tilde\beta$ value,
- and by $60P_o$ only one vortex remains as the growth of these modes are now
- accelerated by planetary influence as apposed to `planet off' simulations.
-%i.e. only the $m=1$ mode remains.
-
-The evolution of the amplitude of the $m=1$ surface density mode
- over the radial range of $[2r_h,10r_h]$ is shown in Fig~\ref{lifetimeplot}
+vortex merging. We now find merging timescales independent of
+$\tilde\beta$, and by $60P_o$ only one vortex remains. 
+%as the growth of these modes are now
+%accelerated by planetary influence as apposed to `planet off' simulations.
+% i.e. only the $m=1$ mode remains. 
+The evolution of the amplitude of the $m=1$ surface density mode,
+averaged over $r-r_p\in[2,10]r_h$ is shown in Fig~\ref{lifetimeplot} 
 for different $\tilde\beta$. 
+{\bf any particular reason for the larger radial range for average?
+  did we get bigger vortices?} 
 % {\bf should state (also for planet-off runs)
@@
   0cm 0cm 1cm]{figures/longterm_stability}
-  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$
-    background mode for longterm non-linear simulations corrsponding
-    to vortex lifetime. The longest living mode correpsonds to
-    $\tilde\beta=5.0$ (the red line). \label{lifetimeplot}} 
+  \caption{Evolution of the $m=1$ surface density component,
+    non-dimenionlised by the initial $m=0$ component, for long term
+    simulations with the planet potential kept on. The case with 
+    $\tilde\beta=5.0$ (red) gave the longest-living vortex. {\bf
+      remember to update to 
+      include vortex dissipation for all cases} \label{lifetimeplot}} 
 \end{figure}
@@
 surface density perturbation in this state. During this stage, the 
-vortex grows in intensity. As the vortex forms it has a
- value of $Ro\approx-0.1$ for all $\tilde\beta$
+vortex intensifies. As the vortex forms it has a
+Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
 indicating anti-cyclonic
-motion. As the $m=1$ amplitude grows the relative spin of the vortex
-continuously grows along with it, reaching values of
-$Ro\approx-0.4$ characteristicly for all cooling rates.
+motion. As the $m=1$ amplitude grows the Rossby number also increases
+in magnitude, 
+%relative spin of the vortex
+%continuously grows along with it
+, reaching a characteristic value of $Ro\approx-0.4$ for all cooling
+rates. {\bf is the the value reached just before decay for all cooling?}
 
-Surface density pertibations relative to the intial disk
+Surface density pertibations relative to $t=0$
 at the center of vortices can reach values of $\Delta\Sigma/\Sigma_0=7$ 
@@
 
-{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
-  taken to reach one vortex as a function of cooling}
-{\bf for what cooling are these values quoted for? do
-  these values depend on cooling?}
- {\bf is the evidence for this the fact that growth is not observed in
-  planet-off simulations?}
-
-{\bf note that there
-  are two origins for the surface density boost relative to $t=0$: 1) gap
-  formation, which increases the surface density at the location of
-  the instability (since they happen at surface density max., and 2)
-  vortex instability). what is the surface density boost relative to
-  the gap profile?}
+%{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
+%  taken to reach one vortex as a function of cooling}
+%{\bf for what cooling are these values quoted for? do
+%  these values depend on cooling?}
+% {\bf is the evidence for this the fact that growth is not observed in
+%  planet-off simulations?}   
+%{\bf note that there
+%  are two origins for the surface density boost relative to $t=0$: 1) gap
+%  formation, which increases the surface density at the location of
+%  the instability (since they happen at surface density max., and 2)
+%  vortex instability). what is the surface density boost relative to
+%  the gap profile?}
 
@@
     \label{Vortex2D}} 
-\end{figure}
-
-\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
-Non-linear observations were also made for the `planet-off' simulations.
-After the linear growth phase of the vortices, vortex merging takes hold but
- now on timescales of up to $150P_0$ as the vortices grow as linear
- instabilities discussed previously. The vortex merging time is depedendent on
- the growth rates of the modes and saturation timescales with the
-slowest growing modes in $\tilde\beta=10$ taking the longest. Similar
- quasi-steady vortices form as in the cases with continuous planet interaction
-but the long-term behaviour is significantly
- different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
-The quasi-steady vortices formed have significantly lower density pertibations
-and now slowly decrease in amplitute in time as apposed to previously
-where their growth was supported by planet interactions. The vortices simply
-dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.
-The vortices are never found to undergo the drastic sudden dissipation effects
-that characterize the planet induced growing cases do.
-
-\begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
-    0cm 0cm 1.1cm]{figures/longterm_planetoff}
-  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' 
-simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
 \end{figure}

### 2014-11-19 paper/results.tex
updating results section - MKL

@@
 increased indicates that for planetary gaps, the importance of $h$ on
-the \emph{linear} RWI is through the setting the gap profile, i.e. basic
+the \emph{linear} RWI is through setting up the gap profile, i.e. basic
 state for the instability (as opposed to the linear response).
@@
 
-\subsection{Non-linear evolution without planet
-  influence} \label{nonlinearplanetoff} 
+\subsection{Long term evolution} \label{nonlinearplanetoff} 
 {\bf typical rossby numbers for these `planet-off' vortices? do
@@
 and $5.0$. The vortices simply decay, on a timescale of $O(10^3)$
-orbits. We will see in the that section that this behaviour is very
+orbits. We will see in the next section that this behaviour is very
 different to when the planet potential is kept on. Interestingly,
@@
 disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
-effects on vortex evolution, and apply an open boundaries at
+effects on vortex evolution, and apply open boundaries at
 $r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
@@
 % i.e. only the $m=1$ mode remains. 
-The evolution of the amplitude of the $m=1$ surface density mode,
+The evolution of the amplitude of the $m=1$ surface density component,
 averaged over $r-r_p\in[2,10]r_h$ is shown in Fig~\ref{lifetimeplot} 
@@
 surface density perturbation in this state. During this stage, the 
-vortex intensifies. As the vortex forms it has a
+vortex intensifies. As the vortex forms it has an initial 
 Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
-indicating anti-cyclonic
-motion. As the $m=1$ amplitude grows the Rossby number also increases
+(indicating anti-cyclonic
+motion). As the $m=1$ amplitude grows the Rossby number also increases
 in magnitude, 
@@
 
-Surface density pertibations relative to $t=0$
-at the center of vortices can reach values of $\Delta\Sigma/\Sigma_0=7$ 
-for all cases of $\tilde\beta$ in this quasi-steady state and
-can be found as high as $\Delta\Sigma/\Sigma_0= 9$ times for the 
-$\tilde\beta=10$ case. 
-The gap profiles formed also apply a surface density boost in the vortex 
-region as the vortices are located at the gap edge maxima.
-The average value of relative density pertibation in the outer gap region of 
-$r=r_{p}+2.5r_h$
-has a value $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$ for all times and
-cooling rates.
-Thus the surface density increases caused by gap formation are small 
-compared to the max amplitudes of the vortices.
-Vortex growth is attributed to
-the continuous generation of vorticity by the planet-disc interactions. 
-This is supported by the evidence that in long term simulations where the
-planet is removed, such as seen in Fig \ref{planetofflifetimeplot},
-the amplitude of
-the vortices are not seen to grow without the planet influence.
+The surface density perturbation (relative to $t=0$)  
+can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
+for all cases of $\tilde\beta$ in quasi-steady state, and
+was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
+$\tilde\beta=10$ case. This large increase in surface density is due 
+vortex growth, since there is continuous generation of vorticity by
+planet-disc interaction. This is supported by the observation that in
+the previous simulations without the planet, the amplitude of
+the post-merger vortex does not grow (Fig
+\ref{planetofflifetimeplot}).  We remark that the increase in the
+vortex surface density directly due to the pile-up of material at the
+gap edge (because of gap-opening) is small: the average surface density
+perturbation near the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
+$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.
 
@@
 the $m=1$ amplitude within the simulation time, which occurs over
-timescales of $t\in[20P_0,70P_0]$, i.e. much shorter than the duration of the
-quasi-steady state. Dissipation timescales do not correlate with $\tilde\beta$
- but instead are paramtrized by vortex Mach numbers.
-%{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
+timescale between $20P_0$ to $70P_0$, i.e. much shorter than the duration of the
+quasi-steady state. We did not find this dissipation timescale to
+correlate with $\tilde\beta$. %but instead are paramtrized by vortex Mach numbers.
+%{\bf is this dissipation timescale independent of cooling? Trend is
+%based on mach values which dont follow linear trend }  
 %Eventually
 %the $m=1$ amplitudes are also characterized by 
-After these characteristic drops, the instability is never seen to
-reform at such large overdensities and surface density bumps are never larger
-than that due to the planetary wakes. We designate the time of these
-amplitude drops as the lifetime 
-of the quasi-steady vortex.
-After the vortices dissipate the RWI is not seen to be excited again and no
-vortices are seen to refrom in the simulations time.
+%After these characteristic drops in vortex amplitude, the instability
+%is never seen to reform at such large overdensities: the vortex surface
+%densities are never larger than that due to the planetary wakes. 
+We designate the time at which these amplitude drops as the lifetime 
+of the quasi-steady vortex. After the vortices decay the RWI is 
+not seen to be excited again and no vortices are seen to refrom within the
+simulations time. 
+{\bf Perhaps a snapshot of what the gap edge looks like after vortex death.
+}
 %The instability is not seen to emerge again after the intial vortex dissipation
@@
 % transfer of density from the once tightly held vortex region to the surrounding disk and gap.  
-{\bf Discussed in next section:  question: does instability
-  actually happen again? if it does, but just weaker, that may be
-  because the first vortex has smoothed out the gap edge, so
-  instabilities following that would be weaker.}
+% {\bf Discussed in next section:  question: does instability
+%   actually happen again? if it does, but just weaker, that may be
+%   because the first vortex has smoothed out the gap edge, so
+%   instabilities following that would be weaker.}
 
@@
     $\tilde\beta=0.1$ case during quasi-steady state with a single
-    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
-    similar.% {\bf is this plot truncated? in the text it says the
-      %domain goes to $r=45$?}
-    %The large scale non-axisyemetric $m=1$ overdensity can be
-    %seen. 
+    vortex. The plots for other values of the cooling time
+    $\tilde{\beta}$ are similar.
+    % {\bf is this plot truncated? in the text it says the
+    % domain goes to $r=45$?}
+    % The large scale non-axisyemetric $m=1$ overdensity can be
+    % seen. 
     \label{Vortex2D}} 
@@
 
-{\bf
 \subsection{Additional analysis on vortex dissipation}
-Conjecture: vortex death due to vortex shocking the system and
-dissipating its energy. discuss one case. here we want to explain or
-suggest why vortices die} 
+{\bf are these high resolution results? vortex lifetime in snapshots
+  not consistent with amplitude plots}
+%Conjecture: vortex death due to vortex shocking the system and
+%dissipating its energy. discuss one case. here we want to explain or
+%suggest why vortices die} 
 %Rossby values have dynamics that are coupled with the lifetime and
 %$m=1$ amplitude of the vortices.
+In this subsection we examine the sudden vortex decay observed in our
+simulations in more detail. Fig. \ref{shockplot} show snapshots
+of the vortex leading up to its decay for the case $\tilde{\beta}=1$. 
+The surface density perturbation and the surface density gradient are
+shown. In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
+more compact as it grows before decaying. {\bf rough estimate of
+  vortex aspect-ratio in the three snapshots? 
+  (length/width, by inspection is ok)}
+
+Notice in Fig. \ref{shockplot} the appearence of wakes extending from
+either side of the vortex at $t=1300P_0$ as it grows. These 
+wakes correlate with large gradients in surface density (bottom
+panel), and are first seen in the later half of the quasi-steady state.       
+% As the vortices grow in intensity wakes are seen to extend from them. 
+During this phase the average value of the surface density gradient
+along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf quote something
+  dimensionless, or say `in code units'}  
+Just before the vortices start to decay, we observe this quantity
+sharply increases to $ \sim 6 $, and remains around this value
+until the vortex dies out, at which point the 
+associated Rossby number quickly approaches zero.
 
@@
   } 
-  \caption{Relative density pertibation of vortices (top) and density gradient
- (bottom) for before (left), during (middle) and after dissipation (right).
- Large density gradients are characteristics of shocks and become prominant
- around vortex during dissipation} \label{shockplot}
+  \caption{Evolution of the vortex in the case $\tilde{\beta}=1$
+    just before its sudden decay. The surface density perturbation
+    (top) and the associated surface density gradient (bottom) are shown.
+    \label{shockplot}}
+  % Large surface density gradients become prominant
+  % around vortex during dissipation} 
 \end{figure}
 
-Gradients of surface density are seen to
-spike up in wake-like features around the vortex during dissipation. 
-Snapshots of relative density pertibation and density gradients during the
-quasi steady state, during disspation and just before death are displayed in
-Fig. \ref{shockplot} for $\tilde\beta=1.0$. Quasi-steady vortices have large 
-elliptical shapes, while dissipating vortices are denser and smaller in spacial
-extent.
-As the vortices grow in intensity wakes are seen to extend from them. These
-wakes correlate with large gradients in surface density and are first seen
-when vortices are in the later half of their quasi steady states. During this
-phase the average value of the gradients along the shock region are
-$|\nabla\Sigma| \approx 4 \Sigma$. Just before the vortices start dissipation
-this quantity sharply increases to a value of $ \approx6 \Sigma$ in which it
- remains for rest of vortex lifetime.  
-As the vortex dies off as described above its associated Rossby number quickly
-approaches $0$ indicating a halting of cyclonic motion.
+We also measured large increases in the Mach number near
+the vortex when it begins to decay. Figure \ref{machplot} plots the
+Mach number $M=|\bm{v} - \bm{v}_\mathrm{vor}|/c_s$, where
+$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
+around the disc. {\bf check - i assume this is the definition of mach
+  number used here}. Values in Fig. \ref{machplot} have been averaged over
+a region within $2H$ of the vortex centre.    
+% Mach numbers with repect to the vortex bulk motion around the vortex
+% region of $|r-r_{vortex}|<2H$
+% are shown in  for various cooling rates over time.
+During the quasi-steady state the Mach number increase
+steadily, and for all cases $M$ reaches a maximum at 
+% with larger growth rate for higher $\tilde\beta$.
+a time coincident with the start of vortex decay.
 
-In addition, we also observed large increases in the Mach number near
-the vortex during dissipation, which indicate susceptibility to
-shocks. Mach numbers with repect to the vortex bulk motion around the vortex
- region of $|r-r_{vortex}|<2H$ {\bf (somewhat arbitrary, large enough to have bulk of vortex mach number increases quickly radially away from vortex)}
-%, where $H$ denotes the gas scale height, 
-are shown in Figure \ref{machplot} for various cooling rates over time.
-% $\approx 0.8$ after the intial formation of the vortices for all $\tilde\beta$. 
-During the quasi-steady state the Mach number increases at a
-steady rate with larger growth rate for higher $\tilde\beta$.
-In all cases Mach numbers are seen to spike up at a time which correlates with
-the start of vortex dissipation.
-% Time values in
-% which Mach numbers reach values of 1 approximately correlate with the start 
-%of dissipation and during dissipation values quickly increase to larger max values.
-The timescales of vortex dissipation also correlates with 
-Mach numbers reached during dissipation. The quickest dissipation occurs for 
-$\tilde\beta=0.5$ which is on order of $20P_0$ and reaches Mach number
- $\approx1.4$ while the $\tilde\beta=0.1$ case which dissipates on scales o
-f $90P_0$ reaches Mach numbers $\approx1.2$.
-Thus, the sudden drop in vortex amplitude may  
-be a result of the development of associated spiral shocks dissipating
-its energy \citep[cf.][who 
-  suggested damping due to planet-induced shocks]{fu14}.
-
-{\bf note that fu et al suggested vortex damping by the planet
-  shocks. here, we are suggesting vortex dissipates energy by shocking
-  the disk. our evidence is that we observe vortex growth before it
-  dissipates, and spiral shocks associated with the vortex always
-  develop before vortex death. once it starts shocking the disc it
-  loses energy. if the planet-induced shocks were responsible, death
-  should be more gradual. can we think of other evidences that it's
-  not the planet shock that's responsible?}
-{\bf show
-  this. e.g. vortex before, during and after death. maybe show
-  snapshots of surface density perturbation (vortex) compared with
-  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
-  changes and shocks develop prior ot death. state whether this is
-  generic for all coolings}. 
-{\bf show mach number plot, or quote values. relevant mach
-  number is fluid flow speed with respect to 
-  vortex bulk motion around the disc. does the typical mach number
-  near the vortex just before its death depend on cooling?}
-
-Vortex growth during the quasi-steady state and dissipation also acts to smooth
-out the gap edge. To characterize the gap, we azimuthally and radially average
-the relative density pertibation within $r\in[r_{planet},r_{out}]$ where
- $r_{out}=$ is defined as the outer gap radius defined as point where
-$\langle\Delta \Sigma(r_{out})/\Sigma(t=0)\rangle_{\phi}=0$. Thus a more 
-negative gap depth parameter correlates with a steeper gap edge. In
- Fig. \ref{gapdepth} the gap depth parameter with respect to simulation time
-for $\tilde\beta=1.0$ is shown. The gap depth is seen to slowly increase as
-vortex reaches its max amplutude at $t=1000P_0$ which is followed by a sharp
-increase which correlates with the time of vortex dissipation.
-%Of note is also that the large overdensity the vortex
-%creates during it's quasi-steady state distorts the background
-%disk.  
-The smoothening out of the gap edge during later half of quasi steady state
- is because the vortex has an associated viscosity $O(10^{-2})$.
-%{\bf quote a typical value of alpha viscosity associated with vortex}.
- This acts
-against gap-opening by the planet, so that the condition for the RWI
-becomes less favourable as the vortex grows and hence is why vortices are not
-seen to form again in simulation timescales.
-The sudden smoothening of gap egde during dissipation is due to the large
-influx of density from the vortex to the gap.
-{\bf not sure what you mean by self-limiting here:self-limiting
-  behaviour?} 
-{\bf show this (smoothening of gap edge due to vortex growth),
-  e.g. general vortensity evolution or surface density 
-  evolution}
+Putting the above observations together, we suggest the sudden decay
+in the vortex amplitude is due to shock formation \emph{by the
+  vortex}. When the vortex reaches large amplitude, it begins
+to induce shocks in the surrounding fluid, as supported by the
+increase in Mach number and the appearence of wakes with large surface
+density gradients. The vortex may then lose energy through shock
+dissipation. In addition, a strong vortex (or shock formation) 
+also smoothes out the gap structure that originally gave rise to the
+RWI, this would oppose vortex growth. We examine this below.
 
@@
     \includegraphics[width=\linewidth]{figures/mach}
- \caption{Running plot of Mach number relative to vortex within $2H$ of the
- vortex. Mach numbers increase at rates correlated with cooling rates than
- spike up at times coinciding with the start of vortex dissipation as seen in
- Fig \ref{lifetimeplot}} \label{machplot}
+ \caption{Running-time average {\bf ?} of the Mach number relative to
+   the vortex, averaged over a region within $2H$ of the  
+   vortex centre. This plot can be compared to the evolution of the
+   vortex amplitude shown in Fig. \ref{lifetimeplot}.
+   {\bf remember to update plot to include full simulation duration}
+   \label{machplot}}
 \end{figure}
+
+\subsection{Effect of vortex dissipation on the gap structure} 
+{\bf are these high resolution results?}
+We find vortex growth during the quasi-steady state and its dissipation
+also modifies the gap structure. Here, we 
+define a gap depth parameter by taking the azimuthal and radial average
+of the relative surface density perturbation within $r\in[r_p,r_e]$ where
+$r_{e}$ is defined as the outer gap radius such that 
+$\langle\Delta \Sigma(r_e)/\Sigma(t=0)\rangle_{\phi}=0$. We are
+interested in the region $r \in[r_p,r_e]$ since the vortex is
+located at the outer gap edge. A more negative gap depth parameter
+correlates with a steeper gap edge. 
+{\bf notation clash: rout already used as outer disc boundary - choose
+different notation. updated to $r_e$ for now}
+
+Fig. \ref{gapdepth} shows the evolution of the gap depth parameter
+for $\tilde\beta=1.0$. The magnitude of the gap depth is seen to 
+slowly increase as vortex reaches its maximum amplutude at
+$t=1000P_0$, followed by a sharp decrease coincident with vortex
+dissipation.
+
+We see that the dissipation of the vortex has a `gap-filling'
+effect, and we find that the gap edge is smoothed during this
+process. This can be interpreted as the vortex providing a
+viscosity, and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
+associated with the vortex. This acts against gap-opening
+by the planet, and   
+%The smoothening out of the gap edge during later half of quasi steady state
+%is because the vortex has an associated viscosity $\alpha \sim O(10^{-2})$.
+the condition for the RWI
+becomes less favourable, which explains why vortices do not
+reform again (at least within the simulation timescale).
+
+%The sudden smoothening of gap egde during dissipation is due to the large
+%influx of density from the vortex to the gap.
 
@@
   \includegraphics[width=\linewidth]{figures/gapdepth}
- \caption{Running average of relative density pertibation within the outer half
- of gap} \label{gapdepth}
+  \caption{Running-time average of the gap depth parameter, defined as
+    the relative surface density perturbation averaged over the
+    outer half of the gap.} \label{gapdepth}
 \end{figure}
@@
 
-{\bf describe dependency on
-cooling. Do we have enough data to have a `lifetime v.s. beta'
-  plot? here we want to explain why vortex lives longer with longer
-  cooling, but very long cooling results in shorter lifetimes. 
-  Consider the following quantities as a function of cooling:
-  initial strength of the single vortex; growth rate of single vortex
-  strength from its formation to just before death; ampiltude of
-  vortex just before death; disc temperature in the vortex region.
+% {\bf describe dependency on
+% cooling. Do we have enough data to have a `lifetime v.s. beta'
+%   plot? here we want to explain why vortex lives longer with longer
+%   cooling, but very long cooling results in shorter lifetimes. 
+%   Consider the following quantities as a function of cooling:
+%   initial strength of the single vortex; growth rate of single vortex
+%   strength from its formation to just before death; ampiltude of
+%   vortex just before death; disc temperature in the vortex region.
 
-  we expect the initial $m=1$ vortex (after merging) to be weaker with
-  increasing cooling time because the instability is weaker. (see
-  meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
-  amplitude after linear growth is correlated with the linear growth
-  rate).
+%   we expect the initial $m=1$ vortex (after merging) to be weaker with
+%   increasing cooling time because the instability is weaker. (see
+%   meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
+%   amplitude after linear growth is correlated with the linear growth
+%   rate).
 
-  suppose the vortex shocks when it reaches a certain amplitude (which
-  probably increases with longer cooling rate because it's more
-  difficult to shock a hotter disc). as we
-  increase the cooling time, the initial $m=1$ vortex is weaker, it
-  also grows more slowly because gap-opening is opposed by a hotter
-  disc. the above implies that it takes longer for the vortex reach
-  the shock-inducing amplitude and dissipate.
+%   suppose the vortex shocks when it reaches a certain amplitude (which
+%   probably increases with longer cooling rate because it's more
+%   difficult to shock a hotter disc). as we
+%   increase the cooling time, the initial $m=1$ vortex is weaker, it
+%   also grows more slowly because gap-opening is opposed by a hotter
+%   disc. the above implies that it takes longer for the vortex reach
+%   the shock-inducing amplitude and dissipate.
 
-  but the above is not consistent with non-monotonic behaviour, which
-  predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
-  because we expect the initial vortex to be weakest (check) which
-  means it should take the longest to grow and shock. check
-  whether there's anything special about this case. one possibility is
-  that vortex growth during the quasi-steady state is actually faster
-  because of the raised temperature (li et al 2000 show increased
-  temperature favours RWI - possibiliy evidenced here by a stronger
-  vortex before death than other cases)
+%   but the above is not consistent with non-monotonic behaviour, which
+%   predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
+%   because we expect the initial vortex to be weakest (check) which
+%   means it should take the longest to grow and shock. check
+%   whether there's anything special about this case. one possibility is
+%   that vortex growth during the quasi-steady state is actually faster
+%   because of the raised temperature (li et al 2000 show increased
+%   temperature favours RWI - possibiliy evidenced here by a stronger
+%   vortex before death than other cases)
 
-  perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
-  non-monotonic behaviour
-}
+%   perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
+%   non-monotonic behaviour
+% }
 
@@
     \includegraphics[width=\linewidth]{figures/betaplot}
- \caption{Lifetime of vortices with repect to cooling paramter $\tilde\beta$ for both resolution of $(N_r,N_\phi)=(256,512)$ and $(512,1024)${\bf should I only show higher resolution points?} } \label{betaplot}
+ \caption{Lifetime of vortices with repect to cooling paramter $\tilde\beta$ for both resolution of $(N_r,N_\phi)=(256,512)$ and $(512,1024)${\bf update to high resolution results} } \label{betaplot}
 \end{figure}
@@
 reproduce jump conditions along shocks and which cause vortex dissipation.
- {\bf this is the best reason I could personally think of, for lower beta runs vortices last slightly longer but for higher values it is quite drasticly different.
-Not sure how numerical dissipation as you mentioned would be a factor in this case}
+ {\bf this is the best reason I could personally think of, for lower
+   beta runs vortices last slightly longer but for higher values it is
+   quite drasticly different. Not sure how numerical dissipation as
+   you mentioned would be a factor in this case}
+
 %be the optimum lifetime for long-term simulations of

### 2014-11-19 paper/results.tex
Updated figures

@@
 {\bf any particular reason for the larger radial range for average?
-  did we get bigger vortices?} 
+  did we get bigger vortices?:Yes vortices get quite large in quais steady,
+ changing the planet off averaging range to [2,10] changes the plots negligibly
+ if we want to just use this value range for consistancy } 
 % {\bf should state (also for planet-off runs)
@@
 , reaching a characteristic value of $Ro\approx-0.4$ for all cooling
-rates. {\bf is the the value reached just before decay for all cooling?}
+rates. {\bf is the the value reached just before decay for all cooling?:yes 
+values peak in small range [-0.35,-0.45] for all beta}
 
@@
 around the disc. {\bf check - i assume this is the definition of mach
-  number used here}. Values in Fig. \ref{machplot} have been averaged over
+  number used here:yes this was the definition I used}. Values in Fig. \ref{machplot} have been averaged over
 a region within $2H$ of the vortex centre.    
@@
 \subsection{Effect of vortex dissipation on the gap structure} 
-{\bf are these high resolution results?}
+{\bf are these high resolution results?:yes}
 We find vortex growth during the quasi-steady state and its dissipation
@@
     outer half of the gap.} \label{gapdepth}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/gapchange}
+  \caption{Azimuthally averaged gap profiles of relative surface change for
+ before and after dissipation for $\tilde\beta=1.0$}
 \end{figure}
@@
    you mentioned would be a factor in this case} 
+{\bf you mentioned resolution and convergence to me being a known issue in a previous email}
 

### 2014-11-21 paper/results.tex
major updates by MKL

--- 2014-11-19 paper/results.tex
@@
 azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
- times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
+ times the local surface density. 
 This procedure allows us to analyse the growth of 
@@
 $r_\mathrm{out}=25r_\mathrm{in}$. We consider   
-%for a time up to $200P_0$ allowing high resolution and detection of high
-%$m$ structures.
-cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
+cases with $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
 and slowly cooled discs. % {\bf boundary condition for these sims?}
@@
 
-%For all cases the gap width is $\approx 5r_{hill}$
-%which is set by torque balances in the disk \citep{crida06}.
-
 Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
@@
 \citep{crida06}. 
-% which is a value directly correlated to temperature in our model. 
-%For a planet to carve a gap in the surrounding material it has been
-%shown that $q>H^3$ is needed \citep{crida06}. 
 This indicates that for hotter discs (higher
@@
 %1$
-
-%This is seen in our simulations as changes in $\tilde{\beta}$
-%result in changes of $h$ by less than $20\%$ yet gap depth almost
-%doubles.
 
@@
 stable with longer cooling times.
-
-%Charactereristicly
-%for such disk-planet Rossby vortices the generalized vortensity
-%minima correspond to density maxima in the disk. 
-%Since generalized vortensity extrema are also
-% correspondingly less pronounced and density gradients become smaller
-% we expect that the stability of gaps to increase with larger
-% $\tilde{\beta}$.
 
@@
 \subsection{Axisymmetric stability}
-%Check that the gap profiles are stable against axisymmetric
-%stability. Can use analytical criteria given in section 3.1 of Li et
-%al (2000, ApJ, 533, 1023). No plot needed, just statements. 
-%}
 The intial planet-disc interaction form bumps and grooves in the gap profiles
@@
 is the square of the Brunt-V\"ais\"al\"a frequency. 
-%for the disk due to entropy
-%variation which is non-zero for non-adiabatic models such as being
-%considered. 
 At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
@@
 $\sim 0.002$ (code units) for all $\tilde\beta$. The
-Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. The Solberg-Hoiland
-criteria is similary satisfied for the entire disc throughout the
-simulations. % with little change of values. 
+Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. 
+{\bf Brunt frequency the same for all cooling?}
+The Solberg-Hoiland criteria is similary satisfied for the entire disc throughout the 
+simulations. {\bf i assume this means the azimuthally averaged
+  values. do you ever get $N^2<0$ anywhere in the 2D disc?}  
 Thus for all values of $\tilde\beta$ the planet-induced gaps are
@@
 \subsection{Non-axisymmetric instability}\label{linear}
-%{\bf present `planet-off' simulations. one `fourier mode v.s. time'
-%  plot to show growth of linear instability. do an adiabatic case (or
-%  extremely long cooling time) to see the effect of heated gap edge
-%  (i.e. temp doesn't go back to t=0 value too quickly)
-%  compare growth rate and dominant
-%  m as function of beta (table). 2D figs to contrast (also used to
-%  show it's the minimum in generalized pv that goes unstable.    
-%  result: increasing cooling time makes the gap
-%  more stable, and favors lower m. note: the `basic state' should be
-%  the system at t=30 after azimuthal average. linear results should
-%  have small perturbations. 
-%}
-
-%After the azimuthal averaging, the small pertibation excites rapidly
-%growing instabilities by the RWI.
-
 We now examine the evolution of the gap for $t>30P_0$, with the
@@
 10$, we observe exponential growth of non-axisymmetric
-structures. % after the 
-%planet potential is switched off and the system subject to random
-%perturbations.  
+structures.  
 An example is shown in Fig. \ref{linearmodes} for 
@@
 
-% which result in hotter gaps since the heating
-%due
-%This is as expected from the
-%intial gap profiles, because 
-% and the gap forming criteria discussed in the
-%previous section.
-
 \begin{figure}
@@
     for the vortices to grow and become visible with increasing cooling time. 
-    % correspondingly increase to capture the longer growth phases of
-    % the instabilities. 
     \label{2Dlinear} 
-%{\bf I'm curious to know if plots clearer if
-%      perturbations are measured relative to t=30? remember t=30 is
-%      our reference state for instability. But we'll probably keep
-%      these for consistency (later non-linear plots are all measure
-%      with respect to t=0).}
   } 
@@
 \label{adiabatic_section}
-%{\bf a case with extrmely long cooling time (or adiabatic run) in
-%  order to look at effect of heated gap edge.  prelim result: not
-%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
-%  purely adiabatic disk. change in $h$ not important for linear
-%  perturbations (but important for setting up the basic state). 
-%}
+
 The above `planet-off' simulations are not formally linear
@@
 than the instability growth time, 
-%
-%as we do not capture the effects of a heated
-%disk on the instability growth. 
 $t_c<\gamma^{-1}$.  
@@
 have a steady basic state to formulate a standard linear stability 
-problem. %This means the disc temperture of $h=0.05$ in the the above cases
+problem.
 
@@
 $\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
-%To compare our 'planet-off' analysis with a true
-%linear stability analysis 
 In this simulation the cooling rate is slow enough that the gap 
 temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
-only marginally over the instability growth timescale. %heat is
-                                %retained/maintained.
+only marginally over the instability growth timescale.
 
@@
 the original temperature of $h=0.05$, so linear growth rates are not expected
-to change significantly %according to the linear theory presented in
-\citep{li00}.
+to change significantly \citep{li00}.
 
@@
 state for the instability (as opposed to the linear response).
-
-%results in shallower gaps
-%which is expected to be
-
-%the difference in growth rates between this and the
-%original disc temperature of $h=0.05$ changes almost insignificantly
-%as shown by linear theory \citep{li00}. 
-%Although linear
-%theory predicts higher growth rates with increasing $h$
-
-%This indicate that a
-%continuously hot gap edge is not as important for development of
-%vortices as the heating effects on the formation of the intial gap
-%state.
-
-%{\bf
-%If there's time, do low resolution planet off simulations but keep the planet on
-%until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
-%are generally deeper (including the nearly adiabatic case). Although
-%with the planet instability may already appear by $t=40P_0$, the
-%azimtuhal average performed prior to perturbations will erase it.}
 
@@
 
-%Similar
-%quasi-steady vortices form as in the cases with continuous planet interaction
-%but the long-term behaviour is significantly
-%different without the planet as seen in 
-%The quasi-steady vortices formed have significantly lower density pertibations
-%and now slowly decrease in amplitute in time as apposed to previously
-%where their growth was supported by planet interactions. 
-%The vortices simply
-%dissipate viscously with dissipation rate decreasing with larger $\tilde\beta$.
-
-%Non-linear observations were also made for the `planet-off'
-%simulations.
-% as the vortices grow as linear
-% instabilities discussed previously.
-
-%The vortices are never found to undergo the drastic sudden dissipation effects
-%that characterize the planet induced growing cases do.
-
 \begin{figure}
@@
   gap-edge vortices with finite cooling time}\label{nonlinear} 
-%{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
-%  to plot vortex lifetime v.s. beta? table: 
-%  averaged quantities over quasi-steady state: aspect-ratio (to
-%  compare with fu at al), rossby number (vortex strength
-%  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
-%  only inviscid cases. describe evolution of one case. main
-%  conclusion: longer vortex lifetime with increasing cooling time (up
-%  to some optimal timescale). vortex death: induced-shock and/or
-%  smoothing the gap edge. describe simulation setup, resolution?
-%  should mention that results consistent with lower-resolution prelim
-%  runs. maybe torques? 
-%}
-
 We now examine long-term simulations of gap-edge vortices for
 $\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
-throughout.  
-% up to a total of
-%$2000P_0$. 
-%Planets were left free to interact dynamiclly with the disk
-%and vortices after $t=30P_0$ as apposed to previous section.  
-We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order for these
+throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
+for these 
 simulations to be computationally feasible. We also use a larger
@@
 
-%where also done for the above $\tilde\beta$ values
-%with the additional inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$.
-%Vortex evolution and behaviour are convergent with
-%higher resolution simulations as discussed in this section.
-% {\bf mention lower resolution preliminary runs - 
-%similar results to those presented here (?). want to demonstrate convergence}
-
-%An open boundry
-%condition was used in the outer edge of
-% allowing significant room for
-%linblad resonances of the vortices.
-
 \subsection{Generic evolution} 
@@
 $\tilde\beta$, and by $60P_o$ only one vortex remains. 
-%as the growth of these modes are now 
-%accelerated by planetary influence as apposed to `planet off' simulations.
-% i.e. only the $m=1$ mode remains. 
 The evolution of the amplitude of the $m=1$ surface density component,
-averaged over $r-r_p\in[2,10]r_h$ is shown in Fig~\ref{lifetimeplot} 
+averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
 for different $\tilde\beta$. 
-{\bf any particular reason for the larger radial range for average?
-  did we get bigger vortices?} 
-% {\bf should state (also for planet-off runs)
-% what the radial range is used for averaging the Fourier amplitudes}
+%{\bf any particular reason for the larger radial range for average?
+%  did we get bigger vortices?:Yes vortices get quite large in quais steady,
+% changing the planet off averaging range to [2,10] changes the plots negligibly
+% if we want to just use this value range for consistancy }
 
@@
     non-dimenionlised by the initial $m=0$ component, for long term
-    simulations with the planet potential kept on. The case with 
-    $\tilde\beta=5.0$ (red) gave the longest-living vortex. {\bf
-      remember to update to 
-      include vortex dissipation for all cases} \label{lifetimeplot}} 
+    simulations with the planet potential kept on.\label{lifetimeplot}}   
 \end{figure}
@@
 %continuously grows along with it
-, reaching a characteristic value of $Ro\approx-0.4$ for all cooling
-rates. {\bf is the the value reached just before decay for all cooling?}
+ reaching a characteristic minimum value of $Ro\approx-0.35$ to $-0.45$ for
+all cooling rates. 
+{\bf Maybe useful to have a Rossby number v.s. time plot to complement
+  mode amplitude plot. 
+}
+%{\bf is the the value reached just before decay for all cooling?:yes 
+%values peak in small range [-0.35,-0.45] for all beta}
 
+We find the vortices can become significantly over-dense. 
 The surface density perturbation (relative to $t=0$)  
@@
 was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
-$\tilde\beta=10$ case. This large increase in surface density is due 
+$\tilde\beta=10$ case. {\bf any trend between maximum over-density
+  achieved and cooling?}
+This large increase in surface density is due to  
 vortex growth, since there is continuous generation of vorticity by
@@
 
-%{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
-%  taken to reach one vortex as a function of cooling}
-%{\bf for what cooling are these values quoted for? do
-%  these values depend on cooling?}
-% {\bf is the evidence for this the fact that growth is not observed in
-%  planet-off simulations?}   
-%{\bf note that there
-%  are two origins for the surface density boost relative to $t=0$: 1) gap
-%  formation, which increases the surface density at the location of
-%  the instability (since they happen at surface density max., and 2)
-%  vortex instability). what is the surface density boost relative to
-%  the gap profile?}
+Fig. \ref{lifetimeplot} shows the duration of the quasi-steady state
+varies with the cooling rate: for
+$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
+$t\sim900P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
+$t\sim1200P_0$. This non-monotonic dependence suggests that there
+exists an optimal cooling rate to maximise the vortex lifetime. We
+will discuss this issue futher in
+\S\ref{lifetime_discuss}. However, the decay timescale decreases with
+increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
+whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
+amplitude to decay significantly after reaching maximum.
 
-Interestingly, in most cases we observe a sudden drop in 
-the $m=1$ amplitude within the simulation time, which occurs over
-timescale between $20P_0$ to $70P_0$, i.e. much shorter than the duration of the
-quasi-steady state. We did not find this dissipation timescale to
-correlate with $\tilde\beta$. %but instead are paramtrized by vortex Mach numbers.
-%{\bf is this dissipation timescale independent of cooling? Trend is
-%based on mach values which dont follow linear trend }  
-%Eventually
-%the $m=1$ amplitudes are also characterized by 
-%After these characteristic drops in vortex amplitude, the instability
-%is never seen to reform at such large overdensities: the vortex surface
-%densities are never larger than that due to the planetary wakes. 
-We designate the time at which these amplitude drops as the lifetime 
-of the quasi-steady vortex. After the vortices decay the RWI is 
-not seen to be excited again and no vortices are seen to refrom within the
-simulations time. 
-{\bf Perhaps a snapshot of what the gap edge looks like after vortex death.
+%These
+%timescales are typically much shorter than the duration of the
+%quasi-steady state.
+
+{\bf question: vortex amplitude plot suggest just after linear growth,
+  vortex amplitudes are smaller with increasing beta, but later on in
+  the quasi-steady state, this trend reverses?  
 }
-%The instability is not seen to emerge again after the intial vortex dissipation
-% due to the smoothening out of the gap edge during dissipation by the
-% transfer of density from the once tightly held vortex region to the surrounding disk and gap.  
-% {\bf Discussed in next section:  question: does instability
-%   actually happen again? if it does, but just weaker, that may be
-%   because the first vortex has smoothed out the gap edge, so
-%   instabilities following that would be weaker.}
+%The
+%decay timescale appears to decrease with increasing cooling time. For
+%$\tilde{}$  
+%Interestingly, in most cases we observe a sudden drop in 
+%the $m=1$ amplitude within the simulation time, which occurs over
+%timescale between $20P_0$ to $70P_0$, i.e. much shorter than the duration of the
+%quasi-steady state. 
+%We did not find this dissipation timescale to
+%correlate with $\tilde\beta$. 
+%We designate the time at which these amplitude drops as the lifetime 
+%of the quasi-steady vortex. After the vortices decay the RWI is 
+%not seen to be excited again and no vortices are seen to refrom within the
+%simulations time. 
+{\bf Perhaps a 2D snapshot of what the gap edge looks like after
+  vortex death (like the one in the poster). defer definition of
+  vortex lifetime to later, when we discuss it. 
+}
 
@@
 
-\subsection{Additional analysis on vortex dissipation}
-{\bf are these high resolution results? vortex lifetime in snapshots
-  not consistent with amplitude plots}
-%Conjecture: vortex death due to vortex shocking the system and
-%dissipating its energy. discuss one case. here we want to explain or
-%suggest why vortices die} 
-%Rossby values have dynamics that are coupled with the lifetime and
-%$m=1$ amplitude of the vortices.
-In this subsection we examine the sudden vortex decay observed in our
+\subsection{Additional analysis on vortex decay}
+In this subsection we examine the vortex decay observed in our
 simulations in more detail. Fig. \ref{shockplot} show snapshots
-of the vortex leading up to its decay for the case $\tilde{\beta}=1$. 
-The surface density perturbation and the surface density gradient are
-shown. In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
-more compact as it grows before decaying. {\bf rough estimate of
+of the vortex for the case $\tilde{\beta}=1$. The plots show the surface
+density perturbation and the surface density gradient during
+quasi-steady state ($t=700P_0$), when the $m=1$ amplitude begins to
+decrease ($t=1300P_0$) and just after the rapid amplitude decay
+($t=1510P_0$).
+
+In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
+more compact as it grows before decaying. 
+{\bf rough estimate of
   vortex aspect-ratio in the three snapshots? 
   (length/width, by inspection is ok)}  
+Notice in Fig. \ref{shockplot} the appearence of wakes extending from
+either side of the vortex at $t=1300P_0$. These 
+wakes correlate with large gradients in surface density (bottom
+panel), and are first seen in the later half of the quasi-steady
+state. We find the time at which 
+the vortex begins to decay coincides with the emergence of these
+wakes. {\bf true statement?}
 
-Notice in Fig. \ref{shockplot} the appearence of wakes extending from
-either side of the vortex at $t=1300P_0$ as it grows. These 
-wakes correlate with large gradients in surface density (bottom
-panel), and are first seen in the later half of the quasi-steady state.       
 % As the vortices grow in intensity wakes are seen to extend from them. 
-During this phase the average value of the surface density gradient
+During quasi-steady state the average value of the surface density gradient
 along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf quote something
   dimensionless, or say `in code units'}  
-Just before the vortices start to decay, we observe this quantity
+Just before the $m=1$ amplitude begins to decrease, we observe this quantity
 sharply increases to $ \sim 6 $, and remains around this value
-until the vortex dies out, at which point the 
-associated Rossby number quickly approaches zero.
+until the vortex dies out, at which point the associated Rossby number
+quickly approaches zero.
 
@@
   } 
-  \caption{Evolution of the vortex in the case $\tilde{\beta}=1$
-    just before its sudden decay. The surface density perturbation
+  \caption{The vortex in the case $\tilde{\beta}=1$
+    during quasi-steady state (left), at maximum amplitude of the
+    $m=1$ surface density component, 
+    (middle), and just after the decay in the $m=1$ amplitude
+    (right). The surface density perturbation
     (top) and the associated surface density gradient (bottom) are shown.
@@
 We also measured large increases in the Mach number near
-the vortex when it begins to decay. Figure \ref{machplot} plots the
-Mach number $M=|\bm{v} - \bm{v}_\mathrm{vor}|/c_s$, where
+the vortex as it reaches maximum $m=1$ amplitude and begins to decay. 
+Figure \ref{machplot} plots the Mach number $M=|\bm{v} -
+\bm{v}_\mathrm{vor}|/c_s$, where 
 $\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
-around the disc. {\bf check - i assume this is the definition of mach
-  number used here}. Values in Fig. \ref{machplot} have been averaged over
+around the disc. Values in Fig. \ref{machplot} have been averaged over
 a region within $2H$ of the vortex centre. 
-% Mach numbers with repect to the vortex bulk motion around the vortex
-% region of $|r-r_{vortex}|<2H$
-% are shown in  for various cooling rates over time.
 During the quasi-steady state the Mach number increase
@@
 % with larger growth rate for higher $\tilde\beta$.
-a time coincident with the start of vortex decay.
+a time coincident with the start of vortex decay. {\bf check if
+  statement is true}
 
-Putting the above observations together, we suggest the sudden decay
+Putting the above observations together, we suggest the decay
 in the vortex amplitude is due to shock formation \emph{by the
@@
 increase in Mach number and the appearence of wakes with large surface
-density gradients. The vortex may then lose energy through shock
+density gradients. The vortex may lose energy through shock
 dissipation. In addition, a strong vortex (or shock formation) 
-also smoothes out the gap structure that originally gave rise to the
-RWI, this would oppose vortex growth. We examine this below.
+can smooth out the gap structure that originally gave rise to the
+RWI, which would oppose vortex growth. We examine this below.
 
@@
     vortex amplitude shown in Fig. \ref{lifetimeplot}.
-   {\bf remember to update plot to include full simulation duration}
     \label{machplot}}
@@
 
-\subsection{Effect of vortex dissipation on the gap structure} 
-{\bf are these high resolution results?}
-We find vortex growth during the quasi-steady state and its dissipation
-also modifies the gap structure. Here, we 
-define a gap depth parameter by taking the azimuthal and radial average
-of the relative surface density perturbation within $r\in[r_p,r_e]$ where
-$r_{e}$ is defined as the outer gap radius such that 
-$\langle\Delta \Sigma(r_e)/\Sigma(t=0)\rangle_{\phi}=0$. We are
-interested in the region $r \in[r_p,r_e]$ since the vortex is
-located at the outer gap edge. A more negative gap depth parameter
-correlates with a steeper gap edge. 
-{\bf notation clash: rout already used as outer disc boundary - choose
-different notation. updated to $r_e$ for now}
-
-Fig. \ref{gapdepth} shows the evolution of the gap depth parameter
-for $\tilde\beta=1.0$. The magnitude of the gap depth is seen to 
-slowly increase as vortex reaches its maximum amplutude at
-$t=1000P_0$, followed by a sharp decrease coincident with vortex
-dissipation.
-
-We see that the dissipation of the vortex has a `gap-filling'
-effect, and we find that the gap edge is smoothed during this
-process. This can be interpreted as the vortex providing a
-viscosity, and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
+\subsection{Vortex decay and gap structure}  
+We find vortex decay also modifies the gap
+structure. Fig. \ref{gap_smoothed} shows the gap profile before and
+after vortex decay for the case $\tilde{\beta}=1$. The vortex resides
+around the local surface density maximum at the outer gap edge ($r\sim r_p +
+6r_h$). We see that after its amplitude has decayed ($t\sim1500P_0$,   
+Fig. \ref{lifetimeplot}), this local surface density maximum is 
+smoothed out.           
+%and we find that the gap edge is smoothed during this
+%process. 
+This can be interpreted as the vortex providing a viscosity;  
+and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
 associated with the vortex. This acts against gap-opening
-by the planet, and   
-%The smoothening out of the gap edge during later half of quasi steady state
-%is because the vortex has an associated viscosity $\alpha \sim O(10^{-2})$.
-the condition for the RWI
+by the planet, so the condition for the RWI
 becomes less favourable, which explains why vortices do not
 reform again (at least within the simulation timescale). 
-
-%The sudden smoothening of gap egde during dissipation is due to the large
-%influx of density from the vortex to the gap.
+%Thus, the
+%time needed for the vortex to grow to sufficient amplitude to modify
+%the gap structure
 
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/gapdepth}
-  \caption{Running-time average of the gap depth parameter, defined as
-    the relative surface density perturbation averaged over the
-    outer half of the gap.} \label{gapdepth}
+  \includegraphics[width=\linewidth]{figures/gapchange}
+  \caption{Azimuthally averaged profiles of the relative surface
+    density perturbations for  $\tilde\beta=1.0$
+    before (solid) and after vortex decay
+    (dashed). {\bf remember to correct the time quoted in the
+      legend} \label{gap_smoothed}}  
 \end{figure}
 
-\subsection{Vortex lifetimes as a function of cooling rate}
+{\bf It's a good idea to display the smoothening of the outer gap
+  edge with time. 
+  The gap depth parameter correlates
+  well with vortex decay, but its sharp decrease in magnitude gives
+  the impression that the gap is 
+  `filled', which doesn't reflect Fig. \ref{gap_smoothed}. Visually,
+  the gap depth is the same, but the local surface density maximum at
+  the outer gap edge is smoothed. Perhaps we can come up with another
+  way to characterizing the sharpness of the peak? then we can show
+  how this peak is smoothed during vortex decay
+}
 
-% {\bf describe dependency on
-% cooling. Do we have enough data to have a `lifetime v.s. beta'
-%   plot? here we want to explain why vortex lives longer with longer
-%   cooling, but very long cooling results in shorter lifetimes. 
-%   Consider the following quantities as a function of cooling:
-%   initial strength of the single vortex; growth rate of single vortex
-%   strength from its formation to just before death; ampiltude of
-%   vortex just before death; disc temperature in the vortex region.
+% Here, we are interested in the region
+% $r>r_p$ since the vortex is located at the outer gap edge. We 
+% define a gap depth parameter by averaging the relative surface density
+% perturbation within $r\in[r_p,r_e]$ where $r_{e}$ is defined as the
+% outer gap radius such that $\langle\Delta
+% \Sigma(r_e)/\Sigma(t=0)\rangle_{\phi}=0$.  A more negative gap depth
+% parameter correlates with a steeper gap edge.
 
-%   we expect the initial $m=1$ vortex (after merging) to be weaker with
-%   increasing cooling time because the instability is weaker. (see
-%   meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
-%   amplitude after linear growth is correlated with the linear growth
-%   rate).
+% Fig. \ref{gapdepth} shows the evolution of the gap depth parameter
+% for $\tilde\beta=1.0$. The magnitude of the gap depth is seen to 
+% slowly increase as vortex reaches its maximum amplutude at
+% $t\sim1000P_0$, followed by a sharp decrease coincident with vortex 
+% decay.
 
-%   suppose the vortex shocks when it reaches a certain amplitude (which
-%   probably increases with longer cooling rate because it's more
-%   difficult to shock a hotter disc). as we
-%   increase the cooling time, the initial $m=1$ vortex is weaker, it
-%   also grows more slowly because gap-opening is opposed by a hotter
-%   disc. the above implies that it takes longer for the vortex reach
-%   the shock-inducing amplitude and dissipate.
+% \begin{figure}
+%   \includegraphics[width=\linewidth]{figures/gapdepth}
+%   \caption{Running-time average of the gap depth parameter, defined as
+%     the relative surface density perturbation averaged over the
+%     outer half of the gap.} \label{gapdepth}
+% \end{figure}
 
-%   but the above is not consistent with non-monotonic behaviour, which
-%   predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
-%   because we expect the initial vortex to be weakest (check) which
-%   means it should take the longest to grow and shock. check
-%   whether there's anything special about this case. one possibility is
-%   that vortex growth during the quasi-steady state is actually faster
-%   because of the raised temperature (li et al 2000 show increased
-%   temperature favours RWI - possibiliy evidenced here by a stronger
-%   vortex before death than other cases)
+\subsection{Vortex lifetimes as a function of cooling
+  rate}\label{lifetime_discuss} 
+{\bf the discussion here may need to be changed, depending on
+  updated/new definitions of vortex lifetime. here, i'm assuming we
+  still get the non-monotonic dependence, which may or may not remain
+  true. 
+}
 
-%   perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
-%   non-monotonic behaviour
-% }
+We now examine vortex lifetimes as function of the imposed cooling
+times. We characterise the vortex lifetime as {\bf insert
+  definition(s).} and plot this in Fig. \ref{betaplot}. 
+Increasing the cooling timescale increases the vortex
+lifetime up to $\tilde{\beta}=1$ {\bf (judging from
+  Fig. \ref{lifetimeplot} without a quantitative definition)}, after
+which vortex lifetimes decrease again.
+
+%Fig. \ref{lifetimeplot} and Fig. \ref{betaplot} shows that increasing 
+%$\tilde{\beta}$
+%increases the lifetime of the vortex up to a critical 
+%value $\tilde{\beta}=5.0$ for which the vortex lasted to a simulation time of
+%$1250P_0$ before starting to dissipate. This is consistent in time with the
+%vortex introducing shocks to the system.
 
@@
   \includegraphics[width=\linewidth]{figures/betaplot}
- \caption{Lifetime of vortices with repect to cooling paramter $\tilde\beta$ for both resolution of $(N_r,N_\phi)=(256,512)$ and $(512,1024)${\bf update to high resolution results} } \label{betaplot}
+  \caption{Lifetime of vortices as a function of the cooling paramter 
+    $\tilde\beta$. {\bf maybe useful to over-plot other timescales,
+      e.g. time taken to reach max mach, time for rossby number to
+      reach zero.}\label{betaplot}} 
 \end{figure}
 
-Fig. \ref{lifetimeplot} and Fig. \ref{betaplot} shows that increasing 
-$\tilde{\beta}$
-increases the lifetime of the vortex up to a critical 
-value $\tilde{\beta}=5.0$ for which the vortex lasted to a simulation time of
- $1250P_0$ before starting to dissipate. This is consistent in time with the
- vortex introducing shocks to the system.
+In the previous section, we observed that vortices began to decay when
+it starts to induce shocks. We therefore suggest that the vortex
+timelife is determined by the time needed for the vortex to grow to
+sufficient amplitude to induce shocks in the surrounding
+fluid. Based on this hypothesis, we suggest below competing
+factors that may result in a non-monotonic dependence of vortex
+lifetimes on the cooling timescale.
 
-Our `planet-off' simulations
-indicate weaker vortices are formed initially with increased cooling
-times. This is because it has been shown that vortex amplitude is dependent on 
-growth rate of the linear modes \citep{meheut2013}.
-At the beginning of the vortex quasi-steady state $t=100P_0$ we find
-that the relative surface density boost at center of vortex 
-is $\Delta\Sigma/\Sigma_0=2.5$ for $\tilde\beta=0.1$ 
-and decreases down to a minimum value of $\Delta\Sigma/\Sigma_0=1.48$ for 
-$\tilde\beta=10$.
+\subsubsection{Factors contributing to longer vortex lifetimes}
+It has been shown that the amplitude at which the RWI saturates 
+increases with the growth rate of the linear instability  
+\citep{meheut2013}. Our `planet-off' simulations yield slower growth
+rates with increasing cooling times, which suggest weaker vortices are
+formed initially with increasing $\tilde{\beta}$. This is consistent
+with the present simulations: at the beginning of the 
+quasi-steady state ($t\sim100P_0$) we find the relative surface density
+boost at centre of vortex is $\Delta\Sigma/\Sigma_0=2.5$ for
+$\tilde\beta=0.1$ and $\Delta\Sigma/\Sigma_0=1.48$ for
+$\tilde\beta=10$.  {\bf rossby numbers?}
 
- After merging, the growth of the single vortex, mediated by disc-planet
-interaction, is expected to be slower with increased cooling time
-because gap-opening becomes more difficult in a hotter
+After merging, the growth of the single vortex is mediated by disc-planet
+interaction. However, gap-opening becomes more difficult in a hotter 
 disc. Furthermore, the increased sound-speed suggest the vortex should    
 reach larger amplitudes in order to induce shocks.
-These expectations
-imply that it takes longer for the vortex to induce shocks and
-dissipate, and hence longer lifetimes with increasing
+
+These considerations suggests that, with increased cooling times, 
+it takes longer for the vortex grow to sufficient amplitude to induce
+shocks and dissipate, implying longer lifetimes with increasing 
 $\tilde{\beta}$.
 
-However, Fig. \ref{lifetimeplot} shows that increasing the cooling
-time further to $\tilde{\beta}=10$ actually results in a
-\emph{shorter} vortex lifetime. Notice for $\tilde{\beta}=10$ the vortex growth 
-during the quasi-steady state is faster than for
-$\tilde{\beta}=5$.
- This may be attributed to the higher temperature
-reached in the latter case ($h\simeq 0.07$) in comparison with the former
-($h\simeq 0.06$)
-% {\bf temperature for $\tilde{\beta}=5$?}
-, since \cite{li00} showed that the RWI grows
-faster with increased disc temperature. At long cooling times, this
-effect may dominate over the weaker gap-opening effect at higher $h$.
+{\bf hotter disc $\to$ gap edges less sharp, still true during quasi
+  steady state?}
 
-This effect can be seen in Mach number around the vortices where the
-growth of Mach number with respect to time is increasing with cooling rate.
-Thus the intially weaker formed vortices for higher cooling rate grow
-substantialy faster than the sound speed and can shock the system in earlier
-times.
+\subsubsection{Factors contributing to shorter vortex lifetimes} 
+%However, Fig. \ref{lifetimeplot} shows that very long cooling
+%times (e.g. $\tilde{\beta}=10$) actually results in a \emph{shorter} vortex
+%lifetime. 
+Notice in Fig. \ref{lifetimeplot}, for $\tilde{\beta}=10$ the vortex 
+growth during the quasi-steady state is in fact faster than for
+$\tilde{\beta}=5$. For example, at $t\sim 500P_0$ the vortex with   
+$\tilde{\beta}=10$ has the largest amplitude. This is also reflected
+in Fig. \ref{machplot}, where the Mach number reaches its maximum
+value sooner for $\tilde{\beta}=10$ than for $\tilde{\beta}=5$.
 
-Thus, unlike our `planet-off' simulations we find that there is
-a cooling time $t_c$ which maximises the vortex lifetime. In the long term
- 'planet-off' simulations we find that the vortices dissipate very gradually
- and visciously dependent on cooling rate.
-% {\bf to make this comparison we need to discuss vortex lifetimes for planet-offsimulations as well} 
-This
-non-monotonic behaviour was also reported by \cite{fu14}, who
-performed isothermal disc-planet simluations with different fixed
-aspectratios.   
-%Cooling rates with
-%$\tilde\beta>5.0$ showed lifetimes that were well below the max
-%lifetime. 
-By analyzing the radial region around the vortex within
- $[2r_{h},10r_{h}]$
-% {\bf define this region}
- we find the
-$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
-$h\approx0.06$ on average throughout its lifetime. This value of $h$ was also
- found to maximize vortex
-lifetimes in the isothermal simulations of \cite{fu14}. Note that $h$ is a fixed
-quantity in \cite{fu14} while in this work $h$ is a dynamical variable dependent
- on the viscous heating and cooling rate as by choice of energy equation.
+The above observation is consistent with the RWI being favoured by higher
+disc temperatures \citep{li00,lin12c}, which corresponds to longer
+cooling times in our case. While our `planet-off' simulations indicate
+this is unimportant for the linear instability, it may contribute 
+significantly to the vortex growth during quasi-steady state, when one
+considers very long cooling times (e.g. $\tilde{\beta}=10$). This
+effect alone would lead to a shorter vortex lifetime by allowing it to
+induce shocks sooner.
 
-These results were reproduced in lower resolution simulations as well. Vortex
-shocks were similarly seen during vortex dissipation and non-monotonic
- dependence on lifetime with $\tilde\beta$ was similarly found.
-Vortex lifetimes are considerably longer in these runs with the maximum
- lifetime longer than the simulation timescale of $2000P_o$ for 
-$\tilde\beta=5.0$ which again has $h\approx0.06$.
-This inconsistancy in timescales, may be due to resolution as a finer 
-grid spacing is required to accurately
-reproduce jump conditions along shocks and which cause vortex dissipation.
- {\bf this is the best reason I could personally think of, for lower
-   beta runs vortices last slightly longer but for higher values it is
-   quite drasticly different. Not sure how numerical dissipation as
-   you mentioned would be a factor in this case}
-
-%be the optimum lifetime for long-term simulations of
-%isothermal disks with varying intial temperature profiles $h$.
-
-%Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
-%the the non-montonic form of the lifetime. As the $\tilde\beta$
-%increases so does the stability of the gap edge as seen in
-%section~\ref{linear}, however as $\tilde\beta$ increases so does the
-%$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
-%instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
-%sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
-%a certain point in which the effects of the more stable gap edge
-%becomes a significant effect.
-
-%The correlation of growth rate and sound
-%speed is non-negligible as in previous `planet-off' case as the disk
-%now can reach values of $h\simeq0.08$ for very slow cooling and the
-%importance of the effect can be imagined to accumulate over the now
-%considerably longer timescale. These effects dictate when then the
-%vortex reaches the critical value in which it shocks the system that
-%subsequently causes disspation and hence the lifetime fo the vortex.
+%This effect can be seen in Mach number around the vortices where the
+%growth of Mach number with respect to time is increasing with cooling rate.
+%Thus the intially weaker formed vortices for higher cooling rate grow
+%substantialy faster than the sound speed and can shock the system in earlier
+%times.
+%While our `planet-off' simulations show 
+%This may be attributed to the higher temperature
+%reached in the latter case 
+%
+%($h\simeq 0.07$) in comparison with the former
+%($h\simeq 0.06$), since \cite{li00} showed that the RWI is favoured by
+%larger disc temperature.

### 2014-11-21 paper/summary.tex
major updates by MKL

@@
 \section{Summary and discussion}\label{summary}
-We have performed customised numerical simulations of  
-three-dimensional self-gravitating discs, in which an embedded 
-satellite or planet has opened a gap. We explicitly verified in 
-3D the main results on gap stability previously obtained from 
-2D calculations \citep{koller03,li05,valborro07,meschiari08}, in
-particular those by \cite{lin11a,lin11b}.
+In this paper, we have carried out numerical simulations of
 
-Planetary gaps are potentially unstable because of the existence of 
-vortensity extrema generated by planet-induced shocks. Disc-satellite
-interaction also occur in other systems such as stars in black hole
-accretion discs \citep{kocsis11,mckernan11,baruteau11b}.  Furthermore, 
-gaps opened by satellites are just one type of structured disc. 
-Other examples 
-include dead zone boundaries mentioned in \S\ref{intro} and
-transition discs \citep{regaly12}.
-
-Thus, although we were motivated by previous works on planetary gap 
-stability, and hence considered disc-planet systems, we expect our
-results to be applicable to discs with radial structure of other    
-origin, provided the vortensity profiles involve stationary points and
-therefore prone to the same instabilities.
-
-\subsection{Confirmed 2D results}
-We demonstrated the development of the vortex instability
-at the outer gap edge opened by a giant planet in 3D 
-discs. We began with a non-self-gravitating disc, in which a few vortices
-develop then quickly merge. The quasi-steady state is a single
-azimuthally extended vortex at the gap edge. This evolution is similar
-to the 2D simulation in \cite{valborro07}.
-
-We also showed the effect of self-gravity on the vortex
-instability observed in 2D \citep{lyra09,lin11a} persists in 3D.
-We observe more vortices as the strength of self-gravity is increased
-by increasing the density scale. These vortices resist merging
-on dynamical timescales. In our disc models with $Q_0=3$, the 
-multi-vortex configuration lasts until the end of the
-simulations.
-
-As expected from 2D linear theory \citep{lin11b}, 
-vortex modes are suppressed in our massive disc models. Instead, a 
-global spiral instability develops which is associated with the local
-vortensity maximum just inside the outer gap edge. These are distinct
-from vortex modes since self-gravity is essential.
-
-Our limited numerical resolution does not permit accurate disc-planet
-torque measurements, but 
-the qualitative effect of the vortex and spiral instabilities,
-previously studied in 2D, have been reproduced in 3D --- oscillatory
-torques of either sign and the tendency for spiral modes to provide on
-average a positive torque. 
-The similarity to 2D results is not surprising since for  
-giant planets $r_h\gtrsim H$, so the razor-thin disc approximation is
-expected to be valid as far as disc-planet interaction is
-concerned. Furthermore, vertical self-gravity increases the midplane
-density while reducing that in the atmosphere (Appendix
-\ref{vertsg_mod}), so that given a fixed temperature profile the 2D
-approximation is even better for self-gravitating discs.
-
- It is worth mentioning here that in shearing sheet
-  simulations,  \cite{mamatsashvili09} found 
-  self-gravity to favour vortices of smaller scale. They initialized
-  a local patch of an unstructured disc with random velocity perturbations.
-  Their gravito-turbulent discs are dominated by small vortices limited by
-  the local Jeans length (which is smaller than the scale-height). Without self-gravity, 
-  they merge to form larger vortices. 
-  These observations are similar to the above results for the vortex mode.  
-  However, the setups are quite different 
-  as we consider radially structured, laminar global discs. Our
-  large-scale vortices develop from a linear instability and have horizontal sizes
-  of a few scale-heights. Thus, confirmation of the above results in 3D is only valid
-  for the edge instabilities considered in this paper.
-
-%torques (expected because giant planets are 'thicker' than disc)
-
-\subsection{Effects of 3D}
-In our simulations the dominant three-dimensional effect is
-vertical self-gravity on the density field. In the
-non-self-gravitating limit, the relative density perturbation
-associated with a vortex is columnar with weak vertical dependence. This is
-consistent with with 3D linear and nonlinear simulations
-\citep{meheut10, meheut12a,  meheut12b, lin12}.
-
-As the strength of self-gravity is increased, vortices become 
-more vertically stratified --- they are condensed towards the
-midplane. For moderately self-gravitating discs, the vortex
-midplane density enhancement can be twice that near the upper
-boundary.  The spiral modes display significant vertical structure
-near the gap edge, while the density waves they launch in the outer
-disc is columnar.  The latter is probably due to the chosen equation of state (see below).
-
-The effect of vertical self-gravity on the vortex mode is seen even in
-our least massive disc model with $Q_0=8$. One can consider an
-initially smooth disc which is justified to be
-non-self-gravitating. However, this approximation may become less good
-with the creation of a vortensity minimum, because it
-can also be a local minimum in the Toomre $Q$
-(Eq. \ref{toomreq_vortensity}).  That is, the Toomre parameter is
-decreased with the development of local radial structure (such as a density bump). 
-The non-self-gravitating approximation worsens further when the vortex
-instability associated with $\mathrm{min}(\eta)$ develops, because the   
-vortices are regions of enhanced density, especially if they merge
-into a single large vortex. Thus, the
-non-self-gravitating approximation is not guaranteed to hold 
-in the perturbed state even if it does in the initial disc.
-
-It is interesting to note that linear calculations of the vortex
-instability in vertically isothermal , non-self-gravitating discs 
-show that the vertical velocity vanishes near the vortex
-centroid  \citep{meheut12a,lin12}, but this is not observed in
-nonlinear calculations \citep[][ and in the present
-  simulations]{meheut12a}.  This contrasts to their anti-cyclonic
-horizontal flow, which can be computed in linear theory and seen in
-hydrodynamic simulations \citep{li00,li01}.  This suggests that vertical 
-motions in this case may be associated with secondary processes
-\citep[e.g][]{goodman87}.
-
-Although we observe somewhat complicated vertical flow for both
-vortex and spiral instabilities, the vertical Mach number is
-at most a few per cent in magnitude. Also, the magnitude of
-vertical motion is at most $\sim 20\%$  of the radial motion on
-average. This suggests that the disturbances at the gap edge is 
-roughly two-dimensional in the present disc models. This would be 
-consistent with early studies which find instabilities associated with
-co-rotation singularities are two-dimensional
-\citep{papaloizou85,goldreich86}.  Recent 3D linear 
-  calculations also find that, even with a vertical temperature gradient, 
-  the vortex mode (without self-gravity) is
-  largely 2D near the vortensity minimum \citep{lin12}.
-
-\subsection{Caveats and outlooks}\label{caveats}
-
-In order to provide 3D examples of previous 2D results, a range of 
-disc models had to be simulated, each for many dynamical 
-timescales with full self-gravity. To maintain reasonable
-computational cost, numerical resolution in the $r\phi$ plane is much
-reduced compared to razor-thin disc simulations ($\sim  4$---6 cells
-per $H$ compared to $\sim 16$ in 2D).  Despite this, our
-plots in the $r\phi$ plane closely resemble those obtained from 2D
-simulations.
-
-On the other hand, the low resolutions adopted here are unlikely to
-capture elliptic instabilities which may destroy 
-vortices in 3D \citep{lesur09b,lesur10}. 
-This might not be a serious
-issue when the condition for the vortex instability is maintained 
-(by a planet in the present context). Also, the vortex 
-grows on dynamical timescales whereas the elliptic instability takes
-much longer \citep{lesur09b}. The initial development of vortices is not expected to be 
-suppressed by the elliptic instability \citep[as found
-  by][]{meheut12a}.
-
-The vortex instability produces smaller vortices with increasing
-self-gravity. According to \cite{lesur09b}, vortices with small
-aspect-ratios ($\lesssim 4$) are strongly unstable in 3D, but note that their  
-model is a local patch of a smooth disc without self-gravity, 
-whereas we considered a gap edge in a global self-gravitating disc.
-
-Inclusion of self-gravity may change the stability properties of
-vortices. In particular, we found that a vortex can flatten somewhat under its own
-weight. \cite{lithwick09} suggested vertical gravity helps to 
-stabilise vortices in 3D. Vertical \emph{self}-gravity 
-can enhance this effect.  \citeauthor{lithwick09} 
-found in local 3D simulations that `tall' vortices are unstable 
-whereas `short' vortices survive as in 2D \citep{godon99}. 
-We may expect the more stratified vortices formed in self-gravitating 
-discs to be more stable than those in non-self-gravitating discs \citep{barranco05}. 
-The elliptic instability is also weakened by stratification \citep{lesur09b}, 
-so vertical self-gravity may also be stabilising in this respect.
-
-%Improved numerical models, including the energy equation, should be 
-%simulated to explore these speculations. 
-% 
-%vortex instability on dynmical timescales. elliptic instability is
-%weaker growth
-
-We have adopted the locally isothermal EOS for simplicity and direct
-comparison with previous 2D results. This EOS limits the 3D 
-structure of density waves compared to thermally 
-stratified discs \citep[e.g][]{lin90,lubow98,ogilvie99}, which can 
-cause refraction of waves out of the midplane.  
-The locally isothermal EOS represents the limit of efficient cooling
-\citep{boss98}. This might apply in optically thin
-regions of a disk\footnote{For example, \cite{cossins10} find optical
-depths $\tau < 0.2$ beyond $\sim100\mathrm{AU}$ in their models of protoplanetary
-discs.}, but is violated if high densities develop, such as 
-self-gravitating clumps \citep{pickett00}. Our edge instabilities only 
-reach moderate over-densities. Nevertheless, enhanced
-vertical stratification of the edge disturbances observed in our
-simulations are likely exaggerated by the EOS. Clearly, it is necessary to
-extend models of edge instabilities in self-gravitating discs to
-include an appropriate energy equation.
-
-Another important issue is vertical boundary conditions. We simulated
-a thin disc and imposed a reflecting upper boundary to prevent mass
-loss from above. This setup  may enhance the two-dimensionality
-of the problem. The vortex instability tends to involve the 
-entire column of fluid, especially in the weak self-gravity regime. It
-is therefore a global instability in $z$. We suspect the spiral mode
-is less affected by vertical boundary conditions because the
-instability tends to concentrate material at the midplane. Future
-simulations will consider varying the vertical domain size and upper
-disc boundary conditions.
-
-Our torque measurements indicate migration will be significantly
-affected by the vortex and spiral instabilities. Preliminary 3D
-simulations with a freely migrating planet have been performed. We
-recover the vortex-planet scattering and the spiral-induced outward
-migration described by \cite{lin10,lin12b}.   
-The disc-planet torque is determined by the density field, and
-the above instabilities have density perturbations that either
-have weak vertical dependence or concentrated at the midplane. Thus, we
-believe that at present, 2D simulations are more advantageous for studies
-focusing on migration, because high resolution is feasible and needed.
+This non-monotonic behaviour of the vortex lifetime was also reported
+by \cite{fu14}, who performed isothermal disc-planet 
+simluations with different values of the disc aspect ratio.   
+% Cooling rates with
+% $\tilde\beta>5.0$ showed lifetimes that were well below the max
+% lifetime. 
+By analyzing the radial region around the vortex within $[2r_{h},10r_{h}]$
+% {\bf define this region}
+we find the 
+$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
+$h\approx0.06$ on average throughout its lifetime. This value of $h$ was also
+found to maximise vortex lifetimes in the simulations of
+\cite{fu14}. Note that $h$ is a fixed parameter in \cite{fu14} while in
+our simulations $h$ is a dynamical variable.

### 2014-11-21 paper/summary.tex
first rough draft - MKL

@@
 \section{Summary and discussion}\label{summary}
-In this paper, we have carried out numerical simulations of
+In this paper, we have carried out numerical simulations of non-isothermal disc-planet interaction. 
+Our simulations were customised to examine the effect of a finite cooling time on the stability of gaps 
+opened by giant planets to the so-called vortex or Rossby wave instability. To do so, we 
+included an energy equation with a cooling term that restores the initial disc temperature on a characteristic timescale $t_c$. 
+We studied the evolution of the gap stability as a function of $t_c$. This is a natural extension to previous
+studies of on gap stability, which employ locally or strictly isothermal equations of state.
 
-This non-monotonic behaviour of the vortex lifetime was also reported
-by \cite{fu14}, who performed isothermal disc-planet 
-simluations with different values of the disc aspect ratio.   
-% Cooling rates with
-% $\tilde\beta>5.0$ showed lifetimes that were well below the max
-% lifetime. 
-By analyzing the radial region around the vortex within $[2r_{h},10r_{h}]$
+We considered two types of numerical experiments. 
+We first used disc-planet interaction to self-consistently set up
+gap profiles, which were then perturbed and evolved without further the influence of the planet potential. This procedure  
+isolates the effect of cooling on gap stability through the set up of the initial gap profile. We find that as the cooling time $t_c$ is increased,
+the gaps became more stable, with lower growth rates of non-axisymmetric modes and the dominant azimuthal wavenumber also decreases. 
+This is consistent with the notion that increasing $t_c$ leads to higher temperatures or $h$, which opposes gap-opening by the planet. This means that
+the gaps opened by the planet in a disc with longer $t_c$ is smoother and therefore  more stable to the RWI.
+
+In the second set of calculations, we included the planet potential throughout the simulations 
+and examined the long-term evolution of the gap-edge vortex that develops from the RWI. The vortices to reach a
+quasi-steady state lasting $O(10^3)$ orbits, during which its amplitude grows until it begins to induce shocks,  
+afterwhich the vortex amplitudes decay. 
+We observe vortex lifetimes to increase with increasing cooling time until
+a cooling timescale on the order of the libration time at the gap edge {\bf (this statement assumes longest lifetime is $\tilde{\beta}=1$, so may change
+depending on new definitions of lifetimes)}, beyond which vortex lifetimes shorten again.
+
+Our results indicate there exists an optimum cooling timescale to maximise the lifetime of gap-edge vortices. We suggest the non-monotonic dependence
+can be attributed to the time required for the vortex to grow to sufficient amplitude to induce shocks in the surrounding fluid, thereby losing energy and also smooth out the gap edge.  
+For short cooling timescales, the planet is able to open a deeper gap which favours the RWI, leading to stronger vortices. However, for long 
+cooling timescales, the vortex grow faster during the quasi-steady state {\bf check if statement true}. 
+We suggest the latter is due to the RWI being favoured with increasing disc temperature, according to previous stability
+calculations. These competing factors imply for both very short and long cooling timescales, the vortex reaches its maximum amplitude 
+, shocks, and begins to decay, sooner than intermediate cooling timescales. 
+%decay timescale depends on cooling time, but there seems to be a monotonic trend: longer cooling times, decay timescale is shorter, so this 
+%doesn't contribute to non-monotonic part
+
+We remark that a non-monotonic dependence of the vortex lifetime was also reported
+by \cite{fu14}, who performed isothermal disc-planet simluations with different values of the 
+disc aspect ratio. In their simulations the optimum aspect ratio is $h=0.06$. In our simulations,
+the aspect ratio is a dynamical variable, but by analysing the region where the vortex is located ($r-r_p\in[2,10]r_h$), 
 % {\bf define this region}
-we find the 
-$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
-$h\approx0.06$ on average throughout its lifetime. This value of $h$ was also
-found to maximise vortex lifetimes in the simulations of
-\cite{fu14}. Note that $h$ is a fixed parameter in \cite{fu14} while in
-our simulations $h$ is a dynamical variable.
+we find for $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
+$h\approx0.06$ on average throughout its lifetime. 
+{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
+Our result is therefore qualitatively consistent with \citep{fu14}. 
+%This value of $h$ was also
+%found to maximise vortex lifetimes in the simulations of
+%\cite{fu14}. Note that $h$ is a fixed parameter in \cite{fu14} while in
+%our simulations $h$ is a dynamical variable.
+
+\subsection{Caveats and outlooks}
+We discuss below several outstanding issues that needs to be addressed in future works:
+
+\emph{Convergence.} Although lower-resolution simulations performed in the early stages of this project gave similar results 
+(most importantly, the non-monotonic dependence of vortex lifetimes on the cooling timescale), we did find the lower-resolution simulations 
+typically yield longer vortex lifetimes than that reported in this paper. This could be due to weaker RWI with low resolution. 
+It will be neccessary to perform even higher resolution  simulations in order to obtain quantitatively converged vortex lifetimes.
+
+\emph{Self-gravity.} We have ignored disc self-gravity in this study. Based on linear calculations, \cite{lovelace13} concluded self-gravity
+to be important for the RWI when $Q<O(1/h)$, or $Q\lesssim 20$ for $h\sim0.05$, as was typially considered in this work. This suggests that 
+self-gravity may affect vortex lifetimes even when the Toomre $Q$ parameter is not small. In particular, given that we observe
+vortices can reach significant over-densities (up to almost an order of magnitude), it will be important to include disc self-gravity 
+in the future. %self-gravitational collapse of vortices
+
+\emph{Three-dimensional (3D) effects.} A vortex in a 3D disc may be subject to secondary instabilities 
+that destroy them \citep{lesur09,railton14}. This may an important factor in determining gap-edge vortex lifetimes
+in realistic discs. For example, if these secondary instabilities sets in before the vortex grows to sufficient amplitude to shock, 
+then the dependence of the vortex lifetime on the cooling timescale will be that through the 3D instability (as opposed to the effect on the RWI itself).  
+This problem needs to be clarified with full 3D disc-planet simulations.
+
+%3d is challenging because high resolution needed in for 3d instabilities
+%However, the role of these instabilities on gap-edge vortices, where disc-planet  
+%interaction can maintain the RWI, has not been clarified.

### 2014-11-21 paper/model.tex
penultimate edit by MKL

@@
 \section{Disc-planet models}\label{model}
-The system is a two-dimensional non-self-gravitating gas disc orbiting
-a central star of mass $M_*$.  
-%self-gravitating gas disc of mass
-%$M_d$ rotating about a central star of mass $M_*$. 
-Cylindrical
-co-ordinates $(r,\phi,z)$ centred on the star is adopted. The frame is   
+The system is a two-dimensional (2D) non-self-gravitating gas disc orbiting
+a central star of mass $M_*$. We adopt cylindrical
+co-ordinates $(r,\phi,z)$ centred on the star. The frame is   
 non-rotating. Computational units are such that 
@@
 $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
-weight. %% \cite{lin12}
+weight.
 
-The disc is governed by the standard fluid equations  
+The disc evolution is governed by the standard fluid equations  
 \begin{align}\label{3d_gov_eq}
@@
 The total potential $\Phi$ includes the stellar potential, planet potential
-(described below) %, and the disc potential $\Phi_d$ which satisfies the
-%Poisson equation \ref{poisson}. 
+(described below) 
 and indirect potentials to account for the non-inertial reference
-frame.%  \citep[see][for explicit expressions]{masset02}
+frame. 
 In the momentum 
@@
 have been chosen to be absent at $t=0$, allowing the disc to be
-initialised close to steady state. The $\mathcal{C}$ function attempts to restore the initial energy density (and
+initialised close to steady state. The $\mathcal{C}$ function attempts
+to restore the initial energy density (and 
 therefore temperature) profile. In practice, this is a cooling term at
@@
 
-%The surface density scale $\Sigma_0$ is set by  
-%\begin{align}
-%\Sigma_0 = \frac{h_i\rout\Omega_k^2(\rout)}{\pi G Q_o},
-%\end{align}
-%where $Q_o$ is the thin-disc Keplerian Toomre parameter at the outer boundary initially.
-
 The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
-pressure, stellar gravity and disc gravity. The initial radial velocity is $v_{r}=3\nu/r$ 
-with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi}\simeq r\Omega_k$ 
-and $\hat{\nu}= 10^{-9}$, so that $|v_{r}/v_{\phi}|\ll1$ and the initial 
-flow is effectively only in the azimuthal direction.  With this choice
-of physical viscosity $\hat{\nu}$ the only heating source is through
-shock-heating (via artificial viscosity) and the $\mathcal{C}$
-function when $e\Sigma<e_i\Sigma_i$.
+pressure, stellar gravity and disc gravity. For a thin disc, 
+$v_{\phi}\simeq r\Omega_k$. The initial radial velocity is
+$v_{r}=3\nu/r$, where $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$, and we
+adopt $\hat{\nu}= 10^{-9}$, so that $|v_{r}/v_{\phi}|\ll1$ and the initial 
+flow is effectively only in the azimuthal direction.  With this value 
+of physical viscosity, the only source of heating is through
+compression, shock-heating (via artificial viscosity) and the
+$\mathcal{C}$ function when $e/\Sigma<e_i/\Sigma_i$.
 
@@
 \end{align}
-where $x_s=2r_h$ is the approximate distance from the planet to its
+where $x_s$ is the distance from the planet to its
 gap edge, and $t_\mathrm{lib}$ is the time interval between successive
@@
 azimuth. That is, we measure the cooling time in units of the time
-taken for a gap edge fluid element to encounter planet-induced
-shocks.
+interval between encounters of a fluid element at the gap edge and the
+planet-induced shock.
 
@@
 Eq. \ref{betatilde}. For a planet mass with $q=10^{-3}$,
-Eq. \ref{betatilde} then gives $\beta \simeq 23.9\tilde{\beta}$.
-
-%giant planet gap is 2.5rh
-%2rh is CR of G.E.I. 
-%\subsection{Simulation procedure}
-%The disc is evolved without a planet for $t\in[0,20]P_0$. The planet potential is smoothly switched on
-%between $t\in[20,30]P_0$.
-%%  In a separate set of calculations the planet
-%% potential is turned off after $t>t_\mathrm{off}$, immediately  
-%% afterwhich the surface density, energy density and velocity fields are
-%% azimuthally averaged and then subject to random surface density  
-%% perturbations of magnitude $10^{-2}$.
+Eq. \ref{betatilde} then gives $\beta \simeq 23.9\tilde{\beta}$. In
+terms of planetary orbital periods, this is
+\begin{align} 
+  t_c(r) = \frac{\beta}{2\pi}\left(\frac{r}{r_p}\right)^{3/2}P_0\simeq
+  3.8 \tilde{\beta}\left(\frac{r}{r_p}\right)^{3/2}P_0. 
+\end{align}
 
@@
 the authors show an extremum in $\teta$ may lead to a dynamical
-instability. In a barotropic disc, the entropy factor is absent
-and the important quantity is the PV.
+instability, the RWI. In a barotropic disc, the entropy factor is
+absent and the important quantity is the PV.
 
 \subsubsection{Fourier modes} 
-The Rossby wave instability is characterized by exponetially
-growing pertibations. Though in this paper we do not formally consider
-linear instability, modal analysis will be useful to
-analyse the growth of different wavenumbers, which is associated with
-the number of vortices initially formed by the RWI.
-
-%characterize 
-%azimuthal vortex numbers and overdensity values.
+The RWI is characterized by exponetially
+growing perturbations. Though in this paper we do not consider a
+formal linear instability calculation, modal analysis will be useful
+to analyse the growth of perturbations with different azimuthal
+wavenumbers, which is associated with the number of vortices initially
+formed by the RWI.
 
@@
 $\langle|\cdot|\rangle_r$ denotes the average of the absolute value
-over a raidal region of interest. By using Eq.~\ref{growth} the growth
+over a radial region of interest. By using Eq.~\ref{growth} the growth
 rates of the unstable modes can be found from successive spatial
@@
 vortices and quantify its intensity.
-
-%{\bf {MKL.} describe vortensity and generalized vortensity, and its
-%  relevance to stability. fourier mode amplitude definition and how
-%  growth rates are obtained. rossby number. density
-%  perturbations. (see lin 2014)
-%}
-
-%% According to previous 2D studies, the key quantitiy in describing the
-%% stability of radially structured non-barotropic discs is the
-%% \emph{generalized vortensity} (GV): 
-%% \begin{align}
-%%   \tilde{\eta}\equiv \left(\frac{\kappa^2}{2\Omega\Sigma}\right)S^{-2/\gamma}, 
-%% \end{align}
-%% where $S \equiv p/\Sigma^\gamma$ is defined to be the entropy. The
-%% quantity in brackets, $\eta\equiv \kappa^2/2\Omega\Sigma$, is the
-%% usual vortensity or potential vorticity (PV).

### 2014-11-21 paper/paper.tex
penultimate edit by MKL

@@
 \begin{abstract}
-%Hydrodynamic simulations of self-gravitating disc-planet interactions are presented. 
   Several observations of transition discs show lopsided
@@
   discs using customised numerical simulations of disc-planet
-systems where the planet open an unstable gap. 
-We include in the energy equation a simple cooling term with
-cooling timescale $t_\mathrm{cool}=\beta\Omega_k^{-1}$, where $\Omega_k$ is
+  systems where the planet opens an unstable gap. 
+  We include in the energy equation a simple cooling function with
+  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
   the Keplerian frequency, and examine the effect of $\beta$ on the
@@
   dominant azimuthal wavenumber $m$ decreases.  
-%gap edges stabilize with increasing $\beta$ as dominate modes and
-%growth rates decrease in value. 
   We find a quasi-steady state consisting of one  
-large-scale overdense vortex circulating the outer gap edge, typically  
+  large-scale, over-dense vortex circulating the outer gap edge, typically  
   lasting $O(10^3)$ orbits. 
-%characterized by
-%sudden disspation on the order of 100's or planet orbits.
   Vortex lifetimes were found to increase with cooling times up to an
-optimal value, after which vortex lifetimes decrease. This
+  optimal value, beyond which vortex lifetimes decrease. This
   non-monotonic dependence is qualitatively consistent with 
-strictly isothermal simulations that differ in disc aspect ratios
-presented in Fu et al..% (2014, ApJL, 788, L41).
-
-%a non-monoticly dependence on $\beta$ and
-%an optimum value for cooling rate was found to maximize the lifetime.
-
-%The simulations are customized to examine the stability of gaps
-%opened by massive planets as a function of $\beta$.
-
+  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
+%  presented in Fu et al. (2014, ApJL, 788, L41). 
 \end{abstract}
@@
 \begin{keywords}
-planetary systems: formation --- planetary systems:
-protoplanetary discs
+  accretion, accretion discs, protoplanetary discs, hydrodynamics, instabilities,
+  planet-disc interactions, methods: numerical 
 \end{keywords}
@@
 influence from the planet. %stability through background state 
-We then examine the lifetime of gap-edge vortices in \S\ref{nonlinear}
+We then perform long-term disc-planet simulations to examine the
+lifetime of gap-edge vortices in \S\ref{nonlinear},  
 as a function of the imposed cooling rate. We conclude and summarize

### 2014-11-21 paper/results.tex
first rough draft - MKL
penultimate edit by MKL

@@
 planet potential and azimuthally average the surface density, energy
-and velocity fields. At this point the planet has carved a partial
-gap and the RWI has not yet occured.
+and velocity fields. (At this point the planet has carved a partial
+gap and the RWI has not yet occured.)
  We then perturb the surface density for $r>r_p$ and continue to 
@@
 restores the initial temperature profile corresponding to $H/r=0.05$,
-rather than the heated gap edge. Linear stability calculations and
-adiabatic cases will be further discussed in
-section~\ref{adiabatic_section}.
+rather than the heated gap edge. However, we will 
+examine a nearly adiabatic simulation in \S\ref{adiabatic_section},
+which is closer to a proper linear problem.
 
@@
 with open boundaries at $r=r_\mathrm{in}$ and
-$r_\mathrm{out}=25r_\mathrm{in}$. We consider   
+$r_\mathrm{out}=25r_\mathrm{in}$. We compare
 cases with $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
-and slowly cooled discs. % {\bf boundary condition for these sims?}
+and slowly cooled discs.
 
 \subsection{Gap structure}
-%Since stability of the gap is of interest 
-We first examine the gap structure formed by planet-disc
+We first compare the gap structures formed by planet-disc
 interaction as a function of the cooling time. The azimuthally-averaged 
-gap profiles are shown in Fig. \ref{intial1D} for varying
+gap profiles are shown in Fig. \ref{intial1D} for different values of 
 $\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling) 
@@
 increase (decrease) the surface density maxima (minima). However, a
-clean gap does not form in this short time period.
+clean gap does not form in this short time period.%
+%\footnote{We find the
+% RWI begins to develop if the disc is evolved beyond $t\sim30P_0$}.
 
-Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
+Increasing $\tilde\beta$ leads to higher disc aspect ratios $h=H/r$,
 i.e. higher temperatures. Heating mostly occur at the gap edges
-due to planet-induced spiral shocks. Decreasing the cooling time
+due to planet-induced spiral shocks. Increasing the cooling timescale 
 implies that this heat is retained in the disc. In the inviscid limit the gap
 opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
-\citep{crida06}. 
-This indicates that for hotter discs (higher
+\citep{crida06}, which indicates that for hotter discs (higher
 $h$), it becomes more difficult for a planet of fixed $q$ to open a
@@
 $\tilde{\beta}$ is increased.
+
 %However, notice the gap structure
@@
 
+However, we remark that the change in the gap structure becomes less 
+significant at long cooling times, as Fig. \ref{intial1D} shows that
+the profiles with $\tilde{\beta}=1$ and $\tilde{\beta}=10$ are 
+similar. This implies that the effect of cooling timescale on the
+RWI through the set up of the gap profile, becomes less important for
+large $\tilde{\beta}$. 
+% or 'suggets that .... becomes unimportant for sufficiently large...'
+
 \begin{figure}
@@
     0.5cm 0cm 1cm]{figures/compare_gvortensity}
-  \caption{Gap profiles at $t=30P_0$ for the intial partial gap opened
-    before instability emerges for fast (solid), moderate
-    (dashed), and slow cooling (dashed-dot). The relative surface density
-    pertibation (top), disc aspectratio (middle) and generalized
-    vortensity pertibation (bottom) are shown. \label{intial1D}}  
+  \caption{Azimuthally averaged gap profiles at $t=30P_0$ for the
+    intial partial gap opened before instability emerges for fast
+    (solid, $\tilde{\beta}=0.1$), moderate
+    (dashed,  $\tilde{\beta}=1$), and slow cooling (dashed-dot,
+    $\tilde{\beta}=10$). The relative surface density 
+    perturbation (top), disc aspect ratio (middle) and generalized
+    vortensity perturbation (bottom) are shown. \label{intial1D}}  
 \end{figure}
@@
 %  m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
- % $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
+ % $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38
+ % & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule 
 %\end{tabularx}
@@
 %  m                    & 1 & 2 & 3 & 4 & 5  \\ 
-%  $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
+%  $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\
+%  \bottomrule 
 %\end{tabularx}
@@
 10$, we observe exponential growth of non-axisymmetric
-structures.  
-An example is shown in Fig. \ref{linearmodes} for 
-$\tilde{\beta}=10$. We charcterize these
+structures. An example is shown in Fig. \ref{linearmodes} for 
+$\tilde{\beta}=10$. We characterise these
 modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
@@
 lists the growth rates measured during 
-linear growth for 5 values of $m$ centred around the maximum.
+linear growth for 5 values of $m$ centred around that with maximum
+growth rate.
 
@@
   \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimenionlized by the initial axisymmetric background mode 
-    $\Sigma_0(t=0)$ for the
-    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
+    density, non-dimenionlised by the initial axisymmetric component 
+    $\Sigma_0(t=0)$, for the
+    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
     to different $m$ values. The $m=3$ component is the fastest growing
@@
 Table \ref{modetable} show that as
-$\tilde{\beta}$ is increases from $ 0.1\rightarrow10$ the dominant
+$\tilde{\beta}$ is increased from $ 0.1\rightarrow10$ the dominant
 azimuthal Fourier mode decreases from $ m=9\rightarrow3$ and the
@@
 longer cooling times. This is expected because larger $\tilde{\beta}$
-result in hotter gap profiles at $t=30P_0$ with less pronounced
-generalized vortensity minima. Stabilization with increased
+results in hotter gap profiles at $t=30P_0$ with less pronounced
+generalized vortensity minima. Stabilisation with increased
 cooling time is therefore due to a smoother basic state for the
@@
 The above `planet-off' simulations are not formally linear
-stability calculations, because the cooling time is shorter
-than the instability growth time, 
-$t_c<\gamma^{-1}$.  
+stability calculations, because the cooling time is comparable or shorter
+than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
 Thus the disk cools back to its intial temperature corresponding to
@@
 
-In order to perform a proper linear stability analysis and capture the
-effect of a heated gap edge during instability growth, we ran a simulation  with
+In order to capture the effect of a heated gap edge, we ran a simulation with 
 $\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
@@
 {\bf typical rossby numbers for these `planet-off' vortices? do
-  stronger vortices (beta=0.1) dissipate faster? if we want to say the
-  vortices dissipate viscously, then need to quote a viscosity
-  (e.g. reynolds stress) and compute a viscous time and compare to the
-  observed decay time. Or attribute to numerical viscosity. It might
-  be simplest to just say the vortices decay without the planet. This
-  section is mainly used to contrast with planet-on simulations, so
-  don't need to go into detail. 
+  stronger vortices (beta=0.1) dissipate faster?
 }
 
-We also extended these `planet-off' simulations well into the non-linear
+We also extended these `planet-off' simulations into the non-linear
 regime. After the linear growth phase of the vortices, vortex merging
@@
 Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
-density amplitude, which represents the post-merger single vortex. For
-completeness we also ran intermediate cases with $\tilde{\beta}=0.5$
-and $5.0$. The vortices simply decay, on a timescale of $O(10^3)$
-orbits. We will see in the next section that this behaviour is very
-different to when the planet potential is kept on. Interestingly,
-though, vortices arising from the stronger instabilities (shorter
-cooling times or lower $\tilde{\beta}$) decays faster.
+density component, which represents the amplitude of the post-merger
+single vortex. For completeness we also ran intermediate cases with
+$\tilde{\beta}=0.5$ and $5.0$. The vortices simply decay on a
+timescale of $O(10^3)$ orbits. We will see in the next section that
+this behaviour is very different to when the planet potential is kept
+on. Interestingly, though, vortices arising from the stronger
+instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
+faster.
+
+%main point: no planet: only get growth during linear instability
+%           with palnet: further growth after linear regime
 
@@
     the gap is set up. The $m=1$ surface density component at the
-    outer gap edge is shown.
+    outer gap edge is shown as a function of the cooling timescale. 
   } \label{planetofflifetimeplot}
@@
 simulations to be computationally feasible. We also use a larger
-disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
+disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimise boundary
 effects on vortex evolution, and apply open boundaries at
@@
 for different $\tilde\beta$.
-%{\bf any particular reason for the larger radial range for average?
-%  did we get bigger vortices?:Yes vortices get quite large in quais steady,
-% changing the planet off averaging range to [2,10] changes the plots negligibly
-% if we want to just use this value range for consistancy }
 
@@
 in magnitude, 
-%relative spin of the vortex
-%continuously grows along with it
  reaching a characteristic minimum value of $Ro\approx-0.35$ to $-0.45$ for
@@
 }
-%{\bf is the the value reached just before decay for all cooling?:yes 
-%values peak in small range [-0.35,-0.45] for all beta}
 
 We find the vortices can become significantly over-dense. 
-The surface density perturbation (relative to $t=0$)  
+The surface density perturbation 
 can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
@@
 was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
-$\tilde\beta=10$ case. {\bf any trend between maximum over-density
-  achieved and cooling?}
+$\tilde\beta=10$ case. %despite beta=10 having the weakest linear
+% instability 
+{\bf any trend between maximum over-density achieved and cooling?} 
 This large increase in surface density is due to  
@@
 gap edge (because of gap-opening) is small: the average surface density
-perturbation near the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
+perturbation at the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
 $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.  
+{\bf may be more appropriate to quote the azimuthally averaged
+  perturbation at the location of the vortex (5-6 Hill radii from the
+  planet?)}
 
-Fig. \ref{lifetimeplot} shows the duration of the quasi-steady state
+Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
 varies with the cooling rate: for 
@@
 will discuss this issue futher in
-\S\ref{lifetime_discuss}. However, the decay timescale decreases with
+\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
 increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
@@
 amplitude to decay significantly after reaching maximum.
-
-%These
-%timescales are typically much shorter than the duration of the
-%quasi-steady state.
 
@@
 }
-%The
-%decay timescale appears to decrease with increasing cooling time. For
-%$\tilde{}$  
-%Interestingly, in most cases we observe a sudden drop in 
-%the $m=1$ amplitude within the simulation time, which occurs over
-%timescale between $20P_0$ to $70P_0$, i.e. much shorter than the duration of the
-%quasi-steady state. 
-%We did not find this dissipation timescale to
-%correlate with $\tilde\beta$. 
-%We designate the time at which these amplitude drops as the lifetime 
-%of the quasi-steady vortex. After the vortices decay the RWI is 
-%not seen to be excited again and no vortices are seen to refrom within the
-%simulations time. 
 {\bf Perhaps a 2D snapshot of what the gap edge looks like after
@@
     $\tilde\beta=0.1$ case during quasi-steady state with a single
-    vortex. The plots for other values of the cooling time
+    vortex at the outer gap edge. 
+    The plots for other values of the cooling time
     $\tilde{\beta}$ are similar.
-    % {\bf is this plot truncated? in the text it says the
-    % domain goes to $r=45$?}
-    % The large scale non-axisyemetric $m=1$ overdensity can be
-    % seen. 
     \label{Vortex2D}} 
@@
 During quasi-steady state the average value of the surface density gradient
-along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf quote something
+along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf this quantity
+  has dimensions of 1/length, preferable to quote something
   dimensionless, or say `in code units'}  
@@
   } 
-  \caption{The vortex in the case $\tilde{\beta}=1$
+  \caption{The vortex in the case with $\tilde{\beta}=1$
     during quasi-steady state (left), at maximum amplitude of the
@@
     (right). The surface density perturbation
-    (top) and the associated surface density gradient (bottom) are shown.
+    (top) and the associated surface density gradient (bottom) are
+    shown. This plot can considered in conjunction with
+    Fig. \ref{lifetimeplot}. 
     \label{shockplot}}
@@
 Putting the above observations together, we suggest the decay
-in the vortex amplitude is due to shock formation \emph{by the
-  vortex}. When the vortex reaches large amplitude, it begins
+in the vortex amplitude is due to shock formation by the
+vortex. When the vortex reaches large amplitude, it begins
 to induce shocks in the surrounding fluid, as supported by the
@@
   \includegraphics[width=\linewidth]{figures/mach}
-  \caption{Running-time average {\bf ?} of the Mach number relative to
+  \caption{Running-time average {\bf (?)} of the Mach number relative to
     the vortex, averaged over a region within $2H$ of the  
@@
 \subsection{Vortex decay and gap structure}  
-We find vortex decay also modifies the gap
+We find vortex decay modifies the gap
 structure. Fig. \ref{gap_smoothed} shows the gap profile before and
@@
 In the previous section, we observed that vortices began to decay when
-it starts to induce shocks. We therefore suggest that the vortex
-timelife is determined by the time needed for the vortex to grow to
-sufficient amplitude to induce shocks in the surrounding
-fluid. Based on this hypothesis, we suggest below competing
-factors that may result in a non-monotonic dependence of vortex
-lifetimes on the cooling timescale.
+it starts to induce shocks. We therefore suggest the vortex lifetime
+is determined by the time needed for the vortex to grow to
+sufficient amplitude to induce shocks in the surrounding fluid, as
+well as the decay timescale itself. The latter was found to shorten
+with increasing $\tilde{\beta}$. However, the former (i.e. the
+duration of the quasi-steady state), depends non-monotonically on the
+cooling time.  
+% \footnote{The 
+%  decay timescale itself is also a factor, but we found this timescale
+%  to be much shorter than the duration of the quasi-steady state
+%  (Fig. \ref{lifetimeplot}), except at very short cooling times.}. 
+We discuss below competing factors that may
+result in a non-monotonic dependence of vortex lifetimes on the cooling timescale.
 
@@
 interaction. However, gap-opening becomes more difficult in a hotter 
-disc. Furthermore, the increased sound-speed suggest the vortex should    
+disc. %this effect probably becomes less important for very long
+% cooling times -> our gap profiles for beta=1, 10 and 100 appear similar
+Furthermore, the increased sound-speed suggest the vortex should    
 reach larger amplitudes in order to induce shocks.
 
-These considerations suggests that, with increased cooling times, 
+These considerations suggest, with increased cooling times, 
 it takes longer for the vortex grow to sufficient amplitude to induce 
-shocks and dissipate, implying longer lifetimes with increasing 
+shocks and dissipate, which imply longer lifetimes with increasing 
 $\tilde{\beta}$.
@@
 \subsubsection{Factors contributing to shorter vortex lifetimes} 
-%However, Fig. \ref{lifetimeplot} shows that very long cooling
-%times (e.g. $\tilde{\beta}=10$) actually results in a \emph{shorter} vortex
-%lifetime. 
+% However, Fig. \ref{lifetimeplot} shows that very long cooling
+% times (e.g. $\tilde{\beta}=10$) actually results in a \emph{shorter} vortex
+% lifetime. 
 Notice in Fig. \ref{lifetimeplot}, for $\tilde{\beta}=10$ the vortex 
 growth during the quasi-steady state is in fact faster than for
-$\tilde{\beta}=5$. For example, at $t\sim 500P_0$ the vortex with   
+$\tilde{\beta}=5$, despite the former case started with a weaker
+vortex. For example, at $t\sim 500P_0$ the vortex with   
 $\tilde{\beta}=10$ has the largest amplitude. This is also reflected
@@
 
-The above observation is consistent with the RWI being favoured by higher
+This observation is consistent with the RWI being favoured by higher
 disc temperatures \citep{li00,lin12c}, which corresponds to longer
 cooling times in our case. While our `planet-off' simulations indicate
-this is unimportant for the linear instability, it may contribute 
+this is unimportant for the linear instability, it may have contributed 
 significantly to the vortex growth during quasi-steady state, when one
 considers very long cooling times (e.g. $\tilde{\beta}=10$). This
-effect alone would lead to a shorter vortex lifetime by allowing it to
-induce shocks sooner.
+effect shortens the vortex lifetime by allowing it to grow
+faster and induce shocks sooner.
 

### 2014-11-21 paper/summary.tex
penultimate edit by MKL

@@
 \section{Summary and discussion}\label{summary}
-In this paper, we have carried out numerical simulations of non-isothermal disc-planet interaction. 
-Our simulations were customised to examine the effect of a finite cooling time on the stability of gaps 
-opened by giant planets to the so-called vortex or Rossby wave instability. To do so, we 
-included an energy equation with a cooling term that restores the initial disc temperature on a characteristic timescale $t_c$. 
-We studied the evolution of the gap stability as a function of $t_c$. This is a natural extension to previous
-studies of on gap stability, which employ locally or strictly isothermal equations of state.
+In this paper, we have carried out numerical simulations of
+non-isothermal disc-planet interaction.  
+Our simulations were customised to examine the effect of a finite 
+cooling time on the stability of gaps  
+opened by giant planets to the so-called vortex or Rossby wave
+instability. To do so, we  
+included an energy equation with a cooling term that restores the 
+disc temperature to its initial profile on a characteristic timescale
+$t_c$. We studied the evolution of the gap stability as a function of 
+$t_c$. This is a natural extension to previous studies of on gap
+stability, which employ locally or strictly isothermal equations of
+state.
 
-We considered two types of numerical experiments. 
-We first used disc-planet interaction to self-consistently set up
-gap profiles, which were then perturbed and evolved without further the influence of the planet potential. This procedure  
-isolates the effect of cooling on gap stability through the set up of the initial gap profile. We find that as the cooling time $t_c$ is increased,
-the gaps became more stable, with lower growth rates of non-axisymmetric modes and the dominant azimuthal wavenumber also decreases. 
-This is consistent with the notion that increasing $t_c$ leads to higher temperatures or $h$, which opposes gap-opening by the planet. This means that
-the gaps opened by the planet in a disc with longer $t_c$ is smoother and therefore  more stable to the RWI.
+We considered two types of numerical experiments. We first used
+disc-planet interaction to self-consistently set up gap profiles,
+which were then perturbed and evolved without further the influence of
+the planet potential. This procedure isolates the effect of cooling on
+gap stability through the set up of the initial gap profile. We find
+that as the cooling time $t_c$ is increased, the gaps became more
+stable, with lower growth rates of non-axisymmetric modes and the
+dominant azimuthal wavenumber also decreases. This is consistent with
+the notion that increasing $t_c$ leads to higher temperatures or $h$,
+which opposes gap-opening by the planet. This means that the gaps
+opened by the planet in a disc with longer $t_c$ is smoother and
+therefore  more stable to the RWI.
 
-In the second set of calculations, we included the planet potential throughout the simulations 
-and examined the long-term evolution of the gap-edge vortex that develops from the RWI. The vortices to reach a
-quasi-steady state lasting $O(10^3)$ orbits, during which its amplitude grows until it begins to induce shocks,  
-afterwhich the vortex amplitudes decay. 
-We observe vortex lifetimes to increase with increasing cooling time until
-a cooling timescale on the order of the libration time at the gap edge {\bf (this statement assumes longest lifetime is $\tilde{\beta}=1$, so may change
-depending on new definitions of lifetimes)}, beyond which vortex lifetimes shorten again.
+In the second set of calculations, we included the planet potential
+throughout the simulations  and examined the long-term evolution of
+the gap-edge vortex that develops from the RWI. The vortex reaches 
+a quasi-steady state lasting $O(10^3)$ orbits. Unlike the `planet-off'
+simulations, in which vortices decay after linear growth and merging,
+we find that with the planet potential kept on, the vortex amplitude
+grows during this quasi-steady state, until it begins to induce
+shocks, afterwhich the vortex amplitude begins to decay.
 
-Our results indicate there exists an optimum cooling timescale to maximise the lifetime of gap-edge vortices. We suggest the non-monotonic dependence
-can be attributed to the time required for the vortex to grow to sufficient amplitude to induce shocks in the surrounding fluid, thereby losing energy and also smooth out the gap edge.  
-For short cooling timescales, the planet is able to open a deeper gap which favours the RWI, leading to stronger vortices. However, for long 
-cooling timescales, the vortex grow faster during the quasi-steady state {\bf check if statement true}. 
-We suggest the latter is due to the RWI being favoured with increasing disc temperature, according to previous stability
-calculations. These competing factors imply for both very short and long cooling timescales, the vortex reaches its maximum amplitude 
-, shocks, and begins to decay, sooner than intermediate cooling timescales. 
-%decay timescale depends on cooling time, but there seems to be a monotonic trend: longer cooling times, decay timescale is shorter, so this 
-%doesn't contribute to non-monotonic part
+We observe the duration of the quasi-steady state increases with
+increasing cooling time until some critical value, beyond which this
+quasi-steady state shortens again. The timescale for the vortex to
+decay, after reaching maximum amplitude, decreases monotonically with
+increasing cooling time.
 
-We remark that a non-monotonic dependence of the vortex lifetime was also reported
-by \cite{fu14}, who performed isothermal disc-planet simluations with different values of the 
-disc aspect ratio. In their simulations the optimum aspect ratio is $h=0.06$. In our simulations,
-the aspect ratio is a dynamical variable, but by analysing the region where the vortex is located ($r-r_p\in[2,10]r_h$), 
-% {\bf define this region}
+%technically we should probably say there is an optimum cooling time
+%to maximise the quasi-steady state
+
+Our results indicate there exists an optimum cooling timescale to
+maximise the lifetime of gap-edge vortices. We suggest the
+non-monotonic dependence of the quasi-steady state can be attributed
+to the time required for the vortex to grow to sufficient amplitude to
+induce shocks in the surrounding fluid, thereby losing energy and also
+smooth out the gap edge.
+
+For short cooling timescales, the planet is able to open a
+deeper gap which favours the RWI, leading to stronger
+vortices. However, for long  cooling timescales, the vortex grow
+faster during the quasi-steady state {\bf check if statement true}. We
+suggest the latter is due to the RWI being favoured with increasing
+disc temperature, according to previous stability calculations. These
+competing factors imply for both very short and long cooling
+timescales, the vortex reaches its maximum amplitude, shock, and
+begins to decay, sooner than intermediate cooling timescales.
+
+%decay timescale depends on cooling time, but there seems to be a
+%monotonic trend: longer cooling times, decay timescale is shorter, so
+%this  
+%doesn't contribute to non-monotonic trend
+
+We remark that a non-monotonic dependence of the vortex lifetime was
+also reported by \cite{fu14}, who performed isothermal disc-planet
+simluations with different values of the   
+disc aspect ratio. In their simulations the optimum aspect ratio is
+$h=0.06$. In our simulations, the disc aspect ratio is a dynamical
+variable, but by analysing the region where the vortex is located
+($r-r_p\in[2,10]r_h$), % {\bf define this region}
 we find for $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
@@
 {\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
-Our result is therefore qualitatively consistent with \citep{fu14}. 
-%This value of $h$ was also
-%found to maximise vortex lifetimes in the simulations of
-%\cite{fu14}. Note that $h$ is a fixed parameter in \cite{fu14} while in
-%our simulations $h$ is a dynamical variable.
+Our result is therefore consistent with \citep{fu14}.
+
+% we don't know how fu defined vortex lifetime
 
 \subsection{Caveats and outlooks}
-We discuss below several outstanding issues that needs to be addressed in future works:
+We discuss below several outstanding issues that needs to be addressed
+in future works:
 
-\emph{Convergence.} Although lower-resolution simulations performed in the early stages of this project gave similar results 
-(most importantly, the non-monotonic dependence of vortex lifetimes on the cooling timescale), we did find the lower-resolution simulations 
-typically yield longer vortex lifetimes than that reported in this paper. This could be due to weaker RWI with low resolution. 
-It will be neccessary to perform even higher resolution  simulations in order to obtain quantitatively converged vortex lifetimes.
+\emph{Convergence.} Although lower-resolution simulations performed in
+the early stages of this project gave similar results  
+(most importantly, the non-monotonic dependence of vortex lifetimes on
+the cooling timescale), we did find the lower-resolution simulations  
+typically yield longer vortex lifetimes than that reported in this
+paper. This could be due to weaker RWI with low resolution.  
+It will be neccessary to perform even higher resolution  simulations
+in order to obtain quantitatively co nverged vortex lifetimes.
 
-\emph{Self-gravity.} We have ignored disc self-gravity in this study. Based on linear calculations, \cite{lovelace13} concluded self-gravity
-to be important for the RWI when $Q<O(1/h)$, or $Q\lesssim 20$ for $h\sim0.05$, as was typially considered in this work. This suggests that 
-self-gravity may affect vortex lifetimes even when the Toomre $Q$ parameter is not small. In particular, given that we observe
-vortices can reach significant over-densities (up to almost an order of magnitude), it will be important to include disc self-gravity 
-in the future. %self-gravitational collapse of vortices
+\emph{Self-gravity.} We have ignored disc self-gravity in this
+study. Based on linear calculations, \cite{lovelace13} concluded
+self-gravity to be important for the RWI when $Q<O(1/h)$, or
+$Q\lesssim 20$ for $h\sim0.05$, as was typially considered in this
+work. This suggests that self-gravity may affect vortex lifetimes even
+when the Toomre $Q$ parameter is not small. In particular, given that
+we observe vortices can reach significant over-densities (up to almost
+an order of magnitude), it will be important to include disc
+self-gravity in the future. %self-gravitational collapse of vortices
 
-\emph{Three-dimensional (3D) effects.} A vortex in a 3D disc may be subject to secondary instabilities 
-that destroy them \citep{lesur09,railton14}. This may an important factor in determining gap-edge vortex lifetimes
-in realistic discs. For example, if these secondary instabilities sets in before the vortex grows to sufficient amplitude to shock, 
-then the dependence of the vortex lifetime on the cooling timescale will be that through the 3D instability (as opposed to the effect on the RWI itself).  
-This problem needs to be clarified with full 3D disc-planet simulations.
+\emph{Three-dimensional (3D) effects.} A vortex in a 3D disc may be
+subject to secondary instabilities that destroy them
+\citep{lesur09,railton14}. This may be an important factor in
+determining gap-edge vortex lifetimes in realistic discs. For example,
+if these secondary instabilities sets in before the vortex grows to
+sufficient amplitude to shock, then the dependence of the vortex
+lifetime on the cooling timescale will be its effect through the 3D
+instability (as opposed to the effect on the RWI itself, which is a 2D
+instability). This problem needs to be clarified with full 3D
+disc-planet simulations.
 

### 2014-12-04 paper/results.tex
Update

@@
 (see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
-$\sim 0.002$ (code units) for all $\tilde\beta$. The
-Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. 
-{\bf Brunt frequency the same for all cooling?}
-The Solberg-Hoiland criteria is similary satisfied for the entire disc throughout the 
-simulations. {\bf i assume this means the azimuthally averaged
-  values. do you ever get $N^2<0$ anywhere in the 2D disc?}  
+$\sim 2\times10^{-4}$ (code units) for all $\tilde\beta$. The
+Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$, decreasing marginally with longer cooling rate. 
+The Solberg-Hoiland criteria is similary satisfied for the entire 2D disc throughout the 
+simulations.
 Thus for all values of $\tilde\beta$ the planet-induced gaps are
 stable to axisymmertic instabilities.
+
+{\bf Brunt frequency the same for all cooling?:small decrease with beta,
+ not more than order of magnitude}
+
+ {\bf i assume this means the azimuthally averaged
+  values. do you ever get $N^2<0$ anywhere in the 2D disc?:
+ This was not for azimuthal average, below 0 inside gap but kappa+N always above 0 everywhere}
 
@@
 {\bf typical rossby numbers for these `planet-off' vortices? do
-  stronger vortices (beta=0.1) dissipate faster?
-}
+  stronger vortices (beta=0.1) dissipate faster?:very complicated
+maybe not worth discussing since just comparison to planet-off}
 
@@
 single vortex. For completeness we also ran intermediate cases with
-$\tilde{\beta}=0.5$ and $5.0$. The vortices simply decay on a
-timescale of $O(10^3)$ orbits. We will see in the next section that
+$\tilde{\beta}=0.5$ and $5.0$. The amplitude of the intially formed vortex was 
+found to decrease with increased cooling rate. The vortices simply decay on a
+timescale of $O(10^3)$ orbits with quicker decay for stronger vortices
+ (quicker cooling rates). We will see in the next section that
 this behaviour is very different to when the planet potential is kept
@@
 averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
-for different $\tilde\beta$.
+for different $\tilde\beta$. The intial vortex amplitude is found to
+be weaker for longer cooling rates (slower linear groth rates).
 
@@
 in magnitude, 
- reaching a characteristic minimum value of $Ro\approx-0.35$ to $-0.45$ for
-all cooling rates. 
-{\bf Maybe useful to have a Rossby number v.s. time plot to complement
-  mode amplitude plot. 
-}
+ reaching a characteristic minimum value of $Ro\approx-0.35$ for
+ $\tilde\beta=.01$ to a max value $-0.45$ for $\tilde\beta=10.0$. 
+%{\bf Maybe useful to have a Rossby number v.s. time plot to complement
+%  mode amplitude plot. }
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+  0cm 0cm 1cm]{figures/rossby}
+  \caption{Rossby numbers at the centers of the vortices for various
+ cooling rates. Boxcar averaging was used to remove periodic planet shock contribution.\label{rossbyplot}}   
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+  0cm 0cm 1cm]{figures/vortex_density}
+  \caption{Overdensity at the center of the vortex for various cooling rates, boxcar averaged with respect to time.\label{overdensity}}   
+\end{figure}
 
@@
 can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
-for all cases of $\tilde\beta$ in quasi-steady state, and
-was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
-$\tilde\beta=10$ case. %despite beta=10 having the weakest linear
-% instability 
-{\bf any trend between maximum over-density achieved and cooling?} 
+for all cases of $\tilde\beta$ in quasi-steady state. Generally the
+max overdensity reached increases with longer cooling rate despite the
+intial vortices formed to be weaker. The max overdensity found was
+ $\Delta\Sigma/\Sigma_0 \sim 11$ for $\tilde\beta=5$.
+% despite beta=10 having the weakest linear instability 
 This large increase in surface density is due to  
@@
 gap edge (because of gap-opening) is small: the average surface density
-perturbation at the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
+perturbation at vortex radial region, $r\sim r_{p}+5r_h$, is 
 $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.
-{\bf may be more appropriate to quote the azimuthally averaged
-  perturbation at the location of the vortex (5-6 Hill radii from the
-  planet?)}
 
@@
 
+{\bf may be more appropriate to quote the azimuthally averaged
+  perturbation at the location of the vortex (5-6 Hill radii from the
+  planet?): yes, this was my original idea, mistyped as 2.5
+}
+
+{\bf any trend between maximum over-density achieved and cooling?: generally increasing with beta, max at beta=5}
+
 {\bf question: vortex amplitude plot suggest just after linear growth,
   vortex amplitudes are smaller with increasing beta, but later on in
-  the quasi-steady state, this trend reverses?  
+  the quasi-steady state, this trend reverses? yes. 
 }
+
 {\bf Perhaps a 2D snapshot of what the gap edge looks like after
-  vortex death (like the one in the poster). defer definition of
-  vortex lifetime to later, when we discuss it. 
+  vortex death (like the one in the poster).: Not sure if this is figure 
+you were thinking (Fig 8)
+% defer definition of  vortex lifetime to later, when we discuss it. 
 }
+
+\begin{figure}
+  \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma100}
+  \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma130}
+  \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma200}
+
+  \caption{2D gap profile for $\tilde\beta=1.0$ during quasisteady state (left), during dissipation (middle), and after dissipation (right).
+    \label{2dgaps}} 
+\end{figure}
 
@@
 
-In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
-more compact as it grows before decaying. 
-{\bf rough estimate of
-  vortex aspect-ratio in the three snapshots? 
-  (length/width, by inspection is ok)}  
+In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
+, but becomes more compact approching a ratio of 2 during its decay ($t=1310$).   
 Notice in Fig. \ref{shockplot} the appearence of wakes extending from
@@
 the vortex begins to decay coincides with the emergence of these
-wakes. {\bf true statement?}
+wakes. {\bf true statement? yes, distortion of vortex shape form wakes happens at same time}
 
@@
 During quasi-steady state the average value of the surface density gradient
-along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf this quantity
-  has dimensions of 1/length, preferable to quote something
-  dimensionless, or say `in code units'}  
+along the wakes is $|\nabla\Sigma \cdot w_s/\Sigma| \sim .4 $, where $w_s$ is the 
+average width of the shock $\simeq0.1$ (in code units).
+
 Just before the $m=1$ amplitude begins to decrease, we observe this quantity
-sharply increases to $ \sim 6 $, and remains around this value
+sharply increases to $ \sim .6 $, and remains around this value
 until the vortex dies out, at which point the associated Rossby number
-quickly approaches zero.
+ begins decreasing to zero
+%{\bf :check}.
+
+{\bf rough estimate of
+  vortex aspect-ratio in the three snapshots? 
+  (length/width, by inspection is ok):trend of decreasing aspect ratio is common for all beta}  
+{\bf this quantity has dimensions of 1/length, preferable to quote something
+  dimensionless, or say `in code units': used shock width since relavant
+ distance, since small may appear to weaken value/argument}
 
@@
   \caption{The vortex in the case with $\tilde{\beta}=1$
-    during quasi-steady state (left), at maximum amplitude of the
-    $m=1$ surface density component, 
+    during quasi-steady state (left),start of dissipation, 
     (middle), and just after the decay in the $m=1$ amplitude
@@
 % with larger growth rate for higher $\tilde\beta$.
-a time coincident with the start of vortex decay. {\bf check if
-  statement is true}
+a time $\sim 100P_0$ after the start of vortex decay.
+% {\bf check if  statement is true}
 
@@
   \includegraphics[width=\linewidth]{figures/mach}
-  \caption{Running-time average {\bf (?)} of the Mach number relative to
+  \caption{Mach number relative to
     the vortex, averaged over a region within $2H$ of the  
-    vortex centre. This plot can be compared to the evolution of the
+    vortex centre with respect to time. This plot can be compared to the evolution of the
     vortex amplitude shown in Fig. \ref{lifetimeplot}.
@@
 smoothed out.
+
+We characterize the smoothness of the outer gap edge with a dimensionless
+gap edge gradient parameter
+
+\begin{align}
+ \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r) \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r) \rangle_{\phi}} \ra_{\Delta r} 
+\end{align}
+where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging spanning
+from center of gap to the maxima radial location.
+Hence a larger $\delta \Sigma$ characterizes a sharper gap edge and larger
+peaked outer gap density maxima. A plot of the gradient parameter over time
+for the $\tilde\beta=1.0$ case is shown in Fig. \ref{smoothnessplot}.
+During dissipation the largely peaked gap edge is found to be
+drasticly smoothed out changinging from a value of $\delta\Sigma=1.2$ during
+quasi-steady state to $0.4$ after dissipation.
 %and we find that the gap edge is smoothed during this
@@
 \begin{figure}
+  \includegraphics[width=\linewidth]{figures/gap_smoothness}
+  \caption{Undimesionalized gap edge gradient for
+    $\tilde\beta=1.0$ boxcar averaged in time. A large $\delta\Sigma$
+    correponds to a more sharply peaked outer gap edge.
+    \label{smoothnessplot}}  
+\end{figure}
+
+\begin{figure}
   \includegraphics[width=\linewidth]{figures/gapchange}
@@
     before (solid) and after vortex decay
-    (dashed). {\bf remember to correct the time quoted in the
-      legend} \label{gap_smoothed}}  
+    (dashed).
+    \label{gap_smoothed}} 
 \end{figure}
@@
 We now examine vortex lifetimes as function of the imposed cooling
-times. We characterise the vortex lifetime as {\bf insert
-  definition(s).} and plot this in Fig. \ref{betaplot}. 
-Increasing the cooling timescale increases the vortex
-lifetime up to $\tilde{\beta}=1$ {\bf (judging from
-  Fig. \ref{lifetimeplot} without a quantitative definition)}, after
-which vortex lifetimes decrease again.
+times. We characterise the vortex lifetime, $t_{\mathrm{life}}$, as the time for the
+ overdensity of the vortex
+% to reach $10\%$ of its max overdensity value
+ to reach $\Delta \Sigma/\Sigma_0~1$ (on order of the intial overdensity caused
+ by the gap formation)
+and plot this with respect to cooling rate in Fig. \ref{betaplot}.
+Also plotted are the times for the times for the vortex to begin dissipation
+($t_{\mathrm{diss}}$) and the time in which max Mach number is reached ($t_{\mathrm{Mach}}$).
+For fast cooling rates a max lifetime is found for $\tilde\beta=0.1$ of 
+$t_{\mathrm{life}} \approx 1300P_0$ with lifetime locally decreasing with higher cooling rate.
+After sufficient increase of cooling the lifetime of the vortices increase
+to a second longer max lifetime for values $\tilde\beta=5.0$ of $t_{\mathrm{life}}\approx 1525P_0$
+%Increasing the cooling timescale increases the vortex
+%lifetime up to $\tilde{\beta}=1$
+% {\bf (judging from  Fig. \ref{lifetimeplot} without a quantitative definition)},
+ after which vortex lifetimes decrease again. 
+This is in comparison with the timescale of intial vortex dissipation which has 
+non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.
 
@@
   \includegraphics[width=\linewidth]{figures/betaplot}
-  \caption{Lifetime of vortices as a function of the cooling paramter 
-    $\tilde\beta$. {\bf maybe useful to over-plot other timescales,
-      e.g. time taken to reach max mach, time for rossby number to
-      reach zero.}\label{betaplot}} 
+  \caption{Various timescales of vortices as a function of the cooling
+    parameter $\tilde\beta$: vortex begins dissipation (black),
+    $\Delta\Sigma/\Sigma < 1$ (red),
+    and max Mach number reached (blue). \label{betaplot}} 
+% {\bf maybe useful to over-plot other timescales,
+%      e.g. time taken to reach max mach, time for rossby number to
+%      reach zero.}
 \end{figure}
@@
 $\tilde\beta=0.1$ and $\Delta\Sigma/\Sigma_0=1.48$ for
-$\tilde\beta=10$.  {\bf rossby numbers?}
+$\tilde\beta=10$, both with $Ro\simeq-0.15$. 
+% {\bf rossby numbers?}
 
@@
 {\bf hotter disc $\to$ gap edges less sharp, still true during quasi
-  steady state?}
+  steady state?: gap edges almost identical in quais-steady but generalized vortensity smoothed out}
 

### 2014-12-04 paper/summary.tex
Update

@@
 vortices. However, for long  cooling timescales, the vortex grow
-faster during the quasi-steady state {\bf check if statement true}. We
+faster during the quasi-steady state {\bf check if statement true:yes}. We
 suggest the latter is due to the RWI being favoured with increasing
@@
 ($r-r_p\in[2,10]r_h$), % {\bf define this region}
-we find for $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
-$h\approx0.06$ on average throughout its lifetime. 
-{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
+we find for the longest  $\tilde\beta=5.0$ case, which has the longest lifetime,
+$h\approx0.06$ on average throughout its lifetimes. 
+%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
 Our result is therefore consistent with \citep{fu14}.

### 2014-12-14 paper/model.tex
final edit by MKL

@@
 
-\subsection{Parametrized cooling}
+\subsection{Cooling prescription}
 In this work we only vary one control parameter: the cooling
@@
 
-\subsection{Diagonistic measures}
+\subsection{Diagnostic measures}
 
-\subsubsection{Generalized potential vorticity}
+\subsubsection{Generalised potential vorticity}
 
-The generalized potential vorticity is defined as
+The generalised potential vorticity is defined as
 \begin{align}
@@
 where $\kappa^2 = r^{-3}\partial_r(r^4\Omega^2)$ is the square of the
-epicycle frequency, $\Omega=v_\phi/r$ is the angular speed, and
+epicyclic frequency, $\Omega=v_\phi/r$ is the angular speed, and
 $S\equiv p/\Sigma^\gamma$ is the entropy. The first factor is the
@@
 
-The generalized PV appears in the description of the linear stability
+The generalised PV appears in the description of the linear stability
 of radially-structured adiabatic discs \citep{lovelace99,li00}, where
@@
 \subsubsection{Fourier modes} 
-The RWI is characterized by exponetially
+The RWI is characterised by exponentially
 growing perturbations. Though in this paper we do not consider a
@@
 \end{align} 
-where $m$ is the aziumuthal wave number. We define the growth rate
+where $m$ is the azimuthal wave number. We define the growth rate
 $\gamma$ of the $m^\mathrm{th}$ component of the surface density

### 2014-12-14 paper/paper.tex
Update
final edit by MKL

@@
 
+\def\la{\left\langle\rule{0pt}{3em}}
+\def\ra{\right\rangle}
+
 \usepackage{array,booktabs,tabularx}
@@
   lasting $O(10^3)$ orbits. 
-  Vortex lifetimes were found to increase with cooling times up to an
+  Vortex lifetimes were found to generally increase with cooling times up to an
   optimal value, beyond which vortex lifetimes decrease. This
@@
 %}
-The interaction between planets and protoplantary discs plays an
+The interaction between planets and protoplanetary discs plays an
 important role in the theory of planet formation and disc 
@@
 equations governing the disc-planet system and initial conditions. Our
-numerical approach, including diagnositic measures, are given in
+numerical approach, including diagnostic measures, are given in
 \S\ref{method}. We present results from two sets of numerical
@@
 lifetime of gap-edge vortices in \S\ref{nonlinear},  
-as a function of the imposed cooling rate. We conclude and summarize
+as a function of the imposed cooling rate. We conclude and summarise
 in \S\ref{summary} with a discussion of important caveats.
@@
 
-%\section*{Acknowledgments}
+\section*{Acknowledgments}
+This project was initiated at the Canadian Institute for Theoretical
+Astrophysics (CITA) 2014 summer student programme.  
+Computations were performed on the GPC supercomputer at the
+SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for
+Innovation under the auspices of Compute Canada; the Government of
+Ontario; Ontario Research Fund - Research Excellence; and the
+University of Toronto.
 

### 2014-12-14 paper/results.tex
final edit by MKL

--- 2014-11-21 paper/results.tex
@@
 In this section, the planet is introduced at $t=20P_0$ and 
-its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
+its potential is switched on over $10P_0$. At $t=30P_0$ we switch off the
 planet potential and azimuthally average the surface density, energy
 and velocity fields. (At this point the planet has carved a partial
-gap and the RWI has not yet occured.)
- We then perturb the surface density for $r>r_p$ and continue to 
+gap and the RWI has not yet occurred.) Effectively, we initialise the
+disc with a gap profile. 
+We then perturb the surface density in the outer disc ($r>r_p$) and continue to 
 evolve the disc. We impose  sinusoidal perturbations with 
@@
 Note that these `planet-off' simulations are not linear stability
-calculations since the cooling term in our energy equation
-restores the initial temperature profile corresponding to $H/r=0.05$,
-rather than the heated gap edge. However, we will 
+calculations because the cooling term in our energy equation
+restores the initial temperature profile corresponding to constant
+$H/r=0.05$, rather than the heated gap edge. However, we will 
 examine a nearly adiabatic simulation in \S\ref{adiabatic_section},
@@
 with open boundaries at $r=r_\mathrm{in}$ and
-$r_\mathrm{out}=25r_\mathrm{in}$. We compare
-cases with $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
+$r_\mathrm{out}=25r_\mathrm{in}$. We compare cases with
+$\tilde{\beta}=0.1,1,10$ corresponding to fast, moderately, 
 and slowly cooled discs.
@@
 Increasing $\tilde\beta$ leads to higher disc aspect ratios $h=H/r$,
-i.e. higher temperatures. Heating mostly occur at the gap edges
+i.e. higher temperatures. Heating mostly occurs at the gap edges
 due to planet-induced spiral shocks. Increasing the cooling timescale 
@@
   \caption{Azimuthally averaged gap profiles at $t=30P_0$ for the
-    intial partial gap opened before instability emerges for fast
+    initial partial gap opened before instability emerges for fast
     (solid, $\tilde{\beta}=0.1$), moderate
@@
     $\tilde{\beta}=10$). The relative surface density 
-    perturbation (top), disc aspect ratio (middle) and generalized
+    perturbation (top), disc aspect ratio (middle) and generalised
     vortensity perturbation (bottom) are shown. \label{intial1D}}  
@@
 \subsection{Axisymmetric stability}
-The intial planet-disc interaction form bumps and grooves in the gap profiles
+The initial planet-disc interaction form bumps and grooves in the gap profiles
 which can potentially be unstable due to axisymmetric instabilities. The
@@
 (see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
-$\sim 0.002$ (code units) for all $\tilde\beta$. The
-Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$. 
-{\bf Brunt frequency the same for all cooling?}
-The Solberg-Hoiland criteria is similary satisfied for the entire disc throughout the 
-simulations. {\bf i assume this means the azimuthally averaged
-  values. do you ever get $N^2<0$ anywhere in the 2D disc?}  
+$\sim 2\times10^{-4}$ (code units) for all $\tilde\beta$. The
+Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$, decreasing marginally with longer cooling rate. 
+The Solberg-Hoiland criteria is similarly satisfied for the entire 2D disc throughout the 
+simulations.
 Thus for all values of $\tilde\beta$ the planet-induced gaps are
 stable to axisymmertic instabilities.
+
+% {\bf Brunt frequency the same for all cooling?:small decrease with beta,
+%  not more than order of magnitude}
+
+%  {\bf i assume this means the azimuthally averaged
+%   values. do you ever get $N^2<0$ anywhere in the 2D disc?:
+%  This was not for azimuthal average, below 0 inside gap but kappa+N always above 0 everywhere}
 
@@
     to different $m$ values. The $m=3$ component is the fastest growing
-    mode during linear growth and has a corresponding
+    mode during linear growth with a growth rate of 
     $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
@@
 results in hotter gap profiles at $t=30P_0$ with less pronounced
-generalized vortensity minima. Stabilisation with increased
+generalised vortensity minima. Stabilisation with increased
 cooling time is therefore due to a smoother basic state for the
@@
   }  
-  \caption{Generalized vortensity perturbation (relative to $t=0$) for
+  \caption{Generalised vortensity perturbation (relative to $t=0$) for
     cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
@@
 than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
-Thus the disk cools back to its intial temperature corresponding to
+Thus the disk cools back to its initial temperature corresponding to
 $h=0.05$ before or during the instability growth, so we do not 
@@
 In order to capture the effect of a heated gap edge, we ran a simulation with 
-$\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
+$\tilde{\beta}=100$, corresponding to an almost adiabatic disc.  
 In this simulation the cooling rate is slow enough that the gap 
@@
 We find very similar gap profiles and mode growth rates for
-$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
+$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. At $t=30P_0$, the disc only heats up to
 values $h\simeq0.06$ in the nearly adiabatic case. This is close to
@@
 of disc-planet interaction, increasing $h$ has a stabilizing effect
-through the gap profile because it results in smoother gap
+through the setting up the gap profile because it results in smoother gap
 edges. The fact that we observe smaller growth rates as $h$ is
@@
 \subsection{Long term evolution} \label{nonlinearplanetoff} 
-{\bf typical rossby numbers for these `planet-off' vortices? do
-  stronger vortices (beta=0.1) dissipate faster?
-}
+% {\bf typical rossby numbers for these `planet-off' vortices? do
+%   stronger vortices (beta=0.1) dissipate faster?:very complicated
+% maybe not worth discussing since just comparison to planet-off}
 
@@
 left. We find the vortex merging
-time is depedendent on the growth rates of the modes and saturation
+time is dependent on the growth rates of the modes and saturation
 timescales, with the slowest growing modes in $\tilde\beta=10$ taking
@@
 single vortex. For completeness we also ran intermediate cases with
-$\tilde{\beta}=0.5$ and $5.0$. The vortices simply decay on a
-timescale of $O(10^3)$ orbits. We will see in the next section that
-this behaviour is very different to when the planet potential is kept
-on. Interestingly, though, vortices arising from the stronger
-instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
-faster.
+$\tilde{\beta}=0.5$ and $5.0$. The amplitude of the initially formed vortex was 
+found to decrease with increased cooling rate. The vortices simply decay on a
+timescale of $O(10^3)$ orbits with faster decay for stronger vortices
+(which are obtained with faster cooling rates). We will see in the
+next section that this behaviour is very different to when the planet
+potential is kept on.
+
+%Interestingly, though, vortices arising from the stronger
+%instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
+%faster.
 
@@
 averaged over $r-r_p\in[2,10]r_h$, is shown in Fig.~\ref{lifetimeplot} 
-for different $\tilde\beta$.
+for different $\tilde\beta$. The initial, post-merger vortex amplitude
+is found to be weaker for longer cooling rates (which have smaller 
+linear growth rates).
 
@@
     non-dimenionlised by the initial $m=0$ component, for long term
-    simulations with the planet potential kept on.\label{lifetimeplot}}   
+    simulations with the planet potential kept on. Notice there is
+    vortex growth after the initial linear growth, which
+    contrasts to Fig. \ref{planetofflifetimeplot}, where vortices
+    decay in the absence of continuous disc-planet interaction. 
+    \label{lifetimeplot}}   
 \end{figure}
@@
 surface density perturbation in this state. During this stage, the 
-vortex intensifies. As the vortex forms it has an initial 
-Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
-(indicating anti-cyclonic 
-motion). As the $m=1$ amplitude grows the Rossby number also increases
-in magnitude, 
- reaching a characteristic minimum value of $Ro\approx-0.35$ to $-0.45$ for
-all cooling rates. 
-{\bf Maybe useful to have a Rossby number v.s. time plot to complement
-  mode amplitude plot. 
-}
+vortex intensifies. This is better shown in Fig. \ref{rossbyplot} as the
+evolution of Rossby numbers measured at the vortex centres.  
+%As the vortex forms it has an initial 
+%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
+%(implying anti-cyclonic 
+%motion). 
+The Rossby number increases in magnitude during quasi-steady state,
+but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
+reaches a characteristic value of $Ro\approx-0.35$ for 
+$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.
 
-We find the vortices can become significantly over-dense. 
-The surface density perturbation 
-can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
-for all cases of $\tilde\beta$ in quasi-steady state, and
-was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
-$\tilde\beta=10$ case. %despite beta=10 having the weakest linear
-% instability 
-{\bf any trend between maximum over-density achieved and cooling?} 
-This large increase in surface density is due to  
-vortex growth, since there is continuous generation of vorticity by
-planet-disc interaction. This is supported by the observation that in
-the previous simulations without the planet, the amplitude of
-the post-merger vortex does not grow (Fig
-\ref{planetofflifetimeplot}).  We remark that the increase in the
-vortex surface density directly due to the pile-up of material at the
-gap edge (because of gap-opening) is small: the average surface density
-perturbation at the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
-$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.  
-{\bf may be more appropriate to quote the azimuthally averaged
-  perturbation at the location of the vortex (5-6 Hill radii from the
-  planet?)}
-
-Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
-varies with the cooling rate: for 
-$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
-$t\sim900P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
-$t\sim1200P_0$. This non-monotonic dependence suggests that there
-exists an optimal cooling rate to maximise the vortex lifetime. We
-will discuss this issue futher in
-\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
-increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
-whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
-amplitude to decay significantly after reaching maximum.
-
-{\bf question: vortex amplitude plot suggest just after linear growth,
-  vortex amplitudes are smaller with increasing beta, but later on in
-  the quasi-steady state, this trend reverses?  
-}
-{\bf Perhaps a 2D snapshot of what the gap edge looks like after
-  vortex death (like the one in the poster). defer definition of
-  vortex lifetime to later, when we discuss it. 
-}
+% {\bf Maybe useful to have a Rossby number v.s. time plot to complement
+% mode amplitude plot. }
 
@@
 
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+  0cm 0cm 1cm]{figures/rossby}
+  \caption{Evolution of Rossby numbers at the centres of the vortices
+    form in in discs with different cooling rates. A negative Rossby
+    number implies anti-cyclonic motion. Boxcar averaging
+    was used to remove contributions from the planet-induced spiral shock.\label{rossbyplot}}    
+\end{figure}
+
+We find the vortices become significantly 
+over-dense. Fig. \ref{overdensity} plots the surface density perturbation measured at the
+vortex centres, showing $\Delta\Sigma/\Sigma_0 \gtrsim 7$ for all
+cases of $\tilde\beta$ in quasi-steady state, and 
+$\mathrm{max}(\Delta\Sigma/\Sigma_0)\sim 11$ for $\tilde\beta=5$.  
+The maximum over-density typically increases with longer cooling
+times, despite the vortices are initially weaker at formation with
+increasing $\tilde{\beta}$. The large increase in the surface density
+is due to vortex growth as there is continuous generation of vorticity by
+planet-disc interaction. This is supported by the observation that in
+the previous simulations without the planet, the amplitude of
+the post-merger vortex does not grow (Fig. 
+\ref{planetofflifetimeplot}).
+
+% We remark that the increase in the
+% vortex surface density directly due to the pile-up of material at the
+% gap edge (because of gap-opening) is small: the average surface density
+% perturbation at vortex radial region, $r\sim r_{p}+5r_h$, is 
+% $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
+  0cm 0cm 1cm]{figures/vortex_density}
+  \caption{Average value of the relative surface density perturbation
+    at vortex centres for various cooling
+    times. \label{overdensity}}     
+\end{figure}
+
+Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
+varies with the cooling rate: for 
+$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
+$t\sim900P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
+$t\sim1200P_0$. This non-monotonic dependence suggests that there
+exists an optimal cooling rate to maximise the vortex lifetime. We
+will discuss this issue further in 
+\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
+increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
+whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
+amplitude to decay significantly after reaching maximum.
+
+% {\bf may be more appropriate to quote the azimuthally averaged
+%   perturbation at the location of the vortex (5-6 Hill radii from the
+%   planet?): yes, this was my original idea, mistyped as 2.5
+% }
+
+% {\bf any trend between maximum over-density achieved and cooling?: generally increasing with beta, max at beta=5}
+
+% {\bf question: vortex amplitude plot suggest just after linear growth,
+%   vortex amplitudes are smaller with increasing beta, but later on in
+%   the quasi-steady state, this trend reverses? yes. 
+% }
+
+% {\bf Perhaps a 2D snapshot of what the gap edge looks like after
+%   vortex death (like the one in the poster).: Not sure if this is figure 
+% you were thinking (Fig 8)
+% % defer definition of  vortex lifetime to later, when we discuss it. 
+% }
+
+% \begin{figure}
+%   \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma100}
+%   \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma130}
+%   \includegraphics[width=.3\linewidth,height=.5\linewidth]{figures/analysis_sigma200}
+
+%   \caption{2D gap profile for $\tilde\beta=1.0$ during quasisteady state (left), during dissipation (middle), and after dissipation (right).
+%     \label{2dgaps}} 
+% \end{figure}
+
 \subsection{Additional analysis on vortex decay}
@@
 
-In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
-more compact as it grows before decaying. 
-{\bf rough estimate of
-  vortex aspect-ratio in the three snapshots? 
-  (length/width, by inspection is ok)}  
-Notice in Fig. \ref{shockplot} the appearence of wakes extending from
+In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
+, but becomes more compact approaching a ratio of 2 during its decay ($t=1310$).   
+Notice in Fig. \ref{shockplot} the appearance of wakes extending from
 either side of the vortex at $t=1300P_0$. These 
@@
 the vortex begins to decay coincides with the emergence of these
-wakes. {\bf true statement?}
+wakes. %{\bf true statement? yes, distortion of vortex shape form wakes happens at same time}
 
@@
 During quasi-steady state the average value of the surface density gradient
-along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf this quantity
-  has dimensions of 1/length, preferable to quote something
-  dimensionless, or say `in code units'}  
+along the wakes is $|w_s\nabla\Sigma/\Sigma| \sim 0.4 $, where 
+$w_s\simeq0.1$ (code units) is a typical length scale of 
+the surface density variation across the wake.   
 Just before the $m=1$ amplitude begins to decrease, we observe this quantity
-sharply increases to $ \sim 6 $, and remains around this value
+sharply increases to $ \sim 0.6 $, and remains around this value
 until the vortex dies out, at which point the associated Rossby number
-quickly approaches zero.
+begins decreasing to zero
+%{\bf :check}.
+
+% {\bf rough estimate of
+%   vortex aspect-ratio in the three snapshots? 
+%   (length/width, by inspection is ok):trend of decreasing aspect ratio is common for all beta}  
+% {\bf this quantity has dimensions of 1/length, preferable to quote something
+%   dimensionless, or say `in code units': used shock width since relavant
+%  distance, since small may appear to weaken value/argument}
 
@@
   \caption{The vortex in the case with $\tilde{\beta}=1$
-    during quasi-steady state (left), at maximum amplitude of the
-    $m=1$ surface density component, 
+    during quasi-steady state (left), start of decay 
     (middle), and just after the decay in the $m=1$ amplitude
@@
     (top) and the associated surface density gradient (bottom) are
-    shown. This plot can considered in conjunction with
+    shown. This plot is to be considered in conjunction with
     Fig. \ref{lifetimeplot}. 
@@
 We also measured large increases in the Mach number near
-the vortex as it reaches maximum $m=1$ amplitude and begins to decay. 
-Figure \ref{machplot} plots the Mach number $M=|\bm{v} -
+the vortex as the $m=1$ surface density amplitude reaches maximum and begins to decay. 
+Fig. \ref{machplot} plots the Mach number $M=|\bm{v} -
 \bm{v}_\mathrm{vor}|/c_s$, where 
@@
 a region within $2H$ of the vortex centre. 
-During the quasi-steady state the Mach number increase
-steadily, and for all cases $M$ reaches a maximum at 
+During the quasi-steady state the Mach number increases 
+steadily, and for all cases $M$ maximises about
 % with larger growth rate for higher $\tilde\beta$.
-a time coincident with the start of vortex decay. {\bf check if
-  statement is true}
+$\sim 100P_0$ after the start the $m=1$ surface density
+amplitude starts to decay. 
+% {\bf check if  statement is true}
 
-Putting the above observations together, we suggest the decay
-in the vortex amplitude is due to shock formation by the
-vortex. When the vortex reaches large amplitude, it begins
+Putting the above observations together, we suggest that vortex decay
+(in the $m=1$ surface density amplitude) is due to shock formation by
+the vortex. When the vortex reaches large amplitude, it begins
 to induce shocks in the surrounding fluid, as supported by the
-increase in Mach number and the appearence of wakes with large surface
+increase in Mach number and the appearance of wakes with large surface
 density gradients. The vortex may lose energy through shock
@@
 6r_h$). We see that after its amplitude has decayed ($t\sim1500P_0$,   
-Fig. \ref{lifetimeplot}), this local surface density maximum is 
+Fig. \ref{lifetimeplot}), this local surface density maximum is also 
 smoothed out.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/gapchange}
+  \caption{Azimuthally averaged profiles of the relative surface
+    density perturbations for  $\tilde\beta=1.0$
+    before (solid) and after vortex decay
+    (dashed).
+    \label{gap_smoothed}} 
+\end{figure}
+
+We characterise the smoothness of the outer gap edge with a dimensionless
+gap edge gradient parameter
+\begin{align}
+  \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r)
+    \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r)
+    \rangle_{\phi}} \ra_{\Delta r}  
+\end{align}
+where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging, spanning
+from centre of the gap to the radius of the surface density maximum.
+A larger $\delta \Sigma$ characterises a sharper gap edge and  
+larger local surface density maxima.
+
+A plot of the gradient parameter over time
+for the $\tilde\beta=1.0$ case is shown in Fig. \ref{smoothnessplot}.
+During vortex decay, the outer gap edge is 
+drastically smoothed out, changing from a value of $\delta\Sigma=1.2$ during
+quasi-steady state to $0.4$ after dissipation. 
 %and we find that the gap edge is smoothed during this
@@
 by the planet, so the condition for the RWI
-becomes less favourable, which explains why vortices do not
+becomes less favourable, which may explain why vortices do not
 reform again (at least within the simulation timescale). 
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/gapchange}
-  \caption{Azimuthally averaged profiles of the relative surface
-    density perturbations for  $\tilde\beta=1.0$
-    before (solid) and after vortex decay
-    (dashed). {\bf remember to correct the time quoted in the
-      legend} \label{gap_smoothed}}  
+  \includegraphics[width=\linewidth]{figures/gap_smoothness}
+  \caption{Non-dimensional measure of the surface density gradient at
+    the outer gap edge $\tilde\beta=1.0$. A large $\delta\Sigma$
+    corresponds to a more sharply peaked outer gap edge.
+    \label{smoothnessplot}}  
 \end{figure}
 
-{\bf It's a good idea to display the smoothening of the outer gap
-  edge with time. 
-  The gap depth parameter correlates
-  well with vortex decay, but its sharp decrease in magnitude gives
-  the impression that the gap is 
-  `filled', which doesn't reflect Fig. \ref{gap_smoothed}. Visually,
-  the gap depth is the same, but the local surface density maximum at
-  the outer gap edge is smoothed. Perhaps we can come up with another
-  way to characterizing the sharpness of the peak? then we can show
-  how this peak is smoothed during vortex decay
-}
+% {\bf It's a good idea to display the smoothening of the outer gap
+%   edge with time. 
+%   The gap depth parameter correlates
+%   well with vortex decay, but its sharp decrease in magnitude gives
+%   the impression that the gap is 
+%   `filled', which doesn't reflect Fig. \ref{gap_smoothed}. Visually,
+%   the gap depth is the same, but the local surface density maximum at
+%   the outer gap edge is smoothed. Perhaps we can come up with another
+%   way to characterizing the sharpness of the peak? then we can show
+%   how this peak is smoothed during vortex decay
+% }
 
@@
   rate}\label{lifetime_discuss} 
-{\bf the discussion here may need to be changed, depending on
-  updated/new definitions of vortex lifetime. here, i'm assuming we
-  still get the non-monotonic dependence, which may or may not remain
-  true. 
-}
+% {\bf the discussion here may need to be changed, depending on
+%   updated/new definitions of vortex lifetime. here, i'm assuming we
+%   still get the non-monotonic dependence, which may or may not remain
+%   true. 
+% }
 
 We now examine vortex lifetimes as function of the imposed cooling
-times. We characterise the vortex lifetime as {\bf insert
-  definition(s).} and plot this in Fig. \ref{betaplot}. 
-Increasing the cooling timescale increases the vortex
-lifetime up to $\tilde{\beta}=1$ {\bf (judging from
-  Fig. \ref{lifetimeplot} without a quantitative definition)}, after
-which vortex lifetimes decrease again.
+times. We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
+at which the over-density of the vortex
+% to reach $10\%$ of its max overdensity value
+returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
+{\bf correct interpretation?} (which is on the order of the
+initial over-density associated with the gap formation).
 
-%Fig. \ref{lifetimeplot} and Fig. \ref{betaplot} shows that increasing 
-%$\tilde{\beta}$
-%increases the lifetime of the vortex up to a critical 
-%value $\tilde{\beta}=5.0$ for which the vortex lasted to a simulation time of
-%$1250P_0$ before starting to dissipate. This is consistent in time with the
-%vortex introducing shocks to the system.
+We plot $t_{\mathrm{life}}$ with respect to cooling times in
+Fig. \ref{betaplot}. We also plot $t_{\mathrm{diss}}$: the time elapsed before the
+vortex to begins to dissipate (when the $m=1$ surface density
+amplitude begins to decay); and $t_{\mathrm{Mach}}$: the time taken
+for the average Mach number around the vortex to maximise.
+
+For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
+vortex lifetime is maximised for $\tilde{\beta}\to0$: we find 
+$t_{\mathrm{life}} \approx 1300P_0$ for $\tilde{\beta}=0.1$ and 
+decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
+$\tilde{\beta}=0.5$. This is due to the longer decay timescale with
+decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
+1$) the vortex lifetimes maximises at $\tilde\beta=5.0$ with 
+$t_{\mathrm{life}}\approx 1525P_0$.  The other timescales $t_\mathrm{diss}$ and
+$t_\mathrm{Mach}$ are less complicated, but still display a
+non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
+$\tilde{\beta}=1$. 
+% his is in comparison with the timescale of intial vortex dissipation which has 
+% non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.
+%non-monotonic dependence for all timescales
 
@@
   \includegraphics[width=\linewidth]{figures/betaplot}
-  \caption{Lifetime of vortices as a function of the cooling paramter 
-    $\tilde\beta$. {\bf maybe useful to over-plot other timescales,
-      e.g. time taken to reach max mach, time for rossby number to
-      reach zero.}\label{betaplot}} 
+  \caption{Characteristic timescales associated with vortex evolution, as a function of the cooling
+    parameter $\tilde\beta$: $t_\mathrm{diss}$ is when the $m=1$
+    surface density amplitude begins to decay (black);
+    $t_\mathrm{life}$ is the time at which the over-density at the
+    vortex centre decreases to $\Delta\Sigma/\Sigma \sim 1$ after
+    reaching maximum (red); and $t_\mathrm{Mach}$ is the time taken
+    for the average Mach number around the vortex to maximise
+    (blue). \label{betaplot}}  
+% {\bf maybe useful to over-plot other timescales,
+%      e.g. time taken to reach max mach, time for rossby number to
+%      reach zero.}
 \end{figure}
@@
 it starts to induce shocks. We therefore suggest the vortex lifetime
-is determined by the time needed for the vortex to grow to
-sufficient amplitude to induce shocks in the surrounding fluid, as
-well as the decay timescale itself. The latter was found to shorten
-with increasing $\tilde{\beta}$. However, the former (i.e. the
-duration of the quasi-steady state), depends non-monotonically on the
-cooling time.  
+is determined the time needed for the vortex to grow to 
+sufficient amplitude to induce shocks in the surrounding fluid, and 
+the decay timescale itself. The latter was found to shorten 
+with increasing $\tilde{\beta}$. However, the former timescale
+($t_\mathrm{diss}$), which may be considered as the 
+duration of the quasi-steady state, reaches a single maximum with
+respect to cooling time. The sum of these two timescales results in the
+double-peaked behaviour of $t_\mathrm{life}$ in Fig. \ref{betaplot}. 
 % \footnote{The 
@@
 %  (Fig. \ref{lifetimeplot}), except at very short cooling times.}. 
-We discuss below competing factors that may
-result in a non-monotonic dependence of vortex lifetimes on the cooling timescale.
+We suggest below competing factors that may result in this
+non-monotonic dependence of $t_\mathrm{diss}$ on the cooling rate.
 
-\subsubsection{Factors contributing to longer vortex lifetimes}
+\subsubsection{Factors that lengthen vortex lifetimes}
 It has been shown that the amplitude at which the RWI saturates 
@@
 with the present simulations: at the beginning of the 
-quasi-steady state ($t\sim100P_0$) we find the relative surface density
-boost at centre of vortex is $\Delta\Sigma/\Sigma_0=2.5$ for
-$\tilde\beta=0.1$ and $\Delta\Sigma/\Sigma_0=1.48$ for
-$\tilde\beta=10$.  {\bf rossby numbers?}
+quasi-steady  state ($t\sim100P_0$) we find the over-density at the
+vortex centre is $\Delta\Sigma/\Sigma_0=2.5$ for $\tilde\beta=0.1$ and
+$\Delta\Sigma/\Sigma_0=1.48$ for $\tilde\beta=10$.%, both with
+%$Ro\simeq-0.15$.   
+% {\bf rossby numbers?}
 
-After merging, the growth of the single vortex is mediated by disc-planet
+The growth of the post-merger single vortex is mediated by disc-planet
 interaction. However, gap-opening becomes more difficult in a hotter 
-disc. %this effect probably becomes less important for very long
+disc, and we find the generalised vortensity 
+profiles are smoother with increasing $\tilde{\beta}$. This opposes
+the RWI. 
+%this effect probably becomes less important for very long
 % cooling times -> our gap profiles for beta=1, 10 and 100 appear similar
-Furthermore, the increased sound-speed suggest the vortex should    
-reach larger amplitudes in order to induce shocks.
+Furthermore, the vortex should reach larger amplitudes to induce
+shocks on account of the increased sound-speed.
 
 These considerations suggest, with increased cooling times, 
-it takes longer for the vortex grow to sufficient amplitude to induce 
-shocks and dissipate, which imply longer lifetimes with increasing 
-$\tilde{\beta}$.
+it takes longer for the post-merger vortex grow to sufficient
+amplitude to induce shocks and dissipate. This factor contributes to a
+longer quasi-steady state with increasing $\tilde{\beta}$.
 
-{\bf hotter disc $\to$ gap edges less sharp, still true during quasi
-  steady state?}
+%{\bf hotter disc $\to$ gap edges less sharp, still true during quasi
+%  steady state?: gap edges almost identical in quais-steady but generalized vortensity smoothed out}
 
-\subsubsection{Factors contributing to shorter vortex lifetimes} 
+\subsubsection{Factors that shorten vortex lifetimes} 
 % However, Fig. \ref{lifetimeplot} shows that very long cooling
@@
 % lifetime. 
-Notice in Fig. \ref{lifetimeplot}, for $\tilde{\beta}=10$ the vortex 
-growth during the quasi-steady state is in fact faster than for
-$\tilde{\beta}=5$, despite the former case started with a weaker
-vortex. For example, at $t\sim 500P_0$ the vortex with   
-$\tilde{\beta}=10$ has the largest amplitude. This is also reflected
-in Fig. \ref{machplot}, where the Mach number reaches its maximum
-value sooner for $\tilde{\beta}=10$ than for $\tilde{\beta}=5$.
+Notice in Fig. \ref{lifetimeplot} and Fig. \ref{overdensity}, the vortex 
+growth during the quasi-steady state is actually faster for $\tilde{\beta}=10$
+than for $\tilde{\beta}=5$. 
+For example, at $t\sim 500P_0$ the vortex with    
+$\tilde{\beta}=10$ has a larger 
+amplitude than for $\tilde{\beta}=5$. This is also reflected in Fig. \ref{machplot}, where the
+Mach number reaches its maximum value sooner for $\tilde{\beta}=10$
+than for $\tilde{\beta}=5$.
 
-This observation is consistent with the RWI being favoured by higher
-disc temperatures \citep{li00,lin12c}, which corresponds to longer
-cooling times in our case. While our `planet-off' simulations indicate
-this is unimportant for the linear instability, it may have contributed 
-significantly to the vortex growth during quasi-steady state, when one
-considers very long cooling times (e.g. $\tilde{\beta}=10$). This
-effect shortens the vortex lifetime by allowing it to grow
-faster and induce shocks sooner.
+This observation is consistent with the RWI being favoured by 
+higher temperatures \citep{li00,lin12c} through the perturbations 
+(as opposed to its effect through the set up of the gap profile
+discussed previously),
+which corresponds to longer cooling times in our case. 
+While our `planet-off' simulations indicate this is unimportant for
+the linear instability, it may have contributed  
+significantly to the vortex growth during quasi-steady state at very
+long cooling times (e.g. $\tilde{\beta}=10$). This effect shortens the
+vortex lifetime by allowing it to grow faster and induce shocks
+sooner.
 

### 2014-12-14 paper/summary.tex
final edit by MKL

@@
+% {\bf question: vortex amplitude plot suggest just after linear growth,
+%   vortex amplitudes are smaller with increasing beta, but later on in
+%   the quasi-steady state, this trend reverses? yes. 
+% }
 \section{Summary and discussion}\label{summary}
@@
 which opposes gap-opening by the planet. This means that the gaps
-opened by the planet in a disc with longer $t_c$ is smoother and
-therefore  more stable to the RWI.
+opened by the planet in a disc with longer $t_c$ are smoother and
+therefore  more stable to the RWI. %generalized vortensity minima less
+                                %pronounced
 
@@
 grows during this quasi-steady state, until it begins to induce
-shocks, afterwhich the vortex amplitude begins to decay.
+shocks, after which the vortex amplitude begins to decay.
 
-We observe the duration of the quasi-steady state increases with
-increasing cooling time until some critical value, beyond which this
+The duration of the quasi-steady state increases with
+increasing cooling timescales until a critical value, beyond which this
 quasi-steady state shortens again. The timescale for the vortex to
 decay, after reaching maximum amplitude, decreases monotonically with
-increasing cooling time.
-
+increasing cooling time. The sum of these two timescales results in a
+double-peaked vortex lifetime as a function of cooling
+timescale. 
 %technically we should probably say there is an optimum cooling time
@@
 
-Our results indicate there exists an optimum cooling timescale to
-maximise the lifetime of gap-edge vortices. We suggest the
-non-monotonic dependence of the quasi-steady state can be attributed
-to the time required for the vortex to grow to sufficient amplitude to
-induce shocks in the surrounding fluid, thereby losing energy and also
-smooth out the gap edge.
+%Our results indicate there exists an optimum cooling timescale to
+%maximise the lifetime of gap-edge vortices.
+
+We suggest the non-monotonic dependence of the quasi-steady state can
+be attributed to the time required for the vortex to grow to 
+sufficient amplitude to induce shocks in the surrounding fluid,
+thereby losing energy and also smooth out the gap edge.
 
@@
 deeper gap which favours the RWI, leading to stronger 
-vortices. However, for long  cooling timescales, the vortex grow
-faster during the quasi-steady state {\bf check if statement true:yes}. We
+vortices. For long cooling timescales, we find the vortex
+grows faster during the quasi-steady state. In accordance with
+previous stability calculations \citep{li00}, we 
 suggest the latter is due to the RWI being favoured with increasing
-disc temperature, according to previous stability calculations. These
-competing factors imply for both very short and long cooling
-timescales, the vortex reaches its maximum amplitude, shock, and
-begins to decay, sooner than intermediate cooling timescales.
+disc temperature, and that this effect overcomes weaker
+gap-opening for sufficiently long cooling times. 
+These competing factors imply
+for both very short and long cooling timescales, the vortex reaches
+its maximum amplitude, shock, and begins to decay, sooner than
+intermediate cooling timescales.
 
@@
 also reported by \cite{fu14}, who performed isothermal disc-planet
-simluations with different values of the   
+simulations with different values of the   
 disc aspect ratio. In their simulations the optimum aspect ratio is
@@
 ($r-r_p\in[2,10]r_h$), % {\bf define this region}
-we find for the longest  $\tilde\beta=5.0$ case, which has the longest lifetime,
-$h\approx0.06$ on average throughout its lifetimes. 
+we find for $\tilde\beta=5.0$, which has the longest lifetime, that 
+$h\approx0.06$ on average.  
 %{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
@@
 \subsection{Caveats and outlooks}
-We discuss below several outstanding issues that needs to be addressed
-in future works:
+There are several outstanding issues that needs to be addressed
+in future work:
 
-\emph{Convergence.} Although lower-resolution simulations performed in
+\emph{Convergence.} Although lower resolution simulations performed in
 the early stages of this project gave similar results  
 (most importantly, the non-monotonic dependence of vortex lifetimes on
-the cooling timescale), we did find the lower-resolution simulations  
+the cooling timescale), we did find the lower resolution 
 typically yield longer vortex lifetimes than that reported in this
 paper. This could be due to weaker RWI with low resolution.  
-It will be neccessary to perform even higher resolution  simulations
-in order to obtain quantitatively co nverged vortex lifetimes.
+It will be necessary to perform even higher resolution  simulations
+in order to obtain quantitatively converged vortex lifetimes.
 
@@
 study. Based on linear calculations, \cite{lovelace13} concluded
-self-gravity to be important for the RWI when $Q<O(1/h)$, or
-$Q\lesssim 20$ for $h\sim0.05$, as was typially considered in this
+self-gravity to be important for the RWI when the Toomre parameter $Q<O(1/h)$, or
+$Q\lesssim 20$ for $h\sim0.05$, as was typically considered in this
 work. This suggests that self-gravity may affect vortex lifetimes even
-when the Toomre $Q$ parameter is not small. In particular, given that
+when the $Q$ is not small. In particular, given that
 we observe vortices can reach significant over-densities (up to almost

### 2014-12-20 paper/results.tex
RL final

@@
   \caption{Long term simulations without the planet potential after
-    the gap is set up. The $m=1$ surface density component at the
+    the gap is set up. The $m=1$ surface density component,
+    non-dimenionlised by the initial $m=0$ component, at the
     outer gap edge is shown as a function of the cooling timescale. 
@@
   0cm 0cm 1cm]{figures/longterm_stability}
-  \caption{Evolution of the $m=1$ surface density component,
-    non-dimenionlised by the initial $m=0$ component, for long term
+  \caption{Evolution of the non-dimesnionalized $m=1$ surface density component
+    for long term
     simulations with the planet potential kept on. Notice there is
@@
   \caption{Evolution of Rossby numbers at the centres of the vortices
-    form in in discs with different cooling rates. A negative Rossby
+    formed in discs with different cooling rates. A negative Rossby
     number implies anti-cyclonic motion. Boxcar averaging
@@
     at vortex centres for various cooling
-    times. \label{overdensity}}     
+    times. Intial vortex overdensities decrease with cooling rate while
+    vortex growth rates increase with cooling rate.
+    \label{overdensity}}     
 \end{figure}
@@
 $\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
-$t\sim900P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
+$t\sim800P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
 $t\sim1200P_0$. This non-monotonic dependence suggests that there
@@
     (top) and the associated surface density gradient (bottom) are
-    shown. This plot is to be considered in conjunction with
+    shown. Wake like features corresponding to large density gradients
+    are found to originate from the vortices during the late phase of 
+    their quasi-steady states and into dissipation times.
+    This plot is to be considered in conjunction with
     Fig. \ref{lifetimeplot}. 
@@
     before (solid) and after vortex decay
-    (dashed).
+    (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
+    smoothens out the sharp outer gap edge.
     \label{gap_smoothed}} 
@@
   \caption{Non-dimensional measure of the surface density gradient at
-    the outer gap edge $\tilde\beta=1.0$. A large $\delta\Sigma$
-    corresponds to a more sharply peaked outer gap edge.
+    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
+    the gap edge is found to have a large gradient and sharp peak while vortex
+    dissipation,
+    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
+    works to smoothen out the gap edge.
+% A large $\delta\Sigma$
+ %   corresponds to a more sharply peaked outer gap edge.
     \label{smoothnessplot}}  
@@
 returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
-{\bf correct interpretation?} (which is on the order of the
+%{\bf correct interpretation?}
+ (which is on the order of the
 initial over-density associated with the gap formation).
@@
   \caption{Characteristic timescales associated with vortex evolution, as a function of the cooling
-    parameter $\tilde\beta$: $t_\mathrm{diss}$ is when the $m=1$
-    surface density amplitude begins to decay (black);
-    $t_\mathrm{life}$ is the time at which the over-density at the
+    parameter $\tilde\beta$: time in which the $m=1$
+    surface density amplitude begins to decay, $t_\mathrm{diss}$ (black);
+    time at which the over-density at the
     vortex centre decreases to $\Delta\Sigma/\Sigma \sim 1$ after
-    reaching maximum (red); and $t_\mathrm{Mach}$ is the time taken
+    reaching maximum, $t_\mathrm{life}$ (red);
+    and $t_\mathrm{Mach}$ is the time taken
     for the average Mach number around the vortex to maximise

### 2014-12-22 paper/model.tex
final edits before submission

@@
 The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
-pressure, stellar gravity and disc gravity. For a thin disc, 
+pressure forces and stellar gravity. For a thin disc, 
 $v_{\phi}\simeq r\Omega_k$. The initial radial velocity is
@@
 %Hill radius chosen so eps doesn't change with time (scale-height does)
-The planet is held on a fixed circular orbit with $ r_p = 10\rin$. This also
+The planet is held on a fixed circular orbit with $ r_p = 10\rin$ and $\phi_p=\Omega_k(r_p)t$. 
+This also
 defines the time unit $P_0\equiv 2\pi/\Omega_k(r_p)$ used to describe results. 
@@
 to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
-algorithm to circumvent the time-step restriction set by the inner disc boundary. 
+algorithm to circumvent the time-step restriction set by the fast rotation speed at the 
+inner disc boundary. 
 The disc is divided into $(N_r,N_\phi)$ zones in the radial and azimuthal directions, 
@@
 the authors show an extremum in $\teta$ may lead to a dynamical
-instability, the RWI. In a barotropic disc, the entropy factor is
+instability, the RWI. In a barotropic disc where $p=p(\Sigma)$, the entropy factor is 
 absent and the important quantity is the PV.

### 2014-12-22 paper/paper.tex
final edits before submission

@@
   by a giant planet. Numerical models of gap-edge vortices have
-  thus far employed locally isothermal discs, but the 
+  so far employed locally isothermal discs, but the 
   theory of this vortex-forming or `Rossby wave' instability was
@@
 In adiabatic discs, the relevant quantity for stability
-becomes a generalisation of the PV that include entropy variations
+becomes a generalisation of the PV that accounts for entropy variations
 \citep{lovelace99}.

### 2014-12-22 paper/results.tex
final edits before submission

@@
 planet potential and azimuthally average the surface density, energy
-and velocity fields. (At this point the planet has carved a partial
-gap and the RWI has not yet occurred.) Effectively, we initialise the
-disc with a gap profile. 
+and velocity fields. At this point the planet has carved a partial
+gap but the RWI has not yet occurred. 
 We then perturb the surface density in the outer disc ($r>r_p$) and continue to 
-evolve the disc. We impose  sinusoidal perturbations with 
+evolve the disc. (Effectively, we initialise the
+disc with a gap profile.) We impose  sinusoidal perturbations with 
 azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
@@
   \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimenionlised by the initial axisymmetric component 
+    density, non-dimensionalised by the initial axisymmetric component 
     $\Sigma_0(t=0)$, for the
@@
     the gap is set up. The $m=1$ surface density component,
-    non-dimenionlised by the initial $m=0$ component, at the
+    non-dimensionalised by the initial $m=0$ component, at the
     outer gap edge is shown as a function of the cooling timescale. 
@@
   0cm 0cm 1cm]{figures/longterm_stability}
-  \caption{Evolution of the non-dimesnionalized $m=1$ surface density component
+  \caption{Evolution of the non-dimensionalised $m=1$ surface density component
     for long term
@@
 In all cases the system remains in a quasi-steady state for
-$\gtrsim800P_0$ with a single vortex circulating 
-the outer gap edge at the local Keplerian  
-frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
-surface density perturbation in this state. During this stage, the 
-vortex intensifies. This is better shown in Fig. \ref{rossbyplot} as the
-evolution of Rossby numbers measured at the vortex centres.  
-%As the vortex forms it has an initial 
-%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
-%(implying anti-cyclonic 
-%motion). 
-The Rossby number increases in magnitude during quasi-steady state,
-but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
-reaches a characteristic value of $Ro\approx-0.35$ for 
-$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.
+$\gtrsim800P_0$ with a single vortex circulating the outer gap edge at
+the local Keplerian frequency. Fig. \ref{Vortex2D} shows a typical
+plot of the relative surface density perturbation in this state.    
+The duration of the quasi-steady state varies with the cooling
+rate. Fig. \ref{lifetimeplot} shows that for $\tilde{\beta}=0.1$ and
+$10$, the vortex amplitude begins to decay around 
+$t\sim800P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
+$t\sim1200P_0$. This non-monotonic dependence suggests that there
+exists an optimal cooling rate to maximise the vortex lifetime. We
+will discuss this issue further in 
+\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
+increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
+whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
+amplitude to decay significantly after reaching maximum.
 
@@
 
+We find the vortices intensify during the quasi-steady state. This is
+better shown in Fig. \ref{rossbyplot} as the evolution of Rossby
+numbers measured at the vortex centres.   
+%As the vortex forms it has an initial 
+%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
+%(implying anti-cyclonic 
+%motion). 
+The Rossby number increases in magnitude during quasi-steady state,
+but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
+reaches a characteristic value of $Ro\approx-0.35$ for 
+$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.
+
 \begin{figure}
@@
     formed in discs with different cooling rates. A negative Rossby
-    number implies anti-cyclonic motion. Boxcar averaging
-    was used to remove contributions from the planet-induced spiral shock.\label{rossbyplot}}    
+    number implies anti-cyclonic motion.%  Boxcar averaging
+    % was used to remove contributions from the planet-induced spiral shock.
+    \label{rossbyplot}}    
 \end{figure}
@@
   0cm 0cm 1cm]{figures/vortex_density}
-  \caption{Average value of the relative surface density perturbation
+  \caption{Evolution of the average value of the relative surface density perturbation
     at vortex centres for various cooling
-    times. Intial vortex overdensities decrease with cooling rate while
-    vortex growth rates increase with cooling rate.
+    times. % The initial vortex over-densities decrease with increasing cooling times while
+    % vortex growth rates increase with cooling rate.
     \label{overdensity}}     
 \end{figure}
-
-Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
-varies with the cooling rate: for 
-$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
-$t\sim800P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
-$t\sim1200P_0$. This non-monotonic dependence suggests that there
-exists an optimal cooling rate to maximise the vortex lifetime. We
-will discuss this issue further in 
-\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
-increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
-whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
-amplitude to decay significantly after reaching maximum.
 
@@
     (top) and the associated surface density gradient (bottom) are
-    shown. Wake like features corresponding to large density gradients
+    shown. Wakes corresponding to large density gradients
     are found to originate from the vortices during the late phase of 
-    their quasi-steady states and into dissipation times.
+    their quasi-steady states, and persists during their decay. 
     This plot is to be considered in conjunction with
@@
 Putting the above observations together, we suggest that vortex decay
-(in the $m=1$ surface density amplitude) is due to shock formation by
+(in the $m=1$ surface density amplitude) is associated with shock formation by
 the vortex. When the vortex reaches large amplitude, it begins
@@
 can smooth out the gap structure that originally gave rise to the
-RWI, which would oppose vortex growth. We examine this below.
+RWI, which is expected to oppose vortex growth. We examine this below.
 
@@
     (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
-    smoothens out the sharp outer gap edge.
+    smooth out out the sharp outer gap edge.
     \label{gap_smoothed}} 
@@
 %process. 
-This can be interpreted as the vortex providing a viscosity;  
+This may be interpreted as the vortex providing a viscosity;  
 and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
@@
     which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
-    works to smoothen out the gap edge.
+    works to smooth out out the gap edge.
 % A large $\delta\Sigma$

### 2014-12-22 paper/summary.tex
final edits before submission

@@
 dominant azimuthal wavenumber also decreases. This is consistent with
-the notion that increasing $t_c$ leads to higher temperatures or $h$,
+the notion that increasing $t_c$ leads to higher temperatures or
+equivalently the disc aspect-ratio $h$,
 which opposes gap-opening by the planet. This means that the gaps
@@
 simulations with different values of the   
-disc aspect ratio. In their simulations the optimum aspect ratio is
-$h=0.06$. In our simulations, the disc aspect ratio is a dynamical
+disc aspect-ratio. In their simulations the optimum aspect-ratio is
+$h=0.06$. In our simulations, $h$ is a dynamical
 variable, but by analysing the region where the vortex is located
 ($r-r_p\in[2,10]r_h$), % {\bf define this region}
-we find for $\tilde\beta=5.0$, which has the longest lifetime, that 
+we find for a dimensionless cooing timescale of $\tilde\beta=5.0$, which has
+the longest vortex lifetime, that  
 $h\approx0.06$ on average.  
@@
 work. This suggests that self-gravity may affect vortex lifetimes even
-when the $Q$ is not small. In particular, given that
+when $Q$ is not small. In particular, given that
 we observe vortices can reach significant over-densities (up to almost

### 2014-12-22 paper/model.tex
final edit before 1st submission

@@
 where the power-law index $s=2$, $H(r) = \ciso\Omega_k $ defines the disc scale-height 
-where $\ciso=\sqrt{p/\Sigma}$ isothermal sound-speed. The disc aspect-ratio is defined as $h\equiv H/r$ and initially
+where $\ciso=\sqrt{p/\Sigma}$ isothermal sound-speed. The disc aspect ratio is defined as $h\equiv H/r$ and initially
 $h=0.05$. For a non-self-gravitating disc, the surface density scale
@@
 where $m$ is the azimuthal wave number. We define the growth rate
-$\gamma$ of the $m^\mathrm{th}$ component of the surface density
+$\sigma$ of the $m^\mathrm{th}$ component of the surface density
 through 
@@
 \begin{align}\label{growth}
-  \frac{d \langle|\Sigma_m|\rangle_r }{dt}= \gamma \langle|\Sigma_m|\rangle_r 
+  \frac{d \langle|\Sigma_m|\rangle_r }{dt}= \sigma \langle|\Sigma_m|\rangle_r 
 \end{align}

### 2014-12-22 paper/results.tex
final edit before 1st submission

@@
   \centering
-  \caption{Dominant mode and growth rates for
+  \caption{Dominant azimuthal wavenumber $m$ and associated growth rate $\sigma$ for
     $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling)
@@
       \toprule
-      $m$ & $10^2\gamma/\Omega(r_p)$ \\
+      $m$ & $10^2\sigma/\Omega(r_p)$ \\
       \midrule
@@
 $\tilde{\beta}=10$. We characterise these
-modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
+modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
 Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
@@
     mode during linear growth with a growth rate of 
-    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
+    $\sigma=0.017\Omega(r_p)$.\label{linearmodes}}
   % {\bf is the $\Sigma_0$ at t=0?} }  
@@
 azimuthal Fourier mode decreases from $ m=9\rightarrow3$ and the
-respective growth rate decreases from $ \gamma/\Omega(r_p)=0.079
+respective growth rate decreases from $ \sigma/\Omega(r_p)=0.079
 \rightarrow 0.017$. However, despite two orders of magnitude increase in the
@@
   \caption{Generalised vortensity perturbation (relative to $t=0$) for
-    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
+    cases of $\tilde{\beta}=0.1,1,10$ (left, middle, right) during
     the growth of non-axisymmetric modes. The planet potential has
@@
 stability calculations, because the cooling time is comparable or shorter
-than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
+than the instability growth time, $t_c\lesssim\sigma^{-1}$.  
 Thus the disk cools back to its initial temperature corresponding to
@@
     (top) and the associated surface density gradient (bottom) are
-    shown. Wakes corresponding to large density gradients
-    are found to originate from the vortices during the late phase of 
-    their quasi-steady states, and persists during their decay. 
+    shown. Wakes corresponding to large surface density gradients
+    are found to originate from the vortex during the late phase of 
+    the quasi-steady state, and the wakes persist during vortex decay. 
     This plot is to be considered in conjunction with
@@
     (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
-    smooth out out the sharp outer gap edge.
+    smooth out the outer gap edge.
     \label{gap_smoothed}} 
@@
   \caption{Non-dimensional measure of the surface density gradient at
-    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
+    the outer gap edge for $\tilde\beta=1.0$. During the vortex quasi-steady state
     the gap edge is found to have a large gradient and sharp peak while vortex
@@
     which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
-    works to smooth out out the gap edge.
+    works to smooth out the gap edge.
 % A large $\delta\Sigma$
@@
   \caption{Characteristic timescales associated with vortex evolution, as a function of the cooling
-    parameter $\tilde\beta$: time in which the $m=1$
+    parameter $\tilde\beta$: time at which the $m=1$
     surface density amplitude begins to decay, $t_\mathrm{diss}$ (black);

### 2014-12-22 paper/summary.tex
final edit before 1st submission

@@
 the notion that increasing $t_c$ leads to higher temperatures or
-equivalently the disc aspect-ratio $h$,
+equivalently the disc aspect ratio $h$,
 which opposes gap-opening by the planet. This means that the gaps
@@
 simulations with different values of the   
-disc aspect-ratio. In their simulations the optimum aspect-ratio is
+disc aspect ratio. In their simulations the optimum aspect ratio is
 $h=0.06$. In our simulations, $h$ is a dynamical

### 2015-02-06 paper/results.tex
Minor addition to referees comments

--- 2014-12-20 paper/results.tex
@@
 At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
-(see below), we find $\kappa^2 + N^2$ reaches local minimum with a value
-$\sim 2\times10^{-4}$ (code units) for all $\tilde\beta$. The
-Brunt-V\"ais\"al\"a frequency is $N\sim 10^{-5}$, decreasing marginally with longer cooling rate. 
+(see below),{\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
+$\sim 0.45 \, \Omega^2(r_p)$
+%$\sim 2\times10^{-4}$ (code units)
+ for all $\tilde\beta$. The
+Brunt-V\"ais\"al\"a frequency at the outer gap edge is
+$N\sim 10^{-2} \, \Omega^2(r_p)$,}
+% $N\sim 10^{-5}$,
+decreasing marginally with longer cooling rate. 
 The Solberg-Hoiland criteria is similarly satisfied for the entire 2D disc throughout the 
@@
 the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
+{\bf Isothermal simulations with set sound speed where also analyzed showing growth rates $\sim 5 \times 10^2 / \Omega(r_p)$.
+In comparison nearly isothermal simulation with $\tilde\beta=0.01$ had growth rates $\sim 8\times 10^2 / \Omega(r_p)$}
 
@@
     $\tilde{\beta}$ are similar.
+    {\bf The decrease in density pertibation at $r>40r_{in}$ arises from the open
+    boundry condition.}
     \label{Vortex2D}} 
@@
 wakes. %{\bf true statement? yes, distortion of vortex shape form wakes happens at same time}
+
+{\bf Radial locations of the vortices were found to migrate over their lifetime.
+During the quasi-steady state the vortices were found to orbit $\sim 6.5r_h$ from the planet radius.
+Simultaneously with the appearence of wakes, the vortices were found to move towards $\sim 5.3r_h$.
+This is a result of the the spiral waves transporting angular momentum to the disk (Is this true?
+Shocks on the inner disk seem stronger so I would think it would migrate outward).
+After the vortex reaches marginal amplitudes $<1 \Delta\Sigma/\Sigma$ the vortices were found to migrate back out towards $\sim 6.5 r_h$ from the planet location }
 

### 2015-02-15 paper/response.txt
revision in the works

- §§4 and 5: would the results of a simulation with a very short
  cooling time (beta << 1) be the same as those of a simulation with a
  locally isothermal equation of state and same unperturbed sound
  speed? 
  ********************************************************************
  We have performed addtional (low resolution) simulations with
  beta=0.01 to compare with a locally isothermal disks. We find a
  smaller linear growth rate in the locally isothermal case.

  ********************************************************************

- §4.4: why does the vortex’s perturbed density decrease on the long
  term? I suppose a related question is: what controls the long-term
  evolution of the vortex’s potential vorticity or entropy-modified
  potential vorticity? (whichever quantity is relevant). 
  ********************************************************************
  In theory a vortex should live forever in 2D inviscid flow. In
  practice, however, there is numerical viscosity which we suspect is
  responsible for the long term decay. As the vortex weakens
  (elongates) its area increases and its surface density drops.

  viscosity and elongation of vortex

  ask robert: does gap shape change significantly?

  ********************************************************************

- §5.2-5.3: here again I’d like to have more details on the vortex
        evolution and decay: 	
        . Do the authors see the vortex migrate? I did a simulation
        with the authors’ setup and a rather long cooling time
        (beta_tilde ~ 8) where I see the vortex moving from ~6.5 rh to
        ~5.3 rh relative to the planet’s orbital radius before the
        vortex starts decaying (rh = Hill radius). The time when the
        vortex starts decaying actually corresponds to a rather sudden
        increase in the planet-vortex distance, which goes up to ~6.8
        rh until the vortex dies off completely. If the authors get a
        similar migration behaviour in their simulations, maybe this
        could give further insight into the vortex evolution (vortex
        getting stronger during the quasi steady state before fading
        away). The connection between migration and the appearance of
        spiral waves launched by the vortex could also be made. 
	********************************************************************
        We thank the reviewer for raising the issue of vortex
        migration. We have re-examined our simulation results. We
        indeed observe vortex migration but our result is somewhat
        different to that of the reviewer.

 	We do not see significant vortex migration during quasi-steady
 	state. This may be due to the fact that the
 	vortex forms at a surface density maximum, which is able to
 	trap vortices (Paardekooper et al 2010).

	Instead, we observe the vortex begins to migrate inwards
	with the appearence of the spiral wakes (cf. planetary
	migration). However, we find the vortex eventually migrates
	outward to about 6.5rh before complete decay.

	We remark that for a vortex formed at a surface density
 	maximum (e.g. from the RWI), that subsequent perturbations may
 	lead to inwards or outwards vortex migration, depending on
 	what side of the maximum the vortex is displaced to
 	(Paardekooper et al 2010).

	Although we do not understand the discrepency between our
	results and the reviewer's, we do agree with the reviewer
	that vortex migration probably plays a role in determining
	the overall vortex lifetime (it may affect the duration of the
	quasi-steady state and/or the decay timescale). We now mention
	this in the discussions and caveats.

        ********************************************************************

        . During the quasi steady state the vortex grows as a result
        of continuous vortensity generation upon crossing the planet’s
        wake. What about possible turbulent diffusion of potential
        vorticity if the planet forms in a region of the disc with
        non-negligible ‘turbulent viscosity’? Have the authors done
        simulations with a non vanishingly small viscosity? 
        ********************************************************************
	Since a low-viscosity disc is needed for the RWI to operate,
        we have chosen to focus on invisicd simulations. This also
        helps to minimize vortex decay directly due to a physical
        viscosity, so that we can examine the influence of
        thermodynamics.

	Thus, we expect that vortex lifetimes in realistic viscous discs
	to be shorter than that seen in our 
	simulations. We mention this in the discussions.

        ********************************************************************
        . I was a bit surprised that the RWI couldn’t be launched
        again after the vortex had decayed, and I’d like the authors
        to be more specific about why it is so. About the vortex decay
        providing an effective alpha viscosity~ 1d-2: isn’t this a
        time-varying quantity, decreasing as the vortex is dying? If
        so is it really the reason why the RWI isn’t launched again?
        What about the disc’s aspect ratio at the outer gap edge after
        the vortex’s full decay?     
        ********************************************************************
	The quoted value of 0.01 is a typical viscosity associated
        with the vortex, averaged over radius and time.

	By the time the vortex dies out that there is
	no longer a `vortex viscosity', but the outer gap edge
	(surface density bump) has already been smoothed out, and the
	site is no longer suitable for the RWI. Note that from
	Fig. 11, smoothing out the surface density bump puts material
	into the outer disc.

	In order to re-launch the RWI, the surface density bump needs
	to reform. However, this seems difficult as the planet already
	resides in a deep gap with little material to clear out
	(c.f. the original surface density bump resulted from
	gap-opening: the planet pushes material to either side of its
	orbit).

        ********************************************************************

- §6.1: two caveats / outlooks could be added: (i) having a more
  realistic cooling model (same caveat as for numerical simulations of
  disc fragmentation by the gravitational instability…) and (ii)
  considering planet migration. 
  ********************************************************************
  These are now added. 
  ********************************************************************
- §2: given the very low shear viscosity used in this work, I would
  emphasize straight after eqs. (2) and (3) that f_v and H include the
  contribution of a (Von Neumann-Ritchmyer) artificial bulk viscosity
  to handle shocks. 
  ********************************************************************
  These comments are now added. 
  ********************************************************************
- §4.2: the minimum value of 2d-4 for kappa^2 + N^2 should rather be
  expressed in units of Omega^2 at the planet’s orbital radius. Same
  advice for N, for which the value of 1d-5 didn’t really speak to me
  until I realised that the planet’s orbital radius wasn’t the code’s
  unit of length. By the way, where is N evaluated? 
  ********************************************************************
  The frequencies are now expressed in units of the orbital frequency
  at the planet's orbital radius. 
  ********************************************************************
- Fig. 6: the colour scale suggests that there is a discontinuous jump
  in perturbed density at orbital radii R > 40 r_in, where the density
  seems to take very small values… why is that? 
  ********************************************************************
  This is due to the open boundary conditions adopted at the outer
  boundary, which leads to mass loss. We now mention this in the
  caption of Fig. 6. 
  ********************************************************************

### 2015-02-15 paper/model.tex
revision in the works

@@
 frame. 
-In the momentum 
-equations, $\bm{f}_\nu$ represent viscous forces whose magnitude is 
+In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
+which includes artificial viscosity to handle shocks, and a
+Navier-Stokes viscosity } whose magnitude is  
 characterised by a constant kinematic viscosity parameter
 $\nu$. However, we will be considering effectively inviscid discs by
-adopting small values of $\nu$ .
+adopting small values of $\nu$.
 
@@
 \end{align}
-where $Q^+$ represents viscous heating and subscript $i$ denotes
+where $Q^+$ represents viscous heating {\bf(from both physical and
+  artificial viscosity)} and subscript $i$ denotes
 evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as

### 2015-02-15 paper/results.tex
revision in the works

@@
 the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
-{\bf Isothermal simulations with set sound speed where also analyzed showing growth rates $\sim 5 \times 10^2 / \Omega(r_p)$.
-In comparison nearly isothermal simulation with $\tilde\beta=0.01$ had growth rates $\sim 8\times 10^2 / \Omega(r_p)$}
+
+{\bf We also simulated a locally isothermal disk where the sound-speed is
+  held fixed, which yield a most unstable growth rate
+  $\sigma/Omega(r_p)\simeq 0.05$, compared with a value of $0.08$ for
+  a case with $\tilde{\beta}=0.01$.}   
+%Isothermal simulations with set sound speed where also analyzed
+%  showing growth rates $\sim 5 \times 10^{-2} / \Omega(r_p)$. }
+%In comparison nearly isothermal simulation with $\tilde\beta=0.01$ had growth rates $\sim 8\times 10^52 / \Omega(r_p)$}
 
@@
 timescale of $O(10^3)$ orbits with faster decay for stronger vortices
-(which are obtained with faster cooling rates). We will see in the
-next section that this behaviour is very different to when the planet
-potential is kept on.
+(which are obtained with faster cooling rates). {\bf This decay is 
+  probably due to numerical viscosity. During the slow decay the
+  vortex elongates (weakens) while its radial width remains $O(H)$, so
+  its surface density decreases. (true?)} 
+We will see in the
+next section that this behaviour {\bf after linear growth} is very
+different to when the planet potential is kept on.
 
@@
 the outer gap edge at the local Keplerian  
-frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
+frequency. 
+Fig. \ref{Vortex2D} shows a typical plot of the relative 
 surface density perturbation in this state. During this stage, the 
@@
     $\tilde\beta=0.1$ case during quasi-steady state with a single
-    vortex at the outer gap edge. 
-    The plots for other values of the cooling time
-    $\tilde{\beta}$ are similar.
-    {\bf The decrease in density pertibation at $r>40r_{in}$ arises from the open
-    boundry condition.}
-    \label{Vortex2D}} 
+    vortex at the outer gap edge. The plot for other values of the cooling time
+    $\tilde{\beta}$ are similar. {\bf The decrease in the surface
+      density near} $r\sim 40 r_\mathrm{in}$ {\bf arises from mass loss due to the open
+      boundary condition imposed at} $r_\mathrm{out}$.
+    \label{Vortex2D} }
 \end{figure}
@@
 
+{\bf For comparison we also ran a locally isothermal simulation to
+  compare with the case with rapid cooling ($\tilde{\beta}=0.01$), but
+  did not observe significant vortex decay within our simulation
+  timescale in the former case. This suggests there could be
+  fundamental differences between the two cases.}
+
 % {\bf may be more appropriate to quote the azimuthally averaged
@@
 
-{\bf Radial locations of the vortices were found to migrate over their lifetime.
-During the quasi-steady state the vortices were found to orbit $\sim 6.5r_h$ from the planet radius.
-Simultaneously with the appearence of wakes, the vortices were found to move towards $\sim 5.3r_h$.
-This is a result of the the spiral waves transporting angular momentum to the disk (Is this true?
-Shocks on the inner disk seem stronger so I would think it would migrate outward).
-After the vortex reaches marginal amplitudes $<1 \Delta\Sigma/\Sigma$ the vortices were found to migrate back out towards $\sim 6.5 r_h$ from the planet location }
+{\bf During the quasi-steady state the vortex 
+  orbits at $r\sim r_p+6r_h$. We do not see significant
+  vortex migration at this stage (true?), since the vortex is located at a
+  surface density maximum, which can trap vortices
+  \citep{paardekooper10}. However, simultaneous with the
+  appearence of the wakes, we observe the vortex begins to migrate 
+  inwards to $r\sim r_p+5r_h$. 
+  % This is a result of the the spiral waves transporting
+  % angular momentum to the disk (Is this true? Shocks on the inner disk
+  % seem stronger so I would think it would migrate outward). 
+  }
 
@@
 until the vortex dies out, at which point the associated Rossby number
-begins decreasing to zero
+begins decreasing to zero. {\bf After the 
+  vortex reaches small amplitudes ($1\lesssim\Delta\Sigma/\Sigma$), it 
+  migrates out to $r\sim r_p+ 6.5 r_h$. 
+}
 %{\bf :check}.
@@
 %process.
+
 This can be interpreted as the vortex providing a viscosity;  
@@
 associated with the vortex. This acts against gap-opening
-by the planet, so the condition for the RWI
-becomes less favourable, which may explain why vortices do not
-reform again (at least within the simulation timescale). 
+by the planet {\bf, and smooths out the outer surface density bump,    
+} so the condition for the RWI 
+becomes less favourable. {\bf In
+  order to re-launch the RWI, the surface density bump should
+  reform. However, this is difficult as there is no more material
+  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}).} {\bf This may} may explain
+why vortices do  not reform again (at least within the simulation timescale). 
+{\bf what is aspect-ratio at outer gap edge after full decay?}
+
 %Thus, the

### 2015-02-15 paper/summary.tex
revision in the works

@@
 stability, which employ locally or strictly isothermal equations of
-state.
+state. {\bf We considered the inviscid limit which favors the RWI
+  \citep{li09,fu14}, so that the vortex lifetimes observed in our
+  simulations are likely longer than in realistic disks with non-zero
+  viscosity. }
 
@@
 decay, after reaching maximum amplitude, decreases monotonically with
-increasing cooling time. The sum of these two timescales results in a
-double-peaked vortex lifetime as a function of cooling
-timescale. 
+increasing cooling time. {\bf We observe vortex migration during this
+  decay, which may contribute to this decay timescale.} The sum of
+these two timescales results in a double-peaked vortex lifetime as a
+function of cooling timescale. 
 %technically we should probably say there is an optimum cooling time
@@
 
+{\bf
+  \emph{Orbital migration.} We have held the planet fixed on a circular
+  orbit. However, gap-edge vortices are known to exert
+  significant, oscillatory torques on the planet \citep{li09}
+  which can lead to complex orbital migration. This will likely affect
+  vortex lifetimes as it may alter the planet-vortex separation, as
+  well as leading to direct vortex-planet interactions
+  \citep{lin10,ataiee14}. Thus, 
+  future simulations should allow the planet to freely
+  migrate. Similarly, the role of vortex migration on its lifetime
+  should be clarified. 
+}
+
+{\bf 
+  \emph{Cooling model.} Our prescription for the disk 
+  heating/cooling is convenient to probe the full range of
+  thermodynamic response of the disk. However, in order to calculate 
+  vortex lifetimes in actual 
+  protoplanetary disks, a more realistic thermodynamics treatment ,
+  e.g. based on radiative cooling, should be used in future work.   
+}
+
 \emph{Self-gravity.} We have ignored disc self-gravity in this

### 2015-02-25 paper/results.tex
Near final update

@@
   vortex elongates (weakens) while its radial width remains $O(H)$, so
-  its surface density decreases. (true?)} 
+  its surface density decreases. (true?) (For isothermal and long cases yes. Spiral waves appear in beta=0.1,0.5,1.0})
 We will see in the
@@
 
-{\bf For comparison we also ran a locally isothermal simulation to
-  compare with the case with rapid cooling ($\tilde{\beta}=0.01$), but
-  did not observe significant vortex decay within our simulation
-  timescale in the former case. This suggests there could be
-  fundamental differences between the two cases.}
+{\bf For comparison we also ran a locally isothermal simulation with results concident with those found in this section.
+}
+
+%{\bf For comparison we also ran a locally isothermal simulation to
+%  compare with the case with rapid cooling ($\tilde{\beta}=0.01$), but
+%  did not observe significant vortex decay within our simulation
+%  timescale in the former case. This suggests there could be
+%  fundamental differences between the two cases.}
 
@@
   orbits at $r\sim r_p+6r_h$. We do not see significant
-  vortex migration at this stage (true?), since the vortex is located at a
+  vortex migration at this stage (true?:yes), since the vortex is located at a
   surface density maximum, which can trap vortices
@@
 why vortices do  not reform again (at least within the simulation timescale). 
-{\bf what is aspect-ratio at outer gap edge after full decay?}
+{\bf After full decay the aspect ratio at the outer gap edge is $h\sim0.05$}
+% (0.06 for beta=5.0,10)}
+
+% what is aspect-ratio at outer gap edge after full decay?
 

### 2015-03-04 paper/model.tex
revised manuscript done

@@
 velocity, $p$ is the vertically-integrated pressure, $e=p/(\gamma-1)$ is the energy
-per unit area and the adiabatic index $\gamma=1.4$ is assumed
-constant.
+per unit area and the adiabatic index $\gamma$ is assumed
+constant. We take $\gamma=1.4$ unless otherwise stated.
 
@@
 which includes artificial viscosity to handle shocks, and a
-Navier-Stokes viscosity } whose magnitude is  
+Navier-Stokes viscosity} whose magnitude is  
 characterised by a constant kinematic viscosity parameter

### 2015-03-04 paper/results.tex
revised manuscript done

@@
 Brunt-V\"ais\"al\"a frequency at the outer gap edge is
-$N\sim 10^{-2} \, \Omega^2(r_p)$,}
+$N\sim 10^{-2} \, \Omega(r_p)$} 
 % $N\sim 10^{-5}$,
@@
 
-{\bf We also simulated a locally isothermal disk where the sound-speed is
-  held fixed, which yield a most unstable growth rate
-  $\sigma/Omega(r_p)\simeq 0.05$, compared with a value of $0.08$ for
-  a case with $\tilde{\beta}=0.01$.}   
 %Isothermal simulations with set sound speed where also analyzed
@@
 \end{figure}
+
+{\bf For completeness we also performed a lower resolution simulation
+  where the sound-speed is effectively held fixed (a locally
+  isothermal disk) by setting $\gamma=1.01$. This simulation
+  yield a most unstable growth rate $\sigma/\Omega(r_p)\simeq 0.05$,
+  compared with a value of $0.08$ for a case with
+  $\tilde{\beta}=0.01$. However, the vortex evolutions were similar.}
 
@@
 timescale of $O(10^3)$ orbits with faster decay for stronger vortices
-(which are obtained with faster cooling rates). {\bf This decay is 
-  probably due to numerical viscosity. During the slow decay the
-  vortex elongates (weakens) while its radial width remains $O(H)$, so
-  its surface density decreases. (true?) (For isothermal and long cases yes. Spiral waves appear in beta=0.1,0.5,1.0})
+(which are obtained with faster cooling rates).
+
+{\bf This decay is probably due to numerical viscosity. During the
+  slow decay the vortex elongates (weakens) while its radial width
+  remains $O(H)$, so its surface density decreases. In addition, for 
+  $\tilde{\beta}=0.1,\,0.5$ and $1.0$ we also observe the appearence of
+  spiral waves associated with the vortex, which may contribute to
+  its dissipation (see below).} 
+% (true?) (For isothermal and long
+% cases yes. Spiral waves appear in beta=0.1,0.5,1.0}) 
 We will see in the
-next section that this behaviour {\bf after linear growth} is very
+next section that this {\bf decay after linear growth} is very
 different to when the planet potential is kept on.
@@
 effects on vortex evolution, and apply open boundaries at
-$r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
-simulations with $(N_r,N_{\phi})=(256,512)$ were initially carried
-out, which show similar behaviour and trends as the high-resolution
-runs reported below.
+$r=r_\mathrm{in},\,r_\mathrm{out}$.
+
+We comment that lower-resolution simulations with
+$(N_r,N_{\phi})=(256,512)$ show similar behaviour and trends as the
+high-resolution runs reported below. {\bf These simulations
+  include a case with $\tilde{\beta}=0.01$ and a locally isothermal
+  disk run, which show similar behavior. In particular, we find vortex
+  lifetimes have a non-monotonic dependence on $\tilde{\beta}$, with
+  longest lifetimes as $\tilde{\beta}\to 0$ (or locally isothermal
+  disks) and for $\tilde{\beta}\sim 1$}.
 
@@
 
-{\bf For comparison we also ran a locally isothermal simulation with results concident with those found in this section.
-}
-
 %{\bf For comparison we also ran a locally isothermal simulation to
@@
   orbits at $r\sim r_p+6r_h$. We do not see significant
-  vortex migration at this stage (true?:yes), since the vortex is located at a
-  surface density maximum, which can trap vortices
-  \citep{paardekooper10}. However, simultaneous with the
-  appearence of the wakes, we observe the vortex begins to migrate 
-  inwards to $r\sim r_p+5r_h$. 
+  vortex migration at this stage, since the vortex is located at a 
+  surface density maximum \citep{paardekooper10}. However,
+  simultaneous with the appearence of the wakes, we observe the vortex
+  begins to migrate inwards to $r\sim r_p+5r_h$.  
   % This is a result of the the spiral waves transporting
@@
   reform. However, this is difficult as there is no more material
-  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}).} {\bf This may} may explain
+  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}) to
+  form a surface density bump outside the gap.} {\bf This may} explain
 why vortices do  not reform again (at least within the simulation timescale). 
-{\bf After full decay the aspect ratio at the outer gap edge is $h\sim0.05$}
+{\bf After full decay the aspect ratio at the outer gap edge is
+  $h\sim0.05$. }
 % (0.06 for beta=5.0,10)}

### 2015-03-04 paper/summary.tex
revised manuscript done

@@
 state. {\bf We considered the inviscid limit which favors the RWI
-  \citep{li09,fu14}, so that the vortex lifetimes observed in our
-  simulations are likely longer than in realistic disks with non-zero
-  viscosity. }
+  \citep{li09,fu14} and avoids complications from viscous
+  heating. However, this means that the vortex lifetimes observed in
+  our simulations are likely longer than in realistic disks with
+  non-zero viscosity. }
 
@@
 we find that with the planet potential kept on, the vortex amplitude
-grows during this quasi-steady state, until it begins to induce
+grows during this quasi-steady state {\bf, during which no vortex migration is
+  observed}, until it begins to induce
 shocks, after which the vortex amplitude begins to decay.
@@
 decay, after reaching maximum amplitude, decreases monotonically with
-increasing cooling time. {\bf We observe vortex migration during this
-  decay, which may contribute to this decay timescale.} The sum of
+increasing cooling time. {\bf We do observe vortex migration during this
+  decay, which may influence this decay timescale. } 
+The sum of
 these two timescales results in a double-peaked vortex lifetime as a
@@
   vortex lifetimes in actual 
-  protoplanetary disks, a more realistic thermodynamics treatment ,
-  e.g. based on radiative cooling, should be used in future work.   
+  protoplanetary disks, an improved thermodynamics treatment ,
+  e.g. radiative cooling based on realistic disk temperature, density,
+  opacity models etc., should be used in future work.   
 }

### 2015-03-04 paper/response.txt
Near final update
revised manuscript done

@@
+We thank the reviewer for a timely report, and apologize for the delay
+in our response. We have performed addition simulations (at lower
+resolution) to address some of the reviewer's questions. The revisions
+to the manuscript are made in bold, and further explained below.
+
 - §§4 and 5: would the results of a simulation with a very short
@@
   ********************************************************************
-  We have performed addtional (low resolution) simulations with
-  beta=0.01 to compare with a locally isothermal disks. We find a
-  smaller linear growth rate in the locally isothermal case.
+  We have performed addtional (lower resolution) simulations with
+  beta=0.01 to compare with a locally isothermal disks, which we obtain 
+  from our numerical models by setting gamma=1.01. We find a slightly
+  smaller linear growth rate in the locally isothermal case. However,
+  the vortex formation and evolution are similar. We    
+  obtain longer gap-edge vortex lifetimes than beta=0.1 (at the same
+  resolution) consistent with our double-peaked lifetime result for
+  high resolution simulations.
 
@@
   responsible for the long term decay. As the vortex weakens
-  (elongates) its area increases and its surface density drops.
-
-  viscosity and elongation of vortex
-
-  ask robert: does gap shape change significantly?
+  (elongates) its area increases and its surface density drops. In
+  addition, for beta=0.1, 0.5, 1.0, for which the initially-formed
+  vortex is stronger, we observe the appearence of spiral waves
+  associated with the vortex, which may contribute to the decay. It
+  can be seen in Fig. 4 that these cases experience a larger decay
+  rate.
 
@@
  	vortex forms at a surface density maximum, which is able to
- 	trap vortices (Paardekooper et al 2010).
+ 	trap vortices (Paardekooper et al 2010). Thus, the growth
+ 	during quasi steady state is not due to vortex migration.
 
@@
         we have chosen to focus on invisicd simulations. This also
-        helps to minimize vortex decay directly due to a physical
-        viscosity, so that we can examine the influence of
-        thermodynamics.
+        helps to minimize complications from vortex decay directly due to a physical
+        viscosity (which also contributes to heating).
 
@@
   The frequencies are now expressed in units of the orbital frequency
-  at the planet's orbital radius. 
+  at the planet's orbital radius. The quoted value of N is evaluated
+  at the outer gap edge. 
   ********************************************************************

### 2015-03-05 paper/results.tex
Final revision
Additonal Data Points

@@
 Brunt-V\"ais\"al\"a frequency at the outer gap edge is
-$N\sim 10^{-2} \, \Omega(r_p)$} 
+$N\sim 0.1 \, \Omega(r_p)$} 
 % $N\sim 10^{-5}$,
@@
 We now examine vortex lifetimes as function of the imposed cooling
-times. We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
+times. {\bf For completeness additional simultions were done for $\tilde\beta=0.01,0.25,0.75,2.5,7.5$. }
+ We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
 at which the over-density of the vortex
@@
 vortex lifetime is maximised for $\tilde{\beta}\to0$: we find 
-$t_{\mathrm{life}} \approx 1300P_0$ for $\tilde{\beta}=0.1$ and 
+$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
 decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
@@
 decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
-1$) the vortex lifetimes maximises at $\tilde\beta=5.0$ with 
-$t_{\mathrm{life}}\approx 1525P_0$.  The other timescales $t_\mathrm{diss}$ and
+1$) the vortex lifetimes maximises at $\tilde\beta=2.5$ with 
+$t_{\mathrm{life}}\approx 1650P_0$.  The other timescales $t_\mathrm{diss}$ and
 $t_\mathrm{Mach}$ are less complicated, but still display a

### 2015-03-05 paper/summary.tex
Additonal Data Points

@@
 ($r-r_p\in[2,10]r_h$), % {\bf define this region}
-we find for a dimensionless cooing timescale of $\tilde\beta=5.0$, which has
+we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
 the longest vortex lifetime, that  
-$h\approx0.06$ on average.  
+$h\approx0.058$ on average.  
 %{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 

### 2015-03-05 paper/response.txt
revised manuscript, submitted

@@
 We thank the reviewer for a timely report, and apologize for the delay
-in our response. We have performed addition simulations (at lower
-resolution) to address some of the reviewer's questions. The revisions
-to the manuscript are made in bold, and further explained below.
+in our response. We have performed additional simulations to address
+some of the reviewer's questions. In particular, Fig. 13 has been
+expanded to include more cases to strengthen our main result 
+of a non-monotonic dependence of the vortex lifetime on beta. The
+revisions to the manuscript are made in bold, and further explained
+below.
 
@@
   ********************************************************************
-  We have performed addtional (lower resolution) simulations with
-  beta=0.01 to compare with a locally isothermal disks, which we obtain 
+  We have performed additional (lower resolution) simulations that
+  compare beta=0.01 with a locally isothermal disks, which we obtain 
   from our numerical models by setting gamma=1.01. We find a slightly
   smaller linear growth rate in the locally isothermal case. However,
-  the vortex formation and evolution are similar. We    
-  obtain longer gap-edge vortex lifetimes than beta=0.1 (at the same
-  resolution) consistent with our double-peaked lifetime result for
-  high resolution simulations.
+  the vortex formation and evolution are similar.
+
+  We obtain longer gap-edge vortex lifetimes with beta=0.01 than 
+  beta=0.1. We now include this result in Fig. 13 (all simulations in
+  this figure employ the same resolution).
 
@@
   addition, for beta=0.1, 0.5, 1.0, for which the initially-formed
-  vortex is stronger, we observe the appearence of spiral waves
+  vortex is stronger, we observe the appearance of spiral waves
   associated with the vortex, which may contribute to the decay. It
-  can be seen in Fig. 4 that these cases experience a larger decay
+  can be seen in Fig. 4 that these cases experience a faster decay
   rate.
@@
  	vortex forms at a surface density maximum, which is able to
- 	trap vortices (Paardekooper et al 2010). Thus, the growth
- 	during quasi steady state is not due to vortex migration.
+ 	trap vortices (Paardekooper et al 2010). Thus, in our
+ 	simulations the growth during quasi steady state is not due to
+ 	vortex migration.
 
 	Instead, we observe the vortex begins to migrate inwards
-	with the appearence of the spiral wakes (cf. planetary
+	with the appearance of the spiral wakes (cf. planetary
 	migration). However, we find the vortex eventually migrates
@@
 
-	Although we do not understand the discrepency between our
+	Although we do not understand the discrepancy between our
 	results and the reviewer's, we do agree with the reviewer
@@
 	Since a low-viscosity disc is needed for the RWI to operate,
-        we have chosen to focus on invisicd simulations. This also
-        helps to minimize complications from vortex decay directly due to a physical
+        we have chosen to focus on inviscid simulations. This also
+        helps to minimize complications from vortex decay directly due
+        to a physical  
         viscosity (which also contributes to heating).

### 2015-03-05 paper/model.tex
revised manuscript, submitted

@@
 In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
-which includes artificial viscosity to handle shocks, and a
+which includes artificial bulk viscosity to handle shocks, and a
 Navier-Stokes viscosity} whose magnitude is  
@@
 systematic way: $\beta\ll1$ is a locally isothermal disc while
-$\beta\gg1$ is an adiabatic disk.
+$\beta\gg1$ is an adiabatic disc.
 

### 2015-03-05 paper/paper.tex
revision in the works
revised manuscript, submitted

@@
   originally developed for adiabatic discs.  
-  We generalise the study of planetary gap stability to non-isothermal
-  discs using customised numerical simulations of disc-planet
+  We generalize the study of planetary gap stability to non-isothermal
+  discs using customized numerical simulations of disc-planet
   systems where the planet opens an unstable gap. 
@@
   large-scale, over-dense vortex circulating the outer gap edge, typically  
-  lasting $O(10^3)$ orbits. 
-  Vortex lifetimes were found to generally increase with cooling times up to an
-  optimal value, beyond which vortex lifetimes decrease. This
+  lasting $O(10^3)$ orbits. {\bf
+    We find vortex lifetimes generally increase with the cooling
+    timescale $t_c$} up to an optimal value {\bf of $t_c\sim 10$
+    orbits}, beyond which vortex lifetimes decrease. This 
   non-monotonic dependence is qualitatively consistent with 
-  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
+  recent studies using strictly isothermal discs that vary the disc
+  aspect ratio.  {\bf The lifetime and observability of
+    vortices in protoplanetary discs is therefore dependent on disc
+    thermodynamics. }
 %  presented in Fu et al. (2014, ApJL, 788, L41). 
@@
 In adiabatic discs, the relevant quantity for stability
-becomes a generalisation of the PV that accounts for entropy variations
+becomes a generalization of the PV that accounts for entropy variations
 \citep{lovelace99}.
@@
 planetary gaps, however, the increased temperature may also act to
-stabilise the disc by making gap-opening more difficult. It is 
+stabilize the disc by making gap-opening more difficult. It is 
 therefore of theoretical interest to clarify the effect of heating and
@@
 lifetime of gap-edge vortices in \S\ref{nonlinear},  
-as a function of the imposed cooling rate. We conclude and summarise
+as a function of the imposed cooling rate. We conclude and summarize
 in \S\ref{summary} with a discussion of important caveats.
@@
 This project was initiated at the Canadian Institute for Theoretical
-Astrophysics (CITA) 2014 summer student programme.  
+Astrophysics (CITA) 2014 summer student programme. {\bf The authors
+  thank an anonymous referee for an insightful report.} 
 Computations were performed on the GPC supercomputer at the

### 2015-03-05 paper/results.tex
revised manuscript, submitted

@@
   \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimenionlised by the initial axisymmetric component 
+    density, non-dimensionalized by the initial axisymmetric component 
     $\Sigma_0(t=0)$, for the
@@
 results in hotter gap profiles at $t=30P_0$ with less pronounced
-generalised vortensity minima. Stabilisation with increased
+generalised vortensity minima. Stabilization with increased
 cooling time is therefore due to a smoother basic state for the
@@
   where the sound-speed is effectively held fixed (a locally
-  isothermal disk) by setting $\gamma=1.01$. This simulation
+  isothermal disc) by setting $\gamma=1.01$. This simulation
   yield a most unstable growth rate $\sigma/\Omega(r_p)\simeq 0.05$,
   compared with a value of $0.08$ for a case with
-  $\tilde{\beta}=0.01$. However, the vortex evolutions were similar.}
+  $\tilde{\beta}=0.01$. However, the vortex evolution were similar.}
 
@@
 than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
-Thus the disk cools back to its initial temperature corresponding to
+Thus the disc cools back to its initial temperature corresponding to
 $h=0.05$ before or during the instability growth, so we do not 
@@
   remains $O(H)$, so its surface density decreases. In addition, for 
-  $\tilde{\beta}=0.1,\,0.5$ and $1.0$ we also observe the appearence of
+  $\tilde{\beta}=0.1,\,0.5$ and $1.0$ we also observe the appearance of
   spiral waves associated with the vortex, which may contribute to
@@
     the gap is set up. The $m=1$ surface density component,
-    non-dimenionlised by the initial $m=0$ component, at the
+    non-dimensionalized by the initial $m=0$ component, at the
     outer gap edge is shown as a function of the cooling timescale. 
@@
 We now examine long-term simulations of gap-edge vortices for
-$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
+$\tilde{\beta}=0.1,0.5,1,5,10$.{\bf(Additional cases are presented in 
+  \S\ref{lifetime_discuss} when examining vortex lifetimes as a
+  function of $\tilde{\beta}$.)} The planet potential is kept on
 throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
@@
 high-resolution runs reported below. {\bf These simulations
-  include a case with $\tilde{\beta}=0.01$ and a locally isothermal
-  disk run, which show similar behavior. In particular, we find vortex
-  lifetimes have a non-monotonic dependence on $\tilde{\beta}$, with
-  longest lifetimes as $\tilde{\beta}\to 0$ (or locally isothermal
-  disks) and for $\tilde{\beta}\sim 1$}.
+  include a comparison between a run with $\tilde{\beta}=0.01$ with a 
+  locally isothermal disc, which show similar behavior. In particular,
+  we find vortex lifetimes have a non-monotonic dependence on
+  $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
+  locally isothermal discs) and for $\tilde{\beta} = O(1)$}.
 
@@
     at vortex centres for various cooling
-    times. Intial vortex overdensities decrease with cooling rate while
+    times. Initial vortex over-densities decrease with cooling rate while
     vortex growth rates increase with cooling rate.
@@
   surface density maximum \citep{paardekooper10}. However,
-  simultaneous with the appearence of the wakes, we observe the vortex
+  simultaneous with the appearance of the wakes, we observe the vortex
   begins to migrate inwards to $r\sim r_p+5r_h$.  
   % This is a result of the the spiral waves transporting
-  % angular momentum to the disk (Is this true? Shocks on the inner disk
+  % angular momentum to the disc (Is this true? Shocks on the inner disc
   % seem stronger so I would think it would migrate outward). 
@@
 During the quasi-steady state the Mach number increases 
-steadily, and for all cases $M$ maximises about
+steadily, and for all cases $M$ maximizes about
 % with larger growth rate for higher $\tilde\beta$.
@@
     (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
-    smoothens out the sharp outer gap edge.
+    smooth out the sharp outer gap edge.
     \label{gap_smoothed}} 
@@
 from centre of the gap to the radius of the surface density maximum.
-A larger $\delta \Sigma$ characterises a sharper gap edge and  
+A larger $\delta \Sigma$ characterizes a sharper gap edge and  
 larger local surface density maxima.
@@
 associated with the vortex. This acts against gap-opening
-by the planet {\bf, and smooths out the outer surface density bump,    
+by the planet{\bf, and smooths out the outer surface density bump,    
 } so the condition for the RWI 
@@
     which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
-    works to smoothen out the gap edge.
+    works to smooth out the gap edge.
 % A large $\delta\Sigma$
@@
 We now examine vortex lifetimes as function of the imposed cooling
-times. {\bf For completeness additional simultions were done for $\tilde\beta=0.01,0.25,0.75,2.5,7.5$. }
+times. {\bf For this study, additional simulations with intermediate
+  values of $\tilde\beta=0.25,0.75,2.5,7.5$ were performed, as well as
+  a nearly (locally) isothermal case with $\tilde{\beta}=0.01$.} 
  We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
@@
 For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
-vortex lifetime is maximised for $\tilde{\beta}\to0$: we find 
+vortex lifetime is maximized for $\tilde{\beta}\to0$: we find 
 $t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
@@
 decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
-1$) the vortex lifetimes maximises at $\tilde\beta=2.5$ with 
+1$) the vortex lifetimes maximizes at $\tilde\beta=2.5$ with 
 $t_{\mathrm{life}}\approx 1650P_0$.  The other timescales $t_\mathrm{diss}$ and

### 2015-03-05 paper/summary.tex
revised manuscript, submitted

--- 2015-03-04 paper/summary.tex
@@
 non-isothermal disc-planet interaction.  
-Our simulations were customised to examine the effect of a finite 
+Our simulations were customized to examine the effect of a finite 
 cooling time on the stability of gaps  
@@
   \citep{li09,fu14} and avoids complications from viscous
-  heating. However, this means that the vortex lifetimes observed in
-  our simulations are likely longer than in realistic disks with
-  non-zero viscosity. }
+  heating other that shock heating. However, this means that the vortex lifetimes observed in
+  our simulations are likely longer than in realistic discs with
+  non-zero physical viscosity.}
 
@@
 we find that with the planet potential kept on, the vortex amplitude
-grows during this quasi-steady state {\bf, during which no vortex migration is
+grows during this quasi-steady state{\bf, during which no vortex migration is
   observed}, until it begins to induce
@@
 
-We suggest the non-monotonic dependence of the quasi-steady state can
+We suggest the non-monotonic dependence \emph{of the quasi-steady state} can
 be attributed to the time required for the vortex to grow to 
@@
 intermediate cooling timescales. 
+%{\bf, for which the quasi-steady state is 
+%the longest.}
 
@@
 $h=0.06$. In our simulations, $h$ is a dynamical
-variable, but by analysing the region where the vortex is located
+variable, but by analyzing the region where the vortex is located
 ($r-r_p\in[2,10]r_h$), % {\bf define this region}
-we find for a dimensionless cooing timescale of $\tilde\beta=5.0$, which has
+we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
 the longest vortex lifetime, that  
-$h\approx0.06$ on average.  
+$h\approx0.058$ on average.  
 %{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
-Our result is therefore consistent with \citep{fu14}.
+Our result is therefore consistent with \cite{fu14}.
 
@@
 {\bf 
-  \emph{Cooling model.} Our prescription for the disk 
+  \emph{Cooling model.} Our prescription for the disc 
   heating/cooling is convenient to probe the full range of
-  thermodynamic response of the disk. However, in order to calculate 
+  thermodynamic response of the disc. However, in order to calculate 
   vortex lifetimes in actual 
-  protoplanetary disks, an improved thermodynamics treatment ,
-  e.g. radiative cooling based on realistic disk temperature, density,
+  protoplanetary discs, an improved thermodynamics treatment ,
+  e.g. radiative cooling based on realistic disc temperature, density,
   opacity models etc., should be used in future work.   

### 2015-03-05 paper/paper.tex
revised manuscript, submitted

@@
   by a giant planet. Numerical models of gap-edge vortices have
-  so far employed locally isothermal discs, but the 
+  so far employed locally isothermal discs {\bf in which the temperature profile is held fixed}, but the 
   theory of this vortex-forming or `Rossby wave' instability was
@@
   recent studies using strictly isothermal discs that vary the disc
-  aspect ratio.  {\bf The lifetime and observability of
+  aspect ratio.  {\bf The lifetime and observability of gap-edge 
     vortices in protoplanetary discs is therefore dependent on disc

### 2015-03-05 paper/results.tex
revised manuscript, submitted

@@
 At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
-(see below),{\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
+(see below), {\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
 $\sim 0.45 \, \Omega^2(r_p)$
@@
 Brunt-V\"ais\"al\"a frequency at the outer gap edge is
-$N\sim 0.1 \, \Omega(r_p)$} 
+$N\sim 0.1 \, \Omega(r_p)$},  
 % $N\sim 10^{-5}$,
@@
   compared with a value of $0.08$ for a case with
-  $\tilde{\beta}=0.01$. However, the vortex evolution were similar.}
+  $\tilde{\beta}=0.01$. However, the vortex evolution is similar.}
 

### 2015-03-26 paper/model.tex
revision 2 done

--- 2015-02-15 paper/model.tex
@@
 frame. 
-In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
-which includes artificial viscosity to handle shocks, and a
-Navier-Stokes viscosity } whose magnitude is  
+In the momentum equations, $\bm{f}_\nu$ represent viscous forces, 
+which includes artificial bulk viscosity to handle shocks, and a
+Navier-Stokes viscosity whose magnitude is  
 characterised by a constant kinematic viscosity parameter
@@
 \end{align}
-where $Q^+$ represents viscous heating {\bf(from both physical and
-  artificial viscosity)} and subscript $i$ denotes
+where $Q^+$ represents viscous heating (from both physical and
+  artificial viscosity) and subscript $i$ denotes
 evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
@@
 is a dimensionless measure of relative vorticity. 
-% the ratio of interial to Corriolis forces. 
  Here $\langle \cdot \rangle_{\phi}$ denotes an azimuthal

### 2015-03-26 paper/paper.tex
revision 2 done

--- 2015-03-05 paper/paper.tex
@@
   by a giant planet. Numerical models of gap-edge vortices have
-  so far employed locally isothermal discs, but the 
+  so far employed locally isothermal discs in which the temperature profile is held fixed, but the 
   theory of this vortex-forming or `Rossby wave' instability was
@@
   large-scale, over-dense vortex circulating the outer gap edge, typically  
-  lasting $O(10^3)$ orbits. {\bf
+  lasting $O(10^3)$ orbits. 
     We find vortex lifetimes generally increase with the cooling
-    timescale $t_c$} up to an optimal value {\bf of $t_c\sim 10$
-    orbits}, beyond which vortex lifetimes decrease. This 
+    timescale $t_c$ up to an optimal value of $t_c\sim 10$
+    orbits, beyond which vortex lifetimes decrease. This 
   non-monotonic dependence is qualitatively consistent with 
   recent studies using strictly isothermal discs that vary the disc
-  aspect ratio.  {\bf The lifetime and observability of
+  aspect ratio.  The lifetime and observability of gap-edge 
     vortices in protoplanetary discs is therefore dependent on disc
-    thermodynamics. }
-%  presented in Fu et al. (2014, ApJL, 788, L41). 
+    thermodynamics. 
 \end{abstract}
@@
 \section{Introduction}\label{intro}
-%{
-%\bf 
-%observational motivation: asymmetric dust distributions. 
-%theoretical motivations: generalize gap stability to include energy equation. 
-%}
 The interaction between planets and protoplanetary discs plays an
@@
 protoplanetary discs  \citep[see][for a recent review]{baruteau13}.
-%\citep[][see \cite{baruteau13} for a recent
-%  review]{goldreich79, lin79, goldreich80,lin86}.
 
@@
 companions within them \citep[e.g.][]{quanz13b,reggiani14}.
-
-%gaps can be unstable - vortices
 
@@
 However, studies of Rossby vortices at planetary gap-edges have 
-adopted locally isothermal discs, %cite lin11, zhu14, lyra
+adopted locally isothermal discs, 
 where the disc temperature is a fixed function of
-position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
-%isothermal valid for outer disc  
-On the other hand, the theory of the RWI was in fact
+position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. On the other hand, the theory of the RWI was in fact
 developed for adiabatic discs \citep{li00}, which permits
-heating. %cite 
+heating. 
 In adiabatic discs, the relevant quantity for stability
@@
 up discs with gaps, but study their stability without further
-influence from the planet. %stability through background state 
+influence from the planet. 
 We then perform long-term disc-planet simulations to examine the
@@
 This project was initiated at the Canadian Institute for Theoretical
-Astrophysics (CITA) 2014 summer student programme. {\bf The authors
-  thank an anonymous referee for an insightful report.} 
+Astrophysics (CITA) 2014 summer student programme. The authors
+  thank an anonymous referee for an insightful report. 
 Computations were performed on the GPC supercomputer at the

### 2015-03-26 paper/results.tex
revision 2 done

@@
 clean gap does not form in this short time period.%
-%\footnote{We find the
-% RWI begins to develop if the disc is evolved beyond $t\sim30P_0$}.
 
@@
 $\tilde{\beta}$ is increased.
-
-%However, notice the gap structure
-%changes most significantly for $\beta =0.1 \to 1.0$. For $\beta \geq
-%1$
 
@@
 large $\tilde{\beta}$.
-% or 'suggets that .... becomes unimportant for sufficiently large...'
 
@@
 At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
-(see below), {\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
+(see below),  we find $\kappa^2 + N^2$ reaches local minimum with a value
 $\sim 0.45 \, \Omega^2(r_p)$
@@
 Brunt-V\"ais\"al\"a frequency at the outer gap edge is
-$N\sim 0.1 \, \Omega(r_p)$},  
+$N\sim 0.1 \, \Omega(r_p)$,  
 % $N\sim 10^{-5}$,
@@
 Thus for all values of $\tilde\beta$ the planet-induced gaps are
-stable to axisymmertic instabilities.
-
-% {\bf Brunt frequency the same for all cooling?:small decrease with beta,
-%  not more than order of magnitude}
-
-%  {\bf i assume this means the azimuthally averaged
-%   values. do you ever get $N^2<0$ anywhere in the 2D disc?:
-%  This was not for azimuthal average, below 0 inside gap but kappa+N always above 0 everywhere}
+stable to axisymmertic {\bf perturbations}.
 
@@
 
-{\bf For completeness we also performed a lower resolution simulation
-  where the sound-speed is effectively held fixed (a locally
-  isothermal disc) by setting $\gamma=1.01$. This simulation
-  yield a most unstable growth rate $\sigma/\Omega(r_p)\simeq 0.05$,
-  compared with a value of $0.08$ for a case with
+{\bf For completeness we also simulated a locally
+  isothermal disk (at lower resolution) with $\epsilon=0.05$ where the
+  sound-speed is kept strictly equal to its initial value.  This
+  simulation yield a most unstable growth rate
+  $\sigma/\Omega(r_p)\simeq 0.05$ at $m=5$,   
+  compared with a value of $0.085$ at $m=?$ for a corresponding
+  simulation that includes the energy equation but with rapid cooling
   $\tilde{\beta}=0.01$. However, the vortex evolution is similar.}
@@
 \subsection{Long term evolution} \label{nonlinearplanetoff}
-% {\bf typical rossby numbers for these `planet-off' vortices? do
-%   stronger vortices (beta=0.1) dissipate faster?:very complicated
-% maybe not worth discussing since just comparison to planet-off}
 
@@
 
-{\bf This decay is probably due to numerical viscosity. During the
+This decay is probably due to numerical viscosity. During the
   slow decay the vortex elongates (weakens) while its radial width
@@
   spiral waves associated with the vortex, which may contribute to
-  its dissipation (see below).} 
-% (true?) (For isothermal and long
-% cases yes. Spiral waves appear in beta=0.1,0.5,1.0}) 
+  its dissipation (see below). 
 We will see in the
-next section that this {\bf decay after linear growth} is very
+next section that this decay after linear growth is very
 different to when the planet potential is kept on.
-
-%Interestingly, though, vortices arising from the stronger
-%instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
-%faster.
-
-%main point: no planet: only get growth during linear instability
-%           with palnet: further growth after linear regime
 
@@
 We now examine long-term simulations of gap-edge vortices for
-$\tilde{\beta}=0.1,0.5,1,5,10$.{\bf(Additional cases are presented in 
+$\tilde{\beta}=0.1,0.5,1,5,10$. (Additional cases are presented in 
   \S\ref{lifetime_discuss} when examining vortex lifetimes as a
-  function of $\tilde{\beta}$.)} The planet potential is kept on
+  function of $\tilde{\beta}$.)  The planet potential is kept on
 throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
@@
 $(N_r,N_{\phi})=(256,512)$ show similar behaviour and trends as the
-high-resolution runs reported below. {\bf These simulations
-  include a comparison between a run with $\tilde{\beta}=0.01$ with a 
-  locally isothermal disc, which show similar behavior. In particular,
-  we find vortex lifetimes have a non-monotonic dependence on
-  $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
-  locally isothermal discs) and for $\tilde{\beta} = O(1)$}.
+high-resolution runs reported below. % {\bf These simulations
+  % include a comparison between a run with $\tilde{\beta}=0.01$ with a 
+  % locally isothermal disc, which show similar behavior. In particular,
+  % we find vortex lifetimes have a non-monotonic dependence on
+  % $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
+  % locally isothermal discs) and for $\tilde{\beta} = O(1)$}.
 
@@
 evolution of Rossby numbers measured at the vortex centres.  
-%As the vortex forms it has an initial 
-%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
-%(implying anti-cyclonic 
-%motion). 
 The Rossby number increases in magnitude during quasi-steady state,
@@
     vortex at the outer gap edge. The plot for other values of the cooling time
-    $\tilde{\beta}$ are similar. {\bf The decrease in the surface
-      density near} $r\sim 40 r_\mathrm{in}$ {\bf arises from mass loss due to the open
-      boundary condition imposed at} $r_\mathrm{out}$.
+    $\tilde{\beta}$ are similar. The decrease in the surface
+      density near $r\sim 40 r_\mathrm{in}$ arises from mass loss due to the open
+      boundary condition imposed at $r_\mathrm{out}$.
     \label{Vortex2D} }
@@
 \ref{planetofflifetimeplot}).
-
-% We remark that the increase in the
-% vortex surface density directly due to the pile-up of material at the
-% gap edge (because of gap-opening) is small: the average surface density
-% perturbation at vortex radial region, $r\sim r_{p}+5r_h$, is 
-% $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.
 
@@
 
-% {\bf may be more appropriate to quote the azimuthally averaged
-%   perturbation at the location of the vortex (5-6 Hill radii from the
-%   planet?): yes, this was my original idea, mistyped as 2.5
-% }
-
-% {\bf any trend between maximum over-density achieved and cooling?: generally increasing with beta, max at beta=5}
-
-% {\bf question: vortex amplitude plot suggest just after linear growth,
-%   vortex amplitudes are smaller with increasing beta, but later on in
-%   the quasi-steady state, this trend reverses? yes. 
-% }
-
-% {\bf Perhaps a 2D snapshot of what the gap edge looks like after
-%   vortex death (like the one in the poster).: Not sure if this is figure 
-% you were thinking (Fig 8)
-% % defer definition of  vortex lifetime to later, when we discuss it. 
-% }
-
 % \begin{figure}
@@
 the vortex begins to decay coincides with the emergence of these
-wakes. %{\bf true statement? yes, distortion of vortex shape form wakes happens at same time}
+wakes.
 
-{\bf During the quasi-steady state the vortex 
+During the quasi-steady state the vortex 
   orbits at $r\sim r_p+6r_h$. We do not see significant
@@
   begins to migrate inwards to $r\sim r_p+5r_h$.
-  % This is a result of the the spiral waves transporting
-  % angular momentum to the disc (Is this true? Shocks on the inner disc
-  % seem stronger so I would think it would migrate outward). 
-  }
 
@@
 until the vortex dies out, at which point the associated Rossby number
-begins decreasing to zero. {\bf After the 
+begins decreasing to zero. After the 
   vortex reaches small amplitudes ($1\lesssim\Delta\Sigma/\Sigma$), it 
   migrates out to $r\sim r_p+ 6.5 r_h$.
-}
+
 %{\bf :check}.
@@
     (top) and the associated surface density gradient (bottom) are
-    shown. Wake like features corresponding to large density gradients
+    shown. Wake-like features corresponding to large density gradients
     are found to originate from the vortices during the late phase of 
@@
 associated with the vortex. This acts against gap-opening
-by the planet{\bf, and smooths out the outer surface density bump,    
-} so the condition for the RWI 
-becomes less favourable. {\bf In
+by the planet, and smooths out the outer surface density bump,    
+ so the condition for the RWI 
+becomes less favourable. In
   order to re-launch the RWI, the surface density bump should
@@
   in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}) to
-  form a surface density bump outside the gap.} {\bf This may} explain
-why vortices do  not reform again (at least within the simulation timescale). 
-{\bf After full decay the aspect ratio at the outer gap edge is
-  $h\sim0.05$. }
-% (0.06 for beta=5.0,10)}
-
-% what is aspect-ratio at outer gap edge after full decay?
-
-%Thus, the
-%time needed for the vortex to grow to sufficient amplitude to modify
-%the gap structure
+  form a surface density bump outside the gap. This may explain
+why vortices do  not reform again (at least within the simulation
+timescale).  After full decay the aspect ratio at the outer gap edge is
+  $h\sim0.05$.
 
@@
     dissipation,
-    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
+    which occurs at $t\approx10^3P_0$ as seen in Fig. \ref{lifetimeplot},
     works to smooth out the gap edge.
@@
 We now examine vortex lifetimes as function of the imposed cooling
-times. {\bf For this study, additional simulations with intermediate
-  values of $\tilde\beta=0.25,0.75,2.5,7.5$ were performed, as well as
-  a nearly (locally) isothermal case with $\tilde{\beta}=0.01$.} 
+times. {\bf For this study, additional simulations with 
+  $\tilde\beta=0.01,\,0.25,\,0.75,\,2.5,\,7.5$ were also performed.} 
 We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
 at which the over-density of the vortex
-% to reach $10\%$ of its max overdensity value
 returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
-%{\bf correct interpretation?}
  (which is on the order of the
@@
 $\tilde{\beta}=1$.
-% his is in comparison with the timescale of intial vortex dissipation which has 
-% non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.
-%non-monotonic dependence for all timescales
+
+{\bf We comment here that a locally isothermal simulation, where the
+  disc sound-speed is kept constant in time, was also
+  performed for comparison. In this case we did not observe
+  significant vortex decay within the simulation timescale
+ (implying $t_\mathrm{diss}\gtrsim2000P_0$).  We expect corresponding vortex
+ lifetime to exceed that for $\tilde{\beta}=0.01$. Including an energy
+ equation, even with rapid cooling (or setting $\gamma$ close to
+ unity), could lead to discrepencies with a locally isothermal disk due to the
+ advection-creation of specific entropy within the planet's horseshoe
+ region in the former case,  thereby affecting the generalized vortensity 
+ and therefore the vortex evolution. Nevertheless, a longer vortex
+ lifetime in a locally isothermal disk would 
+ be consistent with the above trend of increasing vortex lifetimes as
+ $\tilde{\beta}\to0$. }
 
@@
     (blue). \label{betaplot}}  
-% {\bf maybe useful to over-plot other timescales,
-%      e.g. time taken to reach max mach, time for rossby number to
-%      reach zero.}
 \end{figure}
@@
 double-peaked behaviour of $t_\mathrm{life}$ in Fig. \ref{betaplot}. 
-% \footnote{The 
-%  decay timescale itself is also a factor, but we found this timescale
-%  to be much shorter than the duration of the quasi-steady state
-%  (Fig. \ref{lifetimeplot}), except at very short cooling times.}. 
 We suggest below competing factors that may result in this
@@
 $\Delta\Sigma/\Sigma_0=1.48$ for $\tilde\beta=10$.%, both with
-%$Ro\simeq-0.15$.   
-% {\bf rossby numbers?}
 

### 2015-03-26 paper/summary.tex
revision 2 done

@@
 stability, which employ locally or strictly isothermal equations of
-state. {\bf We considered the inviscid limit which favors the RWI
+state. We considered the inviscid limit which favors the RWI
   \citep{li09,fu14} and avoids complications from viscous
@@
   our simulations are likely longer than in realistic discs with
-  non-zero physical viscosity.}
+  non-zero physical viscosity.
 
@@
 we find that with the planet potential kept on, the vortex amplitude
-grows during this quasi-steady state{\bf, during which no vortex migration is
-  observed}, until it begins to induce
+grows during this quasi-steady state, during which no vortex migration is
+  observed, until it begins to induce
 shocks, after which the vortex amplitude begins to decay.
@@
 decay, after reaching maximum amplitude, decreases monotonically with
-increasing cooling time. {\bf We do observe vortex migration during this
-  decay, which may influence this decay timescale. } 
+increasing cooling time. We do observe vortex migration during this
+  decay, which may influence this decay timescale.  
 The sum of
@@
 
-{\bf
   \emph{Orbital migration.} We have held the planet fixed on a circular
@@
   should be clarified.
-}
 
-{\bf 
   \emph{Cooling model.} Our prescription for the disc 
@@
   opacity models etc., should be used in future work.
-}
 

### 2015-03-26 paper/results.tex
Response 2

@@
 {\bf For completeness we also simulated a locally
-  isothermal disk (at lower resolution) with $\epsilon=0.05$ where the
+  isothermal disk with $\epsilon=0.05$ where the
   sound-speed is kept strictly equal to its initial value.  This
@@
   $\sigma/\Omega(r_p)\simeq 0.05$ at $m=5$,   
-  compared with a value of $0.085$ at $m=?$ for a corresponding
+  compared with a value of $0.085$ at $m=6$ for a corresponding
   simulation that includes the energy equation but with rapid cooling

### 2015-03-26 paper/model.tex
revision 2 done

@@
 Navier-Stokes viscosity whose magnitude is  
-characterised by a constant kinematic viscosity parameter
+characterized by a constant kinematic viscosity parameter
 $\nu$. However, we will be considering effectively inviscid discs by
@@
 have been chosen to be absent at $t=0$, allowing the disc to be
-initialised close to steady state. The $\mathcal{C}$ function attempts
+initialized close to steady state. The $\mathcal{C}$ function attempts
 to restore the initial energy density (and 
@@
 \subsubsection{Fourier modes} 
-The RWI is characterised by exponentially
+The RWI is characterized by exponentially
 growing perturbations. Though in this paper we do not consider a

### 2015-03-26 paper/results.tex
revision 2 done

@@
 structures. An example is shown in Fig. \ref{linearmodes} for 
-$\tilde{\beta}=10$. We characterise these
+$\tilde{\beta}=10$. We characterize these
 modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
@@
   \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimensionalized by the initial axisymmetric component 
+    density, non-dimensioqnalized by the initial axisymmetric component 
     $\Sigma_0(t=0)$, for the
@@
 
-We characterise the smoothness of the outer gap edge with a dimensionless
+We characterize the smoothness of the outer gap edge with a dimensionless
 gap edge gradient parameter
@@
   significant vortex decay within the simulation timescale
- (implying $t_\mathrm{diss}\gtrsim2000P_0$).  We expect corresponding vortex
+ (implying $t_\mathrm{diss}\gtrsim2000P_0$).  We expect the corresponding vortex
  lifetime to exceed that for $\tilde{\beta}=0.01$. Including an energy
  equation, even with rapid cooling (or setting $\gamma$ close to 
- unity), could lead to discrepencies with a locally isothermal disk due to the
- advection-creation of specific entropy within the planet's horseshoe
- region in the former case,  thereby affecting the generalized vortensity 
- and therefore the vortex evolution. Nevertheless, a longer vortex
- lifetime in a locally isothermal disk would 
+ unity), could still lead to discrepancies with a locally isothermal disc. This is 
+ due to the advection-creation of specific entropy within the planet's horseshoe
+ region with the former case,  thereby affecting the generalized vortensity 
+ and therefore the instability (\S\ref{linear}) and subsequent vortex evolution. 
+ Nevertheless, a longer vortex 
+ lifetime at the gap edge in a locally isothermal disk would 
  be consistent with the above trend of increasing vortex lifetimes as
- $\tilde{\beta}\to0$. }
+ $\tilde{\beta}\to0$.}
 
@@
 %growth of Mach number with respect to time is increasing with cooling rate.
-%Thus the intially weaker formed vortices for higher cooling rate grow
+%Thus the initially weaker formed vortices for higher cooling rate grow
 %substantialy faster than the sound speed and can shock the system in earlier

### 2015-03-27 paper/model.tex
updating

@@
 where the power-law index $s=2$, $H(r) = \ciso\Omega_k $ defines the disc scale-height 
-where $\ciso=\sqrt{p/\Sigma}$ isothermal sound-speed. The disc aspect ratio is defined as $h\equiv H/r$ and initially
+where $\ciso=\sqrt{p/\Sigma}$ is the isothermal sound-speed. The disc aspect ratio is defined as $h\equiv H/r$ and initially
 $h=0.05$. For a non-self-gravitating disc, the surface density scale

### 2015-03-27 paper/results.tex
updating

@@
 {\bf For completeness we also simulated a locally
-  isothermal disk with $\epsilon=0.05$ where the
+  isothermal disc with $h=0.05$ where the
   sound-speed is kept strictly equal to its initial value.  This
@@
  lifetime to exceed that for $\tilde{\beta}=0.01$. Including an energy
- equation, even with rapid cooling (or setting $\gamma$ close to 
+ equation with rapid cooling (or setting $\gamma$ close to 
  unity), could still lead to discrepancies with a locally isothermal disc. This is 
@@
  Nevertheless, a longer vortex 
- lifetime at the gap edge in a locally isothermal disk would 
+ lifetime in a locally isothermal disc would 
  be consistent with the above trend of increasing vortex lifetimes as

### 2015-03-27 paper/model.tex

--- 2015-03-26 paper/model.tex
@@
 frame. 
-In the momentum equations, $\bm{f}_\nu$ represent viscous forces, 
+In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
 which includes artificial bulk viscosity to handle shocks, and a
-Navier-Stokes viscosity whose magnitude is  
+Navier-Stokes viscosity} whose magnitude is  
 characterised by a constant kinematic viscosity parameter
@@
 \end{align}
-where $Q^+$ represents viscous heating (from both physical and
-  artificial viscosity) and subscript $i$ denotes
+where $Q^+$ represents viscous heating {\bf(from both physical and
+  artificial viscosity)} and subscript $i$ denotes
 evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
@@
 is a dimensionless measure of relative vorticity. 
+% the ratio of interial to Corriolis forces. 
 Here $\langle \cdot \rangle_{\phi}$ denotes an azimuthal

### 2015-03-27 paper/results.tex
revision 3

@@
 
-{\bf For completeness we also simulated a locally
+ For completeness we also simulated a locally
   isothermal disc with $h=0.05$ where the
@@
   simulation that includes the energy equation but with rapid cooling
-  $\tilde{\beta}=0.01$. However, the vortex evolution is similar.}
+  $\tilde{\beta}=0.01$. However, the vortex evolution is similar.
 
@@
 will discuss this issue further in 
-\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
-increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
+\S\ref{lifetime_discuss}. Notice the decay timescale {\bf can be long} with
+rapid cooling: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$ 
 whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
@@
 In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
-, but becomes more compact approaching a ratio of 2 during its decay ($t=1310$).   
+, but becomes more compact approaching a ratio of 2 during its decay ($t=1310P_0$).   
 Notice in Fig. \ref{shockplot} the appearance of wakes extending from
@@
 We now examine vortex lifetimes as function of the imposed cooling
-times. {\bf For this study, additional simulations with 
-  $\tilde\beta=0.01,\,0.25,\,0.75,\,2.5,\,7.5$ were also performed.} 
+times. For this study, additional simulations with 
+  $\tilde\beta=0.01,\,0.25,\,0.75,\,2.5,\,7.5$ were also performed.  
 We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
@@
 decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
-$\tilde{\beta}=0.5$. This is due to the longer decay timescale with
-decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
+$\tilde{\beta}=0.5$. {\bf Note that for very small $\tilde{\beta}$, 
+  there is significant contribution to the overall vortex lifetime due
+  to a long decay timescale. }
+For longer cooling times ($\tilde{\beta}\gtrsim
 1$) the vortex lifetimes maximizes at $\tilde\beta=2.5$ with 
-$t_{\mathrm{life}}\approx 1650P_0$.  The other timescales $t_\mathrm{diss}$ and
-$t_\mathrm{Mach}$ are less complicated, but still display a
-non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
-$\tilde{\beta}=1$.
+$t_{\mathrm{life}}\approx 1650P_0$.  
+%The other timescales $t_\mathrm{diss}$ and
+%$t_\mathrm{Mach}$ are less complicated, but still display a
+%non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
+%$\tilde{\beta}=1$.
 
-{\bf We comment here that a locally isothermal simulation, where the
+ We comment here that a locally isothermal simulation, where the
   disc sound-speed is kept constant in time, was also
@@
  be consistent with the above trend of increasing vortex lifetimes as
- $\tilde{\beta}\to0$.}
+ $\tilde{\beta}\to0$.
 
@@
 
-In the previous section, we observed that vortices began to decay when
-it starts to induce shocks. We therefore suggest the vortex lifetime
-is determined the time needed for the vortex to grow to 
-sufficient amplitude to induce shocks in the surrounding fluid, and 
-the decay timescale itself. The latter was found to shorten 
-with increasing $\tilde{\beta}$. However, the former timescale
-($t_\mathrm{diss}$), which may be considered as the 
-duration of the quasi-steady state, reaches a single maximum with
-respect to cooling time. The sum of these two timescales results in the
-double-peaked behaviour of $t_\mathrm{life}$ in Fig. \ref{betaplot}. 
-We suggest below competing factors that may result in this
+In the previous section, we observed that vortices began to decay when it starts to
+induce shocks. {\bf We thus suggest that the time needed for the vortex to grow to 
+sufficient amplitude to induce shocks in the surrounding fluid,   
+%The former timescale
+%($t_\mathrm{diss}$), 
+which may be considered as the duration of the quasi-steady state or
+$t_\mathrm{diss}$, is an important contribution to the overall vortex
+lifetime.} 
+%reaches a single maximum with
+%respect to cooling time {\bf for $\tilde{\beta}\geq 0.1$ (see also Fig. \ref{lifetimeplot})}. 
+%{\bf (The case with $\tilde{\beta}=0.01$ and the locally isothermal disc, 
+%does not follow this broad trend and may need to be analyzed separately.)} 
+%The sum of these two timescales leads to the 
+%double-peaked behaviour of $t_\mathrm{life}$ in Fig. \ref{betaplot}. 
+We discuss below some competing factors that may result in a 
 non-monotonic dependence of $t_\mathrm{diss}$ on the cooling rate.

### 2015-03-27 paper/summary.tex
updating
revision 3

@@
 
-The duration of the quasi-steady state increases with
+{\bf For our main simulations with $\tilde{\beta}\geq 0.1$}, the duration of the quasi-steady state increases with
 increasing cooling timescales until a critical value, beyond which this
-quasi-steady state shortens again. The timescale for the vortex to
-decay, after reaching maximum amplitude, decreases monotonically with
-increasing cooling time. We do observe vortex migration during this
+quasi-steady state shortens again. {\bf We find the timescale for the vortex to
+decay, after reaching maximum amplitude, can be long for small
+$\tilde{\beta}$ which contributes to a long overall vortex lifetime
+with rapid cooling.} We do observe vortex migration during {\bf its}
 decay, which may influence this decay timescale.
-The sum of
-these two timescales results in a double-peaked vortex lifetime as a
-function of cooling timescale. 
-%technically we should probably say there is an optimum cooling time
-%to maximise the quasi-steady state
 
-%Our results indicate there exists an optimum cooling timescale to
-%maximise the lifetime of gap-edge vortices.
+% The sum of
+% these two timescales results in a double-peaked vortex lifetime as a
+% function of cooling timescale.
 
-We suggest the non-monotonic dependence \emph{of the quasi-steady state} can
+We suggest a non-monotonic dependence of the quasi-steady state {\bf
+on the cooling timescale $\tilde{\beta}$} can
 be attributed to the time required for the vortex to grow to 
@@
 These competing factors imply
-for both very short and long cooling timescales, the vortex reaches
+for both short and long cooling timescales, the vortex reaches
 its maximum amplitude, shock, and begins to decay, sooner than
-intermediate cooling timescales. 
+intermediate cooling timescales. {(\bf However, for very rapid
+  cooling, e.g. $\tilde{\beta}=0.01$, the quasi-steady state is also
+  quite long. In this case other factors may be important, such as
+  vortex migration.)}
+
 %{\bf, for which the quasi-steady state is 
 %the longest.}      
-
 %decay timescale depends on cooling time, but there seems to be a
@@
 we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
-the longest vortex lifetime, that  
+the longest vortex lifetime {\bf in the presence of moderate cooling}, that  
 $h\approx0.058$ on average.  
 %{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
-Our result is therefore consistent with \cite{fu14}.
+Our result is consistent with \cite{fu14}.
 
@@
   vortex lifetimes in actual 
-  protoplanetary discs, an improved thermodynamics treatment ,
+  protoplanetary discs, an improved thermodynamics treatment,
   e.g. radiative cooling based on realistic disc temperature, density,

### 2015-03-27 paper/results.tex
revision 3

@@
 $t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
-decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
-$\tilde{\beta}=0.5$. {\bf Note that for very small $\tilde{\beta}$, 
+decreases to $t_{\mathrm{life}} \approx 950P_0$ for
+$\tilde{\beta}=0.25$. {\bf Note that for very small $\tilde{\beta}$, 
   there is significant contribution to the overall vortex lifetime due

### 2015-03-27 paper/results.tex

@@
 clean gap does not form in this short time period.%
+%\footnote{We find the
+% RWI begins to develop if the disc is evolved beyond $t\sim30P_0$}.
 
@@
 $\tilde{\beta}$ is increased.
+
+%However, notice the gap structure
+%changes most significantly for $\beta =0.1 \to 1.0$. For $\beta \geq
+%1$
 
@@
 large $\tilde{\beta}$. 
+% or 'suggets that .... becomes unimportant for sufficiently large...'
 
@@
 At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
-(see below),  we find $\kappa^2 + N^2$ reaches local minimum with a value
+(see below),{\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
 $\sim 0.45 \, \Omega^2(r_p)$
@@
 Brunt-V\"ais\"al\"a frequency at the outer gap edge is
-$N\sim 0.1 \, \Omega(r_p)$,  
+$N\sim 0.1 \, \Omega(r_p)$} 
 % $N\sim 10^{-5}$,
@@
 Thus for all values of $\tilde\beta$ the planet-induced gaps are
-stable to axisymmertic {\bf perturbations}.
+stable to axisymmertic instabilities.
+
+% {\bf Brunt frequency the same for all cooling?:small decrease with beta,
+%  not more than order of magnitude}
+
+%  {\bf i assume this means the azimuthally averaged
+%   values. do you ever get $N^2<0$ anywhere in the 2D disc?:
+%  This was not for azimuthal average, below 0 inside gap but kappa+N always above 0 everywhere}
 
@@
 structures. An example is shown in Fig. \ref{linearmodes} for 
-$\tilde{\beta}=10$. We characterize these
+$\tilde{\beta}=10$. We characterise these
 modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
@@
   \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimensioqnalized by the initial axisymmetric component 
+    density, non-dimensionalized by the initial axisymmetric component 
     $\Sigma_0(t=0)$, for the
@@
 \subsection{Long term evolution} \label{nonlinearplanetoff} 
+% {\bf typical rossby numbers for these `planet-off' vortices? do
+%   stronger vortices (beta=0.1) dissipate faster?:very complicated
+% maybe not worth discussing since just comparison to planet-off}
 
@@
 
-This decay is probably due to numerical viscosity. During the
+{\bf This decay is probably due to numerical viscosity. During the
   slow decay the vortex elongates (weakens) while its radial width
@@
   spiral waves associated with the vortex, which may contribute to
-  its dissipation (see below). 
+  its dissipation (see below).} 
+% (true?) (For isothermal and long
+% cases yes. Spiral waves appear in beta=0.1,0.5,1.0}) 
 We will see in the
-next section that this decay after linear growth is very
+next section that this {\bf decay after linear growth} is very
 different to when the planet potential is kept on.
+
+%Interestingly, though, vortices arising from the stronger
+%instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
+%faster.
+
+%main point: no planet: only get growth during linear instability
+%           with palnet: further growth after linear regime
 
@@
 We now examine long-term simulations of gap-edge vortices for
-$\tilde{\beta}=0.1,0.5,1,5,10$. (Additional cases are presented in 
+$\tilde{\beta}=0.1,0.5,1,5,10$.{\bf(Additional cases are presented in 
   \S\ref{lifetime_discuss} when examining vortex lifetimes as a
-  function of $\tilde{\beta}$.)  The planet potential is kept on
+  function of $\tilde{\beta}$.)} The planet potential is kept on
 throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
@@
 $(N_r,N_{\phi})=(256,512)$ show similar behaviour and trends as the
-high-resolution runs reported below. % {\bf These simulations
-  % include a comparison between a run with $\tilde{\beta}=0.01$ with a 
-  % locally isothermal disc, which show similar behavior. In particular,
-  % we find vortex lifetimes have a non-monotonic dependence on
-  % $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
-  % locally isothermal discs) and for $\tilde{\beta} = O(1)$}.
+high-resolution runs reported below. {\bf These simulations
+  include a comparison between a run with $\tilde{\beta}=0.01$ with a 
+  locally isothermal disc, which show similar behavior. In particular,
+  we find vortex lifetimes have a non-monotonic dependence on
+  $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
+  locally isothermal discs) and for $\tilde{\beta} = O(1)$}.
 
@@
 evolution of Rossby numbers measured at the vortex centres.  
+%As the vortex forms it has an initial 
+%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
+%(implying anti-cyclonic 
+%motion). 
 The Rossby number increases in magnitude during quasi-steady state,
@@
     vortex at the outer gap edge. The plot for other values of the cooling time
-    $\tilde{\beta}$ are similar. The decrease in the surface
-      density near $r\sim 40 r_\mathrm{in}$ arises from mass loss due to the open
-      boundary condition imposed at $r_\mathrm{out}$.
+    $\tilde{\beta}$ are similar. {\bf The decrease in the surface
+      density near} $r\sim 40 r_\mathrm{in}$ {\bf arises from mass loss due to the open
+      boundary condition imposed at} $r_\mathrm{out}$.
     \label{Vortex2D} }
@@
 \ref{planetofflifetimeplot}).
+
+% We remark that the increase in the
+% vortex surface density directly due to the pile-up of material at the
+% gap edge (because of gap-opening) is small: the average surface density
+% perturbation at vortex radial region, $r\sim r_{p}+5r_h$, is 
+% $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.
 
@@
 
+% {\bf may be more appropriate to quote the azimuthally averaged
+%   perturbation at the location of the vortex (5-6 Hill radii from the
+%   planet?): yes, this was my original idea, mistyped as 2.5
+% }
+
+% {\bf any trend between maximum over-density achieved and cooling?: generally increasing with beta, max at beta=5}
+
+% {\bf question: vortex amplitude plot suggest just after linear growth,
+%   vortex amplitudes are smaller with increasing beta, but later on in
+%   the quasi-steady state, this trend reverses? yes. 
+% }
+
+% {\bf Perhaps a 2D snapshot of what the gap edge looks like after
+%   vortex death (like the one in the poster).: Not sure if this is figure 
+% you were thinking (Fig 8)
+% % defer definition of  vortex lifetime to later, when we discuss it. 
+% }
+
 % \begin{figure}
@@
 the vortex begins to decay coincides with the emergence of these
-wakes.
+wakes. %{\bf true statement? yes, distortion of vortex shape form wakes happens at same time}
 
-During the quasi-steady state the vortex 
+{\bf During the quasi-steady state the vortex 
   orbits at $r\sim r_p+6r_h$. We do not see significant
@@
   begins to migrate inwards to $r\sim r_p+5r_h$.  
+  % This is a result of the the spiral waves transporting
+  % angular momentum to the disc (Is this true? Shocks on the inner disc
+  % seem stronger so I would think it would migrate outward). 
+  }
 
@@
 until the vortex dies out, at which point the associated Rossby number
-begins decreasing to zero. After the 
+begins decreasing to zero. {\bf After the 
   vortex reaches small amplitudes ($1\lesssim\Delta\Sigma/\Sigma$), it 
   migrates out to $r\sim r_p+ 6.5 r_h$. 
-
+}
 %{\bf :check}.
@@
     (top) and the associated surface density gradient (bottom) are
-    shown. Wake-like features corresponding to large density gradients
+    shown. Wake like features corresponding to large density gradients
     are found to originate from the vortices during the late phase of 
@@
 
-We characterize the smoothness of the outer gap edge with a dimensionless
+We characterise the smoothness of the outer gap edge with a dimensionless
 gap edge gradient parameter
@@
 associated with the vortex. This acts against gap-opening
-by the planet, and smooths out the outer surface density bump,    
- so the condition for the RWI 
-becomes less favourable. In
+by the planet{\bf, and smooths out the outer surface density bump,    
+} so the condition for the RWI 
+becomes less favourable. {\bf In
   order to re-launch the RWI, the surface density bump should
@@
   in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}) to
-  form a surface density bump outside the gap. This may explain
-why vortices do  not reform again (at least within the simulation
-timescale).  After full decay the aspect ratio at the outer gap edge is
-  $h\sim0.05$.
+  form a surface density bump outside the gap.} {\bf This may} explain
+why vortices do  not reform again (at least within the simulation timescale). 
+{\bf After full decay the aspect ratio at the outer gap edge is
+  $h\sim0.05$. }
+% (0.06 for beta=5.0,10)}
+
+% what is aspect-ratio at outer gap edge after full decay?
+
+%Thus, the
+%time needed for the vortex to grow to sufficient amplitude to modify
+%the gap structure
 
@@
     dissipation,
-    which occurs at $t\approx10^3P_0$ as seen in Fig. \ref{lifetimeplot},
+    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
     works to smooth out the gap edge.
@@
 at which the over-density of the vortex
+% to reach $10\%$ of its max overdensity value
 returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
+%{\bf correct interpretation?}
  (which is on the order of the
@@
 $t_{\mathrm{life}}\approx 1650P_0$.
-%The other timescales $t_\mathrm{diss}$ and
-%$t_\mathrm{Mach}$ are less complicated, but still display a
-%non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
-%$\tilde{\beta}=1$.
 
@@
     (blue). \label{betaplot}}  
+% {\bf maybe useful to over-plot other timescales,
+%      e.g. time taken to reach max mach, time for rossby number to
+%      reach zero.}
 \end{figure}
@@
 %growth of Mach number with respect to time is increasing with cooling rate.
-%Thus the initially weaker formed vortices for higher cooling rate grow
+%Thus the intially weaker formed vortices for higher cooling rate grow
 %substantialy faster than the sound speed and can shock the system in earlier

### 2015-03-27 paper/summary.tex
backup

@@
 stability, which employ locally or strictly isothermal equations of
-state. We considered the inviscid limit which favors the RWI
+state. {\bf We considered the inviscid limit which favors the RWI
   \citep{li09,fu14} and avoids complications from viscous
@@
   our simulations are likely longer than in realistic discs with
-  non-zero physical viscosity.
+  non-zero physical viscosity.}
 
@@
 we find that with the planet potential kept on, the vortex amplitude
-grows during this quasi-steady state, during which no vortex migration is
-  observed, until it begins to induce
+grows during this quasi-steady state{\bf, during which no vortex migration is
+  observed}, until it begins to induce
 shocks, after which the vortex amplitude begins to decay.
@@
 increasing cooling timescales until a critical value, beyond which this
-quasi-steady state shortens again. {\bf We find the timescale for the vortex to
-decay, after reaching maximum amplitude, can be long for small
-$\tilde{\beta}$ which contributes to a long overall vortex lifetime
-with rapid cooling.} We do observe vortex migration during {\bf its}
-decay, which may influence this decay timescale.
-
-% The sum of
-% these two timescales results in a double-peaked vortex lifetime as a
-% function of cooling timescale.
+quasi-steady state shortens again. The timescale for the vortex to
+decay, after reaching maximum amplitude, decreases monotonically with
+increasing cooling time. {\bf We do observe vortex migration during this
+  decay, which may influence this decay timescale. }
 
@@
 its maximum amplitude, shock, and begins to decay, sooner than
-intermediate cooling timescales. {(\bf However, for very rapid
-  cooling, e.g. $\tilde{\beta}=0.01$, the quasi-steady state is also
-  quite long. In this case other factors may be important, such as
-  vortex migration.)}
+intermediate cooling timescales.
+
+{(\bf However, for very rapid cooling, e.g. $\tilde{\beta}=0.01$, the quasi-steady state is also
+  quite long. This suggests that the above effects themselves do not
+  have a simple dependence on the cooling timescale when considering
+  $\tilde{\beta}\to0$ and/or that other factors become important in
+  this limit. This should be investigated in future works.)}
 
@@
 We remark that a non-monotonic dependence of the vortex lifetime was
-also reported by \cite{fu14}, who performed isothermal disc-planet
+also reported by \cite{fu14}, who performed {\bf locally} isothermal disc-planet
 simulations with different values of the   
@@
 
+{\bf
   \emph{Orbital migration.} We have held the planet fixed on a circular
@@
   should be clarified. 
+}
 
+{\bf 
   \emph{Cooling model.} Our prescription for the disc 
@@
   opacity models etc., should be used in future work.   
+}
 

### 2015-03-27 paper/summary.tex
fixed versions

--- 2015-03-27 paper/summary.tex
@@
 quasi-steady state shortens again. {\bf We find the timescale for the vortex to
-decay, after reaching maximum amplitude, can be long for small
-$\tilde{\beta}$ which contributes to a long overall vortex lifetime
+decay after reaching maximum amplitude can be long for small
+$\tilde{\beta}$, which contributes to a long overall vortex lifetime
 with rapid cooling.} We do observe vortex migration during {\bf its}
